---
title: "Niche Generation and Differential Abundance — Threshold Sweep"
author: "Tazein Shah"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)

suppressPackageStartupMessages({
  library(imcRtools)
  library(tidyverse)
  library(SingleCellExperiment)
  library(SpatialExperiment)
  library(sccomp)
  library(ggrepel)
  library(mclust)   # adjustedRandIndex
})

set.seed(12345)
```

---

## 1. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/spatial_analysis/run_3"
spatial_graph_path <- file.path(analysis_path, "spe_spatial_graph_r3.rds")

spe_graph <- readRDS(spatial_graph_path)

thresholds  <- c(10, 20, 30, 40)
kmeansClust <- 50

# All sweep output lives under its own subdirectory, separate from the
# existing production niche_k50_thresh20/ folder -- this sweep never
# reads from or writes to that folder.
sweep_path <- file.path(analysis_path, "niche_sweep")
dir.create(sweep_path, recursive = TRUE, showWarnings = FALSE)

cat("Cells:", ncol(spe_graph), "\n")
cat("Levels in spatial graph object:\n")
print(levels(spe_graph$final_celltype))
```

---

## 2. Load spatial graph object

```{r load}
spe <- spe_graph

cat("Cells:", ncol(spe), "\n")
cat("Cell types:", nlevels(spe$final_celltype), "\n")
```

---

## 3. Threshold sweep: build graph, aggregate neighbors, generate niches

For each threshold, rebuild the expansion graph from scratch, aggregate
neighbor composition, and run kmeans. Each threshold gets its own
output folder under `sweep_path`.

```{r covariate-fn}
# Patient-level covariates are threshold-invariant -- factored out so
# they're computed identically for every spe_t below.
add_covariates <- function(spe_t) {
  spe_t$stage_group <- sapply(spe_t$patient_id, function(x) {
    if (x %in% c("Patient12", "Patient15")) return("IIB")
    else if (x == "Patient13") return("IIIA")
    else if (x == "Patient19") return("IB")
    else if (x == "Patient11" || grepl("^Patient20", x)) return("IA3")
    else if (x %in% c("Patient14", "Patient16", "Patient17", "Patient18")) return("IA2")
    else if (x == "Patient10") return("Benign")
    else return(NA)
  })
  spe_t$stage_group <- factor(spe_t$stage_group,
                              levels = c("Benign", "IA2", "IA3", "IB", "IIIA", "IIB"))

  spe_t$disease_state <- sapply(spe_t$patient_id, function(x) {
    if (x == "Patient10") return("Benign")
    else if (x %in% c("Patient11", "Patient14", "Patient16", "Patient17", "Patient18") || grepl("^Patient20", x)) return("Lower")
    else if (x %in% c("Patient12", "Patient13", "Patient15", "Patient19")) return("Higher")
    else return(NA)
  })
  spe_t$disease_state <- factor(spe_t$disease_state, levels = c("Benign", "Lower", "Higher"))

  spe_t
}
```

```{r threshold-sweep}
niche_list <- list()

for (t in thresholds) {

  cat("\n=== Building niches at threshold =", t, "===\n")

  thresh_path <- file.path(sweep_path, paste0("k", kmeansClust, "_thresh", t))
  dir.create(thresh_path, recursive = TRUE, showWarnings = FALSE)
  niche_spe_out_path <- file.path(thresh_path, "spe_niche.rds")

  if (file.exists(niche_spe_out_path)) {
    cat("Found existing niche object, loading:", niche_spe_out_path, "\n")
    spe_t <- readRDS(niche_spe_out_path)
  } else {
    spe_t <- buildSpatialGraph(spe, img_id = "sample_id", type = "expansion", threshold = t)

    spe_t <- aggregateNeighbors(spe_t,
                                colPairName = "expansion_interaction_graph",
                                aggregate_by = "metadata",
                                count_by = "final_celltype")

    cn <- kmeans(spe_t$aggregatedNeighbors, centers = kmeansClust, iter.max = 10000)
    spe_t$Niche <- as.factor(cn$cluster)

    spe_t <- add_covariates(spe_t)

    saveRDS(spe_t, niche_spe_out_path)
    cat("Saved:", niche_spe_out_path, "\n")
  }

  niche_list[[as.character(t)]] <- spe_t
}
```

---

## 4. Sweep diagnostics: niche size and stability across thresholds

```{r niche-size-comparison, fig.width = 10, fig.height = 6}
niche_sizes <- purrr::imap_dfr(niche_list, function(spe_t, t) {
  as.data.frame(table(spe_t$Niche)) |>
    setNames(c("Niche", "n_cells")) |>
    mutate(threshold = as.numeric(t))
})

ggplot(niche_sizes, aes(x = factor(threshold), y = n_cells)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 0.8, alpha = 0.4) +
  theme_classic() +
  labs(x = "Expansion threshold (µm)", y = "Cells per niche",
       title = "Niche size distribution by threshold")
```

```{r ari-comparison}
# Niche assignments are only comparable cell-to-cell if cell order is
# identical across thresholds -- true here since all four spe_t objects
# derive from the same spe_graph without any row/column reordering.
ari_mat <- outer(thresholds, thresholds, Vectorize(function(a, b) {
  adjustedRandIndex(niche_list[[as.character(a)]]$Niche,
                    niche_list[[as.character(b)]]$Niche)
}))
rownames(ari_mat) <- colnames(ari_mat) <- thresholds
cat("Pairwise ARI between threshold niche assignments:\n")
print(round(ari_mat, 3))
```

---

## 5. Differential niche abundance across thresholds (sccomp)

Stage model only, fit separately per threshold. `.sample = Patient`
reasoning unchanged (stage constant within patient, not ROI).

```{r sccomp-sweep}
sccomp_stage_list <- list()

for (t in thresholds) {

  cat("\n=== sccomp Stage model at threshold =", t, "===\n")

  spe_t <- niche_list[[as.character(t)]]
  thresh_path <- file.path(sweep_path, paste0("k", kmeansClust, "_thresh", t))

  sccomp_stage_out_path <- file.path(thresh_path, "sccomp_stage.rds")

  sceIMC <- SingleCellExperiment(list(counts = assay(spe_t, "counts")))
  sceIMC$Patient <- factor(spe_t$patient_id)
  sceIMC$Sample  <- factor(spe_t$sample_id)
  sceIMC$Niche   <- factor(spe_t$Niche)
  sceIMC$Stage   <- factor(spe_t$stage_group,
                           levels = c("Benign", "IA2", "IA3", "IB", "IIIA", "IIB"))

  if (file.exists(sccomp_stage_out_path)) {
    cat("Found existing Stage output, loading\n")
    sccomp_s <- readRDS(sccomp_stage_out_path)
  } else {
    sccomp_s <-
      sceIMC |>
      sccomp_estimate(
        formula_composition = ~ Stage + (1 | Patient),
        .sample = Sample,
        .cell_group = Niche,
        bimodal_mean_variability_association = FALSE,
        variational_inference = FALSE,
        output_directory = file.path(thresh_path, "sccomp_draws_stage"),
        cores = 8
      ) |>
      sccomp_test()
    saveRDS(sccomp_s, sccomp_stage_out_path)
  }

  sccomp_stage_list[[as.character(t)]] <- sccomp_s |>
    as.data.frame() |>
    filter(factor == "Stage") |>
    mutate(Niche = as.character(Niche), threshold = t)
}

sccomp_stage_all <- bind_rows(sccomp_stage_list)
```

---

## 6. Significant niche counts by threshold

```{r sig-counts-table}
sig_counts <- sccomp_stage_all |>
  group_by(threshold) |>
  summarise(
    n_fdr10 = sum(c_FDR < 0.1, na.rm = TRUE),
    n_fdr05 = sum(c_FDR < 0.05, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(threshold)

sig_counts |>
  knitr::kable(caption = "Number of significant Stage niche-contrasts by threshold (out of up to 250 rows: 50 niches x 5 contrasts)")
```

```{r sig-counts-plot, fig.width = 8, fig.height = 5}
ggplot(sig_counts, aes(x = factor(threshold), y = n_fdr10)) +
  geom_col(fill = "steelblue") +
  theme_classic() +
  labs(x = "Expansion threshold (µm)", y = "Significant niches (FDR < 0.1)",
       title = "Significant Stage niche-contrast counts across thresholds")
```

---

## 7. Volcano plots by threshold {.tabset}

```{r volcano-sweep, results = "asis", fig.width = 7, fig.height = 5.5}
for (t in thresholds) {
  cat("\n\n### Threshold", t, "\n\n")

  df_sub <- sccomp_stage_all |>
    filter(threshold == t, !is.nan(c_FDR), c_FDR < 0.1) |>
    mutate(
      LogFDR = -log(c_FDR),
      Significant = ifelse(c_FDR < 0.05, "FDR < 0.05", "Not significant")
    )

  print(
    ggplot(df_sub, aes(x = c_effect, y = LogFDR, color = Significant)) +
      geom_point(size = 2) +
      geom_text_repel(aes(label = Niche), size = 3, show.legend = FALSE) +
      scale_color_manual(values = c("FDR < 0.05" = "firebrick", "Not significant" = "grey70")) +
      theme_classic() +
      labs(x = "Effect size", y = "-log(FDR)",
           title = paste0("Stage, threshold = ", t))
  )
  cat("\n\n")
}
```

---
## 8. Choosing among candidate thresholds

Threshold 10 is excluded from here on — it showed a 70,000-cell outlier
niche (Section 4 size plot) and the weakest ARI agreement with every
other threshold (Section 4 stability matrix), both signs the
neighborhood radius is too small to produce stable niches. Thresholds
20, 30, and 40 all show comparable significant-contrast counts and
clean volcano plots (Sections 6-7), so the tiebreaker here is
composition: whether each threshold's significant niches have cell
type profiles that make biological sense, not just low FDR.

### 8a. Composition-based ranking across candidate thresholds {.tabset}

```{r composition-ranking, results = "asis"}
candidate_thresholds <- c(20, 30, 40)

patient_driven_clusters <- tibble::tribble(
  ~cluster,                                                ~dominant_patient,
  "Monocyte (PD1+)",                                       "Patient10",
  "Macrophage, Proliferating (MKI67+)",                    "Patient10",
  "Tumor Epithelial (Neuroendocrine)",                     "Patient17",
  "B Cell (CD19+)",                                        "Patient20_519",
  "Plasma Cell, TIM3+ (CD138+TIM3+CD45RO-)",               "Patient10",
  "DC/mregDC (Epi Adj)",                                   "Patient10"
)

# Built once per candidate threshold and kept for reuse in 8b/8c/8d
# below, so the composition/risk logic isn't computed twice.
niche_patient_driven_by_t <- list()
niche_ranking_by_t <- list()

for (t in candidate_thresholds) {

  cat("\n\n#### Threshold", t, "\n\n")

  spe_t <- niche_list[[as.character(t)]]
  sccomp_t <- sccomp_stage_list[[as.character(t)]]

  # Niche composition: proportion of each cell type within each niche
  niche_composition_t <- as.data.frame(prop.table(table(spe_t$Niche, spe_t$final_celltype), margin = 1))
  colnames(niche_composition_t) <- c("Niche", "CellType", "Proportion")

  niche_top_celltypes_t <- niche_composition_t |>
    group_by(Niche) |>
    slice_max(Proportion, n = 3) |>
    summarise(top_celltypes = paste0(CellType, " (", round(Proportion * 100, 1), "%)", collapse = "; "),
              .groups = "drop")

  # Patient-driven risk, specific to this threshold's niche assignments
  niche_patient_driven_t <- purrr::pmap_dfr(
    patient_driven_clusters,
    function(cluster, dominant_patient) {
      idx <- which(spe_t$final_celltype == cluster)
      data.frame(
        Niche = spe_t$Niche[idx],
        Patient = spe_t$patient_id[idx]
      ) |>
        group_by(Niche) |>
        summarise(
          n_cells = n(),
          pct_dominant_patient = mean(Patient == dominant_patient) * 100,
          .groups = "drop"
        ) |>
        mutate(cluster = cluster, dominant_patient = dominant_patient)
    }
  )
  niche_patient_driven_by_t[[as.character(t)]] <- niche_patient_driven_t

  niche_risk_flag_t <- niche_patient_driven_t |>
    mutate(Niche = as.character(Niche)) |>
    group_by(Niche) |>
    summarise(max_dominant_pct = max(pct_dominant_patient), .groups = "drop") |>
    mutate(patient_risk = ifelse(max_dominant_pct >= 50, "HIGH", "low"))

  niche_ranking_t <- sccomp_t |>
    filter(c_FDR < 0.05) |>
    mutate(Niche = as.character(Niche)) |>
    left_join(niche_top_celltypes_t, by = "Niche") |>
    left_join(niche_risk_flag_t, by = "Niche") |>
    mutate(patient_risk = replace_na(patient_risk, "low")) |>
    arrange(c_FDR) |>
    select(Niche, parameter, c_effect, c_FDR, patient_risk, top_celltypes)

  niche_ranking_by_t[[as.character(t)]] <- niche_ranking_t

  print(
    niche_ranking_t |>
      knitr::kable(digits = 4, caption = paste0("Ranked significant niches, threshold = ", t))
  )
  cat("\n\n")
}
```

### 8b. Select final threshold

Review the ranked tables above (skip any niche flagged `patient_risk =
"HIGH"`) and set the threshold whose top non-flagged niches have the
most coherent cell type composition.

```{r select-threshold}
selected_threshold <- 40

spe <- niche_list[[as.character(selected_threshold)]]
sccomp_niche_stage_sub <- sccomp_stage_list[[as.character(selected_threshold)]]
niche_patient_driven <- niche_patient_driven_by_t[[as.character(selected_threshold)]]

cat("Using threshold:", selected_threshold, "\n")
```

### 8c. Patient-driven clusters, per niche — plot

```{r patient-driven-plot, fig.width = 12, fig.height = 8}
ggplot(niche_patient_driven, aes(x = factor(Niche, levels = sort(unique(as.numeric(as.character(Niche))))),
                                  y = pct_dominant_patient)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~ paste0(cluster, "\n(", dominant_patient, ")"), scales = "free_x", ncol = 2) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
  labs(x = "Niche", y = "% of cluster's cells in niche from dominant patient",
       title = paste0("Patient-driven cluster composition across niches (threshold = ", selected_threshold, ")"))
```
```{r}
niche_patient_driven_by_t[["40"]] |>
  filter(Niche %in% c(11, 23, 30)) |>
  arrange(Niche, desc(pct_dominant_patient))
```

### 8d. High-risk niche x cluster combinations

```{r high-risk-table}
sig_niches_fdr01 <- sccomp_niche_stage_sub |> filter(c_FDR < 0.1) |> pull(Niche)

n_show <- 15
risk_threshold_pct <- 50
risk_threshold_n    <- 5

high_risk_tbl <- niche_patient_driven |>
  mutate(Niche = as.character(Niche)) |>
  filter(pct_dominant_patient >= risk_threshold_pct,
         n_cells >= risk_threshold_n,
         Niche %in% sig_niches_fdr01) |>
  arrange(desc(pct_dominant_patient))

high_risk_tbl |>
  slice_head(n = n_show) |>
  select(Niche, Cluster = cluster, Patient = dominant_patient,
         `% Dominant` = pct_dominant_patient, `N Cells` = n_cells) |>
  knitr::kable(digits = 1, caption = paste0("Top ", n_show, " highest-risk niche x cluster combinations, threshold = ", selected_threshold))
```

### 8e. Boxplots per significant niche {.tabset}

```{r boxplots, results = "asis", fig.width = 5, fig.height = 5.5}
sig_niches <- sccomp_niche_stage_sub |>
  filter(c_FDR < 0.05) |>
  pull(Niche) |>
  as.character() |>
  unique()
sig_niches <- as.character(sort(as.numeric(sig_niches)))

cat("Niches significant (FDR < 0.05), threshold =", selected_threshold, ":", length(sig_niches), "\n")

niche_prop <- table(spe$sample_id, spe$Niche) |>
  prop.table(margin = 1) |>
  as.data.frame()
colnames(niche_prop) <- c("Sample", "Niche", "Value")

niche_prop$Patient <- spe$patient_id[match(niche_prop$Sample, spe$sample_id)]
niche_prop$Stage <- spe$stage_group[match(niche_prop$Patient, spe$patient_id)]

for (n in sig_niches) {
  cat("\n\n#### Niche", n, "\n\n")
  df_sub <- niche_prop[niche_prop$Niche == n, ]
  df_sub$Value <- df_sub$Value * 100
  print(
    ggplot(df_sub, aes(x = Stage, y = Value)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(aes(color = Patient), width = 0.1, size = 1.5) +
      theme_classic() +
      ylab("% within sample") +
      ggtitle(paste0("Niche ", n))
  )
  cat("\n\n")
}
```