---
title: "Niche Annotation: Composition Summary for Significant Niches"
author: "Tazein Shah"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 8)

suppressPackageStartupMessages({
  library(SpatialExperiment)
  library(tidyverse)
  library(pheatmap)
  library(janitor)
  library(kableExtra)
  library(RColorBrewer)
  library(scales)
  library(tidytext)  # reorder_within() / scale_x_reordered() for per-facet sorting in 8a-facet
})
```
---

## 1. Paths

```{r paths}
analysis_path      <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/spatial_analysis/run_3"
spatial_graph_path <- file.path(analysis_path, "spe_spatial_graph_r3.rds")
niche_path         <- file.path(analysis_path, "niche_k50_thresh20")

niche_spe_path      <- file.path(niche_path, "spe_niche.rds")
sccomp_stage_path   <- file.path(niche_path, "sccomp_stage.rds")

composition_out_path   <- file.path(niche_path, "niche_avg_composition.csv")
top_celltypes_out_path <- file.path(niche_path, "niche_top_celltypes.csv")
```

---

## 2. Load objects

```{r load}
spe <- readRDS(niche_spe_path)
sccomp_niche_stage <- readRDS(sccomp_stage_path)
```

---

## 2b. Diagnose cell type naming

Checks whether `spe$aggregatedNeighbors` column names match
`levels(spe$final_celltype)` exactly -- confirmed clean at this stage
(`identical()` returns `TRUE`). The mangling actually originates one
step later: `spe$aggregatedNeighbors` is an S4 `DFrame`
(`S4Vectors`), and its `as.data.frame()` S4 method silently ignores
`check.names = FALSE`, so names got mangled on conversion in Section 4
despite that argument being set. Fixed there by routing the conversion
through `as.matrix()` first. `column_to_rownames()` remains a separate,
secondary source of mangling further downstream (fixed in Sections 4b
and 5 by building matrices manually with `as.matrix()` instead).

```{r diagnose-names}
raw_names   <- colnames(spe$aggregatedNeighbors)
clean_names <- levels(spe$final_celltype)

cat("Raw column names (n =", length(raw_names), "):\n")
print(raw_names)

cat("\nClean cell type levels (n =", length(clean_names), "):\n")
print(clean_names)

# Direct comparison -- raw names should already match clean levels
# exactly (no make.names() mangling at this stage).
cat("\nRaw names identical to clean levels (same set, same order):",
    identical(raw_names, clean_names), "\n")

setdiff_raw_clean <- setdiff(raw_names, clean_names)
setdiff_clean_raw <- setdiff(clean_names, raw_names)

if (length(setdiff_raw_clean) == 0 && length(setdiff_clean_raw) == 0) {
  cat("Confirmed: no mismatch. spe$aggregatedNeighbors column names are clean.\n")
} else {
  cat("Mismatch found:\n")
  cat("  In raw but not in clean levels:\n"); print(setdiff_raw_clean)
  cat("  In clean levels but not in raw:\n"); print(setdiff_clean_raw)
}
```

---

## 3. Identify significant niches

Significant niches (FDR < 0.05) from the Stage model.

```{r significant-niches}
sig_niches <- sccomp_niche_stage |>
  filter(factor == "Stage", c_FDR < 0.05) |>
  pull(Niche) |>
  unique()

cat("Significant niches (Stage):", length(sig_niches), "\n")
print(sort(as.numeric(as.character(sig_niches))))
```

---

## 3a. Significant niches: effect size and FDR

Table of every niche x contrast pair that cleared FDR < 0.05 under the
Stage model, sorted by FDR, with MCMC convergence diagnostics
(`c_rhat`, `c_ess_bulk`, `c_ess_tail`) and `c_pH0` included for
reference.

```{r significant-niches-table}
sig_niches_table <- sccomp_niche_stage |>
  filter(factor == "Stage", c_FDR < 0.05) |>
  select(Niche, factor, parameter, c_effect, c_FDR, c_pH0, c_rhat, c_ess_bulk, c_ess_tail) |>
  arrange(c_FDR)

sig_niches_table |>
  as.data.frame() |>
  knitr::kable(digits = 3, caption = "Significant niches: effect size and FDR (Stage model)")
```

---

## 3b. Patient-dominance check across flagged annotations

Several annotations show a single patient accounting for a
disproportionate share of that cluster cohort-wide, per the diagnostic
review table:

| Annotation                                             | Dominant Patient | % of Cells |
|---------------------------------------------------------|------------------|-----------:|
| Monocyte (PD1+)                                          | Patient10        | 48.4       |
| Macrophage, Proliferating (MKI67+)                       | Patient10        | 44.7       |
| Tumor Epithelial (Neuroendocrine)                         | Patient17        | 39.6       |
| B Cell (CD19+)                                            | Patient20_519    | 34.7       |
| Plasma Cell, TIM3+ (CD138+TIM3+CD45RO-)                   | Patient10        | 32.1       |
| DC/mregDC (Epi Adj)                                       | Patient10        | 28.9       |

Since all of these feed directly into niche composition, this reports,
per niche, what fraction of each flagged cluster's cells come from its
dominant patient -- joined onto the significant-niches table below
(3c) so it's visible alongside FDR/effect size when writing annotation
comments.

```{r patient-dominance-check}
flagged_annotations <- tribble(
  ~celltype_label,                                           ~dominant_patient,
  "Monocyte (PD1+)",                                         "Patient10",
  "Macrophage, Proliferating (MKI67+)",                      "Patient10",
  "Tumor Epithelial (Neuroendocrine)",                        "Patient17",
  "B Cell (CD19+)",                                           "Patient20_519",
  "Plasma Cell, TIM3+ (CD138+TIM3+CD45RO-)",                  "Patient10",
  "DC/mregDC (Epi Adj)",                                      "Patient10"
)

niche_patient_dominance <- purrr::map2_dfr(
  flagged_annotations$celltype_label,
  flagged_annotations$dominant_patient,
  function(label, dom_patient) {
    idx <- which(spe$final_celltype == label)

    if (length(idx) == 0) {
      warning("No cells found for annotation: ", label)
      return(tibble())
    }

    data.frame(
      Niche = spe$Niche[idx],
      Patient = spe$patient_id[idx]
    ) |>
      group_by(Niche) |>
      summarise(
        n_cells = n(),
        pct_dominant_patient = mean(Patient == dom_patient) * 100,
        .groups = "drop"
      ) |>
      mutate(celltype_label = label, dominant_patient = dom_patient)
  }
)
```

---

## 3c. Significant niches: FDR with patient-dominance flag

Lighter companion to 3a/3b -- just Niche and FDR alongside the
patient-dominance summary. Note the percentage here differs from 3b:
`max_patient_share` is the dominant patient's share of *all* cells in
the niche (not just cells of the flagged annotation), so it reflects
how much of the niche's total composition that single patient/
annotation combination actually represents.

```{r significant-niches-table-with-dominance}
# Full long-format diagnostic (within-annotation %, as computed in 3b),
# saved for reference/drill-down
dominance_out_path <- file.path(niche_path, "niche_patient_dominance_full.csv")
write_csv(niche_patient_dominance, dominance_out_path)
cat("Saved full patient-dominance diagnostic to:", dominance_out_path, "\n")

# Total cells per niche -- denominator for the "% of all niche cells" calc
niche_total_cells <- data.frame(Niche = spe$Niche) |>
  group_by(Niche) |>
  summarise(n_total_niche_cells = n(), .groups = "drop") |>
  mutate(Niche = as.character(Niche))

# Recompute dominance as % of all niche cells, then collapse to the
# annotation with the highest share per niche
niche_max_dominance <- niche_patient_dominance |>
  mutate(Niche = as.character(Niche)) |>
  left_join(niche_total_cells, by = "Niche") |>
  mutate(pct_of_niche = n_cells / n_total_niche_cells * 100) |>
  group_by(Niche) |>
  slice_max(pct_of_niche, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(
    Niche,
    n_cells,
    n_total_niche_cells,
    max_share_annotation = celltype_label,
    max_share_patient = dominant_patient,
    max_patient_share = round(pct_of_niche, 1)
  )

# Niche + FDR only (not the full 3a table) -- lowest FDR per niche under
# the Stage model
niche_min_fdr <- sig_niches_table |>
  mutate(Niche = as.character(Niche)) |>
  group_by(Niche) |>
  slice_min(c_FDR, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(Niche, c_FDR)

niche_dominance_summary <- niche_min_fdr |>
  left_join(niche_max_dominance, by = "Niche") |>
  mutate(
    n_cells = replace_na(n_cells, 0),
    n_total_niche_cells = replace_na(n_total_niche_cells, 0),
    max_share_annotation = replace_na(max_share_annotation, "none"),
    max_share_patient = replace_na(max_share_patient, "-"),
    max_patient_share = replace_na(max_patient_share, 0)
  ) |>
  arrange(c_FDR)

niche_dominance_summary |>
  as.data.frame() |>
  knitr::kable(
    digits = 3,
    caption = "Significant niches: FDR with maximum single-patient share of niche cells among flagged annotations",
    booktabs = knitr::is_latex_output(),
    col.names = c("Niche", "FDR", "n cells", "n total\nniche cells",
                  "Max-share annotation", "Max-share patient", "Max patient\nshare (%)")
  ) |>
  {\(kbl) if (knitr::is_latex_output()) {
    kbl |>
      kableExtra::kable_styling(
        latex_options = c("hold_position", "repeat_header"),
        font_size = 9
      ) |>
      kableExtra::column_spec(1, width = "1.2cm") |>
      kableExtra::column_spec(2, width = "1.5cm") |>
      kableExtra::column_spec(3, width = "1.5cm") |>
      kableExtra::column_spec(4, width = "2cm") |>
      kableExtra::column_spec(5, width = "4.5cm") |>
      kableExtra::column_spec(6, width = "3cm") |>
      kableExtra::column_spec(7, width = "2cm")
  } else {
    kbl
  }}()
```

---

## 4. Average neighborhood composition per niche

All 50 niches, saved for reference.

```{r composition}
# spe$aggregatedNeighbors is an S4 DFrame (S4Vectors) -- its as.data.frame
# S4 method does not honor check.names = FALSE the way base
# as.data.frame.matrix does, so names were silently getting mangled here
# even with check.names = FALSE set. Routing through as.matrix() first
# strips the S4 class before any renaming logic can run, and the explicit
# colnames() reassignment guarantees the clean names survive regardless.
raw_neighbors <- spe$aggregatedNeighbors
neighbor_mat  <- as.data.frame(as.matrix(raw_neighbors))
colnames(neighbor_mat) <- colnames(raw_neighbors)
neighbor_mat$Niche <- spe$Niche

niche_composition <- neighbor_mat |>
  group_by(Niche) |>
  summarise(across(everything(), mean), .groups = "drop")

# Sanity check: confirm no mangling crept in before this gets used
# downstream in 4b, 5, 6, and 8.
mangled_check <- grepl("\\.", setdiff(colnames(niche_composition), "Niche"))
cat("niche_composition mangled column check -- any TRUE below means trouble:\n")
print(any(mangled_check))
if (any(mangled_check)) {
  cat("Mangled columns:\n")
  print(colnames(niche_composition)[-1][mangled_check])
} else {
  cat("niche_composition column names look clean.\n")
}

write_csv(niche_composition, composition_out_path)
cat("Saved full niche composition table to:", composition_out_path, "\n")
```

---

## 4b. Detailed composition dot plot (all niches)

Dot size and color both encode mean fraction; niches and cell types
are hierarchically clustered so similar composition profiles sit next
to each other. Long annotation names are wrapped onto multiple lines
rather than truncated, and figure height scales with the number of
cell types shown so labels get fixed vertical spacing instead of
compressing.

```{r composition-dotplot-setup}
niche_mat_full <- niche_composition |> select(-Niche) |> as.matrix()
rownames(niche_mat_full) <- niche_composition$Niche

niche_order    <- rownames(niche_mat_full)[hclust(dist(niche_mat_full))$order]
celltype_order <- colnames(niche_mat_full)[hclust(dist(t(niche_mat_full)))$order]

niche_long <- niche_composition |>
  pivot_longer(-Niche, names_to = "cell_type", values_to = "mean_fraction") |>
  mutate(
    Niche = factor(Niche, levels = niche_order),
    # wrap long cell type names onto multiple lines instead of letting
    # them get clipped or run off the plot area
    cell_type = factor(str_wrap(cell_type, width = 30),
                        levels = str_wrap(celltype_order, width = 30))
  )

# Fixed height per cell type row instead of a flat fig.height -- this is
# the same fix applied to the heatmap in Section 5: a per-row constant
# guarantees labels get consistent room no matter how many rows there are.
n_celltypes_4b <- n_distinct(niche_long$cell_type)
dotplot_height <- max(10, n_celltypes_4b * 0.35)
cat("Cell types:", n_celltypes_4b, "-> computed fig height:", round(dotplot_height, 1), "in\n")
```

```{r composition-dotplot, fig.width=14, fig.height=dotplot_height}
ggplot(niche_long, aes(x = Niche, y = cell_type, size = mean_fraction, color = mean_fraction)) +
  geom_point() +
  scale_color_viridis_c() +
  scale_size_continuous(range = c(0.5, 6)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 7),
    axis.text.y = element_text(size = 8),
    plot.margin = margin(t = 10, r = 20, b = 10, l = 10)
  ) +
  labs(x = "Niche", y = "Cell type", color = "Mean fraction", size = "Mean fraction",
       title = "Average neighborhood composition per niche (all 50 niches)")
```

---

## 5. Heatmap (significant niches only)

Displayed inline only -- not saved to disk.

```{r heatmap}
niche_composition_sig <- niche_composition |>
  filter(Niche %in% sig_niches)

# Drop cell types with near-zero variance across these niches. Column
# z-scoring (scale = "column") divides by each column's SD -- a flat/near-zero
# column has a tiny SD, so a single stray value blows up into an extreme
# z-score. That's what was producing the noisy, extreme-red heatmap.
range_cutoff_sig <- 0.05

celltype_range_sig <- niche_composition_sig |>
  select(-Niche) |>
  summarise(across(everything(), ~ max(.x) - min(.x))) |>
  pivot_longer(everything(), names_to = "cell_type", values_to = "range")

keep_celltypes_sig <- celltype_range_sig |>
  filter(range > range_cutoff_sig) |>
  pull(cell_type)

cat("Keeping", length(keep_celltypes_sig), "of", ncol(niche_composition_sig) - 1,
    "cell types (range >", range_cutoff_sig, ")\n")

niche_composition_sig_filtered <- niche_composition_sig |>
  select(Niche, all_of(keep_celltypes_sig))

niche_mat_sig <- niche_composition_sig_filtered |> select(-Niche) |> as.matrix()
rownames(niche_mat_sig) <- niche_composition_sig_filtered$Niche

# Pre-scale (column z-score) ourselves so we can force symmetric, well-
# contrasted breaks around zero rather than relying on pheatmap's default
# range, which produced the washed-out/inconsistent coloring above.
niche_mat_scaled <- scale(niche_mat_sig)
max_abs <- max(abs(niche_mat_scaled), na.rm = TRUE)
breaks <- seq(-max_abs, max_abs, length.out = 101)
heat_colors <- colorRampPalette(rev(brewer.pal(11, "RdBu")))(100)

# Fixed cell size instead of guessing overall device dimensions from
# string length -- this guarantees every row/column gets the same
# amount of room no matter how many niches or cell types are shown,
# which is what actually prevents condensed/overlapping labels.
cell_w <- 14   # points per column (cell type)
cell_h <- 14   # points per row (niche)

# IMPORTANT: call pheatmap::pheatmap() explicitly with the namespace.
# ComplexHeatmap (loaded elsewhere in this pipeline) also exports a
# function called pheatmap() that wraps its own S4 Heatmap class -- if
# it's loaded after the pheatmap package, it masks pheatmap::pheatmap()
# and silently returns an S4 object instead of the expected list with a
# $gtable slot, causing "$ operator not defined for this S4 class" when
# drawn below. Namespacing the call sidesteps the masking entirely.
#
# angle_col must also be passed as a character string from pheatmap's
# fixed set ("0", "45", "90", "270", "315") -- it's matched internally
# via match.arg(), so a bare numeric 45 fails with "'arg' must be NULL
# or a character vector".
ph <- pheatmap::pheatmap(
  niche_mat_scaled,
  scale = "none",
  color = heat_colors,
  breaks = breaks,
  fontsize_row = 9,
  fontsize_col = 9,
  cellwidth = cell_w,
  cellheight = cell_h,
  labels_col = colnames(niche_mat_scaled),
  border_color = NA,
  angle_col = "45",
  main = "Average neighborhood composition: significant niches only (Stage)",
  silent = TRUE
)

cat("Heatmap grid size:", nrow(niche_mat_scaled), "rows x", ncol(niche_mat_scaled), "cols\n")
cat("Class of ph:", paste(class(ph), collapse = ", "), "\n")
```

```{r heatmap-inline, fig.width=12, fig.height=10}
# Draw directly from the grob rather than relying on print.pheatmap()
# dispatch, which isn't guaranteed to call grid.draw() consistently
# across pheatmap versions -- this is what actually renders the plot.
# Defensive branch: if ph still comes back as an S4 object (e.g. some
# other masking path), fall back to ComplexHeatmap::draw() instead of
# erroring on $gtable.
grid::grid.newpage()

if (isS4(ph)) {
  ComplexHeatmap::draw(ph)
} else {
  grid::grid.draw(ph$gtable)
  grid::grid.text("Z-score", x = 0.98, y = 0.5, rot = 270,
                   gp = grid::gpar(fontsize = 10))
}
```

---

## 6. Top dominant cell types per significant niche

For writing the usual `#Niche N: cell_type1, cell_type2: reasoning`
annotation comments. Long-format table (`niche_top_celltypes`) is still
saved to CSV for reference. The table below collapses each niche to a
single row with its top cell types formatted as `Name (XX%)`; full
names are kept intact (no truncation), with wrapping handled via
`kableExtra` for PDF output.

```{r top-celltypes}
top_n <- 5

niche_top_celltypes <- niche_composition_sig |>
  pivot_longer(-Niche, names_to = "cell_type", values_to = "mean_fraction") |>
  group_by(Niche) |>
  slice_max(mean_fraction, n = top_n) |>
  arrange(Niche, desc(mean_fraction))

# Semicolon separator instead of comma, since annotation names themselves
# contain commas (e.g. "Plasma Cell, TIM3+ (...)"), which made the
# collapsed string ambiguous with a comma separator.
niche_top_celltypes_wide <- niche_top_celltypes |>
  mutate(label = sprintf("%s (%.0f%%)", cell_type, mean_fraction * 100)) |>
  group_by(Niche) |>
  summarise(top_celltypes = paste(label, collapse = "; "), .groups = "drop") |>
  arrange(as.numeric(as.character(Niche)))

# Saving the collapsed/wide version instead of the long-format table --
# this is the one with full, untruncated annotation names per niche,
# which is what's actually useful to read back later. The long-format
# table (niche_top_celltypes) is kept in-memory for the kable below but
# no longer written to CSV.
write_csv(niche_top_celltypes_wide, top_celltypes_out_path)
cat("Saved top-", top_n, " cell types per significant niche to:", top_celltypes_out_path, "\n")

# kable alone doesn't wrap long strings in PDF output -- kableExtra's
# column_spec with an explicit width lets LaTeX wrap the text instead
# of truncating or overflowing the page. HTML wraps naturally, so the
# plain kable() is used there instead.
niche_kable <- niche_top_celltypes_wide |>
  as.data.frame() |>
  knitr::kable(caption = "Top dominant cell types per significant niche (collapsed for annotation)",
               booktabs = knitr::is_latex_output())

if (knitr::is_latex_output()) {
  niche_kable |>
    kableExtra::kable_styling(latex_options = c("hold_position"), font_size = 9) |>
    kableExtra::column_spec(1, width = "2cm") |>
    kableExtra::column_spec(2, width = "16cm")
} else {
  niche_kable
}
```

---

## 7. Auto-drafted annotation comments

Assembles a `#Niche N: ...` comment per significant niche in the
convention used for cluster annotation, combining:

- **Composition** (Section 6) -- the dominant cell type and its %
- **Statistical context** (Section 3a) -- effect direction and FDR
  under the Stage model
- **Patient-dominance flag** (Section 3c) -- whether the dominant
  cell type is one of the six flagged annotations, and how much of
  the niche's total cells that single patient represents

These are drafts, not final annotations -- review each one, especially
any flagged for patient dominance, before committing it to the
pipeline. Note that Benign, IB, and IIA stage contrasts each rest on a
single patient, so any "significant" effect for those specific
contrasts reflects a within-patient ROI comparison rather than a
population-level signal -- flag this explicitly when reviewing drafts
for those stages.

```{r auto-annotation-drafts}
# Effect direction per niche: take the same lowest-FDR row used in 3c
# for consistency, and pull its factor + effect sign.
niche_effect_direction <- sig_niches_table |>
  mutate(Niche = as.character(Niche)) |>
  group_by(Niche) |>
  slice_min(c_FDR, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(
    Niche,
    factor,
    c_FDR,
    direction = ifelse(c_effect > 0, "enriched", "depleted")
  )

# Top cell type + % per niche, from the same data Section 6 uses
niche_top1 <- niche_composition_sig |>
  pivot_longer(-Niche, names_to = "cell_type", values_to = "mean_fraction") |>
  group_by(Niche) |>
  slice_max(mean_fraction, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(
    Niche = as.character(Niche),
    top_celltype = cell_type,
    top_pct = round(mean_fraction * 100)
  )

# Assemble the draft comment per niche
niche_annotation_drafts <- niche_effect_direction |>
  left_join(niche_top1, by = "Niche") |>
  left_join(
    niche_dominance_summary |> select(Niche, max_share_annotation, max_share_patient, max_patient_share),
    by = "Niche"
  ) |>
  mutate(
    is_flagged = !is.na(max_share_annotation) & max_share_annotation == top_celltype,
    draft_comment = case_when(
      is_flagged ~ sprintf(
        "#Niche %s: %s (%d%%): %s in %s (FDR = %.3f) -- CAUTION: dominant cell type is patient-dominance flagged, %s accounts for %.1f%% of this niche's total cells. Review before finalizing.",
        Niche, top_celltype, top_pct, direction, factor, c_FDR, max_share_patient, max_patient_share
      ),
      TRUE ~ sprintf(
        "#Niche %s: %s (%d%%): %s in %s (FDR = %.3f)",
        Niche, top_celltype, top_pct, direction, factor, c_FDR
      )
    )
  ) |>
  arrange(as.numeric(Niche))

annotation_drafts_out_path <- file.path(niche_path, "niche_annotation_drafts.txt")
writeLines(niche_annotation_drafts$draft_comment, annotation_drafts_out_path)
cat("Saved draft annotation comments to:", annotation_drafts_out_path, "\n\n")

cat(paste(niche_annotation_drafts$draft_comment, collapse = "\n\n"))
```

---

## 8. Stacked bar plot: cell type composition per significant niche

Color encodes cell type annotation; bar segment size (within each
niche's stacked column) encodes the mean proportion of that cell type
in that niche. Uses the unfiltered `niche_composition_sig` (not the
range-filtered version from Section 5), so each niche's stack sums to
~100%.

---

## 8a. Stacked bar plot: top 5 cell types per niche (collapsed to "Other")

Same stacked-by-niche layout as before, but each niche's bar only
breaks out its own top 5 cell types individually; everything else is
collapsed into a single "Other" segment. Kept cell types are the
*union* of each niche's top 5 (so the legend is still fewer than the
full ~38 annotations, but a type that's top-5 in one niche keeps its
own color everywhere it appears). Stack order and legend order are
fixed by overall abundance (most abundant first) so segments line up
consistently across niches, with "Other" always pinned to the bottom.
Stack still sums to ~100% per niche.

Colors come from Kelly's 22 colors of maximum contrast (public-domain
reference palette; white/black dropped since they clash with the plot
background/gridlines) rather than an HCL hue sweep -- varying hue
*and* chroma/luminance together gives much better separation between
adjacent legend entries than evenly-spaced hues at fixed chroma/
luminance, which still put similarly-toned colors next to each other.
Falls back to hue-swept HCL only if `n_kept` exceeds the 20-color pool.

```{r stacked-bar-top5, fig.width=14, fig.height=8}
top_n_stack <- 5

# Global top 5 cell types by overall mean abundance across significant
# niches (not a per-niche union) -- this caps the legend at
# top_n_stack + 1 ("Other") entries total, rather than the ~20 that
# resulted from unioning each niche's own top 5. Every niche's bar is
# still built from the same fixed 5 categories, so segment colors are
# directly comparable niche-to-niche, and "Other" now absorbs a niche's
# own dominant cell type whenever that type isn't in the global top 5.
keep_celltypes_top5 <- niche_composition_sig |>
  pivot_longer(-Niche, names_to = "cell_type", values_to = "mean_fraction") |>
  group_by(cell_type) |>
  summarise(overall_mean = mean(mean_fraction), .groups = "drop") |>
  slice_max(overall_mean, n = top_n_stack, with_ties = FALSE) |>
  arrange(desc(overall_mean)) |>
  pull(cell_type)

celltype_rank_top5 <- keep_celltypes_top5  # already ranked by overall_mean above

celltype_levels_top5 <- c(celltype_rank_top5, "Other")

niche_long_sig_top5 <- niche_composition_sig |>
  pivot_longer(-Niche, names_to = "cell_type", values_to = "mean_fraction") |>
  mutate(cell_type = ifelse(cell_type %in% keep_celltypes_top5, cell_type, "Other")) |>
  group_by(Niche, cell_type) |>
  summarise(mean_fraction = sum(mean_fraction), .groups = "drop") |>
  mutate(
    cell_type = factor(cell_type, levels = celltype_levels_top5),
    Niche = factor(as.character(Niche),
                    levels = as.character(sort(as.numeric(as.character(sig_niches)))))
  )

cat("Cell types kept (global top", top_n_stack, "by overall abundance):",
    length(keep_celltypes_top5), "+ Other\n")
print(keep_celltypes_top5)

# Kelly's colors of maximum contrast (public domain reference palette),
# minus white/black since those clash with the plot background and grid
# lines. With only top_n_stack (5) categories now, this comfortably fits
# within the pool -- no HCL fallback needed at this scale, but kept for
# robustness if top_n_stack is increased later.
kelly_colors <- c(
  "#F3C300", "#875692", "#F38400", "#A1CAF1", "#BE0032",
  "#C2B280", "#848482", "#008856", "#E68FAC", "#0067A5",
  "#F99379", "#604E97", "#F6A600", "#B3446C", "#DCD300",
  "#882D17", "#8DB600", "#654522", "#E25822", "#2B3D26"
)

n_kept <- length(celltype_rank_top5)

if (n_kept <= length(kelly_colors)) {
  celltype_colors_top5 <- kelly_colors[seq_len(n_kept)]
} else {
  n_extra <- n_kept - length(kelly_colors)
  hues <- seq(15, 375, length.out = n_extra + 1)[seq_len(n_extra)]
  extra_colors <- grDevices::hcl(h = hues, c = 90, l = 45)
  celltype_colors_top5 <- c(kelly_colors, extra_colors)
}

names(celltype_colors_top5) <- celltype_rank_top5
celltype_colors_top5["Other"] <- "grey85"


ggplot(niche_long_sig_top5, aes(x = Niche, y = mean_fraction, fill = cell_type)) +
  geom_col(position = "stack", color = "white", linewidth = 0.15) +
  scale_fill_manual(values = celltype_colors_top5, breaks = celltype_levels_top5) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                      expand = expansion(mult = c(0, 0.02))) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 8),
    legend.position = "right",
    legend.key.size = unit(0.35, "cm"),
    legend.text = element_text(size = 8),
    panel.grid.major.x = element_blank(),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
  ) +
  guides(fill = guide_legend(ncol = 1)) +
  labs(x = "Niche", y = "Proportion of cells", fill = "Cell type",
       title = "Cell type composition per significant niche (Stage)",
       subtitle = sprintf("Top %d cell types overall shown individually; remainder collapsed to Other", top_n_stack))
```

---

## 8a-facet. Faceted small multiples: top 5 cell types per niche

Same underlying data as 8a (`niche_long_sig_top5`), but laid out as
one horizontal bar panel per niche instead of one stacked column per
niche. Every bar starts at a shared zero baseline, so comparing
proportions within a niche -- and roughly across niches -- is a length
comparison rather than a segment-position comparison, which is what
made 8a hard to read once a niche's top cell type wasn't the first
segment in the stack.

Bars are sorted within each panel (highest proportion at top) via
`tidytext::reorder_within()`; cell type names are wrapped at 25
characters since they now appear directly as axis text in every
panel rather than only once in a shared legend, and long strings
(e.g. "Plasma Cell, TIM3+ (CD138+TIM3+CD45RO-)") would otherwise
overflow narrow facet columns.

Trade-off vs. 8a: no shared legend and no "Other" segment, since each
panel is self-contained and every one of its top-5 bars is already
labeled by name -- there is no unlabeled remainder to fold in the way
a stacked column has room for. If tracking one specific cell type's
presence across niches at a glance matters more than within-niche
readability, the dot plot in Section 4b (filtered to `sig_niches`)
does that job better; this plot and 8a's stacked version aren't built
for it.

```{r stacked-bar-top5-facet-setup}
niche_long_sig_top5_facet <- niche_long_sig_top5 |>
  mutate(cell_type_wrapped = str_wrap(as.character(cell_type), width = 25))
```

```{r stacked-bar-top5-facet, fig.width=14, fig.height=12}
ggplot(niche_long_sig_top5_facet,
       aes(x = reorder_within(cell_type_wrapped, mean_fraction, Niche),
           y = mean_fraction, fill = cell_type)) +
  geom_col(color = "white", linewidth = 0.1, show.legend = FALSE) +
  scale_fill_manual(values = celltype_colors_top5) +
  scale_x_reordered() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  coord_flip() +
  facet_wrap(~ Niche, scales = "free_y", ncol = 5) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 6),
    strip.text = element_text(size = 8, face = "bold"),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
  ) +
  labs(x = NULL, y = "Proportion of cells",
       title = "Cell type composition per significant niche (Stage)",
       subtitle = sprintf("Top %d cell types per niche shown; sorted within each panel", top_n_stack))
```

---

## 8b. Bar plot: top 10 cell types overall, grouped by niche

Instead of stacking every cell type per niche, this restricts to the
10 annotations with the highest peak proportion in any significant
niche, then shows each one's proportion across niches as grouped bars.
Complementary to 8a: 8a shows what each niche is made of, 8b shows how
a given annotation's presence varies across niches.

Niche colors also draw from the Kelly palette (same rationale as 8a)
instead of `hcl.colors(..., "Dark 3")`, since niches are the encoded
category here and benefit from the same separation. Facet fallback
(`use_facet`) still applies when there are too many niches to dodge
readably, in which case per-niche color doesn't matter since each
facet panel is colored by cell type instead.

```{r stacked-bar-top10annot-setup}
top_n_annot <- 10

# Rank cell types by their single highest proportion in any significant niche
top_celltypes_overall <- niche_composition_sig |>
  pivot_longer(-Niche, names_to = "cell_type", values_to = "mean_fraction") |>
  group_by(cell_type) |>
  summarise(max_fraction = max(mean_fraction), .groups = "drop") |>
  slice_max(max_fraction, n = top_n_annot, with_ties = FALSE) |>
  pull(cell_type)

niche_long_sig_top10annot <- niche_composition_sig |>
  pivot_longer(-Niche, names_to = "cell_type", values_to = "mean_fraction") |>
  filter(cell_type %in% top_celltypes_overall) |>
  mutate(
    cell_type = factor(cell_type, levels = rev(top_celltypes_overall)),
    Niche = factor(as.character(Niche),
                    levels = as.character(sort(as.numeric(as.character(sig_niches)))))
  )

n_niches_bar <- n_distinct(niche_long_sig_top10annot$Niche)
cat("Significant niches shown:", n_niches_bar, "\n")

# Same Kelly's colors of maximum contrast used in 8a -- reused here so
# both figures share one distinct-color source instead of two different
# palette strategies.
kelly_colors <- c(
  "#F3C300", "#875692", "#F38400", "#A1CAF1", "#BE0032",
  "#C2B280", "#848482", "#008856", "#E68FAC", "#0067A5",
  "#F99379", "#604E97", "#F6A600", "#B3446C", "#DCD300",
  "#882D17", "#8DB600", "#654522", "#E25822", "#2B3D26"
)

if (n_niches_bar <= length(kelly_colors)) {
  niche_colors <- kelly_colors[seq_len(n_niches_bar)]
} else {
  n_extra <- n_niches_bar - length(kelly_colors)
  hues <- seq(15, 375, length.out = n_extra + 1)[seq_len(n_extra)]
  niche_colors <- c(kelly_colors, grDevices::hcl(h = hues, c = 90, l = 45))
}
names(niche_colors) <- levels(niche_long_sig_top10annot$Niche)

# If there are too many significant niches for dodge to stay readable,
# fall back to faceting by niche instead of coloring by it.
use_facet <- n_niches_bar > 15
cat("Using facet layout instead of dodge:", use_facet, "\n")
```

```{r stacked-bar-top10annot, fig.width=14, fig.height=8}
if (!use_facet) {
  ggplot(niche_long_sig_top10annot, aes(x = cell_type, y = mean_fraction, fill = Niche)) +
    geom_col(position = "dodge", color = "white", linewidth = 0.1) +
    scale_fill_manual(values = niche_colors) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    coord_flip() +
    theme_minimal() +
    theme(
      legend.position = "right",
      legend.key.size = unit(0.35, "cm"),
      legend.text = element_text(size = 7),
      plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
    ) +
    guides(fill = guide_legend(ncol = 1)) +
    labs(x = "Cell type", y = "Proportion of cells", fill = "Niche",
         title = sprintf("Top %d cell type annotations across significant niches (Stage)", top_n_annot),
         subtitle = "Proportion of each annotation within each significant niche")
} else {
  ggplot(niche_long_sig_top10annot, aes(x = cell_type, y = mean_fraction, fill = cell_type)) +
    geom_col(color = "white", linewidth = 0.1, show.legend = FALSE) +
    scale_fill_manual(values = setNames(
      if (top_n_annot <= length(kelly_colors)) kelly_colors[seq_len(top_n_annot)]
      else c(kelly_colors, grDevices::hcl.colors(top_n_annot - length(kelly_colors), palette = "Dark 3")),
      levels(niche_long_sig_top10annot$cell_type)
    )) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    coord_flip() +
    facet_wrap(~ Niche, ncol = 5) +
    theme_minimal() +
    theme(
      axis.text.y = element_text(size = 6),
      strip.text = element_text(size = 8),
      plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
    ) +
    labs(x = "Cell type", y = "Proportion of cells",
         title = sprintf("Top %d cell type annotations across significant niches (Stage)", top_n_annot),
         subtitle = "Faceted by niche -- proportion of each annotation within each niche")
}
```