---
title: "Niche Generation and Differential Abundance"
author: "Tazein Shah"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)

suppressPackageStartupMessages({
  library(imcRtools)
  library(tidyverse)
  library(SingleCellExperiment)
  library(SpatialExperiment)
  library(sccomp)
  library(ggrepel)
})

set.seed(12345)
```

---

## 1. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/spatial_analysis/run_3"
spatial_graph_path <- file.path(analysis_path, "spe_spatial_graph_r3.rds")

spe_graph <- readRDS(spatial_graph_path)

cat("Cells:", ncol(spe_graph), "\n")
cat("Levels in spatial graph object:\n")
print(levels(spe_graph$final_celltype))
```

```{r niche-out-paths}
niche_path <- file.path(analysis_path, "niche_k50_thresh20")
dir.create(niche_path, recursive = TRUE, showWarnings = FALSE)

niche_spe_out_path      <- file.path(niche_path, "spe_niche.rds")
sccomp_nstatus_out_path <- file.path(niche_path, "sccomp_nstatus.rds")
sccomp_stage_out_path   <- file.path(niche_path, "sccomp_stage.rds")
```

---

## 2. Load spatial graph object

`spe_spatial_graph_r3.rds` already carries `final_celltype` built
directly on the r2 annotation set (confirmed: labels match r2's plain
cluster names, e.g. `Macrophage, Proliferating (MKI67+)`, with no
diagnostic-suffix renames from the earlier annotation round), along
with the `expansion_interaction_graph` colPair already built at
`threshold = 20`. No merge step is needed for this run.

```{r load}
spe <- spe_graph

cat("Cells:", ncol(spe), "\n")
cat("Cell types:", nlevels(spe$final_celltype), "\n")
print(table(spe$final_celltype))
```

---

## 3. Aggregate neighbor composition

```{r aggregate-neighbors}
# Reusing the existing graph (threshold = 20) rather than rebuilding it.
# See Overview note above re: 20 vs. 50 discrepancy.
spe <- aggregateNeighbors(spe,
                          colPairName = "expansion_interaction_graph",
                          aggregate_by = "metadata",
                          count_by = "final_celltype")
```

---

## 4. K-means niche clustering

```{r kmeans-niches}
kmeansClust <- 50

cn_1 <- kmeans(spe$aggregatedNeighbors, centers = kmeansClust, iter.max = 10000)
spe$Niche <- as.factor(cn_1$cluster)

table(spe$Niche)
```

---

## 5. Patient-level covariates

```{r covariates}
spe$stage_group <- sapply(spe$patient_id, function(x) {
  if (x %in% c("Patient12", "Patient15")) return("IIB")
  else if (x == "Patient13") return("IIIA")
  else if (x == "Patient19") return("IB")
  else if (x == "Patient11" || grepl("^Patient20", x)) return("IA3")
  else if (x %in% c("Patient14", "Patient16", "Patient17", "Patient18")) return("IA2")
  else if (x == "Patient10") return("Benign")
  else return(NA)
})

spe$stage_group <- factor(spe$stage_group,
                          levels = c("Benign", "IA2", "IA3", "IB", "IIIA", "IIB"))

# Three-tier disease state grouping (Benign / Lower / Higher)
spe$disease_state <- sapply(spe$patient_id, function(x) {
  if (x == "Patient10") return("Benign")
  else if (x %in% c("Patient11", "Patient14", "Patient16", "Patient17", "Patient18") || grepl("^Patient20", x)) return("Lower")
  else if (x %in% c("Patient12", "Patient13", "Patient15", "Patient19")) return("Higher")
  else return(NA)
})

spe$disease_state <- factor(spe$disease_state, levels = c("Benign", "Lower", "Higher"))

# N-status: only IIB (P12, P15) is N1; everything else is N0
# NOTE: flagging for confirmation -- does IIIA (Patient13) carry N1 status?
# AJCC IIIA often implies nodal involvement, but this cohort's original
# stage list only assigned N1 possibility to IIB. Left as-is pending confirmation.
spe$N_status <- factor(ifelse(spe$stage_group == "IIB", "N1", "N0"),
                       levels = c("N0", "N1"))

table(spe$patient_id, spe$N_status)
table(spe$patient_id, spe$stage_group)
table(spe$patient_id, spe$disease_state)
```

### 5b. Patient-driven clusters, per niche (data prep)

Several clusters show a single patient accounting for a large share
of cells cohort-wide (flagged from cohort-wide composition review).
Each of these needs a per-niche check, since any of them could
confound niche-level stage/N-status associations if a niche happens
to be enriched for that cluster. This chunk only builds the per-niche
composition table (`niche_patient_driven`) -- the actual high-risk
summary is deferred to Section 9c, after the sccomp models are fit,
so that it can be restricted to niches that are *actually* significant
(FDR < 0.1) rather than flagging niches that were never going to be
reported anyway.

```{r patient-driven-clusters-check}
# Cluster -> dominant patient pairs flagged from cohort-wide composition review
patient_driven_clusters <- tibble::tribble(
  ~cluster,                                                ~dominant_patient,
  "Monocyte (PD1+)",                                       "Patient10",
  "Macrophage, Proliferating (MKI67+)",                    "Patient10",
  "Tumor Epithelial (Neuroendocrine)",                     "Patient17",
  "B Cell (CD19+)",                                        "Patient20_519",
  "Plasma Cell, TIM3+ (CD138+TIM3+CD45RO-)",               "Patient10",
  "DC/mregDC (Epi Adj)",                                   "Patient10"
)

# Per-niche share of the dominant patient, for each flagged cluster
# (full table -- used downstream by 5c, 7b/9b plots, and the Section
# 9c high-risk summary; not printed in full here)
niche_patient_driven <- purrr::pmap_dfr(
  patient_driven_clusters,
  function(cluster, dominant_patient) {
    idx <- which(spe$final_celltype == cluster)
    data.frame(
      Niche = spe$Niche[idx],
      Patient = spe$patient_id[idx]
    ) |>
      group_by(Niche) |>
      summarise(
        n_cells = n(),
        pct_dominant_patient = mean(Patient == dominant_patient) * 100,
        .groups = "drop"
      ) |>
      mutate(cluster = cluster, dominant_patient = dominant_patient)
  }
)
```

### 5c. Plot: dominant-patient share across niches, by cluster

Broad diagnostic view across all six flagged clusters at once,
faceted by cluster, so any niche at risk from *any* of these
confounds is visible in one plot.

```{r patient-driven-plot, fig.width = 12, fig.height = 8}
ggplot(niche_patient_driven, aes(x = factor(Niche, levels = sort(unique(as.numeric(as.character(Niche))))),
                                  y = pct_dominant_patient)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~ paste0(cluster, "\n(", dominant_patient, ")"), scales = "free_x", ncol = 2) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
  labs(x = "Niche", y = "% of cluster's cells in niche from dominant patient",
       title = "Patient-driven cluster composition across niches")
```

```{r save-niche-spe}
saveRDS(spe, niche_spe_out_path)
cat("Saved object with niches to:", niche_spe_out_path, "\n")
```

---

## 6. Differential niche abundance (sccomp)

`.sample = Patient`, since N-status and stage are constant within
patient (not within ROI) -- mirroring how the original script's `LN`
is the invariant unit for its `TumorStage`/`HighN1` covariates.

```{r sccomp-prep}
sceIMC <- SingleCellExperiment(list(counts = assay(spe, "counts")))
sceIMC$Patient   <- factor(spe$patient_id)
sceIMC$Sample    <- factor(spe$sample_id)
sceIMC$Niche     <- factor(spe$Niche)
sceIMC$N_status  <- factor(spe$N_status, levels = c("N0", "N1"))
sceIMC$Stage     <- factor(spe$stage_group,
                           levels = c("Benign", "IA2", "IA3", "IB", "IIIA", "IIB"))
```

### N-status model (parallel to `HighN1` in the original script)

```{r sccomp-nstatus}
if (file.exists(sccomp_nstatus_out_path)) {
  cat("Found existing output, loading:", sccomp_nstatus_out_path, "\n")
  sccomp_niche_nstatus <- readRDS(sccomp_nstatus_out_path)
} else {
  sccomp_niche_nstatus <-
    sceIMC |>
    sccomp_estimate(
      formula_composition = ~ N_status + (1 | Patient),
      .sample = Sample,
      .cell_group = Niche,
      bimodal_mean_variability_association = FALSE,
      variational_inference = FALSE,
      output_directory = file.path(niche_path, "sccomp_draws_nstatus"),
      cores = 8
    ) |>
    sccomp_test()

  saveRDS(sccomp_niche_nstatus, sccomp_nstatus_out_path)
}
```

### Stage model (parallel to `TumorStage` in the original script)

```{r sccomp-stage}
if (file.exists(sccomp_stage_out_path)) {
  cat("Found existing output, loading:", sccomp_stage_out_path, "\n")
  sccomp_niche_stage <- readRDS(sccomp_stage_out_path)
} else {
  sccomp_niche_stage <-
    sceIMC |>
    sccomp_estimate(
      formula_composition = ~ Stage + (1 | Patient),
      .sample = Sample,
      .cell_group = Niche,
      bimodal_mean_variability_association = FALSE,
      variational_inference = FALSE,
      output_directory = file.path(niche_path, "sccomp_draws_stage"),
      cores = 8
    ) |>
    sccomp_test()

  saveRDS(sccomp_niche_stage, sccomp_stage_out_path)
}
```

---

## 7. Results table

```{r results-table}
sccomp_niche_nstatus <- readRDS(sccomp_nstatus_out_path)
sccomp_niche_nstatus_sub <- sccomp_niche_nstatus[which(sccomp_niche_nstatus$factor == "N_status"), ] |>
  as.data.frame() |>
  mutate(Niche = as.character(Niche))

sccomp_niche_nstatus_sub |>
  filter(c_FDR < 0.1) |>
  arrange(c_FDR) |>
  select(Niche, parameter, c_effect, c_FDR) |>
  knitr::kable(digits = 3, caption = "Niche composition association with N-status (FDR < 0.1)")
```

### 7b. Plot: patient-driven cluster risk for significant niches

Table of the niches significant here (FDR < 0.05), then a plot of
patient-driven cluster shares restricted to those niches -- any niche
flagged with a high dominant-patient share for one of the 6 flagged
clusters should be read as a possible artifact rather than a genuine
N-status association.

```{r sig-niches-table-nstatus}
sig_niches_nstatus_tbl <- sccomp_niche_nstatus_sub |>
  filter(c_FDR < 0.05)

sig_niches_nstatus_tbl |>
  arrange(c_FDR) |>
  select(Niche, parameter, c_effect, c_FDR) |>
  knitr::kable(digits = 3, caption = "Niches significant for N-status (FDR < 0.05)")
```

```{r patient-risk-plot-nstatus, fig.width = 10, fig.height = 6}
risk_plot_df_nstatus <- niche_patient_driven |>
  mutate(Niche = as.character(Niche)) |>
  filter(Niche %in% sig_niches_nstatus_tbl$Niche)

ggplot(risk_plot_df_nstatus, aes(x = cluster, y = pct_dominant_patient, fill = dominant_patient)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "grey40") +
  facet_wrap(~ Niche, nrow = 2) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        strip.text = element_text(size = 9)) +
  labs(x = NULL,
       y = "% of cluster's cells in niche from dominant patient",
       fill = "Dominant patient",
       title = "Patient-driven cluster risk among N-status-significant niches (FDR < 0.05)")
```

---

## 8. Volcano plot

```{r volcano}
plot_df <- sccomp_niche_nstatus_sub |>
  as.data.frame() |>
  filter(!is.nan(c_FDR), c_FDR < 0.1) |>
  mutate(
    LogFDR = -log(c_FDR),
    Significant = ifelse(c_FDR < 0.05, "FDR < 0.05", "Not significant")
  )

ggplot(plot_df, aes(x = c_effect, y = LogFDR, color = Significant)) +
  geom_point(size = 2) +
  geom_text_repel(
    aes(label = Niche), size = 3, show.legend = FALSE
  ) +
  scale_color_manual(values = c("FDR < 0.05" = "firebrick", "Not significant" = "grey70")) +
  theme_classic() +
  labs(x = "Effect size", y = "-log(FDR)",
       title = "Niche composition vs. N-status")
```

---

## 9. Results table (Stage)

`factor == "Stage"` covers all 5 non-reference contrasts (each vs.
`Benign`) at once, distinguished by `parameter`
(`StageIA2`, `StageIA3`, `StageIB`, `StageIIIA`, `StageIIB`) --
250 rows total (50 niches x 5 contrasts).

```{r results-table-stage}
sccomp_niche_stage <- readRDS(sccomp_stage_out_path)
sccomp_niche_stage_sub <- sccomp_niche_stage[which(sccomp_niche_stage$factor == "Stage"), ] |>
  as.data.frame() |>
  mutate(Niche = as.character(Niche))

sccomp_niche_stage_sub |>
  filter(c_FDR < 0.1) |>
  arrange(c_FDR) |>
  select(Niche, parameter, c_effect, c_FDR) |>
  knitr::kable(digits = 3, caption = "Niche composition association with Stage (vs. Benign, FDR < 0.1)")
```

### 9b. Plot: patient-driven cluster risk for significant niches (Stage)

Same approach as 7b, applied to the Stage-significant niches.

```{r sig-niches-table-stage}
sig_niches_stage_tbl <- sccomp_niche_stage_sub |>
  filter(c_FDR < 0.05)

sig_niches_stage_tbl |>
  arrange(c_FDR) |>
  select(Niche, parameter, c_effect, c_FDR) |>
  knitr::kable(digits = 3, caption = "Niches significant for Stage (FDR < 0.05)")
```

```{r patient-risk-plot-stage, fig.width = 10, fig.height = 6}
risk_plot_df_stage <- niche_patient_driven |>
  mutate(Niche = as.character(Niche)) |>
  filter(Niche %in% sig_niches_stage_tbl$Niche)

ggplot(risk_plot_df_stage, aes(x = cluster, y = pct_dominant_patient, fill = dominant_patient)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "grey40") +
  facet_wrap(~ Niche, nrow = 2) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        strip.text = element_text(size = 9)) +
  labs(x = NULL,
       y = "% of cluster's cells in niche from dominant patient",
       fill = "Dominant patient",
       title = "Patient-driven cluster risk among Stage-significant niches (FDR < 0.05)")
```

### 9c. High-risk niche x cluster combinations, restricted to FDR < 0.1 niches

Only niches significant (FDR < 0.1) in *either* the N-status or Stage
model are considered here -- this is the definitive high-risk summary
for the notebook, computed after both sccomp models so that only
niches actually being reported are flagged, rather than every niche
that happens to have a lopsided patient composition regardless of
whether it's statistically significant.

```{r sig-niches-fdr01}
sig_niches_fdr01 <- union(
  sccomp_niche_nstatus_sub |> filter(c_FDR < 0.1) |> pull(Niche),
  sccomp_niche_stage_sub   |> filter(c_FDR < 0.1) |> pull(Niche)
)

cat("Niches significant (FDR < 0.1) in N-status and/or Stage model:", length(sig_niches_fdr01), "\n")
```

```{r high-risk-table}
n_show <- 15
risk_threshold_pct <- 50
risk_threshold_n    <- 5

high_risk_tbl <- niche_patient_driven |>
  mutate(Niche = as.character(Niche)) |>
  filter(pct_dominant_patient >= risk_threshold_pct,
         n_cells >= risk_threshold_n,
         Niche %in% sig_niches_fdr01) |>
  arrange(desc(pct_dominant_patient))

high_risk_tbl |>
  slice_head(n = n_show) |>
  select(Niche, Cluster = cluster, Patient = dominant_patient,
         `% Dominant` = pct_dominant_patient, `N Cells` = n_cells) |>
  knitr::kable(digits = 1, caption = paste0("Top ", n_show, " highest-risk niche x cluster combinations among FDR < 0.1 niches (of ",
                                             nrow(high_risk_tbl), " total meeting threshold)"))
```

---

## 10. Volcano plot (Stage) {.tabset}

Faceted by contrast, since each stage level vs. Benign is a separate
comparison rather than a single binary effect.

```{r volcano-stage, results = "asis", fig.width = 7, fig.height = 5.5}
plot_df_stage <- sccomp_niche_stage_sub |>
  as.data.frame() |>
  filter(!is.nan(c_FDR), c_FDR < 0.1) |>
  mutate(
    LogFDR = -log(c_FDR),
    Significant = ifelse(c_FDR < 0.05, "FDR < 0.05", "Not significant")
  )

for (contrast in unique(plot_df_stage$parameter)) {
  cat("\n\n### ", contrast, "\n\n")
  df_sub <- plot_df_stage[plot_df_stage$parameter == contrast, ]
  print(
    ggplot(df_sub, aes(x = c_effect, y = LogFDR, color = Significant)) +
      geom_point(size = 2) +
      geom_text_repel(aes(label = Niche), size = 3, show.legend = FALSE) +
      scale_color_manual(values = c("FDR < 0.05" = "firebrick", "Not significant" = "grey70")) +
      theme_classic() +
      labs(x = "Effect size", y = "-log(FDR)",
           title = paste0("Niche composition: ", contrast, " vs. Benign"))
  )
  cat("\n\n")
}
```

---

## 11. Boxplots per niche {.tabset}

```{r boxplots, results = "asis", fig.width = 5, fig.height = 5.5}
# Only plot niches significant at FDR < 0.05 in either the N-status or
# Stage model -- matching the "FDR < 0.05" red points on the volcano
# plots (Sections 8 and 10), rather than the wider FDR < 0.1 threshold
# used just to populate the results tables.
sig_niches_nstatus <- sccomp_niche_nstatus_sub |>
  filter(c_FDR < 0.05) |>
  pull(Niche) |>
  as.character()

sig_niches_stage <- sccomp_niche_stage_sub |>
  filter(c_FDR < 0.05) |>
  pull(Niche) |>
  as.character()

sig_niches <- union(sig_niches_nstatus, sig_niches_stage)
sig_niches <- as.character(sort(as.numeric(sig_niches)))

cat("Niches significant (FDR < 0.05) in N-status model:", length(sig_niches_nstatus), "\n")
cat("Niches significant (FDR < 0.05) in Stage model:", length(sig_niches_stage), "\n")
cat("Total unique niches plotted below:", length(sig_niches), "\n")

niche_prop <- table(spe$sample_id, spe$Niche) |>
  prop.table(margin = 1) |>
  as.data.frame()
colnames(niche_prop) <- c("Sample", "Niche", "Value")

niche_prop$Patient <- spe$patient_id[match(niche_prop$Sample, spe$sample_id)]
niche_prop$Stage <- spe$stage_group[match(niche_prop$Patient, spe$patient_id)]

for (n in sig_niches) {
  cat("\n\n### Niche", n, "\n\n")
  df_sub <- niche_prop[niche_prop$Niche == n, ]
  df_sub$Value <- df_sub$Value * 100
  print(
    ggplot(df_sub, aes(x = Stage, y = Value)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(aes(color = Patient), width = 0.1, size = 1.5) +
      theme_classic() +
      ylab("% within sample") +
      ggtitle(paste0("Niche ", n))
  )
  cat("\n\n")
}
```