---
title: "Spatial Graph (Run 3)"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

Builds the cell-cell spatial graph from the current finalized annotated
dataset (`spe_final_celltype.rds`, T-cell-subclustering update), for
handoff to niche analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# 1. Load Libraries

```{r libraries}
#.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(imcRtools)
library(tidyverse)
library(SingleCellExperiment)
library(SpatialExperiment)
library(scater)
```

# 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

final_celltype_path <- file.path(analysis_path, "final_cell_annotation", "r2_annotation", "spe_final_celltype_r2.rds")

spatial_analysis_path <- file.path(analysis_path, "spatial_analysis", "run_3")

spatial_graph_out_path <- file.path(spatial_analysis_path, "spe_spatial_graph_r3.rds")
```

# 3. Load Finalized Annotated Object

```{r load-spe}
spe <- readRDS(final_celltype_path)

cat("Cells loaded:", ncol(spe), "\n")
cat("Cell types:", nlevels(spe$final_celltype), "\n")
cat("Samples/ROIs:", length(unique(spe$sample_id)), "\n")
```

# 4. Build Spatial Graph

Using `type = "expansion"` and `threshold = 20`, matching the convention
used throughout this project (pending final confirmation from Campbell
given the 20 vs. 50 discrepancy between the reference manuscript text and
its code).

```{r build-spatial-graph}
t0 <- Sys.time()
spe <- buildSpatialGraph(
  spe,
  img_id    = "sample_id",
  type      = "expansion",
  threshold = 20,
  name      = "expansion_interaction_graph"
)
t1 <- Sys.time()
cat("Spatial graph built on full dataset in",
    round(as.numeric(t1 - t0, units = "mins"), 2), "minutes\n")

colPairNames(spe)
graph_name <- "expansion_interaction_graph"
```

# 5. Graph Diagnostics

```{r graph-diagnostics}
graph_obj <- colPair(spe, graph_name)
cat("Total edges in graph:", length(graph_obj), "\n")
cat("Mean neighbors per cell:", round(length(graph_obj) / ncol(spe), 2), "\n")
```

# 6. Save Spatial Graph Object

```{r save-spatial-graph}
saveRDS(spe, spatial_graph_out_path)
cat("Saved spatial graph object to:", spatial_graph_out_path, "\n")
```

# Handoff Notes

- Saved to: `spatial_analysis/spe_spatial_graph_r3.rds`
- Built from `spe_final_celltype.rds` (T-cell-subclustering-updated
  annotation, ~39 final cell types, cluster 17 and cluster 15
  low-quality/artifact cells excluded).
- Graph: `type = "expansion"`, `threshold = 20`, stored as
  `"expansion_interaction_graph"` in `colPairNames(spe)`.
- This file includes both the graph and the full `final_celltype`
  annotation -- ready to hand directly to niche analysis
  (`aggregateNeighbors()` + clustering) without needing any further
  merging or relabeling.
