---
title: 'Cluster Resolution Comparison: k=20 vs. k=150 and k=200'
author: "Vaidehi Gupta"
date: "07-07-2026"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
subtitle: Cross-tabulating fine (k=20) clusters against coarser (k=150, k=200) resolutions
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 8)
set.seed(42)
```

## Overview

The 20-cluster (k=20) run is the finest, fully annotated resolution.
The k=150 (12 clusters) and k=200 (11 clusters) runs are much coarser
partitions of the *same underlying cells*. Cross-tabulating which
k=20 clusters collapse into which coarse cluster tells us:

- **If two fine clusters merge into the same coarse cluster** -> they
  likely represent real biology (e.g. activation
  states of one lineage), not two separate cell types.
- **If a fine cluster stays isolated as its own coarse cluster even at k=200** -> it's a genuinely distinct population; any co-expression weirdness inside it (e.g. Cluster 4's LAMP3/CD117 bimodality) is a *local* contamination problem, not a resolution problem, and needs targeted sub-clustering to resolve.
- **If a small/weak fine cluster gets absorbed entirely into a bigger neighbor** at coarser resolution -> it's a borderline population worth extra scrutiny before treating its identity as settled.

This is a per-cell comparison, not a per-cluster-label comparison, so
cells are matched by their unique cell ID/barcode across the three
`.rds` objects before any cross-tab is built.

Rows: which of the 20 clusters a cell belongs to
Columns: which of the 12 (or 11) coarse clusters that same cell belongs to
Each cell: how many cells share that particular combination

---

## 1. Load Libraries

```{r libraries}
.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(tidyverse)
library(SingleCellExperiment)
library(pheatmap)
library(RColorBrewer)
 library(ggplot2)
library(dplyr)
```

---

## 2. Load the Three Clustering Runs

k=20: the fully annotated fine-resolution run (20 clusters)
k=150 and k=200: the more broader resolution runs (12 and 11 clusters, respectively)

```{r load-data}

spe_k20_path  <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/clustering/k20_clusters/harmony_20_cluster_annot.rds"

spe_k150_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/clustering/k150_clusters/harmony_150_cluster_annot.rds"

spe_k200_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/clustering/k200_clusters/harmony_200_cluster_annot.rds"

spe_k20  <- readRDS(spe_k20_path)
spe_k150 <- readRDS(spe_k150_path)
spe_k200 <- readRDS(spe_k200_path)

cat("k=20  run: ", ncol(spe_k20),  "cells,", nlevels(factor(spe_k20$broad_clusters)),  "clusters\n")
cat("k=150 run: ", ncol(spe_k150), "cells,", nlevels(factor(spe_k150$broad_clusters)), "clusters\n")
cat("k=200 run: ", ncol(spe_k200), "cells,", nlevels(factor(spe_k200$broad_clusters)), "clusters\n")
```

---

## 3. Cell-Barcode Alignment Function

Cross-tabulating requires matching the SAME cell across two objects, not just assuming row order lines up. Different clustering runs may have dropped different QC-failed cells or been saved in a different cell order. This function finds the shared cells and returns aligned cluster-assignment vectors for both runs.

```{r align-function}

align_clusters <- function(spe_a, spe_b, label_a = "A", label_b = "B") {

  common_cells <- intersect(colnames(spe_a), colnames(spe_b))

  cat(sprintf(
    "%s: %d cells | %s: %d cells | shared: %d cells (%.1f%% of %s, %.1f%% of %s)\n",
    label_a, ncol(spe_a), label_b, ncol(spe_b), length(common_cells),
    100 * length(common_cells) / ncol(spe_a), label_a,
    100 * length(common_cells) / ncol(spe_b), label_b
  ))

  list(
    cells      = common_cells,
    clusters_a = factor(spe_a$broad_clusters[match(common_cells, colnames(spe_a))]),
    clusters_b = factor(spe_b$broad_clusters[match(common_cells, colnames(spe_b))])
  )
}
```

---

## 4. Cross-Tabulation Function (counts, proportions, purity, heatmap)

Builds everything needed from one pair of aligned cluster vectors:
- raw count cross-tab (fine clusters x coarse clusters)
- row-normalized proportions (what % of each fine cluster's cells land in each coarse cluster)
- a "purity" summary: for each fine cluster, which coarse cluster gets the most of its cells, 
  and what fraction that is. 
      - A fine cluster near 1.0 nests cleanly into one coarse cluster (stable). 
      - A fine cluster split across several coarse clusters at a lower fraction is more ambiguous / composite.
- a heatmap of the proportion table for quick visual scanning

```{r crosstab-function}

# plot_heatmap: adaptive text color heatmap, defined once and reused below
plot_heatmap <- function(mat, title = "", legend_title = "value",
                         digits = 2, text_threshold = 0.5) {
  row_levels <- rownames(mat)
  col_levels <- colnames(mat)

  df <- as.data.frame(as.table(mat))
  colnames(df) <- c("row", "col", "value")
  df$row <- factor(df$row, levels = rev(row_levels))
  df$col <- factor(df$col, levels = col_levels)

  val_range <- range(df$value, na.rm = TRUE)
  df$norm_value <- (df$value - val_range[1]) / diff(val_range)
  df$text_color <- ifelse(df$norm_value > text_threshold, "white", "black")

  ggplot(df, aes(x = col, y = row, fill = value)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = sprintf(paste0("%.", digits, "f"), value), color = text_color),
              size = 3) +
    scale_fill_gradientn(colors = brewer.pal(9, "YlGnBu"), name = legend_title) +
    scale_color_identity() +
    labs(title = title, x = NULL, y = NULL) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid  = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
      axis.text.y = element_text(size = 10),
      plot.title  = element_text(face = "bold")
    )
}

run_crosstab <- function(fine_clusters, coarse_clusters,
                         fine_label = "k=20", coarse_label = "coarse",
                         fine_names = NULL, coarse_names = NULL) {
  count_tab <- table(fine = fine_clusters, coarse = coarse_clusters)
  if (!is.null(fine_names)) {
    rownames(count_tab) <- paste0(rownames(count_tab), " - ", fine_names[rownames(count_tab)])
  }
  if (!is.null(coarse_names)) {
    colnames(count_tab) <- paste0(colnames(count_tab), " - ", coarse_names[colnames(count_tab)])
  }
  prop_tab <- prop.table(count_tab, margin = 1)

  purity_df <- data.frame(
    fine_cluster            = rownames(prop_tab),
    dominant_coarse_cluster = colnames(prop_tab)[apply(prop_tab, 1, which.max)],
    fraction_in_dominant    = round(apply(prop_tab, 1, max), 3),
    row.names               = NULL 
  ) |>
    arrange(fraction_in_dominant)

  list(counts = count_tab, proportions = prop_tab, purity = purity_df)
}
```

---

## 5. Comparison A: k=20 vs. k=150 (12 clusters)

Pull the cell-type name lookup from the k=20 annotated object so the heatmap rows are readable (e.g. "4 - DC / mregDC") instead of just bare cluster numbers.

```{r compare-k20-k150, fig.width=11, fig.height=8}

celltype_lookup_k20 <- setNames(
  spe_k20$broad_celltype[match(levels(factor(spe_k20$broad_clusters)), as.character(spe_k20$broad_clusters))],
  levels(factor(spe_k20$broad_clusters))
)

celltype_lookup_k150 <- setNames(
  spe_k150$broad_celltype[match(levels(factor(spe_k150$broad_clusters)), as.character(spe_k150$broad_clusters))],
  levels(factor(spe_k150$broad_clusters))
)

aligned_150 <- align_clusters(spe_k20, spe_k150, label_a = "k=20", label_b = "k=150")

result_150 <- run_crosstab(
  fine_clusters   = aligned_150$clusters_a,
  coarse_clusters = aligned_150$clusters_b,
  fine_label      = "k=20",
  coarse_label    = "k=150",
  fine_names      = celltype_lookup_k20,
  coarse_names    = celltype_lookup_k150
)

plot_heatmap(result_150$proportions,
            title = "k=20 clusters vs. k=150 clusters (row-normalized)",
            legend_title = "proportion")

cat("\n--- k=20 clusters sorted by how 'split' they are across k=150 clusters ---\n")
cat("(low fraction_in_dominant = split across multiple coarse clusters = more ambiguous)\n\n")
knitr::kable(result_150$purity, digits = 3, caption = "k=20 sorted by purity vs. k=150")
```

---

## 6. Comparison B: k=20 vs. k=200 (11 clusters)

```{r compare-k20-k200, fig.width=11, fig.height=8}

celltype_lookup_k200 <- setNames(
  spe_k200$broad_celltype[match(levels(factor(spe_k200$broad_clusters)), as.character(spe_k200$broad_clusters))],
  levels(factor(spe_k200$broad_clusters))
)

aligned_200 <- align_clusters(spe_k20, spe_k200, label_a = "k=20", label_b = "k=200")

result_200 <- run_crosstab(
  fine_clusters   = aligned_200$clusters_a,
  coarse_clusters = aligned_200$clusters_b,
  fine_label      = "k=20",
  coarse_label    = "k=200",
  fine_names      = celltype_lookup_k20,
  coarse_names    = celltype_lookup_k200
)

plot_heatmap(result_200$proportions,
            title = "k=20 clusters vs. k=200 clusters (row-normalized)",
            legend_title = "proportion")


cat("\n--- k=20 clusters sorted by how 'split' they are across k=200 clusters ---\n")
cat("(low fraction_in_dominant = split across multiple coarse clusters = more ambiguous)\n\n")
knitr::kable(result_200$purity, digits = 3, caption = "k=20 clusters sorted by purity vs. k=200")
```

---

## 7. Bonus: k=150 vs. k=200 Stability Check

k=150 (12 clusters) and k=200 (11 clusters) are very close in cluster count. 
Worth checking directly whether they're actually the same.
Partition with one cluster merged, or genuinely different groupings.
High agreement here would support treating ~11-12 as a stable "natural" number of broad cell types in this tissue.

```{r compare-k150-k200, fig.width=9, fig.height=6}

aligned_150_200 <- align_clusters(spe_k150, spe_k200, label_a = "k=150", label_b = "k=200")

result_150_200 <- run_crosstab(
  fine_clusters   = aligned_150_200$clusters_a,
  coarse_clusters = aligned_150_200$clusters_b,
  fine_label      = "k=150",
  coarse_label    = "k=200",
  fine_names      = celltype_lookup_k150,
  coarse_names    = celltype_lookup_k200
)

plot_heatmap(result_150_200$proportions,
            title = "k=150 clusters vs. k=200 clusters (row-normalized)",
            legend_title = "proportion")

cat("\n--- k=150 clusters sorted by how 'split' they are across k=200 clusters ---\n\n")
knitr::kable(result_150_200$purity, digits = 3, caption = "k=150 clusters sorted by purity vs. k=200")
```

---

## 8. Flagged-Cluster Lookup (quick reference)

Quick pull-out of where our specifically flagged clusters from the violin plot review land at each coarser resolution.

```{r flagged-lookup}
library(kableExtra)

flagged_ids <- c("4", "9", "10", "11", "17", "18", "20")

flagged_summary <- purrr::map_dfr(flagged_ids, function(id) {
  row_150 <- result_150$purity |> filter(str_starts(fine_cluster, paste0(id, " -")))
  row_200 <- result_200$purity |> filter(str_starts(fine_cluster, paste0(id, " -")))

  tibble(
    fine_cluster           = paste0(id, " - ", celltype_lookup_k20[[id]]),
    dominant_k150_cluster  = ifelse(nrow(row_150) > 0, row_150$dominant_coarse_cluster, NA),
    frac_in_k150           = ifelse(nrow(row_150) > 0, row_150$fraction_in_dominant, NA),
    dominant_k200_cluster  = ifelse(nrow(row_200) > 0, row_200$dominant_coarse_cluster, NA),
    frac_in_k200           = ifelse(nrow(row_200) > 0, row_200$fraction_in_dominant, NA)
  )
})

flagged_summary |>
  kbl(caption = "Where the flagged k=20 clusters land at coarser resolution",
      col.names = c("Fine Cluster (k=20)", "Dominant k=150 Cluster", "Frac. in k=150",
                    "Dominant k=200 Cluster", "Frac. in k=200"),
      booktabs = FALSE) |>
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"),
                full_width = FALSE) |>
  row_spec(0, bold = TRUE)
```

---

## Notes

- **Alignment matters:** the `align_clusters()` function intersects
  cell barcodes rather than assuming the same cell order across
  objects. If the shared-cell percentage printed in Section 3/5/6 is
  much less than ~95-100%, that's worth investigating on its own --
  it would mean the two clustering runs were built from meaningfully
  different QC-filtered cell sets, not just different `k` values.
- **Reading the purity table:** `fraction_in_dominant` close to 1.0
  means that fine cluster nests cleanly into one coarse cluster
  (stable, trustworthy). Lower values (e.g. < 0.6) mean that fine
  cluster's cells are genuinely split across multiple coarse groups --
  either a real transitional population, or a cluster whose fine-level
  identity is shakier than it looked in the dot plot / violin plots
  alone.
- **This does not replace the violin plots.** This tells you *which*
  clusters merge or split under different resolutions; it doesn't
  tell you *why* (e.g. doublet vs. real continuum). Use it to decide
  which clusters are worth a targeted subclustering pass, then use
  violin plots on that subclustering result to confirm the biological
  story.
