---
title: "Epithelial Cluster Diagnostics (k20 / k150 / k200 Overlap)"
author: "Vaidehi Gupta"
date: "07-08-2026"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(SpatialExperiment)
library(imcRtools)
library(scran)
library(scater)
library(bluster)
library(dittoSeq)
library(dplyr)
library(tidyverse)
library(viridis)
library(patchwork)
library(ggridges)
```

## Purpose

Same framework as the T cell diagnostics notebook, applied to the
epithelial compartment. Four clusters are in scope here, for different
reasons:

- **Cluster 10 (Transitional Myeloid/Unresolved)** -- this is the most
  important one to resolve. Its violin plot showed TTF1orCD16 and
  EpCAM co-elevated, and the k150/k200 crosstab showed 94%/93% of its
  cells landing in the partner-annotated "Tumor Epithelial (TTF1+)"
  coarse cluster -- not a myeloid group. Two independent resolutions
  both call this population epithelial. This notebook checks whether
  the stable core actually looks epithelial on its own markers, which
  would support relabeling cluster 10 entirely rather than treating it
  as a myeloid cluster with epithelial contamination.
- **Cluster 18 (Tumor Epithelial)** -- only ~60-63% agreement with
  itself across coarser resolutions, small (836 cells), and showed a
  patient enrichment spike (Patient20_425, ~3.2x).
- **Clusters 13 and 16 (both labeled "Epithelial")** -- 13 splits
  roughly evenly between two coarse clusters; 16 showed much weaker
  overall signal than 13 despite being the larger of the two. Included
  here as a lighter screening check on whether they represent two real
  subtypes or one is a lower-quality duplicate of the other.

## Load the SPE object and cluster assignments

```{r load-data}
clustering_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/clustering"

spe <- readRDS("/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/batch_correction/spe_combined_harmony_r3.rds")

k20_spe  <- readRDS(file.path(clustering_path, "k20_clusters",  "harmony_20_cluster_annot.rds"))
k150_spe <- readRDS(file.path(clustering_path, "k150_clusters", "harmony_150_cluster_annot.rds"))
k200_spe <- readRDS(file.path(clustering_path, "k200_clusters", "harmony_200_cluster_annot.rds"))

k20_vec  <- setNames(colData(k20_spe)$broad_clusters,  colnames(k20_spe))
k150_vec <- setNames(colData(k150_spe)$broad_clusters, colnames(k150_spe))
k200_vec <- setNames(colData(k200_spe)$broad_clusters, colnames(k200_spe))

setequal(names(k20_vec), colnames(spe))
setequal(names(k150_vec), colnames(spe))
setequal(names(k200_vec), colnames(spe))

k20_clusters  <- k20_vec[colnames(spe)]
k150_clusters <- k150_vec[colnames(spe)]
k200_clusters <- k200_vec[colnames(spe)]

stopifnot(all(names(k20_clusters) == colnames(spe)))
stopifnot(all(names(k150_clusters) == colnames(spe)))
stopifnot(all(names(k200_clusters) == colnames(spe)))

assay(spe, "exprs") <- asinh(assay(spe, "counts") / 1)

get_celltype_lookup <- function(spe_obj) {
  if ("broad_celltype" %in% colnames(colData(spe_obj))) {
    lv <- levels(factor(colData(spe_obj)$broad_clusters))
    setNames(
      colData(spe_obj)$broad_celltype[match(lv, as.character(colData(spe_obj)$broad_clusters))],
      lv
    )
  } else {
    lv <- levels(factor(colData(spe_obj)$broad_clusters))
    setNames(lv, lv)
  }
}

celltype_k20  <- get_celltype_lookup(k20_spe)
celltype_k150 <- get_celltype_lookup(k150_spe)
celltype_k200 <- get_celltype_lookup(k200_spe)
```

## Build the cross-tabulation table

```{r build-comparison-table}
cluster_compare <- data.frame(
  cell_id = colnames(spe),
  k20  = as.character(k20_clusters),
  k150 = as.character(k150_clusters),
  k200 = as.character(k200_clusters),
  patient_id = spe$patient_id,
  sample_id  = spe$sample_id,
  area = spe$area,
  stringsAsFactors = FALSE
)
```

## Identify epithelial clusters at k20

```{r identify-epithelial-clusters}
# Cluster 10 is NOT labeled "Epithelial" at k20 -- it's included
# explicitly below rather than picked up by the label match, since the
# whole point is testing whether that label is wrong.
epithelial_clusters_k20 <- names(celltype_k20)[grepl("Epithelial", celltype_k20, ignore.case = TRUE)]
epithelial_clusters_k20 <- union(epithelial_clusters_k20, "10")

data.frame(k20_cluster = epithelial_clusters_k20, label = celltype_k20[epithelial_clusters_k20])
```

## Screening pass: where does each cluster land at k150/k200

```{r screening-crosstab}
screen_cluster <- function(k20_id) {
  k150_tab <- table(cluster_compare$k150[cluster_compare$k20 == k20_id])
  k200_tab <- table(cluster_compare$k200[cluster_compare$k20 == k20_id])

  dom_150 <- names(which.max(k150_tab))
  dom_200 <- names(which.max(k200_tab))

  data.frame(
    k20_cluster       = k20_id,
    label             = celltype_k20[[k20_id]],
    n_cells           = sum(cluster_compare$k20 == k20_id),
    dominant_k150     = paste0(dom_150, " - ", celltype_k150[[dom_150]]),
    frac_k150         = round(max(k150_tab) / sum(k150_tab), 3),
    dominant_k200     = paste0(dom_200, " - ", celltype_k200[[dom_200]]),
    frac_k200         = round(max(k200_tab) / sum(k200_tab), 3)
  )
}

screening_results <- bind_rows(lapply(epithelial_clusters_k20, screen_cluster)) |>
  arrange(pmin(frac_k150, frac_k200))

print(screening_results)
```

## Stable-core and marker/composition diagnostic functions

```{r shared-functions}
# identical logic to the T cell notebook -- kept here rather than
# sourced from a shared file so this notebook can run standalone
run_stable_core_diagnostic <- function(k20_id, spe, cluster_compare) {

  cat("\n==============================\n")
  cat("Cluster", k20_id, "-", celltype_k20[[k20_id]], "\n")
  cat("==============================\n")

  cluster_cells <- cluster_compare$cell_id[cluster_compare$k20 == k20_id]

  k150_tab <- table(cluster_compare$k150[cluster_compare$k20 == k20_id])
  k200_tab <- table(cluster_compare$k200[cluster_compare$k20 == k20_id])
  dom_150 <- names(which.max(k150_tab))
  dom_200 <- names(which.max(k200_tab))

  cat("Dominant k150:", dom_150, "-", celltype_k150[[dom_150]],
      sprintf("(%.1f%%)\n", 100 * max(k150_tab) / sum(k150_tab)))
  cat("Dominant k200:", dom_200, "-", celltype_k200[[dom_200]],
      sprintf("(%.1f%%)\n", 100 * max(k200_tab) / sum(k200_tab)))

  stable_core <- cluster_compare |>
    filter(k20 == k20_id, k150 == dom_150, k200 == dom_200) |>
    pull(cell_id)

  disagreement_cells <- setdiff(cluster_cells, stable_core)

  cat("Stable core:", length(stable_core), "/", length(cluster_cells),
      sprintf("(%.1f%%)\n", 100 * length(stable_core) / length(cluster_cells)))

  list(
    k20_id              = k20_id,
    cluster_cells       = cluster_cells,
    stable_core         = stable_core,
    disagreement_cells  = disagreement_cells,
    dominant_k150       = dom_150,
    dominant_k200       = dom_200
  )
}

run_marker_diagnostics <- function(diag_result, spe, top_n = 10) {

  spe$group_label <- case_when(
    colnames(spe) %in% diag_result$stable_core        ~ "stable_core",
    colnames(spe) %in% diag_result$disagreement_cells  ~ "disagreement",
    TRUE                                                ~ "rest"
  )

  markers_vs_rest <- findMarkers(spe, groups = spe$group_label, assay.type = "exprs")
  positive_vs_rest <- as.data.frame(markers_vs_rest$stable_core) |>
    filter(logFC.rest > 0) |>
    arrange(desc(logFC.rest))

  cat("Top markers, stable_core vs rest of dataset:\n")
  print(head(positive_vs_rest, top_n))

  markers_vs_disagreement <- NULL
  if (length(diag_result$disagreement_cells) >= 20) {
    spe_pair <- spe[, spe$group_label %in% c("stable_core", "disagreement")]
    markers_vs_disagreement <- findMarkers(spe_pair, groups = droplevels(factor(spe_pair$group_label)),
                                           assay.type = "exprs")

    cat("\nTop markers, stable_core vs disagreement cells (same original cluster):\n")
    print(head(as.data.frame(markers_vs_disagreement$stable_core) |> arrange(desc(logFC.disagreement)), top_n))
  } else {
    cat("\nToo few disagreement cells (<20) for a meaningful stable_core vs disagreement comparison.\n")
  }

  list(
    positive_vs_rest         = positive_vs_rest,
    markers_vs_disagreement  = markers_vs_disagreement,
    group_label              = spe$group_label
  )
}

run_composition_diagnostics <- function(diag_result, cluster_compare) {

  cat("\nPatient composition of stable core:\n")
  patient_comp <- cluster_compare |>
    filter(cell_id %in% diag_result$stable_core) |>
    count(patient_id) |>
    arrange(desc(n)) |>
    as.data.frame()

  print(patient_comp)

  cat("\nROI (sample_id) composition of stable core:\n")
  print(sort(table(cluster_compare$sample_id[cluster_compare$cell_id %in% diag_result$stable_core]),
             decreasing = TRUE))

  area_df <- cluster_compare |>
    mutate(group = case_when(
      cell_id %in% diag_result$stable_core        ~ "stable_core",
      cell_id %in% diag_result$disagreement_cells  ~ "disagreement",
      TRUE                                          ~ "rest"
    )) |>
    filter(group != "rest")

  cat("\nCell area summary, stable core vs disagreement cells:\n")
  area_summary <- area_df |>
    group_by(group) |>
    summarise(
      mean_area = mean(area, na.rm = TRUE),
      median_area = median(area, na.rm = TRUE),
      sd_area = sd(area, na.rm = TRUE)
    ) |>
    as.data.frame()

  print(area_summary)

  ggplot(area_df, aes(x = group, y = area, fill = group)) +
    geom_violin() +
    theme_classic(base_size = 14) +
    labs(title = paste("Cell area: stable core vs. disagreement cells, cluster", diag_result$k20_id))
}
```

## Ki67 / proliferation check function

```{r ki67-function}
# specific to the epithelial notebook -- tumor epithelial cells are
# where proliferation status is most biologically relevant, per the
# meeting notes ("see if the proliferating cells are high in EpCAM")
run_ki67_check <- function(diag_result, spe, cluster_compare) {

  ki67_vals <- assay(spe, "exprs")["MKI67", ]
  total_counts <- colSums(assay(spe, "counts"))

  group <- case_when(
    colnames(spe) %in% diag_result$stable_core        ~ "stable_core",
    colnames(spe) %in% diag_result$disagreement_cells  ~ "disagreement",
    TRUE                                                ~ "rest"
  )

  cat("Raw Ki67 (MKI67), by group:\n")
  print(tapply(ki67_vals[group != "rest"], group[group != "rest"], summary))

  ki67_normalized <- ki67_vals / total_counts
  cat("\nKi67 normalized by total counts, by group:\n")
  print(tapply(ki67_normalized[group != "rest"], group[group != "rest"], summary))
}
```

---

## Deep dive: Cluster 10 (Transitional Myeloid/Unresolved -> possibly mislabeled)

```{r deep-dive-10}
diag_10 <- run_stable_core_diagnostic("10", spe, cluster_compare)
```

```{r deep-dive-10-markers}
markers_10 <- run_marker_diagnostics(diag_10, spe)
```

```{r deep-dive-10-composition, fig.width=8, fig.height=5}
run_composition_diagnostics(diag_10, cluster_compare)
```

```{r deep-dive-10-ki67}
run_ki67_check(diag_10, spe, cluster_compare)
```

```{r deep-dive-10-ridge, fig.width=10, fig.height=6}
top_markers_10 <- rownames(markers_10$positive_vs_rest)[1:min(10, nrow(markers_10$positive_vs_rest))]

expr_long_10 <- as.data.frame(t(as.matrix(assay(spe, "exprs")[top_markers_10, ]))) |>
  mutate(group_label = markers_10$group_label) |>
  filter(group_label != "rest") |>
  pivot_longer(cols = all_of(top_markers_10), names_to = "marker", values_to = "expression")

ggplot(expr_long_10, aes(x = expression, y = marker, fill = group_label)) +
  geom_density_ridges(alpha = 0.6, scale = 1.2) +
  scale_fill_manual(values = c("stable_core" = "steelblue", "disagreement" = "firebrick")) +
  theme_ridges(font_size = 12) +
  labs(title = "Cluster 10: stable core vs. disagreement cells", x = "asinh expression", y = "Marker")
```

**Observations:**
- TTF1 clearly separates stable core from disagreement cells
- No myeloid marker has their disagreement group popping out. CD14, CD1c, CD117, etc all have largely overlapping distributions between the two groups.


If the stable core's top markers here come back EpCAM/TTF1orCD16-led
with CD68/CD11c/CD14 notably absent or low, that supports relabeling
cluster 10 as epithelial outright (consistent with the k150/k200
crosstab). If myeloid markers are still elevated even within the
stable core, that argues for a genuine hybrid/spillover population
rather than a simple mislabel -- worth a QuPath spatial check either way,
per the earlier discussion, to see whether these cells sit inside tumor
nests or scattered through stroma/macrophage-rich regions.

---

## Deep dive: Cluster 18 (Tumor Epithelial)

```{r deep-dive-18}
diag_18 <- run_stable_core_diagnostic("18", spe, cluster_compare)
```

```{r deep-dive-18-markers}
markers_18 <- run_marker_diagnostics(diag_18, spe)
```

```{r deep-dive-18-composition, fig.width=8, fig.height=5}
run_composition_diagnostics(diag_18, cluster_compare)
```

```{r deep-dive-18-ki67}
run_ki67_check(diag_18, spe, cluster_compare)
```

```{r deep-dive-18-ridge, fig.width=10, fig.height=6}
top_markers_18 <- rownames(markers_18$positive_vs_rest)[1:min(10, nrow(markers_18$positive_vs_rest))]

expr_long_18 <- as.data.frame(t(as.matrix(assay(spe, "exprs")[top_markers_18, ]))) |>
  mutate(group_label = markers_18$group_label) |>
  filter(group_label != "rest") |>
  pivot_longer(cols = all_of(top_markers_18), names_to = "marker", values_to = "expression")

ggplot(expr_long_18, aes(x = expression, y = marker, fill = group_label)) +
  geom_density_ridges(alpha = 0.6, scale = 1.2) +
  scale_fill_manual(values = c("stable_core" = "steelblue", "disagreement" = "firebrick")) +
  theme_ridges(font_size = 12) +
  labs(title = "Cluster 18: stable core vs. disagreement cells", x = "asinh expression", y = "Marker")
```

Given cluster 18 is only 836 cells to begin with, check the stable
core size before over-interpreting -- a stable core under ~50 cells
should be treated as low-confidence, same caveat as cluster 20 in the
T cell notebook.

---

## UMAP check, all four clusters

```{r umap-check}
spe$flagged_group <- case_when(
  colnames(spe) %in% diag_10$stable_core ~ "Cluster 10 - stable core",
  colnames(spe) %in% diag_10$disagreement_cells ~ "Cluster 10 - disagreement",
  colnames(spe) %in% diag_18$stable_core ~ "Cluster 18 - stable core",
  colnames(spe) %in% diag_18$disagreement_cells ~ "Cluster 18 - disagreement",
  TRUE ~ "other"
)

dittoDimPlot(
  spe,
  var = "flagged_group",
  reduction.use = "UMAP_harmony",
  size = 0.2,
  do.label = FALSE
) + ggtitle("Clusters 10 and 18: stable core vs. disagreement cells")
```

```{r umap-check-v2, fig.width=9, fig.height=7}
umap_df <- as.data.frame(reducedDim(spe, "UMAP_harmony"))
colnames(umap_df) <- c("UMAP1", "UMAP2")
umap_df$flagged_group <- spe$flagged_group

group_colors <- c(
  "other" = "grey85",
  "Cluster 10 - stable core"    = "#4C9BE8",
  "Cluster 10 - disagreement"   = "#E8A13B",
  "Cluster 18 - stable core"    = "#F5E642",
  "Cluster 18 - disagreement"   = "#2CA37F"
)

# draw order: background first, then progressively rarer/smaller groups on top
draw_order <- c("other",
                "Cluster 10 - stable core", "Cluster 10 - disagreement",
                "Cluster 18 - stable core", "Cluster 18 - disagreement")

umap_df$flagged_group <- factor(umap_df$flagged_group, levels = draw_order)
umap_df <- umap_df[order(umap_df$flagged_group), ]

point_sizes <- c(
  "other" = 0.15,
  "Cluster 10 - stable core"  = 0.3,
  "Cluster 10 - disagreement" = 0.5,
  "Cluster 18 - stable core"  = 1.2,
  "Cluster 18 - disagreement" = 1.2
)
umap_df$pt_size <- point_sizes[as.character(umap_df$flagged_group)]

ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = flagged_group)) +
  geom_point(aes(size = flagged_group), alpha = 0.8) +
  scale_color_manual(values = group_colors) +
  scale_size_manual(values = point_sizes, guide = "none") +
  theme_bw(base_size = 13) +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 1))) +
  labs(title = "Clusters 10 and 18: stable core vs. disagreement cells",
       color = "Group")
```

## Summary and next steps

* **Cluster 10:** this is the highest-value result in this notebook --
  if the stable core's markers confirm epithelial identity, this
  should be relabeled and merged conceptually with the other
  epithelial clusters rather than kept as "Transitional
  Myeloid/Unresolved." Follow up with the QuPath spatial check
  discussed earlier before finalizing, since marker evidence alone
  can't distinguish "these are epithelial cells near macrophages" from
  "these are macrophages that ate epithelial debris."
* **Cluster 18:** check stable core size first given the small starting
  n. If markers still look cleanly epithelial/tumor-associated (EpCAM,
  GATA3, and ideally MKI67/PDL1 per the meeting notes' recommended
  tumor markers) in the stable core, that's reassuring despite the
  patient enrichment flag. If Patient20_425 dominates the stable core
  specifically, treat this cluster's tumor-epithelial identity as
  patient-specific rather than general until confirmed elsewhere.
* **Clusters 13 and 16:** use the screening table to see whether they
  land in the same or different coarse clusters at both k150 and
  k200. Landing consistently in different coarse groups across both
  supports two real subtypes; landing in the same one, especially with
  16 unable to beat baseline convincingly, supports treating 16 as a
  lower-confidence duplicate of 13.
