---
title: "Spatial Setup: Final Cell Type Annotation Merge"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

This notebook performs the final annotation merge (myeloid + cluster15 +
B cell/plasma subclustering results), strips the tracking-number prefixes
from cell type labels, excludes confirmed low-quality clusters, and saves
the finalized `spe_final_celltype.rds` object.

This does NOT re-derive the cluster 15 / subclustering QC checks themselves
(those were already done and are documented separately) — this notebook
assumes the subcluster label objects already exist and just combines them.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r libraries}
.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(tidyverse)
library(SingleCellExperiment)
library(SpatialExperiment)
library(scater)

# Adjust to your actual project paths
analysis_path <- "/restricted/projectnb/camplab/home/vgupta7/analysis"
```

# Load intermediate subclustering objects

`r # FILL IN: load whatever intermediate objects hold your finalized subcluster`
`r # labels for myeloid, cluster 15, and B cell/plasma -- adjust object/file`
`r # names below to match what you actually saved during subclustering.`

```{r load-intermediates}
spe <- readRDS(file.path(analysis_path, "spe_broad_clusters.rds"))

myeloid_labels    <- readRDS(file.path(analysis_path, "myeloid_subcluster_labels.rds"))
cluster15_labels  <- readRDS(file.path(analysis_path, "cluster15_subcluster_labels.rds"))
bcell_labels      <- readRDS(file.path(analysis_path, "bcell_subcluster_labels.rds"))
```

# Final merge

Order matters: myeloid → cluster15 → bcell (filtered to exclude
cluster-15-origin cells), so cluster 15 cells consistently use the
cluster-15-derived T cell labels rather than being overwritten by the
partner's independently-run B cell/plasma subclustering.

```{r final-merge}
spe$final_celltype <- as.character(spe$broad_celltype)

# Myeloid subcluster labels
spe$final_celltype[match(names(myeloid_labels), colnames(spe))] <- myeloid_labels

# Cluster 15 subcluster labels
spe$final_celltype[match(names(cluster15_labels), colnames(spe))] <- cluster15_labels

# B cell/plasma subcluster labels, EXCLUDING cluster-15-origin cells
# (avoids overwriting the cluster-15 resolution with the disproved
# plasma-cell labeling from the independent B cell subclustering)
bcell_labels_filtered <- bcell_labels[!(names(bcell_labels) %in% names(cluster15_labels))]
spe$final_celltype[match(names(bcell_labels_filtered), colnames(spe))] <- bcell_labels_filtered

spe$final_celltype <- factor(spe$final_celltype)

cat("Cell types after merge:", nlevels(spe$final_celltype), "\n")
cat("NAs:", sum(is.na(spe$final_celltype)), "\n")
```

# Strip tracking-number prefixes from labels

The leading numbers (e.g. `"10. Memory T/Plasma Cell (CD45RO+)"`) were only
ever meant as an internal bookkeeping aid during annotation, but they
interfere with downstream niche annotation, so they're removed here at the
source before saving the final object.

```{r strip-label-numbers}
spe$final_celltype <- factor(gsub("^\\d+\\.\\s*", "", as.character(spe$final_celltype)))

# Sanity check: confirm no unintended matches, and no numbers remain
levels(spe$final_celltype)
```

# Exclude confirmed low-quality clusters

Cluster 17 (broad-level, confirmed low quality) and cluster 15 subclusters
2/4 (same signature as cluster 17 — no markers clear threshold, lower total
counts not explained by cell area) are excluded here. Note: the ~22-cell
tiny leftover categories from the cluster-15/B-cell merge conflict are
intentionally NOT excluded here — they stay in this base annotated object,
and are excluded downstream on an as-needed basis (e.g. before sccomp or
the spatial graph), so this file remains a complete record of every
annotated cell.

```{r exclude-low-quality}
n_before <- ncol(spe)

spe <- spe[, spe$broad_clusters != "17"]

low_quality_labels <- c("Unresolved/Low Quality")
spe <- spe[, !(spe$final_celltype %in% low_quality_labels)]

n_after <- ncol(spe)

cat("Cells before exclusion:", n_before, "\n")
cat("Cells after excluding low-quality clusters:", n_after, "\n")
cat("Cells removed:", n_before - n_after,
    sprintf("(%.1f%% of dataset)\n", 100 * (n_before - n_after) / n_before))

spe$final_celltype <- droplevels(spe$final_celltype)
```

# Final cell type counts

```{r final-celltype-table}
spe$final_celltype |>
  table(useNA = "ifany") |>
  as.data.frame() |>
  setNames(c("CellType", "Count")) |>
  arrange(desc(Count)) |>
  knitr::kable(caption = "Final cell type counts after merge and QC exclusion")
```

# Save finalized object

```{r save-final}
saveRDS(spe, file.path(analysis_path, "spe_final_celltype.rds"))
```

# Handoff notes

- `spe_final_celltype.rds` now has unnumbered cell type labels — any
  downstream code (sccomp exclusion lists, spatial graph exclusion lists,
  niche annotation) that references the old numbered label strings
  (e.g. `"10. Memory T/Plasma Cell (CD45RO+)"`) needs to be updated to the
  new unnumbered versions before re-running.
- This file still contains the small (~1-2 cell) leftover categories from
  the cluster-15/B-cell merge conflict — exclude these explicitly in each
  downstream analysis as needed, rather than assuming they're already gone.
