---
title: "Update Master Annotation: T Cell Subcluster Relabeling (r2 -> r3)"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
---

Recodes the T cell subcluster labels in the master annotated object
following PI feedback on naive/memory/exhaustion naming
(`tcell_subclustering_2.Rmd`). No cell IDs or subcluster boundaries
change here -- only the label text for cells already carrying the old
T cell subcluster names.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# 1. Load Libraries

```{r libraries}
#.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(tidyverse)
library(SpatialExperiment)
```

# 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

r2_path <- file.path(analysis_path, "final_cell_annotation", "r2_annotation", "spe_final_celltype_r2.rds")
r3_dir  <- file.path(analysis_path, "final_cell_annotation", "r3_annotation")
r3_path <- file.path(r3_dir, "spe_final_celltype_r3.rds")
```

# 3. Load and Recode

```{r recode}
spe_final <- readRDS(r2_path)
spe_final$final_celltype <- as.character(spe_final$final_celltype)

cat("Cell counts before recoding (T cell labels only):\n")
print(table(spe_final$final_celltype)[c(
  "CD4+ T Cell, Exhaustion-leaning", "CD4+ T Cell, Activated", "Treg, Activated",
  "Memory T Cell", "CD8+ T Cell, Activated", "Treg, Resting", "CD8+ T Cell, Resting"
)])

# old label (as currently stored) -> new label
label_recode <- c(
  "CD4+ T Cell, Exhaustion-leaning"  = "CD4+ Memory T Cell, Exhaustion-leaning",
  "CD4+ T Cell, Activated"           = "CD4+ Memory T Cell, Partial-exhausion",
  "Treg, Activated"                  = "Treg, Memory",
  "Memory T Cell"                    = "Memory T Cell (Unresolved)",
  "CD8+ T Cell, Activated"           = "CD8+ Memory T Cell",
  "Treg, Resting"                    = "Treg, Naive",
  "CD8+ T Cell, Resting"             = "CD8+ Naive T Cell"
)

still_present <- setdiff(names(label_recode), unique(spe_final$final_celltype))
if (length(still_present) > 0) {
  cat("\nWARNING -- these old labels weren't found, check spelling:\n")
  print(still_present)
}

recode_hits <- spe_final$final_celltype %in% names(label_recode)
spe_final$final_celltype[recode_hits] <- label_recode[spe_final$final_celltype[recode_hits]]
spe_final$final_celltype <- factor(spe_final$final_celltype)

cat("\nCells relabeled:", sum(recode_hits), "\n")
cat("\nCell counts after recoding (new T cell labels only):\n")
print(table(spe_final$final_celltype)[unname(label_recode)])
```

# 4. Save Updated Object

```{r save}
dir.create(r3_dir, recursive = TRUE, showWarnings = FALSE)
saveRDS(spe_final, r3_path)
cat("Saved updated object to:", r3_path, "\n")
```

# Notes

- Only the 7 T cell subcluster labels are recoded here (see
  `tcell_subclustering_2.Rmd` for the analysis behind the naming). All
  other cell type labels are untouched.
- `metadata(spe_final)$color_vectors`, if present, won't have entries for
  the new label strings -- downstream scripts with auto-fill logic (e.g.
  the cytomapper overlay Rmd) handle this gracefully, but anything with a
  hardcoded color vector keyed to the old T cell names will need updating.