---
title: "Tumor IMC - Batch Correction and Dimensionality Reduction (Run 5)"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r libraries}
.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(imcRtools)
library(SpatialExperiment)
library(SingleCellExperiment)
library(ggplot2)
library(dplyr)
library(stringr)
library(harmony)
library(scater)
library(dittoSeq)
library(viridis)
library(cowplot)

set.seed(42)
```

## Load Data

Adding back in Patient20's 5/19 acquisition to increase total cell counts. 
The 230 vs. 430 scale difference reflects a larger capture
area at the same resolution, not a calibration/resolution mismatch,
so cell segmentation and marker intensities should be directly
comparable between the two acquisitions. This is checked explicitly
(not just assumed) in the `check-sample-id-collision` and downstream
overcorrection notebook. Patient17's ROI 1 and ROI 4 are still faulty
and are removed after loading, once `sample_id` is available.

```{r paths}
base_dir <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/steinbock/"

figures_dir <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/figures/batch_correction_umaps/run5"
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)

# Patient20_519 added back in
steinbock_dirs <- list(
  Patient10   = file.path(base_dir, "p10_results"),
  Patient11   = file.path(base_dir, "p11_results"),
  Patient12   = file.path(base_dir, "p12_results"),
  Patient13   = file.path(base_dir, "p13_results"),
  Patient14   = file.path(base_dir, "p14_results"),
  Patient15   = file.path(base_dir, "p15_results"),
  Patient16   = file.path(base_dir, "p16_results"),
  Patient17   = file.path(base_dir, "p17_results"),
  Patient18   = file.path(base_dir, "p18_results"),
  Patient19   = file.path(base_dir, "p19_results"),
  Patient20_425 = file.path(base_dir, "p20_4_25_results"),
  Patient20_519 = file.path(base_dir, "p20_5_19_results")
)
```

```{r read-steinbock}
spe_list <- lapply(names(steinbock_dirs), function(patient) {
  cat("Loading:", patient, "\n")
  spe <- read_steinbock(steinbock_dirs[[patient]])
  spe$patient_id  <- patient
  spe
})
names(spe_list) <- names(steinbock_dirs)

# Check cells loaded per patient
for (p in names(spe_list)) {
  cat(p, ":", ncol(spe_list[[p]]), "cells\n")
}
```

## Metadata

Both Patient20 acquisitions are now in the cohort, so `disease_state`
includes both `Patient20_425` and `Patient20_519` (same disease state --
same patient, just two acquisitions).

```{r disease-state}
# Disease state metadata
disease_state <- c(
  Patient10    = "Benign",
  Patient11    = "Lower",
  Patient12    = "Higher",
  Patient13    = "Higher",
  Patient14    = "Lower",
  Patient15    = "Higher",
  Patient16    = "Lower",
  Patient17    = "Lower",
  Patient18    = "Lower",
  Patient19    = "Higher",
  Patient20_425 = "Lower",
  Patient20_519 = "Lower"
)
```

## Combine All Patients

```{r combine}
spe_combined <- do.call(cbind, spe_list)
spe_combined$disease_state <- disease_state[spe_combined$patient_id]

cat("Total cells before filtering:", ncol(spe_combined), "\n")
cat("Total markers:", nrow(spe_combined), "\n")

# Check cells per patient
table(spe_combined$patient_id)
```

## Check for sample_id Collision Between Patient20 Acquisitions

Harmony's batch variable is `sample_id`, not `patient_id`. It has no
concept that Patient20_425 and Patient20_519 are the same underlying
patient. `sample_id` is assigned automatically by `read_steinbock()`
per directory, so if both acquisitions happen to generate the same ROI
names (e.g. both producing `"Patient20-T_ROI_001_1"`), cells from two
genuinely different acquisitions would collapse onto one `sample_id`
after `cbind()` -- meaning Harmony would treat them as a single batch
and never separately correct for any technical difference between the
two scans. This check must run BEFORE Harmony, since fixing a collision
after the fact would require rerunning the correction anyway.

```{r check-sample-id-collision}
p20_ids <- unique(spe_combined$sample_id[spe_combined$patient_id %in% c("Patient20_425", "Patient20_519")])

p20_check <- data.frame(
  sample_id  = p20_ids,
  patient_id = sapply(p20_ids, function(s) {
    pids <- unique(spe_combined$patient_id[spe_combined$sample_id == s])
    paste(pids, collapse = " & ")  # collapse shows the collision directly if one exists
  })
)
print(p20_check)

# n_patients_per_sample_id > 1 anywhere below means a real collision
collision_check <- sapply(p20_ids, function(s) {
  length(unique(spe_combined$patient_id[spe_combined$sample_id == s]))
})

cat("\nsample_id values shared between Patient20_425 and Patient20_519:",
    sum(collision_check > 1), "out of", length(collision_check), "\n")

if (any(collision_check > 1)) {
  cat("\nCOLLISION DETECTED -- disambiguating sample_id by patient_id before Harmony.\n")
  spe_combined$sample_id <- paste(spe_combined$patient_id, spe_combined$sample_id, sep = "_")
  # NOTE: this changes sample_id strings for ALL patients, not just
  # Patient20, to keep the column internally consistent. If you rely on
  # exact sample_id string matches elsewhere (e.g. the Patient17 ROI
  # exclusion filter below), update those strings to match the new format.
} else {
  cat("\nNo collision -- Patient20_425 and Patient20_519 already have distinct sample_id values.\n")
  cat("Harmony will correctly treat their ROIs as separate batches without any changes needed.\n")
}
```

## Remove Faulty ROIs (Patient17 ROI 1 and ROI 4)

`sample_id` follows the pattern `Patient_<n>T_ROI_<roi#>_<index>`, e.g.
`Patient_17T_ROI_001_1`. We exclude ROI 001 and ROI 004 from Patient17.

```{r remove-faulty-rois}
rois_exclude <- c("Patient_17T_ROI_001_1", "Patient_17T_ROI_004_4")

cat("Cells in excluded Patient17 ROIs:\n")
print(table(spe_combined$sample_id[spe_combined$sample_id %in% rois_exclude]))

spe_combined <- spe_combined[, !spe_combined$sample_id %in% rois_exclude]

cat("Total cells after removing faulty Patient17 ROIs:", ncol(spe_combined), "\n")
```

## Filter by Cell Area

```{r area-filter}
spe_combined <- spe_combined[, spe_combined$area >= 20 & spe_combined$area < 500]

cat("Total cells after area filtering:", ncol(spe_combined), "\n")
```

## Select Markers

```{r markers}
# Exclude structural/QC markers
markers_exclude <- c("DNA191", "DNA193", "Membrane")
markers_use <- rownames(spe_combined)[!rownames(spe_combined) %in% markers_exclude]

cat("Markers used for batch correction:", length(markers_use), "\n")
print(markers_use)
```

## PCA and UMAP Before Batch Correction

No asinh transformation is applied - steinbock already performed 0-100
normalization/filtering on all patients, so PCA is run directly on `counts`.

```{r pca-uncorrected}
spe_combined <- runPCA(spe_combined,
                       assay.type = "counts",
                       subset_row = markers_use,
                       ncomponents = min(30, length(markers_use) - 1),
                       name = "PCA_uncorrected")

cat("PCA complete\n")
```

```{r umap-uncorrected}
spe_combined <- runUMAP(spe_combined,
                        dimred = "PCA_uncorrected",
                        name = "UMAP_uncorrected")

cat("UMAP complete\n")
```

```{r plot-uncorrected, fig.width=12, fig.height=5}
# Color by patient
p1 <- plotReducedDim(spe_combined,
                     dimred = "UMAP_uncorrected",
                     colour_by = "patient_id",
                     point_size = 0.3,
                     point_alpha = 0.3) +
  labs(title = "UMAP Before Batch Correction",
       subtitle = "Colored by patient") +
  theme_bw() +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 1)))

# Color by sample_id
p2 <- plotReducedDim(spe_combined,
                     dimred = "UMAP_uncorrected",
                     colour_by = "sample_id",
                     point_size = 0.3,
                     point_alpha = 0.3) +
  labs(title = "UMAP Before Batch Correction",
       subtitle = "Colored by sample_id (ROI)") +
  theme_bw() +
  theme(legend.position = "none")

plot_grid(p1, p2, ncol = 2)

ggsave(file.path(figures_dir, "umap_before_harmony.png"),
       width = 12, height = 10, dpi = 300)
```

## Run Harmony

```{r harmony}
# Extract PCA embeddings
pca_mat <- reducedDim(spe_combined, "PCA_uncorrected")

# Run Harmony using sample_id as batch variable
harmony_embeddings <- harmony::RunHarmony(
  pca_mat,
  meta_data = as.data.frame(colData(spe_combined)),
  vars_use  = "sample_id",
  verbose   = TRUE
)
```

```{r store-harmony}
# Store corrected embeddings
reducedDim(spe_combined, "Harmony") <- harmony_embeddings

cat("Harmony batch correction complete\n")
```

## UMAP After Batch Correction

```{r umap-harmony}
spe_combined <- runUMAP(spe_combined,
                        dimred = "Harmony",
                        name = "UMAP_harmony")

cat("UMAP on Harmony embeddings complete\n")
```

```{r plot-harmony, fig.width=12, fig.height=5}
# Color by patient
p3 <- plotReducedDim(spe_combined,
                     dimred = "UMAP_harmony",
                     colour_by = "patient_id",
                     point_size = 0.3,
                     point_alpha = 0.3) +
  labs(title = "UMAP After Harmony Batch Correction",
       subtitle = "Colored by patient") +
  theme_bw() +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 1)))

# Color by sample_id
p4 <- plotReducedDim(spe_combined,
                     dimred = "UMAP_harmony",
                     colour_by = "sample_id",
                     point_size = 0.3,
                     point_alpha = 0.3) +
  labs(title = "UMAP After Harmony Batch Correction",
       subtitle = "Colored by sample_id (ROI)") +
  theme_bw() +
  theme(legend.position = "none")

plot_grid(p3, p4, ncol = 2)

ggsave(file.path(figures_dir, "umap_after_harmony.png"),
       width = 14, height = 10, dpi = 300)
```

```{r plot-disease-state, fig.width=6, fig.height=5}
p5 <- plotReducedDim(spe_combined,
                     dimred = "UMAP_harmony",
                     colour_by = "disease_state",
                     point_size = 0.3,
                     point_alpha = 0.3) +
  labs(title = "UMAP After Harmony Batch Correction",
       subtitle = "Colored by disease state") +
  theme_bw() +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 1)))

p5

ggsave(file.path(figures_dir, "umap_harmony_disease_state.png"),
       plot = p5,
       width = 8, height = 6, dpi = 300)
```

```{r plot-disease-state-separated, fig.width=12, fig.height=5}
# Same UMAP, but split into one panel per disease state for easier comparison
p5_split <- plotReducedDim(spe_combined,
                     dimred = "UMAP_harmony",
                     colour_by = "disease_state",
                     point_size = 0.3,
                     point_alpha = 0.5) +
  facet_wrap(~ spe_combined$disease_state) +
  labs(title = "UMAP After Harmony Batch Correction",
       subtitle = "Split by disease state") +
  theme_bw() +
  theme(legend.position = "none")

p5_split

ggsave(file.path(figures_dir, "umap_harmony_disease_state_split.png"),
       plot = p5_split,
       width = 12, height = 5, dpi = 300)
```

```{r plot-comparison, fig.width=12, fig.height=10}
comparison_plot <- plot_grid(p1, p2, p3, p4, ncol = 2)
comparison_plot

ggsave(file.path(figures_dir, "umap_before_after_harmony.png"),
       plot = comparison_plot,
       width = 14, height = 10, dpi = 300)
```

```{r plot-per-patient, fig.width=12, fig.height=10}
# Check if patients mix well after correction
# Good batch correction = patients mix in UMAP
# Bad batch correction = patients still cluster separately
per_patient_plot <- plotReducedDim(spe_combined,
               dimred = "UMAP_harmony",
               colour_by = "patient_id",
               point_size = 0.3,
               point_alpha = 0.5) +
  facet_wrap(~ spe_combined$patient_id) +
  labs(title = "UMAP After Harmony - Per Patient") +
  theme_bw() +
  theme(legend.position = "none")

per_patient_plot

ggsave(file.path(figures_dir, "umap_harmony_per_patient.png"),
       plot = per_patient_plot,
       width = 14, height = 10, dpi = 300)
```

## Save Output

```{r save}
output_path <- file.path(base_dir, "../analysis/spe_combined_harmony_r5.rds")
saveRDS(spe_combined, output_path)
cat("Saved combined SPE object to:", output_path, "\n")
```

## Summary

```{r summary}
cat("=== Batch Correction Summary ===\n")
cat("Total cells:", ncol(spe_combined), "\n")
cat("Total markers:", nrow(spe_combined), "\n")
cat("Markers used for correction:", length(markers_use), "\n")
cat("Batch variable: sample_id\n")
cat("Method: Harmony\n")
cat("Included: Patient20 5/19 acquisition (added back per PI request for more cells); Patient17 ROI 1 and ROI 4 still excluded (faulty)\n")

cat("\nCells per patient:\n")
print(table(spe_combined$patient_id))

cat("\nCells per ROI (sample_id):\n")
print(table(spe_combined$sample_id))
```
