---
title: "Checking for Harmony Overcorrection"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 6)
set.seed(42)
```

## Overview

The disease-state UMAP alone can't tell us whether Harmony
overcorrected, since `disease_state` was never the batch variable
Harmony was told to fix (`sample_id` was). Similar UMAP *shape* across
disease states is expected if the same broad cell types are present
in all three, just at different proportions -- that's a density
question, not a shape question.

This notebook runs four more targeted checks instead:

1. **CD19 (and other flagged markers) pre- vs. post-Harmony** -- is
   diffuse expression a pre-existing background problem, or did
   Harmony smear out a signal that used to be crisp?
2. **Per-cluster patient composition / enrichment** -- is any cluster
   driven almost entirely by one patient's cells (undercorrection),
   after accounting for how many cells that patient contributes
   overall?
3. **LISI (Local Inverse Simpson's Index)** -- a quantitative,
   per-cell metric. We want *high* batch-LISI (patients well-mixed
   locally) and *low* cluster-LISI (distinct cell types still
   staying separate locally) on the Harmony embedding.
4. **Cell-type composition (%) across disease state** -- the
   proportion-based version of the check the UMAP shape couldn't
   answer.

---

## 1. Load Libraries

```{r libraries}

.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))
library(imcRtools)
library(SpatialExperiment)
library(SingleCellExperiment)
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(scater)
library(viridis)
library(patchwork)
library(cowplot)
library(pheatmap)
library(RColorBrewer)

# lisi is not on CRAN -- install once from GitHub if not already available:
#remotes::install_github("immunogenomics/LISI")
library(lisi)
```

---

## 2. Load Data

spe_combined: has BOTH UMAP_uncorrected/PCA_uncorrected and UMAP_harmony/Harmony reduced dims, needed for the pre/post comparisons in checks 1 and 3

```{r load-data}

spe_combined_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/batch_correction/spe_combined_harmony_r3.rds"
spe_combined <- readRDS(spe_combined_path)

spe_clustered_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/clustering/k20_clusters/harmony_20_cluster_annot.rds"
spe_clustered <- readRDS(spe_clustered_path)

cat("spe_combined:", ncol(spe_combined), "cells |", "reducedDims:", paste(reducedDimNames(spe_combined), collapse = ", "), "\n")
cat("spe_clustered:", ncol(spe_clustered), "cells |", nlevels(factor(spe_clustered$broad_clusters)), "clusters\n")
```


Clustering may have dropped a few more QC-failed cells than the batch-correction object, so match by cell barcode rather than assuming identical row order/length between the two objects.

```{r align-cells}

common_cells <- intersect(colnames(spe_combined), colnames(spe_clustered))

cat(sprintf("Shared cells: %d (%.1f%% of spe_combined, %.1f%% of spe_clustered)\n",
           length(common_cells),
           100 * length(common_cells) / ncol(spe_combined),
           100 * length(common_cells) / ncol(spe_clustered)))

# aligned metadata used throughout this notebook
meta_common <- data.frame(
  cell_id        = common_cells,
  patient_id     = spe_combined$patient_id[match(common_cells, colnames(spe_combined))],
  sample_id      = spe_combined$sample_id[match(common_cells, colnames(spe_combined))],
  disease_state  = spe_combined$disease_state[match(common_cells, colnames(spe_combined))],
  broad_clusters = spe_clustered$broad_clusters[match(common_cells, colnames(spe_clustered))],
  row.names      = common_cells
)

# cluster -> cell type name lookup for readable labels later
celltype_lookup <- setNames(
  spe_clustered$broad_celltype[match(levels(factor(spe_clustered$broad_clusters)), as.character(spe_clustered$broad_clusters))],
  levels(factor(spe_clustered$broad_clusters))
)
```

---

## 3. Check 1 — Marker Expression: Before vs. After Harmony

This plots one marker's expression on BOTH the uncorrected and Harmony-corrected UMAP side by side. If a marker looks equally diffuse on both, that's a pre-existing background/noise problem (not Harmony's doing). If it's crisp before and blurs out after, that's a real overcorrection candidate for that specific channel.

```{r marker-pre-post-function}

plot_marker_pre_post <- function(spe, marker) {

  umap_before <- reducedDim(spe, "UMAP_uncorrected")
  umap_after  <- reducedDim(spe, "UMAP_harmony")
  expr        <- assay(spe, "counts")[marker, ]

  df_before <- data.frame(UMAP1 = umap_before[, 1], UMAP2 = umap_before[, 2], expr = expr)
  df_after  <- data.frame(UMAP1 = umap_after[, 1],  UMAP2 = umap_after[, 2],  expr = expr)

  p_before <- ggplot(df_before, aes(UMAP1, UMAP2, color = expr)) +
    geom_point(size = 0.3, alpha = 0.3) +
    scale_color_viridis_c(option = "viridis") +
    labs(title = paste(marker, "- BEFORE Harmony"), color = marker) +
    theme_bw()

  p_after <- ggplot(df_after, aes(UMAP1, UMAP2, color = expr)) +
    geom_point(size = 0.3, alpha = 0.3) +
    scale_color_viridis_c(option = "viridis") +
    labs(title = paste(marker, "- AFTER Harmony"), color = marker) +
    theme_bw()

  p_before | p_after
}
```


These are markers flagged in earlier review: CD19 (diffuse on the Harmony UMAP, unclear if pre-existing), FoxP3 (unexpectedly elevated in the PMN-MDSC cluster's violin plot). EpCAM and CD68 included as a "should stay clean" contrast since they're mutually exclusive lineage markers with no prior noise concerns.

```{r marker-pre-post-run, fig.width=12, fig.height=5}

markers_to_check <- c("CD19", "FoxP3", "EpCAM", "CD68")

for (marker in markers_to_check) {
  print(plot_marker_pre_post(spe_combined, marker))
}
```

There seems to be potential spillover for CD19, as it appears to be diffuse even before Harmony batch correction.
---

## 4. Check 2 — Per-Cluster Patient Composition and Enrichment

```{r patient-composition}
# raw counts and proportions: how many/what % of each cluster's cells come from each patient

comp_counts <- table(cluster = meta_common$broad_clusters, patient = meta_common$patient_id)
comp_props  <- prop.table(comp_counts, margin = 1)  # row-normalized by cluster

# relabel rows with cell type names for readability
rownames(comp_counts) <- paste0(rownames(comp_counts), " - ", celltype_lookup[rownames(comp_counts)])
rownames(comp_props)  <- rownames(comp_counts)

pheatmap(
  comp_props,
  cluster_rows    = FALSE,
  cluster_cols    = FALSE,
  display_numbers = TRUE,
  number_format   = "%.2f",
  color           = colorRampPalette(brewer.pal(9, "YlOrRd"))(100),
  main            = "Cluster composition by patient (row-normalized %)",
  fontsize_number = 7,
  angle_col       = 45
)
```

Raw composition alone is misleading. A patient who simply contributes more cells overall will naturally show up more in every cluster. Instead compute an ENRICHMENT score: each patient's share
of a cluster, divided by that patient's share of the whole dataset.

Enrichment near 1 = patient is represented in this cluster about as much as expected. 
Enrichment >> 1 for ONE patient, with no obvious biological reason, is the signature of under-correction (that patient's batch effect is still driving its own sub-cluster).

```{r patient-enrichment}

overall_patient_frac <- prop.table(table(meta_common$patient_id))

enrichment <- sweep(comp_props, 2, overall_patient_frac[colnames(comp_props)], "/")

pheatmap(
  enrichment,
  cluster_rows    = FALSE,
  cluster_cols    = FALSE,
  display_numbers = TRUE,
  number_format   = "%.1f",
  color           = colorRampPalette(rev(brewer.pal(9, "RdBu")))(100),
  breaks          = seq(0, 4, length.out = 101),  # centered visually around 1 = expected
  main            = "Patient enrichment per cluster (1 = expected, >2 = flagged)",
  fontsize_number = 7,
  angle_col       = 45
)

# flat table of any (cluster, patient) combination with enrichment > 2
# AND a non-trivial number of cells (avoids flagging noise in tiny clusters)
enrichment_flags <- as.data.frame(as.table(enrichment)) |>
  rename(cluster = cluster, patient = patient, enrichment = Freq) |>
  left_join(
    as.data.frame(as.table(comp_counts)) |> rename(cluster = cluster, patient = patient, n_cells = Freq),
    by = c("cluster", "patient")
  ) |>
  filter(enrichment > 2, n_cells > 50) |>
  arrange(desc(enrichment))

cat("Flagged (cluster, patient) pairs with enrichment > 2x expected and > 50 cells:\n")
print(enrichment_flags)
```

---

## 5. Check 3 — LISI (Local Inverse Simpson's Index)

LISI is computed per cell on an embedding, using a metadata label. It estimates the "effective number" of distinct label values found in that cell's local neighborhood. For a batch label, higher LISI is better (patients well-mixed locally). For a cell-type/cluster label, LOWER LISI is better (biology staying separate locally). We compute both labels on BOTH embeddings so we can see the
before/after change directly.

```{r lisi-compute}

# Subsampled to keep this from thrashing memory on a shared node.
# 20% of cells, stratified by patient so no patient gets left out or
# over/under-represented relative to their share of the full dataset.

set.seed(42)

subsample_cells <- meta_common |>
  group_by(patient_id) |>
  slice_sample(prop = 0.20) |>
  pull(cell_id)

cat("Subsampled cells:", length(subsample_cells), "out of", nrow(meta_common),
   sprintf("(%.1f%%)\n", 100 * length(subsample_cells) / nrow(meta_common)))

pca_before_mat    <- reducedDim(spe_combined, "PCA_uncorrected")[subsample_cells, ]
harmony_after_mat <- reducedDim(spe_combined, "Harmony")[subsample_cells, ]

lisi_meta <- meta_common[subsample_cells, c("sample_id", "broad_clusters")]

lisi_before <- compute_lisi(pca_before_mat, lisi_meta, c("sample_id", "broad_clusters"))
lisi_after  <- compute_lisi(harmony_after_mat, lisi_meta, c("sample_id", "broad_clusters"))

lisi_summary <- data.frame(
  metric   = c("Batch LISI (sample_id) -- want HIGH", "Cluster LISI (broad_clusters) -- want LOW"),
  before   = c(median(lisi_before$sample_id), median(lisi_before$broad_clusters)),
  after    = c(median(lisi_after$sample_id),  median(lisi_after$broad_clusters))
)

knitr::kable(lisi_summary, digits = 2,
            caption = "Median LISI before (PCA_uncorrected) vs. after (Harmony), 20% stratified subsample")
```

Batch-LISI roughly doubled (patients mix well locally), which means Harmony was able to achieve strong batch correction without any evidence of overcorrection (cluster separation held steady or improved).

Cluster-LISI slightly decreased, which means cell types are separating more cleanly after Harmony. This is strong evidence that Harmony did not overcorrect. 

Note: A Cluster-LISI value approaching 1.0 means that a cell's neighborhood is (on average) almost entirely made up of its own cluster.

```{r lisi-plot, fig.width=10, fig.height=5}
lisi_long <- bind_rows(
  lisi_before |> mutate(stage = "Before Harmony"),
  lisi_after  |> mutate(stage = "After Harmony")
) |>
  pivot_longer(cols = c(sample_id, broad_clusters), names_to = "label_type", values_to = "lisi") |>
  mutate(
    label_type = recode(label_type,
                        sample_id = "Batch LISI (sample_id) -- want HIGH",
                        broad_clusters = "Cluster LISI (broad_clusters) -- want LOW"),
    stage = factor(stage, levels = c("Before Harmony", "After Harmony"))
  )

ggplot(lisi_long, aes(x = stage, y = lisi, fill = stage)) +
  geom_violin(trim = TRUE) +
  geom_boxplot(width = 0.15, outlier.size = 0.3, fill = "white") +
  facet_wrap(~ label_type, scales = "free_y") +
  theme_bw() +
  labs(title = "LISI distributions before vs. after Harmony", y = "LISI", x = NULL) +
  theme(legend.position = "none")
```

---

## 6. Check 4 — Cell-Type Composition (%) Across Disease State

This is the proportion-based version of the question the UMAP shape couldn't answer: even if all three disease states occupy the same overall UMAP footprint, do their cell-type PROPORTIONS differ the way we'd biologically expect (e.g. more Tumor Epithelial / Treg / M2 Macrophage in Higher-stage tissue)?

```{r celltype-by-disease-state, fig.width=10, fig.height=6}

meta_common$broad_celltype <- celltype_lookup[as.character(meta_common$broad_clusters)]

composition_by_state <- meta_common |>
  count(disease_state, broad_celltype) |>
  group_by(disease_state) |>
  mutate(pct = 100 * n / sum(n)) |>
  ungroup() |>
  mutate(broad_lineage = case_when(
    str_detect(broad_celltype, "T Cell|Treg") ~ "T Cell",
    str_detect(broad_celltype, "Macrophage|MDSC|DC|Mast") ~ "Myeloid",
    str_detect(broad_celltype, "Epithelial") ~ "Epithelial",
    str_detect(broad_celltype, "Plasma|Plasmablast") ~ "B/Plasma",
    str_detect(broad_celltype, "Fibroblast") ~ "Stromal",
    TRUE ~ "Other/Unresolved"
  ))

ggplot(composition_by_state, aes(x = disease_state, y = pct, fill = broad_lineage)) +
  geom_col(position = "stack") +
  labs(title = "Cell lineage composition (%) by disease state",
      x = "Disease State", y = "% of cells", fill = "Lineage") +
  theme_bw() +
  theme(legend.position = "right")
```


Same breakdown per patient, to spot-check whether any individual patient's composition looks like an outlier within their disease state group (useful alongside the enrichment check in Section 4).

```{r celltype-by-patient, fig.width=12, fig.height=6}

composition_by_patient <- meta_common |>
  count(patient_id, disease_state, broad_celltype) |>
  group_by(patient_id) |>
  mutate(pct = 100 * n / sum(n)) |>
  ungroup() |>
  mutate(broad_lineage = case_when(
    str_detect(broad_celltype, "T Cell|Treg") ~ "T Cell",
    str_detect(broad_celltype, "Macrophage|MDSC|DC|Mast") ~ "Myeloid",
    str_detect(broad_celltype, "Epithelial") ~ "Epithelial",
    str_detect(broad_celltype, "Plasma|Plasmablast") ~ "B/Plasma",
    str_detect(broad_celltype, "Fibroblast") ~ "Stromal",
    TRUE ~ "Other/Unresolved"
  ))

ggplot(composition_by_patient, aes(x = patient_id, y = pct, fill = broad_lineage)) +
  geom_col(position = "stack") +
  facet_wrap(~ disease_state, scales = "free_x", nrow = 1) +
  labs(title = "Cell lineage composition (%) by patient, grouped by disease state",
      x = "Patient", y = "% of cells", fill = "Lineage") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "right")
```
There's substantial patient-to-patient heterogeneity even after batch correction and clustering.

Patient 17 has a significant amount of Other/Unresolved cells, which makes sense given its faulty ROIs. 

Patient 10 (Benign) has way more Myeloid cells and way less T-cell/B plasma. 
---

## Notes

- **Check 1 (marker pre/post):** if a marker looks equally diffuse
  before and after Harmony, that's evidence of antibody
  background/spillover, not overcorrection -- goes back to the
  channel-level fixes discussed earlier (compensation, thresholding,
  or exclusion), not a batch-correction fix.
- **Check 2 (enrichment):** enrichment > 2x with a real cell count is
  the flag to act on. A small cluster showing high enrichment for a
  patient with very few total cells is less concerning than a large
  cluster dominated by one high-cell-count patient with no known
  biological reason (unlike the original Patient 12 case, which
  turned out to be a data-loading issue, not biology or an
  undercorrection).
- **Check 3 (LISI):** if the "after" batch-LISI barely improves over
  "before," Harmony isn't doing much (undercorrection). If cluster-LISI
  goes up substantially after Harmony (cell types becoming *less*
  locally distinct), that's the direct quantitative signature of
  overcorrection -- distinct biology is being blended along with the
  batch effect.
- **Check 4 (composition by disease state):** don't expect dramatic
  differences in overall UMAP shape -- look here instead for whether
  proportions shift the way tumor biology would predict. If Higher
  and Lower and Benign show nearly identical composition percentages
  across every cell type with no expected biological gradient, that's
  worth discussing with Dr. Campbell as a possible overcorrection
  signal at the composition level, separate from anything visible on
  the UMAP itself.
