---
title: 'All Patients: Broad Cell Type Clustering — Harmony V4'
author: "Tazein Shah"
date: "07-13-2026"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
subtitle: 20 Clusters
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)
set.seed(42)
```

## Overview

Broad cell type clustering on the harmonized combined SPE
(`spe_combined_harmony_r3.rds`). This Harmony run was computed on raw
counts directly (PCA on `counts`, no asinh, no cap/Z-score) before
Harmony. Clustering runs on the Harmony corrected embeddings
(analogous to Xi et al.'s PCA on top 2,000 variable features). The
cap + Z-score assay used here is for visualization only and is
computed fresh from raw counts; it was not part of how Harmony was
run. After inspecting the dot plot and UMAP, broad cell type labels
are assigned manually via a lookup table.

---

## 1. Load Libraries

```{r libraries}
library(imcRtools)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(scater)
library(scran)
library(bluster)
library(patchwork)
library(RColorBrewer)
```

---

## 2. Paths

All paths for this notebook are built off a single base path below.
Change `analysis_path` here if the project moves; everything else is
derived with `file.path()`.

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

# input
spe_path <- file.path(analysis_path, "batch_correction", "spe_combined_harmony_r4.rds")

# clustering outputs
clustering_path   <- file.path(analysis_path, "clustering")
intermediate_path <- file.path(clustering_path, "h4_k20_clusters", "harmony_20_clustered_intermediate.rds")
findmarkers_path  <- file.path(clustering_path, "h4_k20_clusters", "harmony_20_cluster_findmarkers.csv")
out_path          <- file.path(clustering_path, "h4_k20_clusters", "harmony_20_cluster_annot.rds")
```

---

## 3. Load Harmonized Data

```{r load-data}
spe <- readRDS(spe_path)

cat("Cells:", ncol(spe), "\n")
cat("Patients:", n_distinct(spe$patient_id), "\n")
cat("ROIs:", n_distinct(spe$sample_id), "\n")
cat("Channels:", nrow(spe), "\n")
cat("Reduced dims available:", paste(reducedDimNames(spe), collapse = ", "), "\n")
cat("Assays available:", paste(assayNames(spe), collapse = ", "), "\n")
```

---

## 4. Marker Sets

```{r markers}
# core markers for broad annotation and UMAP overlays
core_markers <- c("SMA", "CD4", "CD8a", "CD19", "CD15",
                  "CD14", "CD86", "LAMP3", "CD163", "EpCAM")
core_markers <- core_markers[core_markers %in% rownames(spe)]
cat("Core markers found:", paste(core_markers, collapse = ", "), "\n")
cat("Missing:", paste(setdiff(
  c("SMA", "CD4", "CD8a", "CD19", "CD15", "CD14", "CD86", "LAMP3", "CD163", "EpCAM"),
  rownames(spe)), collapse = ", "), "\n")

# full panel, used to investigate clusters that don't separate on core markers
full_markers <- c(
  "SMA", "CD4", "CD8a", "CD19", "CD15", "CD14", "CD86", "LAMP3",
  "Ki67", "MKI67", "CD138", "PD1", "PDL1", "FoxP3", "CD68", "CD163",
  "TIM3", "CD141", "CD11c", "CD56", "CD117", "EpCAM", "CD38",
  "CD45RO", "LAG3", "CD1c", "GATA3", "TTF1orCD16"
)
full_markers <- full_markers[full_markers %in% rownames(spe)]
cat("Full panel markers found:", length(full_markers), "\n")
cat("Missing:", paste(setdiff(
  c("SMA", "CD4", "CD8a", "CD19", "CD15", "CD14", "CD86", "LAMP3",
    "Ki67", "MKI67", "CD138", "PD1", "PDL1", "FoxP3", "CD68", "CD163",
    "TIM3", "CD141", "CD11c", "CD56", "CD117", "EpCAM", "CD38",
    "CD45RO", "LAG3", "CD1c", "GATA3", "TTF1orCD16"),
  rownames(spe)), collapse = ", "), "\n")
```

---

## 5. Preprocessing: 99th Percentile Cap + Z-score

```{r helpers}
cap_99 <- function(mat) {
  t(apply(mat, 1, function(x) {
    q99 <- quantile(x, 0.99)
    pmin(x, q99)
  }))
}

zscore_rows <- function(mat) {
  t(scale(t(mat)))
}
```

```{r preprocess}
mat_0100 <- counts(spe) |> cap_99() |> zscore_rows()
assay(spe, "scaled_0100") <- mat_0100

cat("Assays in SPE:", paste(assayNames(spe), collapse = ", "), "\n")
```

---

## 6. Louvain Clustering on Harmony Embeddings

Clustering runs on the Harmony corrected embeddings, which serve the
same role as PCA on top variable features in the Xi et al. approach.
Resolution 1.0 as per the preprint's first pass. The clustered SPE is
saved immediately after so clustering does not need to rerun on every
render.

```{r cluster}
if (file.exists(intermediate_path)) {
  cat("Loading existing clustered SPE from intermediate file.\n")
  spe <- readRDS(intermediate_path)
  cat("Number of broad clusters:", nlevels(spe$broad_clusters), "\n")
  print(table(spe$broad_clusters))
} else {
  set.seed(220225)

  harmony_mat <- reducedDim(spe, "Harmony")
  cat("Harmony embedding dimensions:", dim(harmony_mat), "\n")

  g     <- buildSNNGraph(t(harmony_mat), k = 20, type = "rank")
  clust <- igraph::cluster_louvain(g)
  spe$broad_clusters <- factor(clust$membership)

  cat("Number of broad clusters:", nlevels(spe$broad_clusters), "\n")
  print(table(spe$broad_clusters))

  # ensure the parent directory exists before saving
  dir.create(dirname(intermediate_path), recursive = TRUE, showWarnings = FALSE)
  saveRDS(spe, intermediate_path)
  cat("Saved clustered intermediate SPE to:", intermediate_path, "\n")
}
```

---

## 7. UMAP on Harmony Embeddings

`UMAP_harmony` already exists in this object from the batch correction
step, so it is reused here instead of recomputed.

```{r umap}
if ("UMAP_harmony" %in% reducedDimNames(spe)) {
  cat("Reusing existing UMAP_harmony from batch correction step.\n")
} else {
  set.seed(220225)
  spe <- runUMAP(spe, dimred = "Harmony", name = "UMAP_harmony")
  cat("UMAP computed.\n")
}
```

---

## 8. Subset for Visualization

20,000 cells per ROI cap to keep plots manageable.

```{r subset}
set.seed(220225)
cells_per_roi  <- split(seq_len(ncol(spe)), spe$sample_id)
sampled_cells  <- unlist(lapply(cells_per_roi, function(idx) {
  sample(idx, min(20000, length(idx)))
}))
spe_sub <- spe[, sampled_cells]
cat("Cells in subset:", ncol(spe_sub), "\n")
```

---

## 9. UMAP Colored by Patient

```{r umap-patient, fig.width=10, fig.height=7}
n_patients    <- n_distinct(spe$patient_id)
patient_cols  <- setNames(
  colorRampPalette(brewer.pal(9, "Set1"))(n_patients),
  unique(spe$patient_id)
)

p_patient <- dittoDimPlot(spe_sub, var = "patient_id",
                          reduction.use = "UMAP_harmony", size = 0.2) +
  scale_color_manual(values = patient_cols) +
  ggtitle("UMAP colored by Patient ID (Harmony, no asinh)")
print(p_patient)
```

## 11. UMAP Colored by Broad Cluster

```{r umap-clusters, fig.width=10, fig.height=7}
p_clusters <- dittoDimPlot(spe_sub, var = "broad_clusters",
                           reduction.use = "UMAP_harmony",
                           size = 0.2, do.label = TRUE, labels.size = 3) +
  ggtitle("UMAP colored by Broad Louvain Cluster") +
  theme(legend.position = "right")
print(p_clusters)
```

---

## 12. UMAP Per Core Marker

Used to visually match clusters to known marker expression before
annotating.

```{r umap-markers, fig.width=8, fig.height=6}
for (marker in core_markers) {
  p <- dittoDimPlot(spe_sub, var = marker,
                    reduction.use = "UMAP_harmony",
                    assay = "scaled_0100", size = 0.2) +
    scale_color_viridis(name = marker) +
    ggtitle(paste("UMAP:", marker))
  print(p)
}
```

---

## 13. Dot Plot: Core Marker Expression per Cluster

The primary tool for annotating clusters. Each row is a marker, each
column is a cluster. Look for which clusters show high expression of
the core lineage markers.

```{r make-dotplot}
make_dotplot <- function(spe, assay_name, cluster_col, marker_set, title) {
  expr_mat   <- assay(spe, assay_name)[marker_set, ]
  cluster_id <- spe[[cluster_col]]

  mean_expr <- sapply(levels(cluster_id), function(cl) {
    idx <- which(cluster_id == cl)
    rowMeans(expr_mat[, idx, drop = FALSE])
  })

  global_med <- apply(expr_mat, 1, median)
  frac_expr  <- sapply(levels(cluster_id), function(cl) {
    idx <- which(cluster_id == cl)
    rowMeans(expr_mat[, idx, drop = FALSE] > global_med)
  })

  mean_df <- as.data.frame(mean_expr) |>
    rownames_to_column("marker") |>
    pivot_longer(-marker, names_to = "cluster", values_to = "mean_exprs")

  frac_df <- as.data.frame(frac_expr) |>
    rownames_to_column("marker") |>
    pivot_longer(-marker, names_to = "cluster", values_to = "frac_expressing")

  plot_df <- left_join(mean_df, frac_df, by = c("marker", "cluster"))
  plot_df$cluster <- factor(plot_df$cluster, levels = levels(cluster_id))

  marker_order <- rownames(mean_expr)[hclust(dist(mean_expr))$order]
  plot_df$marker <- factor(plot_df$marker, levels = marker_order)

  ggplot(plot_df, aes(x = cluster, y = marker)) +
    geom_point(aes(size = frac_expressing, color = mean_exprs)) +
    scale_color_viridis(option = "magma", name = "Mean\nZ-score expr") +
    scale_size_continuous(range = c(0.5, 6), name = "Fraction\nexpressing") +
    labs(title = title, x = "Cluster", y = "Marker") +
    theme_bw(base_size = 12) +
    theme(
      axis.text.x      = element_text(angle = 0, hjust = 0.5),
      axis.text.y      = element_text(size = 9),
      panel.grid.major = element_line(color = "grey90"),
      legend.position  = "right"
    )
}
```

```{r dotplot-core, fig.width=16, fig.height=6}
p_dot_core <- make_dotplot(spe, "scaled_0100", "broad_clusters",
                           core_markers,
                           "Harmony (no asinh): Core Marker Expression per Broad Louvain Cluster")
print(p_dot_core)
```

---

## 14. Full Panel Dot Plot for Low-Signal Clusters

Several clusters may not show a clear standout marker on the 10 core
markers above, this can happen for macrophage/myeloid populations
where markers like CD14, CD86, CD163 are broadly and continuously
expressed rather than sharply spiking in one cluster, or because the
cluster is actually separated by markers outside the core panel
(e.g. PD1, Ki67, FoxP3, CD45RO), since this Harmony run was computed
on raw counts across the full panel, not just the 10 core markers.
Borderline or low quality segmented cells are also possible. Use this
expanded dot plot to investigate clusters that look flat above before
calling them "Unknown."

```{r dotplot-full, fig.width=18, fig.height=10}
p_dot_full <- make_dotplot(spe, "scaled_0100", "broad_clusters",
                           full_markers,
                           "Harmony (no asinh): Full Panel Marker Expression per Broad Louvain Cluster")
print(p_dot_full)
```

---

## 15. Statistical Marker Discovery (findMarkers)

For clusters with no obvious standout marker on either dot plot,
`findMarkers()` ranks markers by how distinctly they separate that
cluster from all others, accounting for variance rather than just
mean difference. This can surface subtle but consistent shifts that
don't show up clearly in a Z-score dot plot averaged across thousands
of cells.

```{r findmarkers}
markers_de <- findMarkers(spe,
                          groups     = spe$broad_clusters,
                          assay.type = "scaled_0100",
                          subset.row = full_markers)

# save all clusters to CSV
all_markers <- lapply(levels(spe$broad_clusters), function(cl) {
  df <- as.data.frame(markers_de[[cl]])
  data.frame(
    cluster = cl,
    marker  = rownames(df),
    Top     = df$Top,
    FDR     = df$FDR,
    logFC   = df$summary.logFC
  )
})

all_markers_df <- do.call(rbind, all_markers)

write.csv(all_markers_df, findmarkers_path, row.names = FALSE)

# print top 5 positive markers per cluster
df <- read.csv(findmarkers_path)

# summary table: top positive marker per cluster at a glance
top1_summary <- do.call(rbind, lapply(levels(spe$broad_clusters), function(cl) {
  sub <- df[df$cluster == cl & df$logFC > 1.0, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  if (nrow(sub) == 0) {
    data.frame(cluster = cl, top_markers = "none above logFC 1.0")
  } else {
    data.frame(cluster = cl,
               top_markers = paste(head(sub$marker, 5), collapse = ", "))
  }
}))

cat("Top positive markers (logFC > 1.0) per cluster:\n")
print(knitr::kable(top1_summary, col.names = c("Cluster", "Top Positive Markers")))

cat("\n\nDetailed tables:\n")

# detailed table per cluster, positive logFC only
for (cl in levels(spe$broad_clusters)) {
  sub <- df[df$cluster == cl & df$Top <= 5 & df$logFC > 1.0, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  if (nrow(sub) == 0) {
    cat("Cluster", cl, ": no markers with logFC > 1.0 in top 5\n\n")
  } else {
    print(knitr::kable(sub[, c("marker", "Top", "FDR", "logFC")],
                       caption = paste("Cluster", cl),
                       digits  = 3))
  }
}
```
--- 

## 15b. Trying to Find "Unknown" Clusters

```{r}
for (cl in c("3","5","21","22")) {
  sub <- df[df$cluster == cl, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  print(knitr::kable(sub[, c("marker", "Top", "FDR", "logFC")],
                     caption = paste("Cluster", cl, "(all markers)"),
                     digits  = 3))
}
```

## 16. Diagnostic: Area and Patient Composition for Ambiguous Clusters

```{r diagnostic-unknown, fig.width=10, fig.height=5}
ambiguous_clusters <- c("3","5","21","22")

# cell area boxplot
df_area <- data.frame(
  area    = spe$area,
  cluster = as.character(spe$broad_clusters)
)

# reference: median area across all annotated clusters
df_area_ref <- df_area[!df_area$cluster %in% ambiguous_clusters, ]
ref_median   <- median(df_area_ref$area)

p_area <- ggplot(df_area[df_area$cluster %in% ambiguous_clusters, ],
                 aes(x = cluster, y = area, fill = cluster)) +
  geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.3) +
  geom_hline(yintercept = ref_median, linetype = "dashed", color = "grey40") +
  annotate("text", x = 0.6, y = ref_median + 5,
           label = "median (annotated clusters)", size = 3, color = "grey40") +
  labs(title = "Cell Area Distribution for Ambiguous Clusters",
       x = "Cluster", y = "Cell Area (pixels)") +
  theme_bw() +
  theme(legend.position = "none")
print(p_area)
```

```{r diagnostic-patient, fig.width=12, fig.height=5}
# patient composition per ambiguous cluster
comp_df <- as.data.frame(
  table(cluster   = as.character(spe$broad_clusters),
        patient   = spe$patient_id)
)
comp_df <- comp_df[comp_df$cluster %in% ambiguous_clusters & comp_df$Freq > 0, ]

p_comp <- ggplot(comp_df, aes(x = patient, y = Freq, fill = patient)) +
  geom_col() +
  facet_wrap(~ cluster, scales = "free_y", nrow = 2) +
  labs(title = "Patient Composition for Ambiguous Clusters",
       x = "Patient", y = "Cell Count") +
  theme_bw() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        legend.position = "none")
print(p_comp)
```

## 17. Manual Broad Annotation

After inspecting the dot plots and `findMarkers()` output above, fill
in the lookup table below. Map each cluster number to one of:

```{r annotate, echo=TRUE}
broad_annotation <- c(
  "1"  = "cDC1", #CD141 canonical cDC1 marker, TIM3 suggests checkpoint expressing subset
  "2"  = "Stroma", #SMA marks myofibroblasts/pericytes, PDL1/TIM3 checkpoint markers, CD19 likely spillover
  "3"  = "Low Stain", #no markers clear logFC 1.0, matches k150 cluster 2 suppression pattern
  "4"  = "Mixed DC/Mast/Treg", #LAMP3 mature migratory DC, CD117 mast cell, PD1/FoxP3 Treg or exhausted T contact, TTF1orCD16 near tumor
  "5"  = "Epithelial", #low confidence, only EpCAM and FoxP3 clear at logFC 0.5, FoxP3 likely Treg admixture
  "6"  = "CD8 T Cell", #CD8a/CD45RO core, TIM3/FoxP3 exhaustion or Treg admixture, CD138 likely spillover
  "7"  = "Myeloid", #CD14/CD11c/CD68/CD86 monocyte macrophage markers, MKI67 proliferation
  "8"  = "Tumor Epithelial", #TTF1orCD16 and EpCAM core tumor signature, FoxP3/CD15/CD1c likely infiltrating immune cells
  "9"  = "Memory T/Plasma Cell", #CD45RO and CD138 suggest boundary between memory T zone and plasma cell niche
  "10" = "Mast Cell", #CD117 defining mast cell marker, CD11c/TIM3 suggest myeloid DC proximity
  "11" = "Treg", #FoxP3 defines Treg, CD15/PDL1/TTF1orCD16 suggest tumor associated granulocyte proximity
  "12" = "Granulocyte", #CD15 defines granulocyte lineage, CD45RO/PDL1/CD138 mixed lymphoid signal
  "13" = "B Cell/Plasmablast", #matches CD38/FoxP3 leaning B cell subtype from notebook 07
  "14" = "Monocyte", #CD14 alone, clean monocyte signature
  "15" = "CD4 T Cell", #CD4/CD45RO memory helper T cells, TIM3/FoxP3 exhaustion or Treg identity, CD138 likely plasma cell admixture
  "16" = "Plasma Cell", #CD138/CD19 define lineage, CD86/PDL1 suggest activated state
  "17" = "Memory T Cell", #CD45RO defines memory T cells, CD1c admixture suggests nearby cDC2
  "18" = "cDC2", #CD1c alone, clean cDC2 signature
  "19" = "Proliferating Cells", #MKI67 only, no lineage marker to assign identity
  "20" = "NK Cell", #CD56 alone, clean NK signature
  "21" = "T Cells", #low confidence
  "22" = "Low Stain" #same suppression pattern as cluster 3, only LAG3 clears zero at rank 2, not enough to call a distinct population
)

if (length(broad_annotation) > 0) {
  spe$broad_celltype <- broad_annotation[as.character(spe$broad_clusters)]
  cat("Annotation applied.\n")
  print(table(spe$broad_celltype, useNA = "ifany"))
} else {
  cat("No annotation provided yet. Fill in broad_annotation above after inspecting plots.\n")
}


if (length(broad_annotation) > 0) {
  spe$broad_celltype <- broad_annotation[as.character(spe$broad_clusters)]
  cat("Annotation applied.\n")
  print(table(spe$broad_celltype))
} else {
  cat("No annotation provided yet. Fill in broad_annotation above after inspecting plots.\n")
}
```
---

## 18. UMAP Colored by Broad Cell Type

Run this chunk after filling in the annotation above.

```{r umap-celltype, fig.width=10, fig.height=7}
if ("broad_celltype" %in% colnames(colData(spe))) {
  spe_sub$broad_celltype <- spe$broad_celltype[match(colnames(spe_sub), colnames(spe))]

  p_celltype <- dittoDimPlot(spe_sub, var = "broad_celltype",
                             reduction.use = "UMAP_harmony",
                             size = 0.2, do.label = TRUE, labels.size = 3) +
    ggtitle("UMAP colored by Broad Cell Type") +
    theme(legend.position = "right")
  print(p_celltype)
} else {
  cat("Run chunk 17 first to apply annotations.\n")
}
```

---

## 19. Save Annotated SPE

```{r save}
if ("broad_celltype" %in% colnames(colData(spe))) {
  dir.create(dirname(out_path), recursive = TRUE, showWarnings = FALSE)
  saveRDS(spe, out_path)
  cat("Saved annotated SPE to:", out_path, "\n")
} else {
  cat("Annotation not yet applied. Fill in chunk 17 before saving.\n")}
```

---

## Notes

- All paths in this notebook are built from a single `analysis_path`
  base variable in chunk 2, using `file.path()`, matching the pattern
  used in the other clustering notebooks
- Input `spe_combined_harmony_r3.rds` has Harmony embeddings computed
  via PCA on raw `counts`, no asinh and no cap/Z-score; the cap +
  Z-score assay used in this notebook is for visualization only and
  was not part of how Harmony was run
- Because raw counts can have a long right tail, this PCA/Harmony run
  is more exposed to outlier driven distortion than the asinh or
  cap+Z-score versions; worth keeping in mind when comparing UMAPs
  and cluster quality across the three conditions
- Preprocessing: counts already 0 to 100 filtered; 99th percentile cap + Z-score applied and stored as `scaled_0100` assay
- Clustering uses Harmony corrected embeddings (`reducedDim(spe, "Harmony")`),
  which play the same role as PCA on top 2,000 variable features in Xi et al.
- Dot plot and UMAP marker overlays use `scaled_0100` assay
- `igraph::cluster_louvain` does not accept a resolution parameter; if the
  number of clusters is too coarse or too fine, switch to
  `igraph::cluster_leiden(g, resolution_parameter = 1.0)` from the `leiden`
  package
- Doublet like clusters (hybrid marker expression across lineages) should be
  assigned "Unknown" and flagged for removal before subclustering
- Note: rerunning this notebook may shuffle Louvain cluster numbers; if you
  need to preserve a finalized annotation across reruns, save the clustered
  SPE before annotating and load it separately
