---
title: "B Cell Cluster Comparison Across Resolutions (k20 / k150 / k200)"
author: "Tazein Shah"
date: "07-08-2026"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(SpatialExperiment)
library(imcRtools)
library(scran)
library(scater)
library(bluster)
library(dittoSeq)
library(dplyr)
library(tidyverse)
library(viridis)
library(patchwork)
library(ggridges)
```

## Purpose

B cell and plasma cell related clusters have been annotated at all three resolutions: clusters 9 and 11 at k150, clusters 3 and 15 at k20, and cluster 5 at k200. This notebook compares what is being expressed across these clusters, and any subtypes, to see whether the same broad population splits, merges, or stays consistent across resolutions and whether the marker profile supports the subtype calls already made at each one.

This follows the same cell level, ID matched comparison approach as `06_unknown_cluster_diagnostics.Rmd`, but the focus here is on expression content within B cell clusters rather than on resolving a single Unknown cluster.

Note: these are post segmentation SPE objects, so no additional filtering or normalization is applied here. All marker expression in this notebook is on raw counts, not asinh transformed.

Order of operations: confirm the cluster IDs with CD19 first, run `findMarkers` on those clusters right away to see the full profile (not just CD19), and only then move on to the cross resolution consistency checks, UMAP, ridge plots, and composition diagnostics.

## Load the SPE object and cluster assignments

```{r load-data}
clustering_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/clustering"

spe <- readRDS("/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/batch_correction/spe_combined_harmony_r3.rds")

k20_spe  <- readRDS(file.path(clustering_path, "k20_clusters",  "harmony_20_cluster_annot.rds"))
k150_spe <- readRDS(file.path(clustering_path, "k150_clusters", "harmony_150_cluster_annot.rds"))
k200_spe <- readRDS(file.path(clustering_path, "k200_clusters", "harmony_200_cluster_annot.rds"))

k20_vec  <- setNames(colData(k20_spe)$broad_clusters,  colnames(k20_spe))
k150_vec <- setNames(colData(k150_spe)$broad_clusters, colnames(k150_spe))
k200_vec <- setNames(colData(k200_spe)$broad_clusters, colnames(k200_spe))

setequal(names(k20_vec), colnames(spe))
setequal(names(k150_vec), colnames(spe))
setequal(names(k200_vec), colnames(spe))

k20_clusters  <- k20_vec[colnames(spe)]
k150_clusters <- k150_vec[colnames(spe)]
k200_clusters <- k200_vec[colnames(spe)]

stopifnot(all(names(k20_clusters) == colnames(spe)))
stopifnot(all(names(k150_clusters) == colnames(spe)))
stopifnot(all(names(k200_clusters) == colnames(spe)))

cluster_compare <- data.frame(
  cell_id = colnames(spe),
  k20  = as.character(k20_clusters),
  k150 = as.character(k150_clusters),
  k200 = as.character(k200_clusters),
  patient_id = spe$patient_id,
  sample_id  = spe$sample_id,
  area = spe$area,
  stringsAsFactors = FALSE
)

spe$k20  <- as.character(k20_clusters)
spe$k150 <- as.character(k150_clusters)
spe$k200 <- as.character(k200_clusters)
```

## Plot helper

`dittoDotPlot` needs at least two markers, since it is built as a marker by group bubble grid and errors out with just one. `bcell_markers` starts as just CD19, so this helper falls back to a plain box plot across the same groups whenever there is only a single marker, and switches to the dot plot once more markers are added. No jitter, since with the full cell counts here it just overplots and doesn't add information beyond what the box already shows. Defined once here; every plot chunk below calls this same function rather than redefining it.

```{r plot-helper}
plot_marker_by_cluster <- function(spe_obj, markers, group_col, assay = "counts", title = "") {
  # cluster IDs are stored as characters, so a default sort puts "10" and "11"
  # before "2". Reorder the factor levels numerically before plotting so the
  # x axis reads 1, 2, 3 ... instead of 1, 10, 11, 12 ...
  cluster_vals <- colData(spe_obj)[[group_col]]
  ordered_levels <- as.character(sort(as.numeric(unique(cluster_vals))))
  colData(spe_obj)[[group_col]] <- factor(as.character(cluster_vals), levels = ordered_levels)

  if (length(markers) >= 2) {
    dittoDotPlot(spe_obj, vars = markers, group.by = group_col, assay = assay) +
      ggtitle(title) +
      theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
  } else {
    dittoPlot(spe_obj, var = markers, group.by = group_col, plots = c("boxplot"),
              assay = assay, legend.show = FALSE) +
      ggtitle(title) +
      theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
  }
}
```

## Step 1: Confirm B cell / plasma cell clusters at k20, k150, and k200 with CD19

CD19 is the anchor marker for B cells on this panel (CD20 is absent). Each box plot below maps every cluster at that resolution, not just the candidate B cell ones, so you can see how the candidates compare to everything else before trusting the labels.

```{r cd19-boxplot-k20, fig.width=9, fig.height=6}
print(plot_marker_by_cluster(spe, "CD19", "k20", title = "CD19, k20 clusters"))
```

```{r cd19-boxplot-k150, fig.width=9, fig.height=6}
print(plot_marker_by_cluster(spe, "CD19", "k150", title = "CD19, k150 clusters"))
```

```{r cd19-boxplot-k200, fig.width=9, fig.height=6}
print(plot_marker_by_cluster(spe, "CD19", "k200", title = "CD19, k200 clusters"))
```

## Step 2: Define the B cell / plasma cell cluster IDs at each resolution

```{r define-bcell-clusters}
# k150 is fixed from the finalized broad annotation
# 9 = Plasma Cell / Activated B Cell, 11 = Plasma Cell / B Cell
k150_bcell_clusters <- c("9", "11")

# k20 broad annotation
# 3 = Plasmablast / Activated B Cell, 15 = Plasma Cell
k20_bcell_clusters <- c("3", "15")

# k200 broad annotation
# 5 = Plasma Cell / B Cell
k200_bcell_clusters <- c("5")

cluster_compare <- cluster_compare %>%
  mutate(
    is_bcell_k20  = k20  %in% k20_bcell_clusters,
    is_bcell_k150 = k150 %in% k150_bcell_clusters,
    is_bcell_k200 = k200 %in% k200_bcell_clusters
  )

spe_k150_bcell <- spe[, spe$k150 %in% k150_bcell_clusters]
spe_k20_bcell  <- spe[, spe$k20  %in% k20_bcell_clusters]
spe_k200_bcell <- spe[, spe$k200 %in% k200_bcell_clusters]
```

## Step 3: findMarkers results per resolution, right after the box plot check

Rather than recomputing `findMarkers`, this reads in the CSVs already generated for each k sweep run and filters down to just the B cell cluster rows. Doing this right after Step 1 gives the full panel profile for each candidate cluster while the CD19 result is still fresh, rather than waiting until later in the notebook.

```{r findmarkers-per-resolution}
# Read in the findMarkers CSVs already generated for each k sweep run,
# rather than recomputing findMarkers here. Each CSV has columns:
# cluster, marker, Top, FDR, logFC
k150_markers_csv <- read.csv(file.path(clustering_path, "k150_clusters", "harmony_150_cluster_findmarkers.csv"), stringsAsFactors = FALSE)
k20_markers_csv  <- read.csv(file.path(clustering_path, "k20_clusters",  "harmony_20_cluster_findmarkers.csv"), stringsAsFactors = FALSE)
k200_markers_csv <- read.csv(file.path(clustering_path, "k200_clusters", "harmony_200_cluster_findmarkers.csv"), stringsAsFactors = FALSE)

# Pull out just the B cell / plasma cell cluster rows, positive logFC only,
# and sort so the top markers per cluster are easy to read
filter_bcell_markers <- function(markers_csv, bcell_ids) {
  markers_csv %>%
    filter(as.character(cluster) %in% bcell_ids, logFC > 0) %>%
    arrange(cluster, desc(logFC)) %>%
    group_by(cluster) %>%
    slice_head(n = 10) %>%
    ungroup()
}

markers_k150_bcell <- filter_bcell_markers(k150_markers_csv, k150_bcell_clusters)
markers_k20_bcell  <- filter_bcell_markers(k20_markers_csv,  k20_bcell_clusters)
markers_k200_bcell <- filter_bcell_markers(k200_markers_csv, k200_bcell_clusters)

print(markers_k150_bcell)
print(markers_k20_bcell)
print(markers_k200_bcell)
```

If CD38 and CD19 are consistently the top two markers, and something unexpected (for example FoxP3) also shows up, check whether it appears across all three resolutions before deciding whether it is a real spatial contamination pattern or a resolution specific artifact.

## Step 4: How consistently do the same cells fall into B cell clusters across resolutions

Raw counts per resolution don't tell you whether it's the same cells being called B cell each time. This checks actual cell level overlap, plus subtype level agreement between k150 and k20 given the two distinct subtypes found in Step 3 (CD38/FoxP3 leaning: k150 cluster 9, k20 cluster 3; CD138/CD15 leaning: k150 cluster 11, k20 cluster 15).

```{r bcell-overlap-counts}
n_k20  <- sum(cluster_compare$is_bcell_k20)
n_k150 <- sum(cluster_compare$is_bcell_k150)
n_k200 <- sum(cluster_compare$is_bcell_k200)

n_union <- sum(cluster_compare$is_bcell_k20 | cluster_compare$is_bcell_k150 | cluster_compare$is_bcell_k200)
n_intersect <- sum(cluster_compare$is_bcell_k20 & cluster_compare$is_bcell_k150 & cluster_compare$is_bcell_k200)

cat("k20 B cell candidate cells: ", n_k20, "\n")
cat("k150 B cell cells: ", n_k150, "\n")
cat("k200 B cell candidate cells: ", n_k200, "\n\n")

cat("Cells called B cell in at least one resolution (union): ", n_union, "\n")
cat("Cells called B cell in all three resolutions (intersection): ", n_intersect, "\n")
cat("Percent of the union that all three agree on: ", round(n_intersect / n_union * 100, 1), "%\n\n")

# Full breakdown, not just the two extremes above
print(
  cluster_compare %>%
    count(is_bcell_k20, is_bcell_k150, is_bcell_k200) %>%
    arrange(desc(n)) %>%
    mutate(pct_of_union = round(n / n_union * 100, 1))
)
```

A high intersection relative to the union supports a stable population that all three resolutions agree on. A large block that is TRUE at only one or two resolutions means that resolution is pulling in extra cells the others don't, or missing cells the others include, rather than all three describing the same population.

### Subtype level agreement between k150 and k20

```{r bcell-subtype-overlap}
subtype_compare <- cluster_compare %>%
  filter(is_bcell_k150, is_bcell_k20) %>%
  mutate(
    subtype_k150 = case_when(
      k150 == "9"  ~ "CD38_FoxP3",
      k150 == "11" ~ "CD138_CD15",
      TRUE ~ NA_character_
    ),
    subtype_k20 = case_when(
      k20 == "3"  ~ "CD38_FoxP3",
      k20 == "15" ~ "CD138_CD15",
      TRUE ~ NA_character_
    )
  )

print(
  subtype_compare %>%
    count(subtype_k150, subtype_k20) %>%
    arrange(desc(n)) %>%
    mutate(pct = round(n / sum(n) * 100, 1))
)
```

If most cells land on the diagonal (CD38_FoxP3 matching CD38_FoxP3, CD138_CD15 matching CD138_CD15), that means k150 and k20 agree not just on "this is a B cell" but on which of the two subtypes each cell belongs to, which is a stronger result than the broad overlap above. Off diagonal cells are ones where the two resolutions disagree on subtype specifically, worth a closer look if that number is large.

### Is the 80% agreement spread evenly, or concentrated in one subtype

80% three way agreement isn't unusual for unsupervised clustering across different resolutions on its own, but it's worth checking whether that 20% disagreement is spread evenly across the population or concentrated somewhere specific.

```{r pairwise-overlap}
jaccard <- function(a, b) sum(a & b) / sum(a | b)

cat("k150 vs k20 Jaccard: ", round(jaccard(cluster_compare$is_bcell_k150, cluster_compare$is_bcell_k20), 3), "\n")
cat("k150 vs k200 Jaccard: ", round(jaccard(cluster_compare$is_bcell_k150, cluster_compare$is_bcell_k200), 3), "\n")
cat("k20 vs k200 Jaccard: ", round(jaccard(cluster_compare$is_bcell_k20, cluster_compare$is_bcell_k200), 3), "\n")
```

k150 and k200 agree with each other almost perfectly, while k20 disagrees with both by a similar, much larger margin. That means k20 is the outlier here, not k200, so the more informative check is where k150 and k200 agree but k20 breaks from both, rather than the other way around.

```{r disagreement-by-subtype}
# Cells both k150 and k200 agree are B cell, but k20 does not include
disagree_k20 <- cluster_compare %>%
  filter(is_bcell_k150, is_bcell_k200, !is_bcell_k20) %>%
  mutate(
    subtype_k150 = case_when(
      k150 == "9"  ~ "CD38_FoxP3",
      k150 == "11" ~ "CD138_CD15",
      TRUE ~ NA_character_
    )
  )

cat("Cells k150 and k200 both call B cell, but k20 does not: ", nrow(disagree_k20), "\n\n")

# print() is required here, not just relying on auto-print, since this
# notebook gets run via source() (per the imcRtools/gdal conflict on SCC),
# and source() defaults to print.eval = FALSE, so bare expressions like a
# piped tibble render nothing unless explicitly printed.
print(
  disagree_k20 %>%
    count(subtype_k150) %>%
    mutate(pct = round(n / sum(n) * 100, 1))
)

cat("\nWhere do these cells actually land in k20:\n")
print(disagree_k20 %>% count(k20) %>% arrange(desc(n)))
```

```{r extra-in-k20}
# The reverse direction: cells k20 calls B cell that the other two do not.
# This distinguishes k20 missing cells the others catch (checked above) from
# k20 pulling in extra cells the others exclude, which are different failure
# modes for a coarser resolution to make.
extra_k20 <- cluster_compare %>%
  filter(is_bcell_k20, !is_bcell_k150, !is_bcell_k200)

cat("Cells k20 calls B cell that k150 and k200 do not: ", nrow(extra_k20), "\n\n")

cat("Where do these cells land in k150 and k200:\n")
print(extra_k20 %>% count(k150) %>% arrange(desc(n)))
print(extra_k20 %>% count(k200) %>% arrange(desc(n)))
```

If `disagree_k20` is heavily skewed toward `CD38_FoxP3`, that supports k20 under capturing the weaker, CD19 lower subtype rather than randomly dropping cells from both subtypes equally. Given k20 has only 20 total clusters compared to 150 and 200, it has less room to draw a precise boundary, so I would treat the near identical k150/k200 agreement as the more trustworthy signal here and use k20 mainly as a coarse sanity check rather than an equal partner in this specific comparison.

## Step 5: Where do k150 B cell cells land at k20 and k200

```{r crosstab-bcell}
print(cluster_compare %>% filter(is_bcell_k150) %>% count(k20) %>% arrange(desc(n)))
print(cluster_compare %>% filter(is_bcell_k150) %>% count(k200) %>% arrange(desc(n)))
```

## Step 6: UMAP check across resolutions

```{r umap-bcell-check, fig.width=7, fig.height=5}
# Color by subtype rather than raw cluster ID, so each panel only shows the
# CD38/FoxP3 vs CD138/CD15 split (k200 gets one merged label since it never
# split into two), instead of every cluster the bcell_any union happens to
# touch. No do.label, since that's what was drawing the boxed cluster number
# callouts.
spe$subtype_k150 <- case_when(
  spe$k150 == "9"  ~ "CD38_FoxP3",
  spe$k150 == "11" ~ "CD138_CD15",
  TRUE ~ NA_character_
)

spe$subtype_k20 <- case_when(
  spe$k20 == "3"  ~ "CD38_FoxP3",
  spe$k20 == "15" ~ "CD138_CD15",
  TRUE ~ NA_character_
)

spe$subtype_k200 <- case_when(
  spe$k200 == "5" ~ "CD38_FoxP3_CD138_CD15_merged",
  TRUE ~ NA_character_
)

p1 <- dittoDimPlot(spe, var = "subtype_k20", reduction.use = "UMAP_harmony",
                    cells.use = !is.na(spe$subtype_k20), size = 0.3, do.label = FALSE) +
  ggtitle("k20 B cell subtypes")

p2 <- dittoDimPlot(spe, var = "subtype_k150", reduction.use = "UMAP_harmony",
                    cells.use = !is.na(spe$subtype_k150), size = 0.3, do.label = FALSE) +
  ggtitle("k150 B cell subtypes")

p3 <- dittoDimPlot(spe, var = "subtype_k200", reduction.use = "UMAP_harmony",
                    cells.use = !is.na(spe$subtype_k200), size = 0.3, do.label = FALSE) +
  ggtitle("k200 B cell subtypes")

print(p1)
print(p2)
print(p3)
```

### Where the k150/k200 vs k20 disagreement cells sit relative to Treg

`disagree_k20` (built in Step 4) marks the cells k150 and k200 both call B cell related but k20 does not, 63% of which land in k20's own Treg cluster. Plotting those cells directly on the UMAP, against the B cell subtype region and k20's Treg cluster for context, is a more direct visual test of whether that overlap is a real boundary/spillover zone or scattered noise than the earlier continuous FoxP3 overlay was.

```{r umap-disagreement-check, fig.width=7, fig.height=5}
spe$is_disagree <- ifelse(colnames(spe) %in% disagree_k20$cell_id, "disagree", "agree")

# context cells: the B cell subtype region plus k20's Treg cluster, so the
# disagreement points have something to sit against, everything else stays
# grey rather than getting its own color category
context_cells <- !is.na(spe$subtype_k150) | spe$k20 == "11"

print(
  dittoDimPlot(spe, var = "is_disagree", reduction.use = "UMAP_harmony",
               cells.use = context_cells, size = 0.4,
               color.panel = c("grey50", "red")) +
    ggtitle("Disagreement cells (k150 & k200 call B cell, k20 does not)")
)
```

If the red disagreement points cluster tightly right where the B cell subtype region meets k20's Treg cluster, that supports a real boundary population or spillover zone. If they're scattered throughout the B cell region instead, that argues against a clean spatial explanation and more toward individual cell level doublets rather than a contiguous boundary zone.

## Step 7: Ridge plots of marker distributions by resolution

Ridge plots preferred over heatmaps for distribution level comparisons. `bcell_markers` now includes CD38, FoxP3, CD138, and CD15 alongside CD19, so each plot facets by marker to show whether FoxP3 and CD15 are bimodal within their respective clusters or a smooth low level smear, not just the mean captured by the findMarkers logFC values in Step 3.

```{r ridge-marker-list}
# Now includes the markers that actually define each subtype (CD38/FoxP3 for
# the k150 cluster 9 / k20 cluster 3 subtype, CD138/CD15 for the k150
# cluster 11 / k20 cluster 15 subtype), not just CD19, so the ridge plots can
# show whether FoxP3 and CD15 are bimodal or a smooth low level smear within
# each cluster.
bcell_markers <- c("CD19", "CD38", "FoxP3", "CD138", "CD15")
```

```{r ridge-k150, fig.width=7, fig.height=14}
expr_long_k150 <- as.data.frame(t(as.matrix(assay(spe_k150_bcell, "counts")[bcell_markers, , drop = FALSE]))) %>%
  mutate(cluster = spe_k150_bcell$k150) %>%
  pivot_longer(cols = all_of(bcell_markers), names_to = "marker", values_to = "expression")

print(
  ggplot(expr_long_k150, aes(x = expression, fill = cluster)) +
    geom_density(alpha = 0.5, color = "black") +
    facet_wrap(~marker, ncol = 1, scales = "free") +
    theme_classic(base_size = 12) +
    labs(title = "k150 B cell clusters, marker distribution", x = "raw counts", y = "Density", fill = "Cluster") +
    theme(
      axis.text.x = element_text(size = 10, color = "black"),
      axis.text.y = element_text(size = 10, color = "black"),
      strip.text = element_text(size = 11, face = "bold"),
      strip.background = element_rect(fill = "grey85"),
      legend.position = "right",
      panel.spacing = unit(1.2, "lines")
    )
)
```

```{r ridge-k20, fig.width=7, fig.height=14}
expr_long_k20 <- as.data.frame(t(as.matrix(assay(spe_k20_bcell, "counts")[bcell_markers, , drop = FALSE]))) %>%
  mutate(cluster = spe_k20_bcell$k20) %>%
  pivot_longer(cols = all_of(bcell_markers), names_to = "marker", values_to = "expression")

print(
  ggplot(expr_long_k20, aes(x = expression, fill = cluster)) +
    geom_density(alpha = 0.5, color = "black") +
    facet_wrap(~marker, ncol = 1, scales = "free") +
    theme_classic(base_size = 12) +
    labs(title = "k20 B cell clusters, marker distribution", x = "raw counts", y = "Density", fill = "Cluster") +
    theme(
      axis.text.x = element_text(size = 10, color = "black"),
      axis.text.y = element_text(size = 10, color = "black"),
      strip.text = element_text(size = 11, face = "bold"),
      strip.background = element_rect(fill = "grey85"),
      legend.position = "right",
      panel.spacing = unit(1.2, "lines")
    )
)
```

```{r ridge-k200, fig.width=7, fig.height=14}
expr_long_k200 <- as.data.frame(t(as.matrix(assay(spe_k200_bcell, "counts")[bcell_markers, , drop = FALSE]))) %>%
  mutate(cluster = spe_k200_bcell$k200) %>%
  pivot_longer(cols = all_of(bcell_markers), names_to = "marker", values_to = "expression")

print(
  ggplot(expr_long_k200, aes(x = expression, fill = cluster)) +
    geom_density(alpha = 0.5, color = "black") +
    facet_wrap(~marker, ncol = 1, scales = "free") +
    theme_classic(base_size = 12) +
    labs(title = "k200 B cell clusters, marker distribution", x = "raw counts", y = "Density", fill = "Cluster") +
    theme(
      axis.text.x = element_text(size = 10, color = "black"),
      axis.text.y = element_text(size = 10, color = "black"),
      strip.text = element_text(size = 11, face = "bold"),
      strip.background = element_rect(fill = "grey85"),
      legend.position = "right",
      panel.spacing = unit(1.2, "lines")
    )
)
```

## Step 8: Patient and ROI composition per B cell cluster, per resolution

### Patient level

```{r patient-composition-bcell}
# Patient level: total B cell cells per patient per cluster, most cells first.
# as_tibble() is required before print(n = Inf): cluster_compare is a plain
# data.frame, so count() returns a plain data.frame back, and print.data.frame
# has no n argument, n = Inf would otherwise get misrouted into print.default's
# na.print argument and error out.
print(cluster_compare %>% filter(is_bcell_k150) %>% count(k150, patient_id) %>% arrange(k150, desc(n)) %>% as_tibble(), n = Inf)
print(cluster_compare %>% filter(is_bcell_k20)  %>% count(k20,  patient_id) %>% arrange(k20,  desc(n)) %>% as_tibble(), n = Inf)
print(cluster_compare %>% filter(is_bcell_k200) %>% count(k200, patient_id) %>% arrange(k200, desc(n)) %>% as_tibble(), n = Inf)
```

### ROI level

```{r roi-composition-bcell}
# ROI level: total B cell cells per sample_id per cluster, most cells first.
# patient_id is kept in the count alongside sample_id so each row is self
# contained, since a sample_id like Patient20_425 isn't identifiable on its
# own without knowing which patient it belongs to.
print(cluster_compare %>% filter(is_bcell_k150) %>% count(k150, patient_id, sample_id) %>% arrange(k150, desc(n)) %>% as_tibble(), n = Inf)
print(cluster_compare %>% filter(is_bcell_k20)  %>% count(k20,  patient_id, sample_id) %>% arrange(k20,  desc(n)) %>% as_tibble(), n = Inf)
print(cluster_compare %>% filter(is_bcell_k200) %>% count(k200, patient_id, sample_id) %>% arrange(k200, desc(n)) %>% as_tibble(), n = Inf)
```

A subtype that is concentrated in one or two patients, or one or two specific ROIs within a patient, rather than spread across the cohort is worth flagging the same way as the neutrophil neighbor contamination already noted for clusters 10 and 11 at k150.

## Summary and next steps

* Compare the `findMarkers` output across k20, k150, and k200 for each B cell cluster. If the same marker set defines a given subtype regardless of resolution, that subtype is likely real rather than a resolution artifact.
* If k20 collapses the plasma cell and activated B cell subtypes seen at k150 into a single cluster, that supports k20 simply being too coarse rather than the k150 split being wrong.
* If k200 splits a k150 B cell cluster into two or more clusters with clearly different marker profiles, consider whether that split should be carried into the final annotation as a subtype rather than treated as noise.
* Cross reference any B cell subtype with unusual patient or ROI concentration against the neighbor contamination pattern already flagged for clusters 10 and 11, since plasma cell and B cell clusters were noted as having neutrophil neighbor contamination at k150.
* If a marker like FoxP3 shows up consistently across all three resolutions in Step 3, treat it the same way as the existing neutrophil contamination notes rather than as coding noise, and consider running the spatial neighbor check from `06_unknown_cluster_diagnostics.Rmd` on these clusters specifically.
* Once `bcell_markers` is confirmed against panel.csv and the Step 3 findMarkers output, rerun Step 7's ridge plots with the full marker list rather than CD19 alone.
