---
title: 'All Patients: Broad Clustering k Sweep — Harmony (no asinh)'
author: 'Tazein Shah'
date: "06-29-2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)
set.seed(42)
```

## Overview

Sweeps k = 50 and k = 75 in `buildSNNGraph` to find a clustering
resolution that produces roughly 8 broad clusters, matching the
preprint's broad annotation categories. The 20-cluster notebook
(`05_broad_clustering_harmony_0100.Rmd`) is kept intact as a reference.

---

## 1. Load Libraries

```{r libraries}
library(imcRtools)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(scater)
library(scran)
library(bluster)
library(patchwork)
library(RColorBrewer)
```

---

## 2. Load Harmonized Data

```{r load-data}
spe_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/spe_combined_harmony_r3.rds"

spe <- readRDS(spe_path)

cat("Cells:", ncol(spe), "\n")
cat("Reduced dims available:", paste(reducedDimNames(spe), collapse = ", "), "\n")
cat("Assays available:", paste(assayNames(spe), collapse = ", "), "\n")
```

---

## 3. Marker Sets

```{r markers}
core_markers <- c("SMA", "CD4", "CD8a", "CD19", "CD15",
                  "CD14", "CD86", "LAMP3", "CD163", "EpCAM")
core_markers <- core_markers[core_markers %in% rownames(spe)]

full_markers <- c(
  "SMA", "CD4", "CD8a", "CD19", "CD15", "CD14", "CD86", "LAMP3",
  "Ki67", "MKI67", "CD138", "PD1", "PDL1", "FoxP3", "CD68", "CD163",
  "TIM3", "CD141", "CD11c", "CD56", "CD117", "EpCAM", "CD38",
  "CD45RO", "LAG3", "CD1c", "GATA3", "TTF1orCD16"
)
full_markers <- full_markers[full_markers %in% rownames(spe)]

cat("Core markers:", paste(core_markers, collapse = ", "), "\n")
cat("Full panel markers found:", length(full_markers), "\n")
```

---

## 4. Preprocessing: 99th Percentile Cap + Z-score

```{r helpers}
cap_99 <- function(mat) {
  t(apply(mat, 1, function(x) {
    q99 <- quantile(x, 0.99)
    pmin(x, q99)
  }))
}

zscore_rows <- function(mat) {
  t(scale(t(mat)))
}
```

```{r preprocess}
mat_0100 <- counts(spe) |> cap_99() |> zscore_rows()
assay(spe, "scaled_0100") <- mat_0100
cat("Assays in SPE:", paste(assayNames(spe), collapse = ", "), "\n")
```

---

## 5. UMAP (reuse existing)

```{r umap}
if ("UMAP_harmony" %in% reducedDimNames(spe)) {
  cat("Reusing existing UMAP_harmony.\n")
} else {
  set.seed(220225)
  spe <- runUMAP(spe, dimred = "Harmony", name = "UMAP_harmony")
  cat("UMAP computed.\n")
}
```

---

## 6. Subset for Visualization

```{r subset}
set.seed(220225)
cells_per_roi <- split(seq_len(ncol(spe)), spe$sample_id)
sampled_cells <- unlist(lapply(cells_per_roi, function(idx) {
  sample(idx, min(20000, length(idx)))
}))
spe_sub <- spe[, sampled_cells]
cat("Cells in subset:", ncol(spe_sub), "\n")
```

---

## 7. k Sweep: k = 50 and k = 75

Each value of k produces a different number of clusters. The goal is
roughly 8, matching the preprint's broad annotation categories. Results
are saved as separate columns in `colData(spe)` so both can be
inspected before committing to one.

```{r sweep-k50}
cat("--- k = 50 ---\n")
set.seed(220225)
harmony_mat <- reducedDim(spe, "Harmony")

g50    <- buildSNNGraph(t(harmony_mat), k = 50, type = "rank")
clust50 <- igraph::cluster_louvain(g50)
spe$clusters_k50 <- factor(clust50$membership)

cat("k = 50 clusters:", nlevels(spe$clusters_k50), "\n")
print(table(spe$clusters_k50))
```

```{r sweep-k75}
cat("--- k = 75 ---\n")
set.seed(220225)

g75    <- buildSNNGraph(t(harmony_mat), k = 75, type = "rank")
clust75 <- igraph::cluster_louvain(g75)
spe$clusters_k75 <- factor(clust75$membership)

cat("k = 75 clusters:", nlevels(spe$clusters_k75), "\n")
print(table(spe$clusters_k75))
```

---

## 8. UMAP Colored by Cluster: k = 50 vs k = 75

```{r umap-k50, fig.width=10, fig.height=7}
spe_sub$clusters_k50 <- spe$clusters_k50[match(colnames(spe_sub), colnames(spe))]

p_k50 <- dittoDimPlot(spe_sub, var = "clusters_k50",
                      reduction.use = "UMAP_harmony",
                      size = 0.2, do.label = TRUE, labels.size = 3) +
  ggtitle("UMAP: Louvain Clusters (k = 50)") +
  theme(legend.position = "right")
print(p_k50)
```

```{r umap-k75, fig.width=10, fig.height=7}
spe_sub$clusters_k75 <- spe$clusters_k75[match(colnames(spe_sub), colnames(spe))]

p_k75 <- dittoDimPlot(spe_sub, var = "clusters_k75",
                      reduction.use = "UMAP_harmony",
                      size = 0.2, do.label = TRUE, labels.size = 3) +
  ggtitle("UMAP: Louvain Clusters (k = 75)") +
  theme(legend.position = "right")
print(p_k75)
```

---

## 9. Pick Winning k

After inspecting cluster counts and UMAPs above, set `winning_k` to
whichever is closer to 8 clusters and looks more biologically
interpretable.

```{r pick-k}
# set to "k50" or "k75"
winning_k <- "k50"  # update after inspecting results above

if (winning_k == "k50") {
  spe$broad_clusters <- spe$clusters_k50
} else {
  spe$broad_clusters <- spe$clusters_k75
}

cat("Using", winning_k, "with", nlevels(spe$broad_clusters), "clusters.\n")

# save intermediate so clustering doesn't rerun
intermediate_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/harmony_0100_8broad_intermediate.rds"
saveRDS(spe, intermediate_path)
cat("Saved intermediate to:", intermediate_path, "\n")
```

---

## 10. Dot Plot: Core Markers

```{r make-dotplot}
make_dotplot <- function(spe, assay_name, cluster_col, marker_set, title) {
  expr_mat   <- assay(spe, assay_name)[marker_set, ]
  cluster_id <- spe[[cluster_col]]

  mean_expr <- sapply(levels(cluster_id), function(cl) {
    idx <- which(cluster_id == cl)
    rowMeans(expr_mat[, idx, drop = FALSE])
  })

  global_med <- apply(expr_mat, 1, median)
  frac_expr  <- sapply(levels(cluster_id), function(cl) {
    idx <- which(cluster_id == cl)
    rowMeans(expr_mat[, idx, drop = FALSE] > global_med)
  })

  mean_df <- as.data.frame(mean_expr) |>
    rownames_to_column("marker") |>
    pivot_longer(-marker, names_to = "cluster", values_to = "mean_exprs")

  frac_df <- as.data.frame(frac_expr) |>
    rownames_to_column("marker") |>
    pivot_longer(-marker, names_to = "cluster", values_to = "frac_expressing")

  plot_df <- left_join(mean_df, frac_df, by = c("marker", "cluster"))
  plot_df$cluster <- factor(plot_df$cluster, levels = levels(cluster_id))

  marker_order <- rownames(mean_expr)[hclust(dist(mean_expr))$order]
  plot_df$marker <- factor(plot_df$marker, levels = marker_order)

  ggplot(plot_df, aes(x = cluster, y = marker)) +
    geom_point(aes(size = frac_expressing, color = mean_exprs)) +
    scale_color_viridis(option = "magma", name = "Mean\nZ-score expr") +
    scale_size_continuous(range = c(0.5, 6), name = "Fraction\nexpressing") +
    labs(title = title, x = "Cluster", y = "Marker") +
    theme_bw(base_size = 12) +
    theme(
      axis.text.x      = element_text(angle = 0, hjust = 0.5),
      axis.text.y      = element_text(size = 9),
      panel.grid.major = element_line(color = "grey90"),
      legend.position  = "right"
    )
}
```

```{r dotplot-core, fig.width=14, fig.height=6}
p_dot_core <- make_dotplot(spe, "scaled_0100", "broad_clusters",
                           core_markers,
                           paste0("Harmony (no asinh): Core Markers per Broad Cluster (", winning_k, ")"))
print(p_dot_core)
```

---

## 11. Dot Plot: Full Panel

```{r dotplot-full, fig.width=18, fig.height=10}
p_dot_full <- make_dotplot(spe, "scaled_0100", "broad_clusters",
                           full_markers,
                           paste0("Harmony (no asinh): Full Panel per Broad Cluster (", winning_k, ")"))
print(p_dot_full)
```

---

## 12. findMarkers

```{r findmarkers}
markers_de <- findMarkers(spe,
                          groups     = spe$broad_clusters,
                          assay.type = "scaled_0100",
                          subset.row = full_markers)

all_markers <- lapply(levels(spe$broad_clusters), function(cl) {
  df <- as.data.frame(markers_de[[cl]])
  data.frame(
    cluster = cl,
    marker  = rownames(df),
    Top     = df$Top,
    FDR     = df$FDR,
    logFC   = df$summary.logFC
  )
})

all_markers_df <- do.call(rbind, all_markers)

write.csv(all_markers_df,
          "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/findmarkers_8broad_clusters.csv",
          row.names = FALSE)

cat("Saved to findmarkers_8broad_clusters.csv\n")

# summary table
df <- read.csv("/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/findmarkers_8broad_clusters.csv")

top1_summary <- do.call(rbind, lapply(levels(spe$broad_clusters), function(cl) {
  sub <- df[df$cluster == cl & df$logFC > 1.0, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  if (nrow(sub) == 0) {
    data.frame(cluster = cl, top_markers = "none above logFC 1.0")
  } else {
    data.frame(cluster = cl,
               top_markers = paste(head(sub$marker, 5), collapse = ", "))
  }
}))

print(knitr::kable(top1_summary, col.names = c("Cluster", "Top Positive Markers")))

for (cl in levels(spe$broad_clusters)) {
  sub <- df[df$cluster == cl & df$Top <= 5 & df$logFC > 1.0, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  if (nrow(sub) == 0) {
    cat("Cluster", cl, ": no markers with logFC > 1.0 in top 5\n\n")
  } else {
    print(knitr::kable(sub[, c("marker", "Top", "FDR", "logFC")],
                       caption = paste("Cluster", cl),
                       digits  = 3))
  }
}
```

---

## 13. Manual Broad Annotation

Fill in after inspecting dot plots and findMarkers output above.
Target 8 categories matching the preprint:
T/NK Cell, B Cell, Plasma Cell, DC, Other Myeloid, Neutrophil,
Mast Cell, Epithelial. Add Stroma if needed for tumor tissue context.

```{r annotate}
broad_annotation <- c(
  # fill in after inspecting plots
  # "1" = "T/NK Cell",
  # "2" = "Epithelial",
  # ...
)

if (length(broad_annotation) > 0) {
  spe$broad_celltype <- broad_annotation[as.character(spe$broad_clusters)]
  cat("Annotation applied.\n")
  print(table(spe$broad_celltype, useNA = "ifany"))
} else {
  cat("No annotation provided yet.\n")
}
```

---

## 14. UMAP Colored by Broad Cell Type

```{r umap-celltype, fig.width=10, fig.height=7}
if ("broad_celltype" %in% colnames(colData(spe))) {
  spe_sub$broad_celltype <- spe$broad_celltype[match(colnames(spe_sub), colnames(spe))]

  p_celltype <- dittoDimPlot(spe_sub, var = "broad_celltype",
                             reduction.use = "UMAP_harmony",
                             size = 0.2, do.label = TRUE, labels.size = 3) +
    ggtitle("UMAP colored by Broad Cell Type") +
    theme(legend.position = "right")
  print(p_celltype)
} else {
  cat("Fill in broad_annotation in chunk 13 first.\n")
}
```

---

## 15. Save Annotated SPE

```{r save}
out_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/steinbock/harmony_0100_8broad_cluster.rds"

if ("broad_celltype" %in% colnames(colData(spe))) {
  saveRDS(spe, out_path)
  cat("Saved annotated SPE to:", out_path, "\n")
} else {
  cat("Annotation not yet applied. Fill in chunk 13 before saving.\n")
}
```

---

## Notes

- k = 50 and k = 75 both tested; winning k set in chunk 9
- Intermediate saved to `harmony_0100_8broad_intermediate.rds` after
  winning k is picked; rerun chunk 9 onward only if annotation changes
- Target 8 broad categories matching Xi et al. preprint; Stroma added
  as 9th if needed for tumor tissue context
- findMarkers results saved to `findmarkers_8broad_clusters.csv`
