---
title: "Patient 12 - Cell Segmentation QC"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
    code_folding: show
  pdf_document:
    toc: true
editor_options:
  markdown:
    wrap: 72
---


# Setup
```{r setup, message=FALSE, warning=FALSE}
.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(imcRtools)
library(SpatialExperiment)
library(SingleCellExperiment)
library(ggplot2)
library(dplyr)
library(reshape2)
library(stringr)

set.seed(42)
```

# Load Steinbock Output into SpatialExperiment
```{r load-data}
steinbock_dir <- "/restricted/projectnb/camplab/home/vgupta7/tumor_imc/patient12_steinbock"

spe <- read_steinbock(steinbock_dir)

# Preview the object
spe

# How many cells per ROI?
table(spe$sample_id)

# What markers are available?
rownames(spe)

# What cell-level metadata do we have?
colnames(colData(spe))
```

# Add metadata
```{r}
colnames(spe) <- paste0(spe$sample_id, "_", spe$ObjectNumber)
spe$image_id  <- spe$sample_id
spe$patient_id <- "Patient12"
spe$ROI <- str_extract(spe$sample_id, "00[1-4]")

cat("Before area filtering:", ncol(spe), "cells\n")
```

# Plot area distribution before filtering
```{r}
ggplot(as.data.frame(colData(spe)), aes(x = area)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "white") +
  geom_vline(xintercept = c(20, 500), color = "red", linetype = "dashed") +
  labs(title = "Cell Area Distribution (red lines = filter thresholds)",
       x = "Area (pixels)", y = "Count") +
  theme_classic()
```

Tumor tissue can have larger cells compared to lymph node tissue, so it's important to check the cell area distribution first. 

A small number of cells fall below the 20 pixel threshold. These are likely segmentation artifacts, debris, or noise. These will be filtered out. 

The largest cells reach about 250 pixels, which means that no true doublets were detected at the upper end. 

# Filter by area (same thresholds as preprint)
```{r}
spe <- spe[, spe$area >= 20]
# Removes abnormal cells
spe <- spe[, spe$area < 500]

cat("After area filtering:", ncol(spe), "cells\n")
cat("Cells removed:", 37375 - ncol(spe), "\n")
```


# Apply asinh transformation
```{r}
assay(spe, "asin") <- asinh(counts(spe) / 1)

# Save the spe object (so I don't have to rerun the filtering steps next time)
saveRDS(spe, "/restricted/projectnb/camplab/home/vgupta7/tumor_imc/patient12_steinbock/p12_spe_filtered.rds")
```


# Load cytomapper for segmentation mask overlay
```{r}
library(cytomapper)
```

# Load images and masks
```{r}
images <- loadImages(
  "/restricted/projectnb/camplab/home/vgupta7/tumor_imc/patient12_steinbock/img/"
)

masks <- loadImages(
  "/restricted/projectnb/camplab/home/vgupta7/tumor_imc/patient12_steinbock/masks_deepcell/",
  as.is = TRUE  # keeps masks as integer values (cell IDs)
)

# Check what loaded
images
masks

```

# Set channel names and sample_id metadata 
```{r}
# Set channel names on images to match rownames of spe
channelNames(images) <- rownames(spe)

# Set sample_id metadata on both images and masks so cytomapper can match them to spe
mcols(images)$sample_id <- names(images)
mcols(masks)$sample_id  <- names(masks)

cat("Images loaded:", length(images), "\n")
cat("Masks loaded:", length(masks), "\n")
cat("\nChannel names:\n")
print(channelNames(images))

```

# Normalize images
```{r}
cur_images <- cytomapper::normalize(images, separateImages = TRUE)
# Clip at 0.2 to reduce hot pixel effect
cur_images <- cytomapper::normalize(cur_images, inputRange = c(0, 0.2))

```

# Plot segmentation overlay for all 4 ROIs
```{r fig.height=12}
img_ids <- seq_len(length(cur_images))

plotPixels(
  cur_images,
  mask        = masks,
  img_id      = "sample_id",
  missing_colour = "white",
  colour_by   = c("DNA191", "EpCAM", "CD68", "Membrane"),
  colour      = list(
    DNA191   = c("black", "blue"),
    EpCAM    = c("black", "green"),
    CD68     = c("black", "yellow"),
    Membrane = c("black", "red")
  ),
  legend = list(
    colour_by.title.cex  = 1,
    colour_by.labels.cex = 1
  )
)

```

pink = nucleus + membrane together --> correctly segmented cell

# Plot 1: DNA + EpCAM (tumor structure)
```{r fig.height=12}
plotPixels(cur_images, mask = masks, img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("DNA191", "EpCAM"),
           colour = list(DNA191 = c("black", "blue"),
                         EpCAM  = c("black", "green")),
           legend = list(
            colour_by.title.cex  = 1.5,
            colour_by.labels.cex = 1.5
            )
          )
```


# Plot 2: DNA + CD68 (macrophages)
```{r fig.height=12}
plotPixels(cur_images, mask = masks, img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("DNA191", "CD68"),
           colour = list(DNA191 = c("black", "blue"),
                         CD68   = c("black", "yellow")),
           legend = list(
            colour_by.title.cex  = 1.5,
            colour_by.labels.cex = 1.5
            )
          )
```


# Plot 3: DNA + Membrane (segmentation boundaries)
```{r fig.height=12}
plotPixels(cur_images, mask = masks, img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("DNA191", "Membrane"),
           colour = list(DNA191   = c("black", "blue"),
                         Membrane = c("black", "red")),
           legend = list(
            colour_by.title.cex  = 1.5,
            colour_by.labels.cex = 1.5
            )
          )
```

```{r}
.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))
BiocManager::install(c("dittoSeq", "scuttle"), lib = "/restricted/projectnb/camplab/home/vgupta7/R_libs")

library(dittoSeq)
library(scuttle)
library(viridis)
```


```{r}
# Load filtered SPE object
spe <- readRDS("/restricted/projectnb/camplab/home/vgupta7/tumor_imc/patient12_steinbock/p12_spe_filtered.rds")

# Set channel names and sample metadata
channelNames(images)      <- rownames(spe)
mcols(images)$sample_id  <- names(images)
mcols(masks)$sample_id   <- names(masks)

# Confirm everything loaded correctly
spe
cat("Images loaded:", length(images), "\n")
cat("Masks loaded:", length(masks), "\n")

```
```{r fig.height=10, fig.width=10}
# Subsample to 2000 cells for visualization (same as workflow)
set.seed(42)
cur_cells <- sample(seq_len(ncol(spe)), 2000)

# Define which markers to show — exclude QC/structural markers
markers_to_plot <- rownames(spe)[!rownames(spe) %in% 
                                   c("DNA191", "DNA193", "Membrane")]

dittoHeatmap(
  spe[, cur_cells],
  genes        = markers_to_plot,
  assay        = "asin",
  cluster_cols = TRUE,
  scale        = "none",
  heatmap.colors = viridis(100),
  annot.by     = "sample_id",
  main         = "Patient 12 - Single Cell Expression Heatmap\nasinh-transformed, 2000 subsampled cells"
)

```
This heatmap is clustering the 2000 individual cells by their marker expression profiles using hierarchical clustering. 

Columns (cells) are grouped by similarity. Cells that express similar markers end up in the same block. 

Rows (markers) are clustered by co-expression. 


```{r fig.height=10, fig.width=10}
# Aggregate to ROI level
image_mean <- aggregateAcrossCells(spe,
                                   ids        = spe$sample_id,
                                   statistics = "mean",
                                   use.assay.type = "asin")

# Heatmap of mean expression per ROI
dittoHeatmap(
  image_mean,
  genes          = markers_to_plot,
  assay          = "asin",
  cluster_cols   = TRUE,
  scale          = "none",
  heatmap.colors = viridis(100),
  annot.by       = "sample_id",
  main           = "Patient 12 - Mean Marker Expression per ROI\nasinh-transformed"
)

```


