---
title: 'Patient 14 Cell Segmentation Mask Visualization'
author: 'Tazein Shah'
date: "06-10-2026"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 14, fig.height = 10)
```

## Overview

This script visualizes cell segmentation mask overlays for Patient 14T across
all ROIs. Each section uses a channel combination matched to one annotated cell
type to assess segmentation quality per compartment. Cell type annotations
follow Xi et al. (2026).

1. Nuclear + structural anchors (DNA191 + DNA193 + Membrane)
2. Tumor / Epithelial (EpCAM)
3. Fibroblast (SMA)
4. B Cell (CD19)
5. CD4+ T Cell (CD4)
6. CD8+ T Cell (CD8a)
7. PMN-MDSC (CD15)
8. Macrophage (CD14 + CD86)
9. M-MDSC (CD14 + CD163)
10. DC / mregDC (LAMP3 + CD1c + CD11c)

---

## 1. Load Libraries

```{r libraries}
library(imcRtools)
library(cytomapper)
library(tidyverse)
```

---

## 2. Load Images, Masks, and Panel

```{r load-data}
steinbock_path <- "/restricted/projectnb/camplab/home/tshah20/tumor_IMC/patient14_steinbock"

images <- loadImages(file.path(steinbock_path, "img/"))
masks  <- loadImages(file.path(steinbock_path, "masks_deepcell/"), as.is = TRUE)

panel         <- read.csv(file.path(steinbock_path, "panel_GATA3.csv"))
kept_channels <- panel$name[panel$keep == 1]

channelNames(images) <- kept_channels

mcols(images)$sample_id <- names(images)
mcols(masks)$sample_id  <- names(masks)

cat("Images loaded:", length(images), "\n")
cat("Masks loaded:", length(masks), "\n")
cat("Channels:\n")
print(channelNames(images))
```

---

## 3. Normalize Images for Display

Pixel intensities are normalized per image to [0, 1], then clipped at
0.2 to bring out dim signals. Does not affect downstream analysis.

```{r normalize}
images_norm <- cytomapper::normalize(images, separateImages = TRUE)
images_norm <- cytomapper::normalize(images_norm, inputRange = c(0, 0.2))

img_ids <- seq_len(length(images_norm))
```

---

## 4. Nuclear + Structural Anchors

DNA191 and DNA193 mark nuclei, Membrane marks cell boundaries. This is the
baseline check that individual nuclei are cleanly outlined by their masks
without obvious clumping or splitting.

```{r nuclear-structural, fig.height=12}
plotPixels(
  images_norm[img_ids],
  mask           = masks[img_ids],
  img_id         = "sample_id",
  missing_colour = "white",
  colour_by = c("DNA191", "DNA193", "Membrane"),
  colour    = list(
    DNA191   = c("black", "blue"),
    DNA193   = c("black", "cyan"),
    Membrane = c("black", "red")
  ),
  image_title = list(position = "topleft", cex = 1),
  legend      = list(colour_by.title.cex = 0.8, colour_by.labels.cex = 0.8)
)

dev.copy(png, "images/mask_01_nuclear_structural.png", width = 2800, height = 2800, res = 200)
dev.off()
```

---

## 5. Tumor / Epithelial Cells

EpCAM marks epithelial and tumor cells. Overlaid with DNA191 and Membrane to
check that tumor cell nests are segmented as individual cells rather than merged
into a single large mask.

```{r tumor-epithelial, fig.height=12}
plotPixels(
  images_norm[img_ids],
  mask           = masks[img_ids],
  img_id         = "sample_id",
  missing_colour = "white",
  colour_by = c("EpCAM", "DNA191", "Membrane"),
  colour    = list(
    EpCAM    = c("black", "green"),
    DNA191   = c("black", "blue"),
    Membrane = c("black", "red")
  ),
  image_title = list(position = "topleft", cex = 1),
  legend      = list(colour_by.title.cex = 0.8, colour_by.labels.cex = 0.8)
)

dev.copy(png, "images/mask_02_tumor_epithelial.png", width = 2800, height = 2800, res = 200)
dev.off()
```

---

## 6. Fibroblasts

SMA (alpha-smooth muscle actin) marks fibroblasts and myofibroblasts in the
tumor stroma. Fibroblasts tend to be elongated, so this panel checks whether
the masks correctly capture their irregular shape rather than splitting them
across multiple objects.

```{r fibroblast, fig.height=12}
plotPixels(
  images_norm[img_ids],
  mask           = masks[img_ids],
  img_id         = "sample_id",
  missing_colour = "white",
  colour_by = c("SMA", "DNA191", "Membrane"),
  colour    = list(
    SMA      = c("black", "magenta"),
    DNA191   = c("black", "blue"),
    Membrane = c("black", "red")
  ),
  image_title = list(position = "topleft", cex = 1),
  legend      = list(colour_by.title.cex = 0.8, colour_by.labels.cex = 0.8)
)

dev.copy(png, "images/mask_03_fibroblast.png", width = 2800, height = 2800, res = 200)
dev.off()
```

---

## 7. B Cells

CD19 is the canonical B cell surface marker. B cells are small and round and
tend to cluster in follicle-like structures in lymph node tissue; this panel
checks for over-segmentation in densely packed B cell regions.

```{r bcell, fig.height=12}
plotPixels(
  images_norm[img_ids],
  mask           = masks[img_ids],
  img_id         = "sample_id",
  missing_colour = "white",
  colour_by = c("CD19", "DNA191", "Membrane"),
  colour    = list(
    CD19     = c("black", "yellow"),
    DNA191   = c("black", "blue"),
    Membrane = c("black", "red")
  ),
  image_title = list(position = "topleft", cex = 1),
  legend      = list(colour_by.title.cex = 0.8, colour_by.labels.cex = 0.8)
)

dev.copy(png, "images/mask_04_bcell.png", width = 2800, height = 2800, res = 200)
dev.off()
```

---

## 8. CD4+ T Cells

CD4 marks helper T cells. CD4+ T cells are small and abundant in LN tissue;
this panel assesses whether individual T cells are individually resolved or
merged in dense infiltrates.

```{r cd4-tcell, fig.height=12}
plotPixels(
  images_norm[img_ids],
  mask           = masks[img_ids],
  img_id         = "sample_id",
  missing_colour = "white",
  colour_by = c("CD4", "DNA191", "Membrane"),
  colour    = list(
    CD4      = c("black", "orange"),
    DNA191   = c("black", "blue"),
    Membrane = c("black", "red")
  ),
  image_title = list(position = "topleft", cex = 1),
  legend      = list(colour_by.title.cex = 0.8, colour_by.labels.cex = 0.8)
)

dev.copy(png, "images/mask_05_cd4_tcell.png", width = 2800, height = 2800, res = 200)
dev.off()
```

---

## 9. CD8+ T Cells

CD8a marks cytotoxic T cells. CD8+ T cells are the primary effector population
in anti-tumor immunity; their segmentation quality is especially important for
downstream proximity and neighborhood analyses.

```{r cd8-tcell, fig.height=12}
plotPixels(
  images_norm[img_ids],
  mask           = masks[img_ids],
  img_id         = "sample_id",
  missing_colour = "white",
  colour_by = c("CD8a", "DNA191", "Membrane"),
  colour    = list(
    CD8a     = c("black", "orange"),
    DNA191   = c("black", "blue"),
    Membrane = c("black", "red")
  ),
  image_title = list(position = "topleft", cex = 1),
  legend      = list(colour_by.title.cex = 0.8, colour_by.labels.cex = 0.8)
)

dev.copy(png, "images/mask_06_cd8_tcell.png", width = 2800, height = 2800, res = 200)
dev.off()
```

---

## 10. PMN-MDSCs

CD15 marks polymorphonuclear myeloid-derived suppressor cells (PMN-MDSCs),
a granulocytic immunosuppressive population. PMN-MDSCs are typically
multi-lobed and larger than lymphocytes; this panel checks for mask merging
in regions of granulocyte infiltration.

```{r pmn-mdsc, fig.height=12}
plotPixels(
  images_norm[img_ids],
  mask           = masks[img_ids],
  img_id         = "sample_id",
  missing_colour = "white",
  colour_by = c("CD15", "DNA191", "Membrane"),
  colour    = list(
    CD15     = c("black", "yellow"),
    DNA191   = c("black", "blue"),
    Membrane = c("black", "red")
  ),
  image_title = list(position = "topleft", cex = 1),
  legend      = list(colour_by.title.cex = 0.8, colour_by.labels.cex = 0.8)
)

dev.copy(png, "images/mask_07_pmn_mdsc.png", width = 2800, height = 2800, res = 200)
dev.off()
```

---

## 11. Macrophages

CD14 marks monocytes and macrophages broadly; CD86 is a co-stimulatory marker
expressed on activated macrophages and distinguishes them from M-MDSCs. Macrophages
are large and irregularly shaped, making them prone to segmentation artifacts.

```{r macrophage, fig.height=12}
plotPixels(
  images_norm[img_ids],
  mask           = masks[img_ids],
  img_id         = "sample_id",
  missing_colour = "white",
  colour_by = c("CD14", "CD86", "DNA191"),
  colour    = list(
    CD14   = c("black", "yellow"),
    CD86   = c("black", "magenta"),
    DNA191 = c("black", "blue")
  ),
  image_title = list(position = "topleft", cex = 1),
  legend      = list(colour_by.title.cex = 0.8, colour_by.labels.cex = 0.8)
)

dev.copy(png, "images/mask_08_macrophage.png", width = 2800, height = 2800, res = 200)
dev.off()
```

---

## 12. M-MDSCs

CD14 and CD163 together mark monocytic MDSCs (M-MDSCs), which are
CD14+ but lack the CD86 co-stimulatory signal of activated macrophages.
CD163 additionally marks M2-polarized immunosuppressive macrophages.

```{r m-mdsc, fig.height=12}
plotPixels(
  images_norm[img_ids],
  mask           = masks[img_ids],
  img_id         = "sample_id",
  missing_colour = "white",
  colour_by = c("CD14", "CD163", "DNA191"),
  colour    = list(
    CD14   = c("black", "yellow"),
    CD163  = c("black", "orange"),
    DNA191 = c("black", "blue")
  ),
  image_title = list(position = "topleft", cex = 1),
  legend      = list(colour_by.title.cex = 0.8, colour_by.labels.cex = 0.8)
)

dev.copy(png, "images/mask_09_m_mdsc.png", width = 2800, height = 2800, res = 200)
dev.off()
```

---

## 13. DC / mregDC

LAMP3 is the canonical marker for mature regulatory dendritic cells (mregDCs),
a tolerogenic DC subset. CD1c marks cDC2-like conventional DCs. CD11c is a
broad DC and myeloid marker used to confirm the dendritic lineage. DCs are
rare and morphologically variable; this panel checks that sparse LAMP3+
cells are captured as distinct individual masks.

```{r dc-mregdc, fig.height=12}
plotPixels(
  images_norm[img_ids],
  mask           = masks[img_ids],
  img_id         = "sample_id",
  missing_colour = "white",
  colour_by = c("LAMP3", "CD1c", "CD11c"),
  colour    = list(
    LAMP3 = c("black", "cyan"),
    CD1c  = c("black", "green"),
    CD11c = c("black", "yellow")
  ),
  image_title = list(position = "topleft", cex = 1),
  legend      = list(colour_by.title.cex = 0.8, colour_by.labels.cex = 0.8)
)

dev.copy(png, "images/mask_10_dc_mregdc.png", width = 2800, height = 2800, res = 200)
dev.off()
```

---

## Session Info

```{r session-info}
sessionInfo()
```
