---
title: 'Patient 14 Post-Steinbock Visualization -> Arcsinh'
author: ''
date: "06-10-2026"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 14, fig.height = 8)

set.seed(42)
```

## Overview

This notebook is to visualize the post-steinbock IMC results, checking the cell segmentation, single-cell expression of the ROIs across patient 14, and the pixel visualization.

**With arcsinh transformation**

1.  Visual overlay of mask outlines on composite channel images
2.  Spatial pixel-level heatmaps of selected markers across all ROIs
3.  Distribution of cell areas per ROI (to catch over/under-segmentation)
4.  Fraction of image area covered by cells per ROI
5.  Single-cell expression of all ROIs in heatmap form

------------------------------------------------------------------------

## 1. Load Libraries

```{r libraries}
library(imcRtools)
library(cytomapper)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(scater)
library(patchwork)
library(RColorBrewer)
```

------------------------------------------------------------------------

## 2. Read in Steinbock Data

```{r read-data}
steinbock_path <- "/restricted/projectnb/camplab/home/tshah20/tumor_IMC/patient14_steinbock"

spe <- read_steinbock(steinbock_path)
spe
```

```{r process-spe}
# set informative column names: image_cellID
colnames(spe) <- paste0(spe$sample_id, "_", spe$ObjectNumber)

# extract patient and ROI from sample_id
spe$patient_id <- str_extract(spe$sample_id, "^[^_]+")
spe$ROI        <- spe$sample_id

# arcsinh transform (cofactor = 1, standard for IMC)
assay(spe, "exprs") <- asinh(counts(spe) / 1)

# flag biological channels (exclude DNA and membrane channels used for segmentation)
rowData(spe)$use_channel <- !grepl("DNA191|DNA193|Membrane", rownames(spe))

cat("Cells loaded:", ncol(spe), "\n")
cat("ROIs:", n_distinct(spe$sample_id), "\n")
cat("Channels:", nrow(spe), "\n")

# confirm colData column names for image dimensions
cat("\ncolData columns:\n")
print(colnames(colData(spe)))
```

------------------------------------------------------------------------

## 3. Load Multichannel Images and Segmentation Masks

```{r load-images}
images <- loadImages(file.path(steinbock_path, "img/"))
masks  <- loadImages(file.path(steinbock_path, "masks_deepcell/"), as.is = TRUE)

# read panel to get only the channels written to the image stack (keep == 1)
panel         <- read.csv(file.path(steinbock_path, "panel_GATA3.csv"))
kept_channels <- panel$name[panel$keep == 1]

cat("Channels in images:", length(channelNames(images)), "\n")
cat("Kept channels in panel:", length(kept_channels), "\n")

# assign channel names from panel kept channels only
channelNames(images) <- kept_channels

# link images and masks to SPE via sample_id
mcols(images)$sample_id <- names(images)
mcols(masks)$sample_id  <- names(masks)

cat("Images loaded:", length(images), "\n")
cat("Masks loaded:", length(masks), "\n")
cat("\nChannel names:\n")
print(channelNames(images))
```

------------------------------------------------------------------------

## 4. Normalize Images for Display

Pixel intensities are normalized per image to [0, 1], then clipped at
0.2 to bring out dim signals. (Does not affect downstream, only visuals)

```{r normalize}
images_norm <- cytomapper::normalize(images, separateImages = TRUE)
images_norm <- cytomapper::normalize(images_norm, inputRange = c(0, 0.2))
```

------------------------------------------------------------------------

## 5. Overlay Cell Masks on Composite Images

For patient 14, we load all the ROIs with the mask outlines overlaid:

- EpCAM marks tumor/epithelial cells
- CD68 marks macrophages
- Membrane marks cell boundaries
- DNA191 anchors nuclei

```{r plot-masks, fig.height=12}
img_ids <- seq_len(length(images_norm))

plotPixels(
  images_norm[img_ids],
  mask           = masks[img_ids],
  img_id         = "sample_id",
  missing_colour = "white",
  colour_by = c("DNA191", "EpCAM", "CD68", "Membrane"),
  colour    = list(
    DNA191   = c("black", "blue"),
    EpCAM    = c("black", "green"),
    CD68     = c("black", "yellow"),
    Membrane = c("black", "red")
  ),
  image_title = list(position = "topleft", cex = 1),
  legend      = list(colour_by.title.cex = 0.8, colour_by.labels.cex = 0.8)
)

dev.copy(png, "images/05_mask_overlay_all_rois.png", width = 2800, height = 2800, res = 200)
dev.off()
```

My plot looks very different from patient 12, but the cell segmentation worked and the differences seem more biological vs technical. Cells of similar types also cluster together which is a good sign.

------------------------------------------------------------------------

## 6. Spatial Pixel Heatmaps: All ROIs

The post-steinbock files (in raw/ folder) are scaled 0-100 through the steinbock.sh code. Here we plot them at their X/Y coordinates, which provides a spatial map of marker expression across the tissue for each ROI.

Pixels are downsampled to every 3rd coordinate to reduce rendering time.

Markers are selected to highlight the major cell compartments: tumor (EpCAM), stroma (SMA),
myeloid (CD68, CD163, CD1c), T cell exhaustion (FoxP3, TIM3, PD1), and a combined lineage marker (TTF1orCD16).

```{r spatial-pixel-setup}
raw_filtered_path <- "/restricted/projectnb/camplab/home/tshah20/tumor_IMC/patient14_steinbock/raw/"

# read all ROI TXT files and combine
txt_files <- list.files(raw_filtered_path, pattern = "\\.txt$", full.names = TRUE)

all_rois_filtered <- map_dfr(txt_files, function(f) {
  df <- read.delim(f, check.names = FALSE)
  df$roi <- tools::file_path_sans_ext(basename(f))
  df
})

cat("Total pixels loaded:", nrow(all_rois_filtered), "\n")
cat("ROIs found:", n_distinct(all_rois_filtered$roi), "\n")
```

```{r spatial-pixel-pivot}
focused_markers_spatial <- c(
  "EpCAM(Nd144Di)",
  "SMA(Pr141Di)",
  "FoxP3(Dy162Di)",
  "CD68(Tb159Di)",
  "CD163(Sm147Di)",
  "CD1c(Sm154Di)",
  "TIM3(Nd143Di)",
  "PD1(Ho165Di)",
  "TTF1orCD16(Pt194Di)"
)

# downsample to every 3rd pixel per ROI for rendering
set.seed(42)
all_rois_sub <- all_rois_filtered %>%
  filter(X %% 3 == 0, Y %% 3 == 0)

cat("Downsampled pixels:", nrow(all_rois_sub), "\n")

# pivot to long format and clean marker names for facet labels
all_rois_long <- all_rois_sub %>%
  select(X, Y, roi, all_of(focused_markers_spatial)) %>%
  pivot_longer(
    cols      = -c(X, Y, roi),
    names_to  = "marker",
    values_to = "intensity"
  ) %>%
  mutate(marker = str_extract(marker, "^[^(]+"))
```

```{r spatial-pixel-plot, fig.height=14, fig.width=12}
rois <- unique(all_rois_long$roi)

plots <- map(rois, function(r) {
  ggplot(filter(all_rois_long, roi == r),
         aes(x = X, y = Y, fill = intensity)) +
    geom_tile() +
    scale_fill_viridis_c(option = "viridis") +
    facet_wrap(~marker, ncol = 3) +
    labs(
      title    = paste0("Patient 14T — ", r, " (Filtered Pixels)"),
      subtitle = "Filtered values, 0-100 scaled, downsampled every 3rd pixel",
      fill     = "Intensity\n(0-100)"
    ) +
    theme_void(base_size = 9) +
    theme(
      strip.text      = element_text(size = 8, face = "bold"),
      legend.position = "bottom"
    )
})

# print all plots in notebook
walk(plots, print)

# save each ROI as its own PNG
walk2(plots, rois, function(p, r) {
  ggsave(
    paste0("images/06_", r, "_spatial_pixel_heatmap.png"),
    plot   = p,
    width  = 12,
    height = 14,
    dpi    = 150
  )
})
```

CD163 and PD1 seems pretty noisy across all the ROIs --> is that expected?

------------------------------------------------------------------------

## 7. Expression Heatmap: Sampled Cells Across All ROIs

A heatmap of arcsinh-transformed marker expression across a random
sample of cells, annotated by ROI. This gives a high-level overview of
marker co-expression patterns and whether any ROIs show systematically
different expression profiles before phenotyping.

The color scale represents arcsinh-transformed mean pixel intensity per cell, ranging from 0 (dark purple) to 5 (yellow).

```{r heatmap, fig.height=10, fig.width=10}
cur_cells  <- sample(seq_len(ncol(spe)), min(2000, ncol(spe)))
cell_order <- order(spe[, cur_cells]$sample_id)

p_heatmap <- dittoHeatmap(
  spe[, cur_cells[cell_order]],
  genes          = rownames(spe)[rowData(spe)$use_channel],
  assay          = "exprs",
  cluster_cols   = FALSE,
  scale          = "none",
  heatmap.colors = viridis(100),
  annot.by       = "sample_id",
  annotation_colors = list(
    sample_id = setNames(
      colorRampPalette(brewer.pal(4, "Set1"))(n_distinct(spe$sample_id)),
      unique(spe$sample_id)
    )
  )
)

png("images/07_expression_heatmap.png", width = 2400, height = 3000, res = 200)
grid::grid.newpage()
grid::grid.draw(p_heatmap$gtable)
dev.off()
```

------------------------------------------------------------------------

## 8. Cell Area Distribution per ROI

Small cells are likely segmentation artifacts — partial cell slices or noise. The distribution below helps select a sensible area cutoff.

**How to determine the exact cutoff?**

```{r cell-area, fig.height=6}
cat("Cells before area filter:", ncol(spe), "\n")

dittoPlot(spe,
          var      = "area",
          group.by = "sample_id",
          plots    = "boxplot") +
  ylab("Cell area (pixels)") +
  xlab("") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Cell area distribution per ROI (pre-filter)")
```

```{r filter-cells, fig.height=6}
area_threshold <- 5
spe <- spe[, spe$area >= area_threshold]

cat("Cells after area filter (>=", area_threshold, "px):", ncol(spe), "\n")

dittoPlot(spe,
          var      = "area",
          group.by = "sample_id",
          plots    = "boxplot") +
  ylab("Cell area (pixels)") +
  xlab("") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Cell area distribution per ROI (post-filter)")
```

------------------------------------------------------------------------

## 9. Fraction of Image Area Covered by Cells

High cell coverage (~70-90%) is expected for dense tissue sections. Sparse coverage may indicate tissue loss or large acellular regions (e.g., vessels, necrosis).

```{r cell-coverage, fig.height=5}
cell_density <- colData(spe) %>%
  as.data.frame() %>%
  group_by(sample_id) %>%
  summarise(
    cell_area    = sum(area),
    no_pixels    = unique(width_px) * unique(height_px),
    covered_area = cell_area / no_pixels,
    .groups      = "drop"
  )

ggplot(cell_density, aes(x = sample_id, y = covered_area)) +
  geom_point(size = 3, color = "steelblue") +
  geom_hline(yintercept = 0.7, linetype = "dashed", color = "grey50") +
  ylim(0, 1) +
  ylab("Fraction of image area covered by cells") +
  xlab("") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Image area covered by cells per ROI")
```

------------------------------------------------------------------------

## 10. Dimensionality Reduction: UMAP and tSNE

UMAP/tSNE projects cells to 2D --> looking to see if the ROIs mix well

Marker expression is overlaid to verify that the embeddings separate biologically meaningful populations aka the differences between patient expressions

```{r dimred, fig.height=12, fig.width=14}
set.seed(220225)
cell_sub <- sample(seq_len(ncol(spe)), min(10000, ncol(spe)))
spe_sub  <- spe[, cell_sub]

spe_sub <- runUMAP(spe_sub, subset_row = rowData(spe_sub)$use_channel, exprs_values = "exprs")
spe_sub <- runTSNE(spe_sub, subset_row = rowData(spe_sub)$use_channel, exprs_values = "exprs")

# define ROI colors
roi_colors <- setNames(
  colorRampPalette(brewer.pal(4, "Set1"))(n_distinct(spe$sample_id)),
  unique(spe$sample_id)
)

p1 <- dittoDimPlot(spe_sub, var = "sample_id", reduction.use = "UMAP", size = 0.2) +
  scale_color_manual(values = roi_colors) +
  ggtitle("ROI on UMAP")

p2 <- dittoDimPlot(spe_sub, var = "sample_id", reduction.use = "TSNE", size = 0.2) +
  scale_color_manual(values = roi_colors) +
  ggtitle("ROI on tSNE")

(p1 + p2)
```

```{r dimred-markers, fig.height=12, fig.width=14}
p3 <- dittoDimPlot(spe_sub, var = "EpCAM", reduction.use = "UMAP",
                   assay = "exprs", size = 0.2) +
  scale_color_viridis(name = "EpCAM") +
  ggtitle("EpCAM on UMAP")

p4 <- dittoDimPlot(spe_sub, var = "SMA", reduction.use = "UMAP",
                   assay = "exprs", size = 0.2) +
  scale_color_viridis(name = "SMA") +
  ggtitle("SMA on UMAP")

p5 <- dittoDimPlot(spe_sub, var = "CD45RO", reduction.use = "UMAP",
                   assay = "exprs", size = 0.2) +
  scale_color_viridis(name = "CD45RO") +
  ggtitle("CD45RO on UMAP")

p6 <- dittoDimPlot(spe_sub, var = "CD68", reduction.use = "UMAP",
                   assay = "exprs", size = 0.2) +
  scale_color_viridis(name = "CD68") +
  ggtitle("CD68 on UMAP")

(p3 + p4) / (p5 + p6)
```

------------------------------------------------------------------------

## 11. Save Filtered SPE

```{r save-spe}
saveRDS(spe, file.path(steinbock_path, "spe_segmentation_qc.rds"))
cat("Saved filtered SPE to", file.path(steinbock_path, "spe_segmentation_qc.rds"), "\n")
```

------------------------------------------------------------------------

## Session Info

```{r session-info}
sessionInfo()
```
