---
title: "Patient 20 (4/25 and 5/19) QC + Cell Segmentation"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

# Setup
```{r setup, message=FALSE, warning=FALSE}
.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(ggplot2)
library(ggridges)
library(dplyr)
library(reshape2)
library(stringr)
library(imcRtools)
library(SpatialExperiment)
library(SingleCellExperiment)
library(cytomapper)
library(dittoSeq)
library(scuttle)
library(viridis)

set.seed(42)
```

# Define Paths
```{r paths}
# Raw (pre-filtration) pixel data
data_dir_raw_425  <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/steinbock/4-25-2024_Patient20-T/"
data_dir_raw_519  <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/Data/Imaging/IMC/Tumor/PardeeTumorAblations/Patient_20T_5-19-2024/"

# Filtered pixel data (steinbock raw/ folder, which is actually filtered)
data_dir_filt_425 <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/steinbock/p20_4_25_results/filtered/"
data_dir_filt_519 <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/steinbock/p20_5_19_results/filtered/"

# Steinbock output directories
steinbock_dir_425 <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/steinbock/p20_4_25_results/"
steinbock_dir_519 <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/steinbock/p20_5_19_results/"
```

# List Files
```{r list-files}
txt_files_raw_425  <- list.files(data_dir_raw_425,  pattern = "\\.txt$", full.names = TRUE)
txt_files_raw_519  <- list.files(data_dir_raw_519,  pattern = "\\.txt$", full.names = TRUE)
txt_files_filt_425 <- list.files(data_dir_filt_425, pattern = "\\.txt$", full.names = TRUE)
txt_files_filt_519 <- list.files(data_dir_filt_519, pattern = "\\.txt$", full.names = TRUE)

cat("Raw files (4-25):", length(txt_files_raw_425), "\n")
cat("Raw files (5-19):", length(txt_files_raw_519), "\n")
cat("Filtered files (4-25):", length(txt_files_filt_425), "\n")
cat("Filtered files (5-19):", length(txt_files_filt_519), "\n")
```

# Set Marker Columns
```{r marker-cols}
# Read from first available raw file to get marker columns
marker_cols <- colnames(read.delim(txt_files_raw_425[1], check.names = FALSE))[7:47]
cat("Markers found:", length(marker_cols), "\n")
```

# Load, Sample, and Reshape ROI txt Files
```{r load-function}
load_rois <- function(txt_files, label, n_sample = 50000) {
  all_rois <- lapply(seq_along(txt_files), function(i) {
    df         <- read.delim(txt_files[i], check.names = FALSE)
    df_markers <- df[, marker_cols]
    n          <- min(n_sample, nrow(df_markers))
    df_sample  <- df_markers[sample(nrow(df_markers), n), ]
    df_sample$ROI <- paste0("ROI_", i)
    df_sample
  })
  pixels <- do.call(rbind, all_rois)
  pixels_long <- melt(pixels, id.vars = "ROI",
                      variable.name = "marker",
                      value.name    = "intensity")
  pixels_long$dataset <- label
  pixels_long
}
```


```{r load-data}
pixels_raw_425  <- load_rois(txt_files_raw_425,  label = "Raw_425")
pixels_raw_519  <- load_rois(txt_files_raw_519,  label = "Raw_519")
pixels_filt_425 <- load_rois(txt_files_filt_425, label = "Filtered_425")
pixels_filt_519 <- load_rois(txt_files_filt_519, label = "Filtered_519")

# Apply asinh transformation to raw datasets
pixels_raw_425  <- pixels_raw_425  %>% mutate(intensity_asinh = asinh(intensity / 1))
pixels_raw_519  <- pixels_raw_519  %>% mutate(intensity_asinh = asinh(intensity / 1))
pixels_filt_425 <- pixels_filt_425 %>% mutate(intensity_asinh = asinh(intensity / 1))
pixels_filt_519 <- pixels_filt_519 %>% mutate(intensity_asinh = asinh(intensity / 1))

cat("Raw pixels (4-25):", nrow(pixels_raw_425), "\n")
cat("Raw pixels (5-19):", nrow(pixels_raw_519), "\n")
cat("Filtered pixels (4-25):", nrow(pixels_filt_425), "\n")
cat("Filtered pixels (5-19):", nrow(pixels_filt_519), "\n")
```

```{r}
# Quick sanity check lol
cat("Max raw intensity (4-25):", max(pixels_raw_425$intensity), "\n")
cat("Max filtered intensity (4-25):", max(pixels_filt_425$intensity), "\n")
```

# Focused Markers
```{r focused-markers}
focused_markers <- c(
  "TIM3(Nd143Di)",
  "EpCAM(Nd144Di)",
  "SMA(Pr141Di)",
  "CD19(Nd142Di)",
  "CD15(Sm149Di)",
  "CD1c(Sm154Di)",
  "CD103(Gd158Di)",
  "CD68(Tb159Di)",
  "FoxP3(Dy162Di)",
  "LAMP3(Dy163Di)",
  "HLADRB(Yb171Di)",
  "CD28(Yb174Di)",
  "CD14(Lu175Di)",
  "CD38(Pt198Di)",
  "TTF1orCD16(Pt194Di)",
  "CD163(Sm147Di)"
)
```

---

# SECTION 1: Raw Pixels, No Transformation

## Histograms (ROI 1) — 4-25 Acquisition
```{r raw-hist-425, fig.width=14, fig.height=16}
roi1_raw_425 <- pixels_raw_425 %>% filter(ROI == "ROI_1")

ggplot(roi1_raw_425, aes(x = intensity)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", size = 0.1, alpha = 0.8) +
  facet_wrap(~ marker, scales = "free", ncol = 5) +
  labs(
    title    = "Pixel Intensity Landscape: Patient 20 - 4/25 Acquisition (ROI 1)",
    subtitle = "Raw values, no transformation",
    x        = "Raw Intensity",
    y        = "Pixel Count"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(size = 8),
    axis.text.y      = element_text(size = 8)
  )
```

## Histograms (ROI 1) — 5-19 Acquisition
```{r raw-hist-519, fig.width=14, fig.height=16}
roi1_raw_519 <- pixels_raw_519 %>% filter(ROI == "ROI_1")

ggplot(roi1_raw_519, aes(x = intensity)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", size = 0.1, alpha = 0.8) +
  facet_wrap(~ marker, scales = "free", ncol = 5) +
  labs(
    title    = "Pixel Intensity Landscape: Patient 20 - 5/19 Acquisition (ROI 1)",
    subtitle = "Raw values, no transformation",
    x        = "Raw Intensity",
    y        = "Pixel Count"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(size = 8),
    axis.text.y      = element_text(size = 8)
  )
```

## Density Ridge Plot (All ROIs) — 4-25
```{r raw-ridge-425, fig.width=14, fig.height=22}
ggplot(pixels_raw_425, aes(x = intensity, y = ROI, fill = ROI)) +
  geom_density_ridges(alpha = 0.8, scale = 0.9, rel_min_height = 0.01, show.legend = FALSE) +
  facet_wrap(~ marker, scales = "free_x", ncol = 5) +
  labs(
    title    = "Patient 20 (4/25) - Pixel Intensity Distributions per Marker",
    subtitle = "Raw values, no transformation",
    x        = "Raw Intensity",
    y        = "Region of Interest (ROI)"
  ) +
  theme_ridges(font_size = 10) +
  theme(
    strip.text   = element_text(face = "bold", size = 9),
    axis.text.y  = element_text(size = 8),
    axis.title   = element_text(size = 11, face = "bold")
  )
```

## Density Ridge Plot (All ROIs) — 5-19
```{r raw-ridge-519, fig.width=14, fig.height=22}
ggplot(pixels_raw_519, aes(x = intensity, y = ROI, fill = ROI)) +
  geom_density_ridges(alpha = 0.8, scale = 0.9, rel_min_height = 0.01, show.legend = FALSE) +
  facet_wrap(~ marker, scales = "free_x", ncol = 5) +
  labs(
    title    = "Patient 20 (5/19) - Pixel Intensity Distributions per Marker",
    subtitle = "Raw values, no transformation",
    x        = "Raw Intensity",
    y        = "Region of Interest (ROI)"
  ) +
  theme_ridges(font_size = 10) +
  theme(
    strip.text   = element_text(face = "bold", size = 9),
    axis.text.y  = element_text(size = 8),
    axis.title   = element_text(size = 11, face = "bold")
  )
```

---

# SECTION 2: Raw Pixels, asinh Transformed (All Markers)

## Histograms (ROI 1, asinh) — 4-25
```{r raw-asinh-hist-425, fig.width=14, fig.height=16}
roi1_raw_425 <- pixels_raw_425 %>% filter(ROI == "ROI_1")

ggplot(roi1_raw_425, aes(x = intensity_asinh)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", size = 0.1, alpha = 0.8) +
  facet_wrap(~ marker, scales = "free", ncol = 5) +
  labs(
    title    = "Pixel Intensity Landscape: Patient 20 - 4/25 (ROI 1)",
    subtitle = "Raw values, asinh transformed",
    x        = "asinh(Intensity)",
    y        = "Pixel Count"
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold", size = 9),
        panel.grid.minor = element_blank())
```

## Histograms (ROI 1, asinh) — 5-19
```{r raw-asinh-hist-519, fig.width=14, fig.height=16}
roi1_raw_519 <- pixels_raw_519 %>% filter(ROI == "ROI_1")

ggplot(roi1_raw_519, aes(x = intensity_asinh)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", size = 0.1, alpha = 0.8) +
  facet_wrap(~ marker, scales = "free", ncol = 5) +
  labs(
    title    = "Pixel Intensity Landscape: Patient 20 - 5/19 (ROI 1)",
    subtitle = "Raw values, asinh transformed",
    x        = "asinh(Intensity)",
    y        = "Pixel Count"
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold", size = 9),
        panel.grid.minor = element_blank())
```

## Density Ridge Plot (All ROIs, asinh) — 4-25
```{r raw-asinh-ridge-425, fig.width=14, fig.height=22}
ggplot(pixels_raw_425, aes(x = intensity_asinh, y = ROI, fill = ROI)) +
  geom_density_ridges(alpha = 0.8, scale = 0.9, rel_min_height = 0.01, show.legend = FALSE) +
  facet_wrap(~ marker, scales = "free_x", ncol = 5) +
  labs(
    title    = "Patient 20 (4/25) - Pixel Intensity Distributions",
    subtitle = "Raw values, asinh transformed",
    x        = "asinh(Intensity)", y = "ROI"
  ) +
  theme_ridges(font_size = 10) +
  theme(strip.text = element_text(face = "bold", size = 9))
```

## Density Ridge Plot (All ROIs, asinh) — 5-19
```{r raw-asinh-ridge-519, fig.width=14, fig.height=22}
ggplot(pixels_raw_519, aes(x = intensity_asinh, y = ROI, fill = ROI)) +
  geom_density_ridges(alpha = 0.8, scale = 0.9, rel_min_height = 0.01, show.legend = FALSE) +
  facet_wrap(~ marker, scales = "free_x", ncol = 5) +
  labs(
    title    = "Patient 20 (5/19) - Pixel Intensity Distributions",
    subtitle = "Raw values, asinh transformed",
    x        = "asinh(Intensity)", y = "ROI"
  ) +
  theme_ridges(font_size = 10) +
  theme(strip.text = element_text(face = "bold", size = 9))
```


---

# SECTION 3: Filtered Pixels (0-100 Scaled), No Transformation

## Histograms (ROI 1) — 4-25
```{r filt-hist-425, fig.width=14, fig.height=16}
roi1_filt_425 <- pixels_filt_425 %>% filter(ROI == "ROI_1")

ggplot(roi1_filt_425, aes(x = intensity)) +
  geom_histogram(bins = 50, fill = "darkorange", color = "white", size = 0.1, alpha = 0.8) +
  facet_wrap(~ marker, scales = "free", ncol = 5) +
  labs(
    title    = "Pixel Intensity Landscape: Patient 20 - 4/25 (ROI 1)",
    subtitle = "Filtered values, 0-100 scaled",
    x        = "Intensity (0-100)",
    y        = "Pixel Count"
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold", size = 9),
        panel.grid.minor = element_blank())
```

## Histograms (ROI 1) — 5-19
```{r filt-hist-519, fig.width=14, fig.height=16}
roi1_filt_519 <- pixels_filt_519 %>% filter(ROI == "ROI_1")

ggplot(roi1_filt_519, aes(x = intensity)) +
  geom_histogram(bins = 50, fill = "darkorange", color = "white", size = 0.1, alpha = 0.8) +
  facet_wrap(~ marker, scales = "free", ncol = 5) +
  labs(
    title    = "Pixel Intensity Landscape: Patient 20 - 5/19 (ROI 1)",
    subtitle = "Filtered values, 0-100 scaled",
    x        = "Intensity (0-100)",
    y        = "Pixel Count"
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold", size = 9),
        panel.grid.minor = element_blank())
```

## Compare Max Intensity Before vs After Filtering — 4-25
```{r max-compare-425}
pixels_raw_425 %>%
  group_by(marker) %>%
  summarise(max_raw = max(intensity)) %>%
  left_join(
    pixels_filt_425 %>%
      group_by(marker) %>%
      summarise(max_filtered = max(intensity)),
    by = "marker"
  ) %>%
  mutate(diff = max_filtered - max_raw) %>%
  arrange(desc(diff))
```

## Compare Max Intensity Before vs After Filtering — 5-19
```{r max-compare-519}
pixels_raw_519 %>%
  group_by(marker) %>%
  summarise(max_raw = max(intensity)) %>%
  left_join(
    pixels_filt_519 %>%
      group_by(marker) %>%
      summarise(max_filtered = max(intensity)),
    by = "marker"
  ) %>%
  mutate(diff = max_filtered - max_raw) %>%
  arrange(desc(diff))
```

---

# SECTION 4: Spatial Visualization (Filtered, 0-100 Scaled)

```{r spatial-markers}
# Added CD8a
focused_markers_spatial <- c(
  "EpCAM(Nd144Di)",
  "SMA(Pr141Di)",
  "FoxP3(Dy162Di)",
  "CD68(Tb159Di)",
  "CD163(Sm147Di)",
  "CD1c(Sm154Di)",
  "TIM3(Nd143Di)",
  "PD1(Ho165Di)",
  "TTF1orCD16(Pt194Di)",
  "CD8a(Gd160Di)"
)
```

## Spatial Plot (ROI 1) — 4-25
```{r spatial-425, fig.width=12, fig.height=14}
roi1_filt_raw_425 <- read.delim(txt_files_filt_425[1], check.names = FALSE)

set.seed(42)
roi1_filt_sub_425 <- roi1_filt_raw_425 %>%
  filter(X %% 3 == 0, Y %% 3 == 0)

roi1_filt_long_425 <- roi1_filt_sub_425 %>%
  select(X, Y, all_of(focused_markers_spatial)) %>%
  melt(id.vars = c("X", "Y"),
       variable.name = "marker",
       value.name    = "intensity")

ggplot(roi1_filt_long_425, aes(x = X, y = Y, fill = intensity)) +
  geom_tile() +
  scale_fill_viridis_c(option = "viridis") +
  facet_wrap(~ marker, ncol = 3) +
  labs(
    title    = "Patient 20 (4/25) ROI 1 - Selected Immune Markers (Filtered)",
    subtitle = "Filtered values, 0-100 scaled, downsampled every 3rd pixel",
    fill     = "Intensity\n(0-100)"
  ) +
  theme_void(base_size = 9) +
  theme(strip.text = element_text(size = 8, face = "bold"),
        legend.position = "bottom")
```

## Spatial Plot (ROI 1) — 5-19
```{r spatial-519, fig.width=12, fig.height=14}
roi1_filt_raw_519 <- read.delim(txt_files_filt_519[1], check.names = FALSE)

set.seed(42)
roi1_filt_sub_519 <- roi1_filt_raw_519 %>%
  filter(X %% 3 == 0, Y %% 3 == 0)

roi1_filt_long_519 <- roi1_filt_sub_519 %>%
  select(X, Y, all_of(focused_markers_spatial)) %>%
  melt(id.vars = c("X", "Y"),
       variable.name = "marker",
       value.name    = "intensity")

ggplot(roi1_filt_long_519, aes(x = X, y = Y, fill = intensity)) +
  geom_tile() +
  scale_fill_viridis_c(option = "viridis") +
  facet_wrap(~ marker, ncol = 3) +
  labs(
    title    = "Patient 20 (5/19) ROI 1 - Selected Immune Markers (Filtered)",
    subtitle = "Filtered values, 0-100 scaled, downsampled every 3rd pixel",
    fill     = "Intensity\n(0-100)"
  ) +
  theme_void(base_size = 9) +
  theme(strip.text = element_text(size = 8, face = "bold"),
        legend.position = "bottom")
```

---

# SECTION 5: Cell Segmentation QC

## Load Steinbock Output into SpatialExperiment — 4-25
```{r load-spe-425}
spe_425 <- read_steinbock(steinbock_dir_425)
spe_425

table(spe_425$sample_id)
```

## Load Steinbock Output into SpatialExperiment — 5-19
```{r load-spe-519}
spe_519 <- read_steinbock(steinbock_dir_519)
spe_519

table(spe_519$sample_id)
```

## Add Metadata
```{r metadata}
# 4-25
colnames(spe_425) <- paste0(spe_425$sample_id, "_", spe_425$ObjectNumber)
spe_425$image_id   <- spe_425$sample_id
spe_425$patient_id <- "Patient20"
spe_425$acquisition <- "4-25"

# 5-19
colnames(spe_519) <- paste0(spe_519$sample_id, "_", spe_519$ObjectNumber)
spe_519$image_id   <- spe_519$sample_id
spe_519$patient_id <- "Patient20"
spe_519$acquisition <- "5-19"

cat("Cells before filtering (4-25):", ncol(spe_425), "\n")
cat("Cells before filtering (5-19):", ncol(spe_519), "\n")
```

5-19 acquisition has more cells (makes sense as it has 6 ROIs, while 4-25 has 5 ROIs). 


## Cell Count Per ROI

# 4-25
```{r cellcount-425}
cell_counts_425 <- as.data.frame(table(spe_425$sample_id))
colnames(cell_counts_425) <- c("ROI", "cell_count")

ggplot(cell_counts_425, aes(x = ROI, y = cell_count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = 100, color = "red", linetype = "dashed") +
  labs(title = "Cell Count per ROI — Patient 20 (4/25)",
       subtitle = "Red line = minimum threshold (100 cells)",
       x = "ROI", y = "Cell Count") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# 5-19
```{r cellcount-519}
cell_counts_519 <- as.data.frame(table(spe_519$sample_id))
colnames(cell_counts_519) <- c("ROI", "cell_count")

ggplot(cell_counts_519, aes(x = ROI, y = cell_count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = 100, color = "red", linetype = "dashed") +
  labs(title = "Cell Count per ROI — Patient 20 (5/19)",
       subtitle = "Red line = minimum threshold (100 cells)",
       x = "ROI", y = "Cell Count") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Area Distribution Before Filtering
```{r area-dist, fig.width=10, fig.height=4}
area_df <- bind_rows(
  as.data.frame(colData(spe_425)) %>% mutate(acquisition = "4-25"),
  as.data.frame(colData(spe_519)) %>% mutate(acquisition = "5-19")
)

ggplot(area_df, aes(x = area, fill = acquisition)) +
  geom_histogram(bins = 100, color = "white", alpha = 0.7, position = "identity") +
  geom_vline(xintercept = c(20, 500), color = "red", linetype = "dashed") +
  facet_wrap(~ acquisition) +
  labs(title = "Cell Area Distribution — Both Acquisitions (red lines = filter thresholds)",
       x = "Area (pixels)", y = "Count") +
  theme_classic()
```

## Filter by Area
```{r area-filter}
spe_425 <- spe_425[, spe_425$area >= 20 & spe_425$area < 500]
spe_519 <- spe_519[, spe_519$area >= 20 & spe_519$area < 500]

cat("Cells after filtering (4-25):", ncol(spe_425), "\n")
cat("Cells after filtering (5-19):", ncol(spe_519), "\n")
```

## Apply asinh Transformation and Save
```{r asinh-save}
assay(spe_425, "asin") <- asinh(counts(spe_425) / 1)
assay(spe_519, "asin") <- asinh(counts(spe_519) / 1)

saveRDS(spe_425, file.path(steinbock_dir_425, "p20_425_spe_filtered.rds"))
saveRDS(spe_519, file.path(steinbock_dir_519, "p20_519_spe_filtered.rds"))
```

## SNR per Marker/ROI
```{r}
# SNR = mean signal / sd of signal per marker per ROI
compute_snr <- function(spe, label) {
  mat <- assay(spe, "asin")
  snr_df <- lapply(unique(spe$sample_id), function(roi) {
    cells <- mat[, spe$sample_id == roi]
    data.frame(
      ROI    = roi,
      marker = rownames(mat),
      mean   = rowMeans(cells),
      sd     = apply(cells, 1, sd)
    )
  })
  snr_df <- do.call(rbind, snr_df) %>%
    mutate(SNR = mean / (sd + 1e-6),
           acquisition = label)
  snr_df
}

snr_425 <- compute_snr(spe_425, "4-25")
snr_519 <- compute_snr(spe_519, "5-19")
```

# SNR Plot (asinh)
```{r}
# Combine both acquisitions for comparison
snr_combined <- bind_rows(snr_425, snr_519)

ggplot(snr_combined, aes(x = mean, y = SNR, color = acquisition)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_text(aes(label = marker), size = 2.5, vjust = -0.5, check_overlap = TRUE) +
  facet_wrap(~ acquisition) +
  scale_color_manual(values = c("4-25" = "steelblue", "5-19" = "darkorange")) +
  labs(
    title    = "Signal-to-Noise Ratio per Marker — Patient 20",
    subtitle = "Each point represents a marker averaged across all ROIs",
    x        = "Mean Signal (asinh)",
    y        = "SNR (mean / sd)"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.text      = element_text(face = "bold", size = 10),
    legend.position = "none",
    panel.grid.minor = element_blank()
  )
```

4/25 Dataset:
- CD38, CD127, CD15, CTLA4, TIM3 — high mean signal and high SNR
- CD68, MKI67, LAG3, EpCAM — low mean signal and low SNR

5/19 Dataset:
CD38, CD15, DNA193, DNA191, TTF1orCD16 — very high SNR
LAG3, LAMP3, HLADRB - low mean signal and low SNR

# SNR plot (log)
```{r}
ggplot(snr_combined, aes(x = mean, y = SNR, color = acquisition)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_text(aes(label = marker), size = 2.5, vjust = -0.5, check_overlap = TRUE, color = "black") +
  facet_wrap(~ acquisition) +
  scale_color_manual(values = c("4-25" = "steelblue", "5-19" = "darkorange")) +
  scale_x_continuous(trans = "log2") +
  scale_y_continuous(trans = "log2") +
  labs(
    title    = "Signal-to-Noise Ratio per Marker — Patient 20",
    subtitle = "Each point represents a marker averaged across all ROIs",
    x        = "Mean Signal (asinh, log2 scale)",
    y        = "SNR (log2 scale)"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.text       = element_text(face = "bold", size = 10),
    legend.position  = "none",
    panel.grid.minor = element_blank()
  )
```

---

# SECTION 6: Segmentation Mask Overlay

## Load Images and Masks — 4-25
```{r load-images-425}
images_425 <- loadImages(file.path(steinbock_dir_425, "img/"))
masks_425  <- loadImages(file.path(steinbock_dir_425, "masks_deepcell/"), as.is = TRUE)

channelNames(images_425)     <- rownames(spe_425)
mcols(images_425)$sample_id  <- names(images_425)
mcols(masks_425)$sample_id   <- names(masks_425)

cat("Images (4-25):", length(images_425), "\n")
cat("Masks (4-25):", length(masks_425), "\n")
```

## Load Images and Masks — 5-19
```{r load-images-519}
images_519 <- loadImages(file.path(steinbock_dir_519, "img/"))
masks_519  <- loadImages(file.path(steinbock_dir_519, "masks_deepcell/"), as.is = TRUE)

channelNames(images_519)     <- rownames(spe_519)
mcols(images_519)$sample_id  <- names(images_519)
mcols(masks_519)$sample_id   <- names(masks_519)

cat("Images (5-19):", length(images_519), "\n")
cat("Masks (5-19):", length(masks_519), "\n")
```

## Normalize Images
```{r normalize}
cur_images_425 <- cytomapper::normalize(images_425, separateImages = TRUE)
cur_images_425 <- cytomapper::normalize(cur_images_425, inputRange = c(0, 0.2))

cur_images_519 <- cytomapper::normalize(images_519, separateImages = TRUE)
cur_images_519 <- cytomapper::normalize(cur_images_519, inputRange = c(0, 0.2))
```

## Plot 1: DNA + EpCAM + CD68 + SMA — 4-25
```{r overlay-425, fig.height=12}
plotPixels(
  cur_images_425,
  mask           = masks_425,
  img_id         = "sample_id",
  missing_colour = "white",
  colour_by      = c("DNA191", "EpCAM", "CD68", "SMA"),
  colour         = list(
    DNA191   = c("black", "blue"),
    EpCAM    = c("black", "green"),
    CD68     = c("black", "yellow"),
    SMA      = c("black", "red")
  ),
  legend = list(colour_by.title.cex = 1, colour_by.labels.cex = 1),
  image_title = list(text = mcols(cur_images_425)$sample_id,
                     position = "top",
                     colour = "white",
                     margin = c(5, 5),
                     cex = 1.5)
)
```

## Plot 2: DNA + EpCAM + CD68 + SMA — 5-19
```{r overlay-519, fig.height=12}
plotPixels(
  cur_images_519,
  mask           = masks_519,
  img_id         = "sample_id",
  missing_colour = "white",
  colour_by      = c("DNA191", "EpCAM", "CD68", "SMA"),
  colour         = list(
    DNA191   = c("black", "blue"),
    EpCAM    = c("black", "green"),
    CD68     = c("black", "yellow"),
    SMA      = c("black", "red")
  ),
  legend = list(colour_by.title.cex = 1, colour_by.labels.cex = 1),
  image_title = list(text = mcols(cur_images_519)$sample_id,
                     position = "top",
                     colour = "white",
                     margin = c(5, 5),
                     cex = 1.5)
)
```

## Plot 3: DNA + EpCAM — 4-25
```{r epcam-425, fig.height=12}
plotPixels(cur_images_425, mask = masks_425, img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("DNA191", "EpCAM"),
           colour = list(DNA191 = c("black", "blue"), EpCAM = c("black", "green")
                         ),
           legend = list(colour_by.title.cex = 1.5, colour_by.labels.cex = 1.5),
           image_title = list(text = mcols(cur_images_425)$sample_id,
                     position = "top",
                     colour = "white",
                     margin = c(5, 5),
                     cex = 1.5)
          )
```

## Plot 4: DNA + EpCAM — 5-19
```{r epcam-519, fig.height=12}
plotPixels(cur_images_519, mask = masks_519, img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("DNA191", "EpCAM"),
           colour = list(DNA191 = c("black", "blue"), EpCAM = c("black", "green")),
           legend = list(colour_by.title.cex = 1.5, colour_by.labels.cex = 1.5),
           image_title = list(text = mcols(cur_images_519)$sample_id,
                     position = "top",
                     colour = "white",
                     margin = c(5, 5),
                     cex = 1.5)
           )
```

## Plot 5: DNA + CD68 — 4-25
```{r cd68-425, fig.height=12}
plotPixels(cur_images_425, mask = masks_425, img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("DNA191", "CD68"),
           colour = list(DNA191 = c("black", "blue"), CD68 = c("black", "yellow")),
           legend = list(colour_by.title.cex = 1.5, colour_by.labels.cex = 1.5),
           image_title = list(text = mcols(cur_images_425)$sample_id,
                     position = "top",
                     colour = "white",
                     margin = c(5, 5),
                     cex = 1.5))
```

## Plot 6: DNA + CD68 — 5-19
```{r cd68-519, fig.height=12}
plotPixels(cur_images_519, mask = masks_519, img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("DNA191", "CD68"),
           colour = list(DNA191 = c("black", "blue"), CD68 = c("black", "yellow")),
           legend = list(colour_by.title.cex = 1.5, colour_by.labels.cex = 1.5),
           image_title = list(text = mcols(cur_images_519)$sample_id,
                     position = "top",
                     colour = "white",
                     margin = c(5, 5),
                     cex = 1.5))
```

---

# SECTION 7: Signal Quality — Heatmaps

```{r markers-to-plot}
markers_to_plot <- rownames(spe_425)[!rownames(spe_425) %in% c("DNA191", "DNA193", "Membrane")]
```

## Single Cell Heatmap (2000 subsampled cells) — 4-25
```{r heatmap-cells-425, fig.height=10, fig.width=10}
set.seed(42)
cur_cells_425 <- sample(seq_len(ncol(spe_425)), 2000)

dittoHeatmap(
  spe_425[, cur_cells_425],
  genes          = markers_to_plot,
  assay          = "asin",
  cluster_cols   = TRUE,
  scale          = "row",
  heatmap.colors = viridis(100),
  annot.by       = "sample_id",
  main           = "Patient 20 (4/25) - Single Cell Expression Heatmap\nasinh-transformed, 2000 subsampled cells"
)
```

## Single Cell Heatmap (2000 subsampled cells) — 5-19
```{r heatmap-cells-519, fig.height=10, fig.width=10}
set.seed(42)
cur_cells_519 <- sample(seq_len(ncol(spe_519)), 2000)

dittoHeatmap(
  spe_519[, cur_cells_519],
  genes          = markers_to_plot,
  assay          = "asin",
  cluster_cols   = TRUE,
  scale          = "row",
  heatmap.colors = viridis(100),
  annot.by       = "sample_id",
  main           = "Patient 20 (5/19) - Single Cell Expression Heatmap\nasinh-transformed, 2000 subsampled cells"
)
```

Marker ordering is different between the two acquisition datasets. 
This suggests that the two acquisitions captured different tissue regions with
different cell type compositions, which is expected because they have different ROIs.

## Mean Marker Expression per ROI — 4-25
```{r heatmap-roi-425, fig.height=10, fig.width=10}
image_mean_425 <- aggregateAcrossCells(spe_425,
                                       ids = spe_425$sample_id,
                                       statistics = "mean",
                                       use.assay.type = "asin")

dittoHeatmap(
  image_mean_425,
  genes          = markers_to_plot,
  assay          = "asin",
  cluster_cols   = TRUE,
  scale          = "none",
  heatmap.colors = viridis(100),
  annot.by       = "sample_id",
  main           = "Patient 20 (4/25) - Mean Marker Expression per ROI\nasinh-transformed"
)
```

## Mean Marker Expression per ROI — 5-19
```{r heatmap-roi-519, fig.height=10, fig.width=10}
image_mean_519 <- aggregateAcrossCells(spe_519,
                                       ids = spe_519$sample_id,
                                       statistics = "mean",
                                       use.assay.type = "asin")

dittoHeatmap(
  image_mean_519,
  genes          = markers_to_plot,
  assay          = "asin",
  cluster_cols   = TRUE,
  scale          = "none",
  heatmap.colors = viridis(100),
  annot.by       = "sample_id",
  main           = "Patient 20 (5/19) - Mean Marker Expression per ROI\nasinh-transformed"
)
```

---

# Summary: Data Quality Assessment

**4-25 acquisition**

- Has 5 ROIs with highly variable cell counts (ROI 1 is much larger at about 15,600 cells vs 1300-5100 for the others). 

- Pixel intensity distributions show good signal with clear bimodal distributions for key immune markers. 

- Spatial plots show clear tissue architecture with well-defined EpCAM and SMA structures. 

- ROI 4 shows distinct marker expression profile compared to other ROIs in the mean expression heatmap, which suggests biological heterogeneity across the tissue. 

- Segmentation quality appears reasonable but ROI size variability may introduce technical noise. 

**5-19 acquisition**

- Has 6 ROIs with more consistent cell counts (2000-12000 cells) and more balanced ROI representation

- Density plots show consistent distributions across all 6 ROIs, not a lot of variability between ROIs. 

- SNR scatter plot generally shows higher mean signal compared to 4/25

- Mean expression heatmap shows more uniform marker expression across ROIs 

- White lines in spatial plots indicate stitching artifacts from acquisition but don't affect the data quality.

**Conclusion**
Both acquisitions pass QC with no ROIs requiring removal. Both datasets can be included in downstream analysis to maximize tissue coverage and cell numbers for Patient 20. However, given that the ROIs are more consistent with each other in the 5/19 acquisition across multiple QC metrics, it will be used for assessing batch effects with all patients.
