---
title: "B Cell/Plasma Cell Subclustering — Run5 Extraction - w/o Cluster 15"
author: "Tazein Shah"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(42)
```

## Overview

Extract B cell/plasma cell clusters (3, 10, 16) from the
annotated run5 SPE (`h5_cluster_annot.rds`) and recluster on the
existing Harmony embedding (Louvain, k = 20). This run has no
Unknown/Low Quality or Treg/T Cell admixture subclusters, so the
pipeline goes straight from subclustering to annotation to save.

---

## 1. Load Libraries

```{r libraries}
library(imcRtools)
library(scran)
library(bluster)
library(BiocParallel)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(scater)
library(patchwork)
library(RColorBrewer)
```

---

## 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

spe_path <- file.path(analysis_path, "final_cell_annotation", "h5_cluster_annot.rds")

bcell_clustering_path <- file.path(analysis_path, "final_cell_annotation", "subclustering","bcells")

out_path <- file.path(bcell_clustering_path, "bcell_subclustering_wo15.rds")
```

---

## 3. Load Annotated SPE

```{r load-data}
spe <- readRDS(spe_path)

print(table(spe$broad_clusters))
```

---

## 4. Extract B Cell/Plasma Clusters

```{r subset}
bcell_clusters <- c("3", "10", "16")

spe_bcell <- spe[, as.character(spe$broad_clusters) %in% bcell_clusters]

cat("Cells in bcell subset:", ncol(spe_bcell), "\n")
cat("From cluster 3 (B Cell CD38+):", sum(spe_bcell$broad_clusters == "3"), "\n")
cat("From cluster 10 (Memory T/Plasma Cell):", sum(spe_bcell$broad_clusters == "10"), "\n")
cat("From cluster 16 (Plasma Cell CD19+):", sum(spe_bcell$broad_clusters == "16"), "\n")
```

---

## 5. Recluster: Louvain on Existing Harmony Embedding (k = 20)

```{r recluster}
snn_graph <- buildSNNGraph(spe_bcell, use.dimred = "Harmony", k = 20,
                            BPPARAM = SerialParam())

clusters <- igraph::cluster_louvain(snn_graph)$membership

spe_bcell$bcell_subclusters <- factor(clusters)

cat("Number of subclusters:", nlevels(spe_bcell$bcell_subclusters), "\n")
print(table(spe_bcell$bcell_subclusters))
```

---

## 6. UMAP on the Subset

```{r umap-sub}
set.seed(220225)
spe_bcell <- runUMAP(spe_bcell, dimred = "Harmony", name = "UMAP_sub")
```

```{r plot-umap-sub-original, fig.width=10, fig.height=7}
p_sub_original <- dittoDimPlot(spe_bcell, var = "broad_celltype",
                               reduction.use = "UMAP_sub",
                               size = 0.4) +
  ggtitle("B Cell/Plasma Subset: UMAP colored by Original Broad Label") +
  theme(legend.position = "right")
print(p_sub_original)
```

```{r plot-patient-composition, fig.width=10, fig.height=5}
patient_counts <- as.data.frame(table(patient_id = spe_bcell$patient_id))

p_patient_comp <- ggplot(patient_counts, aes(x = patient_id, y = Freq, fill = patient_id)) +
  geom_col() +
  labs(title = "B Cell/Plasma Subset: Cell Count by Patient",
       x = "Patient", y = "Cell Count") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
print(p_patient_comp)

cat("B cell/plasma subset composition by patient:\n")
print(patient_counts[order(-patient_counts$Freq), ])
```

---

## 7. Marker Sets for B Cell/Plasma Subtyping

```{r markers}
# core B cell/plasma lineage markers
lineage_markers <- c("CD19", "CD38", "CD138")

# activation / functional state markers
state_markers <- c("CD45RO", "MKI67", "PD1", "PDL1", "TIM3", "HLADRB")

# admixture / spillover flags (not B-lineage markers, but useful for
# catching doublets or spatial bleed from other lineages)
admixture_markers <- c("FoxP3", "CD1c", "CD15")

bcell_markers <- c(lineage_markers, state_markers, admixture_markers)
bcell_markers <- bcell_markers[bcell_markers %in% rownames(spe_bcell)]

cat("Markers found:", paste(bcell_markers, collapse = ", "), "\n")
```

---

## 8. Violin Plots per Subcluster

```{r make-violinplot}
make_violinplot <- function(spe, assay_name, cluster_col, marker_set, title, ncol = 3) {
  expr_mat   <- assay(spe, assay_name)[marker_set, , drop = FALSE]
  cluster_id <- spe[[cluster_col]]

  mean_expr <- sapply(levels(cluster_id), function(cl) {
    idx <- which(cluster_id == cl)
    rowMeans(expr_mat[, idx, drop = FALSE])
  })

  if (is.matrix(mean_expr) && nrow(mean_expr) > 1) {
    marker_order <- rownames(mean_expr)[hclust(dist(mean_expr))$order]
  } else {
    marker_order <- marker_set
  }

  plot_df <- as.data.frame(t(expr_mat)) |>
    mutate(cluster = cluster_id) |>
    pivot_longer(-cluster, names_to = "marker", values_to = "expr")

  plot_df$marker  <- factor(plot_df$marker, levels = marker_order)
  plot_df$cluster <- factor(plot_df$cluster, levels = levels(cluster_id))

  ggplot(plot_df, aes(x = cluster, y = expr, fill = cluster)) +
    geom_violin(scale = "width", trim = TRUE, linewidth = 0.2) +
    stat_summary(fun = median, geom = "point", size = 0.6, color = "black") +
    facet_wrap(~ marker, scales = "free_y", ncol = ncol) +
    labs(title = title, x = "Subcluster", y = "Z-score expression") +
    theme_bw(base_size = 11) +
    theme(
      axis.text.x      = element_text(angle = 0, hjust = 0.5, size = 9),
      strip.text        = element_text(face = "bold"),
      panel.grid.major.x = element_blank(),
      legend.position   = "none"
    )
}
```

```{r violin-bcell-global-lineage, fig.width=10, fig.height=5}
p_violin_lineage_global <- make_violinplot(spe_bcell, "scaled_0100", "bcell_subclusters",
                                           lineage_markers,
                                           "B Cell/Plasma Subset: Lineage Markers per Subcluster (Global Z-score)",
                                           ncol = 3)
print(p_violin_lineage_global)
```

```{r violin-bcell-global-state, fig.width=12, fig.height=8}
p_violin_state_global <- make_violinplot(spe_bcell, "scaled_0100", "bcell_subclusters",
                                         state_markers,
                                         "B Cell/Plasma Subset: State Markers per Subcluster (Global Z-score)",
                                         ncol = 3)
print(p_violin_state_global)
```

```{r violin-bcell-global-admixture, fig.width=10, fig.height=5}
p_violin_admixture_global <- make_violinplot(spe_bcell, "scaled_0100", "bcell_subclusters",
                                             admixture_markers,
                                             "B Cell/Plasma Subset: Admixture Markers per Subcluster (Global Z-score)",
                                             ncol = 3)
print(p_violin_admixture_global)
```

Subset-relative renormalization: since `scaled_0100` was capped and Z-scored
over the full run5 dataset, this recomputes the same cap+Z-score recipe
from raw counts using only this B cell/plasma subset's own distribution --
one normalization layer, just a different denominator.

```{r renorm-subset}
cap_99 <- function(mat) {
  t(apply(mat, 1, function(x) {
    q99 <- quantile(x, 0.99)
    pmin(x, q99)
  }))
}

zscore_rows <- function(mat) {
  t(scale(t(mat)))
}

mat_sub_0100 <- counts(spe_bcell) |> cap_99() |> zscore_rows()
assay(spe_bcell, "scaled_0100_sub") <- mat_sub_0100
```

```{r violin-bcell-subset-lineage, fig.width=10, fig.height=5}
p_violin_lineage_sub <- make_violinplot(spe_bcell, "scaled_0100_sub", "bcell_subclusters",
                                        lineage_markers,
                                        "B Cell/Plasma Subset: Lineage Markers per Subcluster (Subset-Renormalized Z-score)",
                                        ncol = 3)
print(p_violin_lineage_sub)
```

```{r violin-bcell-subset-state, fig.width=12, fig.height=8}
p_violin_state_sub <- make_violinplot(spe_bcell, "scaled_0100_sub", "bcell_subclusters",
                                      state_markers,
                                      "B Cell/Plasma Subset: State Markers per Subcluster (Subset-Renormalized Z-score)",
                                      ncol = 3)
print(p_violin_state_sub)
```

```{r violin-bcell-subset-admixture, fig.width=10, fig.height=5}
p_violin_admixture_sub <- make_violinplot(spe_bcell, "scaled_0100_sub", "bcell_subclusters",
                                          admixture_markers,
                                          "B Cell/Plasma Subset: Admixture Markers per Subcluster (Subset-Renormalized Z-score)",
                                          ncol = 3)
print(p_violin_admixture_sub)
```

---

## 9. Manual Subcluster Annotation

No Unknown/Low Quality or Treg/T Cell admixture subclusters showed up
in this run's lineage/state markers, so there's no removal or
reclustering step -- this annotation is applied directly to `spe_bcell`
and carried through to the save step.

```{r annotate-sub, echo=TRUE}
bcell_subannotation <- c(
  "1" = "Plasmablast/Activated B Cell (CD38+)",  # CD38 ~2.0 sole standout, everything else negative
  "2" = "Plasma Cell, Memory-like (Weak) (CD45RO+)",  # only CD45RO positive (~0.55), rest flat -- dim echo
  "3" = "Plasmablast, Strongly Activated (CD38++)",  # CD38 ~4.0, a distinctly separate high-compressed distribution from cluster 1's ~2.0
  "4" = "Plasma Cell, TIM3+ (CD138+TIM3+CD45RO-)",  # CD138 ~0.95, TIM3 ~0.5 co-elevated, but CD45RO (-0.45) and PD1 (-0.4) both negative -- atypical, doesn't match the earlier CD45RO+PD1+ exhausted/effector pattern; worth checking against admixture markers
  "5" = "Transitional B Cell/Plasmablast (CD19+CD138+HLADRB+)",  # CD19 ~1.4, CD138 ~1.3, HLADRB ~0.35, CD45RO ~0.55 co-elevated
  "6" = "B Cell, Weak (CD19+)",  # CD19 ~0.5 modestly positive, everything else near baseline
  "7" = "B Cell/Plasma Cell, HLADRB-high (CD19+CD138+HLADRB+)",  # CD19 ~2.3, CD138 ~1.6, HLADRB ~0.85 (highest of any cluster) -- stronger version of cluster 5's co-expression pattern; could reflect transitional biology or APC-lineage admixture riding on HLADRB, needs the admixture panel to confirm
  "8" = "B Cell (CD19+)"  # CD19 ~2.85, sole standout, CD138 near baseline -- cleanest B cell signature
)

if (length(bcell_subannotation) > 0) {
  spe_bcell$sub_celltype <- bcell_subannotation[as.character(spe_bcell$bcell_subclusters)]
  cat("Subcluster annotation applied.\n")
  print(table(spe_bcell$sub_celltype, useNA = "ifany"))
} else {
  cat("No subcluster annotation yet. Fill in bcell_subannotation above.\n")
}
```

---

## 10. UMAP Colored by Subcluster Cell Type

```{r umap-sub-celltype, fig.width=10, fig.height=5}
if ("sub_celltype" %in% colnames(colData(spe_bcell))) {

  spe_bcell$sub_celltype <- factor(spe_bcell$sub_celltype, levels = bcell_subannotation)

  p_sub_celltype <- dittoDimPlot(spe_bcell, var = "sub_celltype",
                                 reduction.use = "UMAP_sub",
                                 size = 0.4) +
    ggtitle("B Cell/Plasma Subset: UMAP colored by Subcluster Cell Type") +
    theme(legend.position = "right")
  print(p_sub_celltype)

} else {
  cat("Run the annotate-sub chunk first to apply annotations.\n")
}
```

---

## 11. Cell Type Composition by Patient

```{r celltype-patient-composition, fig.width=14, fig.height=16}
celltype_patient_counts <- as.data.frame(
  table(patient_id = spe_bcell$patient_id,
        cell_type  = spe_bcell$sub_celltype)
)

p_celltype_patient <- ggplot(celltype_patient_counts, aes(x = patient_id, y = Freq, fill = patient_id)) +
  geom_col() +
  facet_wrap(~ cell_type, scales = "free", ncol = 2,
             labeller = labeller(cell_type = label_wrap_gen(width = 35))) +
  labs(title = "B Cell/Plasma Subset: Cell Type Composition by Patient",
       x = "Patient", y = "Cell Count") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        strip.text = element_text(face = "bold", size = 8),
        legend.position = "none")

print(p_celltype_patient)

# RStudio's Plots pane doesn't respect fig.width/fig.height when running
# chunks interactively (those only apply to a knitted document) -- save to
# file with fixed dimensions so the full plot is viewable without cutoff
celltype_patient_plot_path <- file.path(bcell_clustering_path, "celltype_composition_by_patient.png")
ggsave(celltype_patient_plot_path, p_celltype_patient, width = 14, height = 16, dpi = 300)
cat("Saved full-size plot to:", celltype_patient_plot_path, "\n")

#cat("Cell type composition by patient:\n")
#celltype_patient_nonzero <- celltype_patient_counts[celltype_patient_counts$Freq > 0, ]
#print(celltype_patient_nonzero[order(celltype_patient_nonzero$cell_type, -celltype_patient_nonzero$Freq), ])
```

---

## 12. Save Subclustered/Annotated Object

```{r save}
dir.create(dirname(out_path), recursive = TRUE, showWarnings = FALSE)
saveRDS(spe_bcell, out_path)
cat("Saved B cell subcluster object (annotated) to:", out_path, "\n")
```
