---
title: "Subclustering: Cluster 15 (T Cell/Plasma Cell, Mixed)"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)
set.seed(42)
```

## Overview

Subclustering cluster 15 (currently labeled "T Cell/Plasma Cell, Mixed
(CD4+/CD138+)") from the broad k=20 annotation. Unlike the myeloid
subclustering (which combined two broad clusters that looked related),
this is a single broad cluster that showed markers from two different
lineages at once. The goal here is to test whether it splits into a
real CD4+ T cell population and a real plasma cell population, or
whether it holds together as one population with some background/
spillover on one of the two marker sets.

## 1. Load Libraries

```{r libraries}
# Only run this if you start a new RStudio session
.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(imcRtools)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(scater)
library(scran)
library(bluster)
library(patchwork)
library(RColorBrewer)
library(ComplexHeatmap)
library(circlize)
```

## 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

spe_path <- file.path(analysis_path, "final_cell_annotation", "h5_cluster_annot.rds")
out_path <- file.path(analysis_path, "final_cell_annotation", "subclustering",
                      "cluster15_subcluster.rds")
findmarkers_path <- file.path(analysis_path, "final_cell_annotation", "subclustering",
                              "cluster15_subcluster_findmarkers.csv")
```

## 3. Load Annotated Data and Subset

```{r load-and-subset}
spe <- readRDS(spe_path)

spe_15 <- spe[, spe$broad_clusters == "15"]

cat("Cells in cluster 15 subset:", ncol(spe_15), "\n")
```

## 4. Re-cluster the Subset on the Existing Harmony Embedding

Starting with k = 10, same as the myeloid subclustering, to encourage
finer splits than the broad k = 20 pass.

```{r subcluster}
harmony_sub <- reducedDim(spe_15, "Harmony")
cat("Harmony embedding dimensions (subset):", dim(harmony_sub), "\n")

set.seed(220225)
g_sub     <- buildSNNGraph(t(harmony_sub), k = 10, type = "rank")
clust_sub <- igraph::cluster_louvain(g_sub)
spe_15$sub_clusters <- factor(clust_sub$membership)

cat("Number of subclusters:", nlevels(spe_15$sub_clusters), "\n")
print(table(spe_15$sub_clusters))
```

## 5. UMAP on the Subset

```{r umap-sub}
set.seed(220225)
spe_15 <- runUMAP(spe_15, dimred = "Harmony", name = "UMAP_sub")
```

```{r plot-umap-sub-clusters, fig.width=10, fig.height=7}
p_sub_clusters <- dittoDimPlot(spe_15, var = "sub_clusters",
                               reduction.use = "UMAP_sub",
                               size = 0.4, do.label = TRUE, labels.size = 4) +
  ggtitle("Cluster 15 Subset: UMAP colored by Subcluster") +
  theme(legend.position = "right")
print(p_sub_clusters)
```

## 6. Marker Sets for T Cell / Plasma Cell Disambiguation

T/checkpoint markers, plasma/B markers, and a few unrelated-lineage
markers included specifically to catch any surprise contamination
(similar to what flagged cluster 13's FoxP3/CD15 mix at the broad
level).

```{r markers}
disambiguation_markers <- c(
  "CD4", "CD8a", "CD45RO", "FoxP3", "TIM3", "PD1",
  "CD138", "CD38", "CD19",
  "CD14", "CD15", "CD117", "CD11c"
)
disambiguation_markers <- disambiguation_markers[disambiguation_markers %in% rownames(spe_15)]
cat("Markers found:", paste(disambiguation_markers, collapse = ", "), "\n")
```

## 7. Violin Plots per Subcluster

```{r make-violinplot}
make_violinplot <- function(spe, assay_name, cluster_col, marker_set, title, ncol = 3) {
  expr_mat   <- assay(spe, assay_name)[marker_set, , drop = FALSE]
  cluster_id <- spe[[cluster_col]]

  mean_expr <- sapply(levels(cluster_id), function(cl) {
    idx <- which(cluster_id == cl)
    rowMeans(expr_mat[, idx, drop = FALSE])
  })

  if (is.matrix(mean_expr) && nrow(mean_expr) > 1) {
    marker_order <- rownames(mean_expr)[hclust(dist(mean_expr))$order]
  } else {
    marker_order <- marker_set
  }

  plot_df <- as.data.frame(t(expr_mat)) |>
    mutate(cluster = cluster_id) |>
    pivot_longer(-cluster, names_to = "marker", values_to = "expr")

  plot_df$marker  <- factor(plot_df$marker, levels = marker_order)
  plot_df$cluster <- factor(plot_df$cluster, levels = levels(cluster_id))

  ggplot(plot_df, aes(x = cluster, y = expr, fill = cluster)) +
    geom_violin(scale = "width", trim = TRUE, linewidth = 0.2) +
    stat_summary(fun = median, geom = "point", size = 0.6, color = "black") +
    facet_wrap(~ marker, scales = "free_y", ncol = ncol) +
    labs(title = title, x = "Subcluster", y = "Z-score expression") +
    theme_bw(base_size = 11) +
    theme(
      axis.text.x      = element_text(angle = 0, hjust = 0.5, size = 9),
      strip.text        = element_text(face = "bold"),
      panel.grid.major.x = element_blank(),
      legend.position   = "none"
    )
}
```

```{r violin-cluster15, fig.width=16, fig.height=12}
p_violin_15 <- make_violinplot(spe_15, "scaled_0100", "sub_clusters",
                               disambiguation_markers,
                               "Cluster 15 Subset: Marker Expression per Subcluster",
                               ncol = 3)
print(p_violin_15)
```

## 8. Statistical Marker Discovery (findMarkers) on Subclusters

```{r findmarkers}
markers_de <- findMarkers(spe_15,
                          groups     = spe_15$sub_clusters,
                          assay.type = "scaled_0100",
                          subset.row = disambiguation_markers)

all_markers <- lapply(levels(spe_15$sub_clusters), function(cl) {
  df <- as.data.frame(markers_de[[cl]])
  data.frame(
    subcluster = cl,
    marker     = rownames(df),
    Top        = df$Top,
    FDR        = df$FDR,
    logFC      = df$summary.logFC
  )
})

all_markers_df <- do.call(rbind, all_markers)
dir.create(dirname(findmarkers_path), recursive = TRUE, showWarnings = FALSE)
write.csv(all_markers_df, findmarkers_path, row.names = FALSE)

top1_summary <- do.call(rbind, lapply(levels(spe_15$sub_clusters), function(cl) {
  sub <- all_markers_df[all_markers_df$subcluster == cl & all_markers_df$logFC > 1.0, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  if (nrow(sub) == 0) {
    data.frame(subcluster = cl, top_markers = "none above logFC 1.0")
  } else {
    data.frame(subcluster = cl,
               top_markers = paste(head(sub$marker, 5), collapse = ", "))
  }
}))

cat("Top positive markers (logFC > 1.0) per subcluster:\n")
print(knitr::kable(top1_summary, col.names = c("Subcluster", "Top Positive Markers")))
```

## 9. Composition Check: Patient Composition of Each Subcluster

Cluster 15 is a single broad cluster (not a merge of two, like the
myeloid subclustering), so there's no "original broad cluster" split
to check here, only whether any subcluster is patient-driven rather
than a real shared population.

```{r composition-patient-heatmap, fig.width=12, fig.height=6}
comp_patient <- table(subcluster = spe_15$sub_clusters,
                      patient = spe_15$patient_id)
comp_patient_props <- prop.table(comp_patient, margin = 1)

col_fun_patient <- colorRamp2(c(0, 0.5, 1), rev(brewer.pal(3, "RdBu")))

ht_patient <- Heatmap(
  comp_patient_props,
  name              = "Proportion",
  col               = col_fun_patient,
  cluster_rows      = FALSE,
  cluster_columns   = FALSE,
  column_title      = "Patient",
  column_title_side = "bottom",
  row_title         = "Subcluster",
  column_names_rot  = 45,
  cell_fun = function(j, i, x, y, width, height, fill) {
    val <- comp_patient_props[i, j]
    text_col <- if (val < 0.2 || val > 0.8) "white" else "black"
    grid.text(sprintf("%.2f", val), x, y, gp = gpar(col = text_col, fontsize = 7))
  }
)

draw(ht_patient, column_title = "Subcluster composition by patient (row-normalized %)")
```

## 10. Manual Subcluster Annotation

```{r annotate-sub, echo=TRUE}
cluster15_subannotation <- c(
  # "1"  = "",
  # "2"  = "",
  # ...fill in after reviewing violin plots + findMarkers output
)

if (length(cluster15_subannotation) > 0) {
  spe_15$sub_celltype <- cluster15_subannotation[as.character(spe_15$sub_clusters)]
  spe_15$sub_celltype <- factor(spe_15$sub_celltype, levels = cluster15_subannotation)
  cat("Subcluster annotation applied.\n")
  print(table(spe_15$sub_celltype, useNA = "ifany"))
} else {
  cat("No subcluster annotation yet. Fill in cluster15_subannotation above.\n")
}
```

```{r umap-sub-celltype, fig.width=10, fig.height=7}
if ("sub_celltype" %in% colnames(colData(spe_15))) {

  n_types <- nlevels(spe_15$sub_celltype)
  subtype_cols <- setNames(
    colorRampPalette(brewer.pal(12, "Paired"))(n_types),
    levels(spe_15$sub_celltype)
  )

  p_sub_celltype <- dittoDimPlot(spe_15, var = "sub_celltype",
                                 reduction.use = "UMAP_sub",
                                 size = 0.4, do.label = TRUE, labels.size = 3.5) +
    scale_color_manual(values = subtype_cols) +
    ggtitle("Cluster 15 Subset: UMAP colored by Subcluster Cell Type") +
    theme(legend.position = "right")
  print(p_sub_celltype)

} else {
  cat("Run the annotate-sub chunk first to apply subcluster annotations.\n")
}
```

## 11. Save Subclustered/Annotated Object

```{r save}
dir.create(dirname(out_path), recursive = TRUE, showWarnings = FALSE)
saveRDS(spe_15, out_path)
cat("Saved cluster 15 subcluster object to:", out_path, "\n")
```

