---
title: 'All Patients: Broad Cell Type Clustering: Harmony (no asinh) (k=200)'
author: "Tazein Shah"
date: "07-07-2026"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
subtitle: k=150 Clusters
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)
set.seed(42)
```

## 1. Load Libraries

```{r libraries}
library(imcRtools)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(scater)
library(scran)
library(bluster)
library(patchwork)
library(RColorBrewer)
```
---

## 2. Load Clustered SPE

```{r}
spe_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/clustering/k200_clusters/harmony_k200_clusters.rds"
spe <- readRDS(spe_path)
```

---

## 3. Marker Sets

```{r markers}
# core markers for broad annotation and UMAP overlays
core_markers <- c("SMA", "CD4", "CD8a", "CD19", "CD15",
                  "CD14", "CD86", "LAMP3", "CD163", "EpCAM")
core_markers <- core_markers[core_markers %in% rownames(spe)]
cat("Core markers found:", paste(core_markers, collapse = ", "), "\n")
cat("Missing:", paste(setdiff(
  c("SMA", "CD4", "CD8a", "CD19", "CD15", "CD14", "CD86", "LAMP3", "CD163", "EpCAM"),
  rownames(spe)), collapse = ", "), "\n")

# full panel, used to investigate clusters that don't separate on core markers
full_markers <- c(
  "SMA", "CD4", "CD8a", "CD19", "CD15", "CD14", "CD86", "LAMP3",
  "Ki67", "MKI67", "CD138", "PD1", "PDL1", "FoxP3", "CD68", "CD163",
  "TIM3", "CD141", "CD11c", "CD56", "CD117", "EpCAM", "CD38",
  "CD45RO", "LAG3", "CD1c", "GATA3", "TTF1orCD16"
)
full_markers <- full_markers[full_markers %in% rownames(spe)]
cat("Full panel markers found:", length(full_markers), "\n")
cat("Missing:", paste(setdiff(
  c("SMA", "CD4", "CD8a", "CD19", "CD15", "CD14", "CD86", "LAMP3",
    "Ki67", "MKI67", "CD138", "PD1", "PDL1", "FoxP3", "CD68", "CD163",
    "TIM3", "CD141", "CD11c", "CD56", "CD117", "EpCAM", "CD38",
    "CD45RO", "LAG3", "CD1c", "GATA3", "TTF1orCD16"),
  rownames(spe)), collapse = ", "), "\n")
```

---

## 4. Preprocessing: 99th Percentile Cap + Z-score

```{r helpers}
cap_99 <- function(mat) {
  t(apply(mat, 1, function(x) {
    q99 <- quantile(x, 0.99)
    pmin(x, q99)
  }))
}

zscore_rows <- function(mat) {
  t(scale(t(mat)))
}
```

```{r preprocess}
mat_0100 <- counts(spe) |> cap_99() |> zscore_rows()
assay(spe, "scaled_0100") <- mat_0100

cat("Assays in SPE:", paste(assayNames(spe), collapse = ", "), "\n")
```

---

## 5. UMAP on Harmony Embeddings

`UMAP_harmony` already exists in this object from the batch correction
step, so it is reused here instead of recomputed.

```{r umap}
if ("UMAP_harmony" %in% reducedDimNames(spe)) {
  cat("Reusing existing UMAP_harmony from batch correction step.\n")
} else {
  set.seed(220225)
  spe <- runUMAP(spe, dimred = "Harmony", name = "UMAP_harmony")
  cat("UMAP computed.\n")
}
```

---

## 6. Subset for Visualization

20,000 cells per ROI cap to keep plots manageable.

```{r subset}
set.seed(220225)
cells_per_roi  <- split(seq_len(ncol(spe)), spe$sample_id)
sampled_cells  <- unlist(lapply(cells_per_roi, function(idx) {
  sample(idx, min(20000, length(idx)))
}))
spe_sub <- spe[, sampled_cells]
cat("Cells in subset:", ncol(spe_sub), "\n")
```

---

## 7. UMAP Colored by Broad Cluster

```{r umap-clusters, fig.width=10, fig.height=7}
p_clusters <- dittoDimPlot(spe_sub, var = "broad_clusters",
                           reduction.use = "UMAP_harmony",
                           size = 0.2, do.label = TRUE, labels.size = 3) +
  ggtitle("UMAP colored by Broad Louvain Cluster (k=150)") +
  theme(legend.position = "right")
print(p_clusters)
```

---

## 8. UMAP Per Core Marker

Used to visually match clusters to known marker expression before
annotating.

```{r umap-markers, fig.width=8, fig.height=6}
markers_to_plot <- union(core_markers, "FoxP3")
markers_to_plot <- markers_to_plot[markers_to_plot %in% rownames(spe)]

for (marker in markers_to_plot) {
  p <- dittoDimPlot(spe_sub, var = marker,
                    reduction.use = "UMAP_harmony",
                    assay = "scaled_0100", size = 0.2) +
    scale_color_viridis(name = marker) +
    ggtitle(paste("UMAP:", marker))
  print(p)
}
```

---

## 9. Dot Plot: Core Marker Expression per Cluster

The primary tool for annotating clusters. Each row is a marker, each
column is a cluster. Look for which clusters show high expression of
the core lineage markers.

```{r make-dotplot}
make_dotplot <- function(spe, assay_name, cluster_col, marker_set, title) {
  expr_mat   <- assay(spe, assay_name)[marker_set, ]
  cluster_id <- spe[[cluster_col]]

  mean_expr <- sapply(levels(cluster_id), function(cl) {
    idx <- which(cluster_id == cl)
    rowMeans(expr_mat[, idx, drop = FALSE])
  })

  global_med <- apply(expr_mat, 1, median)
  frac_expr  <- sapply(levels(cluster_id), function(cl) {
    idx <- which(cluster_id == cl)
    rowMeans(expr_mat[, idx, drop = FALSE] > global_med)
  })

  mean_df <- as.data.frame(mean_expr) |>
    rownames_to_column("marker") |>
    pivot_longer(-marker, names_to = "cluster", values_to = "mean_exprs")

  frac_df <- as.data.frame(frac_expr) |>
    rownames_to_column("marker") |>
    pivot_longer(-marker, names_to = "cluster", values_to = "frac_expressing")

  plot_df <- left_join(mean_df, frac_df, by = c("marker", "cluster"))
  plot_df$cluster <- factor(plot_df$cluster, levels = levels(cluster_id))

  marker_order <- rownames(mean_expr)[hclust(dist(mean_expr))$order]
  plot_df$marker <- factor(plot_df$marker, levels = marker_order)

  ggplot(plot_df, aes(x = cluster, y = marker)) +
    geom_point(aes(size = frac_expressing, color = mean_exprs)) +
    scale_color_viridis(option = "magma", name = "Mean\nZ-score expr") +
    scale_size_continuous(range = c(0.5, 6), name = "Fraction\nexpressing") +
    labs(title = title, x = "Cluster", y = "Marker") +
    theme_bw(base_size = 12) +
    theme(
      axis.text.x      = element_text(angle = 0, hjust = 0.5),
      axis.text.y      = element_text(size = 9),
      panel.grid.major = element_line(color = "grey90"),
      legend.position  = "right"
    )
}
```

```{r dotplot-core, fig.width=16, fig.height=6}
p_dot_core <- make_dotplot(spe, "scaled_0100", "broad_clusters",
                           core_markers,
                           "Harmony (no asinh, k=150): Core Marker Expression per Broad Louvain Cluster")
print(p_dot_core)
```

---

## 10. Full Panel Dot Plot for Low Signal Clusters

Several clusters may not show a clear standout marker on the 10 core
markers above, this can happen for macrophage/myeloid populations
where markers like CD14, CD86, CD163 are broadly and continuously
expressed rather than sharply spiking in one cluster, or because the
cluster is actually separated by markers outside the core panel
(e.g. PD1, Ki67, FoxP3, CD45RO), since this Harmony run was computed
on raw counts across the full panel, not just the 10 core markers.
Borderline or low quality segmented cells are also possible. Use this
expanded dot plot to investigate clusters that look flat above before
calling them "Unknown."

```{r dotplot-full, fig.width=18, fig.height=10}
p_dot_full <- make_dotplot(spe, "scaled_0100", "broad_clusters",
                           full_markers,
                           "Harmony (no asinh, k=150): Full Panel Marker Expression per Broad Louvain Cluster")
print(p_dot_full)
```

---

## 11. Statistical Marker Discovery (findMarkers)

For clusters with no obvious standout marker on either dot plot,
`findMarkers()` ranks markers by how distinctly they separate that
cluster from all others, accounting for variance rather than just
mean difference. This can surface subtle but consistent shifts that
don't show up clearly in a Z-score dot plot averaged across thousands
of cells.

```{r findmarkers}
markers_de <- findMarkers(spe,
                          groups     = spe$broad_clusters,
                          assay.type = "scaled_0100",
                          subset.row = full_markers)

# save all clusters to CSV
all_markers <- lapply(levels(spe$broad_clusters), function(cl) {
  df <- as.data.frame(markers_de[[cl]])
  data.frame(
    cluster = cl,
    marker  = rownames(df),
    Top     = df$Top,
    FDR     = df$FDR,
    logFC   = df$summary.logFC
  )
})

all_markers_df <- do.call(rbind, all_markers)

write.csv(all_markers_df,
          "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/clustering/k200_clusters/harmony_200_cluster_findmarkers.csv",
          row.names = FALSE)

# print top 5 positive markers per cluster
df <- read.csv("/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/clustering/k200_clusters/harmony_200_cluster_findmarkers.csv")

# summary table: top positive marker per cluster at a glance
top1_summary <- do.call(rbind, lapply(levels(spe$broad_clusters), function(cl) {
  sub <- df[df$cluster == cl & df$logFC > 1.0, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  if (nrow(sub) == 0) {
    data.frame(cluster = cl, top_markers = "none above logFC 1.0")
  } else {
    data.frame(cluster = cl,
               top_markers = paste(head(sub$marker, 5), collapse = ", "))
  }
}))

cat("Top positive markers (logFC > 1.0) per cluster:\n")
print(knitr::kable(top1_summary, col.names = c("Cluster", "Top Positive Markers")))

cat("\n\nDetailed tables:\n")

# detailed table per cluster, positive logFC only
for (cl in levels(spe$broad_clusters)) {
  sub <- df[df$cluster == cl & df$Top <= 5 & df$logFC > 1.0, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  if (nrow(sub) == 0) {
    cat("Cluster", cl, ": no markers with logFC > 1.0 in top 5\n\n")
  } else {
    print(knitr::kable(sub[, c("marker", "Top", "FDR", "logFC")],
                       caption = paste("Cluster", cl),
                       digits  = 3))
  }
}
```
---

## 12. Trying to Find "Unknown" Clusters

Cluster 2 was the one with no markers above logFC 1.0, so the
threshold is lowered here to look for anything subtler.

```{r}
for (cl in c("2")) {
  sub <- df[df$cluster == cl & df$logFC > 0, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  print(knitr::kable(sub[, c("marker", "Top", "FDR", "logFC")],
                     caption = paste("Cluster", cl, "(logFC > 0)"),
                     digits  = 3))
}
```
---

## 13. Diagnostic: Area and Patient Composition for Ambiguous Clusters

Clusters 2, 3, 10, and 11 are the ones flagged as clusters of concern
from the top positive marker table: cluster 2 had no marker above
logFC 1.0, cluster 3 mixed a DC signature (LAMP3) with epithelial and
Treg markers (EpCAM, FoxP3), and clusters 10 and 11 both mixed
lymphoid/plasma markers with the granulocyte marker CD15. Cell area
and patient composition are checked here to build evidence before
deciding whether any of these are doublets, low quality segmentation,
or real populations.

```{r diagnostic-unknown, fig.width=10, fig.height=5}
ambiguous_clusters <- c("2")

# cell area boxplot
df_area <- data.frame(
  area    = spe$area,
  cluster = as.character(spe$broad_clusters)
)

if (length(ambiguous_clusters) > 0) {
  # reference: median area across all annotated clusters
  df_area_ref <- df_area[!df_area$cluster %in% ambiguous_clusters, ]
  ref_median   <- median(df_area_ref$area)

  p_area <- ggplot(df_area[df_area$cluster %in% ambiguous_clusters, ],
                   aes(x = cluster, y = area, fill = cluster)) +
    geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.3) +
    geom_hline(yintercept = ref_median, linetype = "dashed", color = "grey40") +
    annotate("text", x = 0.6, y = ref_median + 5,
             label = "median (annotated clusters)", size = 3, color = "grey40") +
    labs(title = "Cell Area Distribution for Ambiguous Clusters",
         x = "Cluster", y = "Cell Area (pixels)") +
    theme_bw() +
    theme(legend.position = "none")
  print(p_area)
} else {
  cat("Fill in ambiguous_clusters above after inspecting the k=150 dot plots.\n")
}
```

```{r diagnostic-patient, fig.width=12, fig.height=5}
if (length(ambiguous_clusters) > 0) {
  # patient composition per ambiguous cluster
  comp_df <- as.data.frame(
    table(cluster   = as.character(spe$broad_clusters),
          patient   = spe$patient_id)
  )
  comp_df <- comp_df[comp_df$cluster %in% ambiguous_clusters & comp_df$Freq > 0, ]

  p_comp <- ggplot(comp_df, aes(x = patient, y = Freq, fill = patient)) +
    geom_col() +
    facet_wrap(~ cluster, scales = "free_y", nrow = 2) +
    labs(title = "Patient Composition for Ambiguous Clusters",
         x = "Patient", y = "Cell Count") +
    theme_bw() +
    theme(axis.text.x  = element_text(angle = 45, hjust = 1),
          legend.position = "none")
  print(p_comp)
} else {
  cat("Fill in ambiguous_clusters above after inspecting the k=150 dot plots.\n")
}
```

---

## 14. Manual Broad Annotation

Filled in below based on the top positive marker table (`top1_summary`
from chunk 11) plus the diagnostics in chunk 13. Clusters 2, 10, and
11 are left as Unknown pending the area and patient composition
results; cluster 3 is tentatively called DC / mregDC but flagged as a
possible doublet population between DCs, tumor cells, and Tregs.

```{r annotate}
broad_annotation <- c(
  "1"  = "Fibroblast / Stromal (CD141+ FRC)",
  "2"  = "Unknown",
  "3"  = "DC / mregDC",
  "4"  = "Tumor / Epithelial",
  "5"  = "Plasma Cell / B Cell",
  "6"  = "CD4+ T Cell",
  "7"  = "Neutrophil",
  "8"  = "Macrophage",
  "9"  = "Tumor Epithelial (TTF1+)",
  "10" = "CD8+ T Cell (exhausted/memory)",
  "11" = "Proliferating Cells"
)
#Cluster 1: SMA, TIM3, CD141, PDL1: CD141+ fibroblastic reticular cells per Xi et al, SMA anchors stromal identity, TIM3/PDL1 likely stromal checkpoint expression
#Cluster 2: none above logFC 1.0, revisit in subclustering
#Cluster 3: LAMP3, CD117, PD1, EpCAM, FoxP3: LAMP3 anchors mregDC, EpCAM and FoxP3 likely neighbor contamination from tumor and Treg per Xi et al, CD117 unclear
#Cluster 4: EpCAM, clean single marker call
#Cluster 5: CD19, CD15, CD138, CD38, FoxP3: CD19/CD138/CD38 anchor plasma or activated B cell identity, CD15 likely neutrophil neighbor bleed, FoxP3 likely Treg contamination
#Cluster 6: CD4, CD45RO, CD138, TIM3, FoxP3: FoxP3 present but ranked low, possible Treg admixture, CD138 likely contamination, worth a FoxP3 specific UMAP check
#Cluster 7: CD45RO, CD15: CD15 anchors neutrophil lineage, CD45RO likely neighbor bleed from adjacent T cells
#Cluster 8: CD14, CD11c, CD86, CD68, CD4: classic myeloid markers, CD86 suggests more activated/M1 leaning phenotype than prior M2 call, worth revisiting with CD163/CD206 for M2 or iNOS for M1
#Cluster 9: TTF1orCD16, EpCAM, FoxP3, CD117, CD1c: TTF1 plus EpCAM is a lung adenocarcinoma signature, FoxP3/CD117/CD1c likely DC and Treg neighbor contamination
#Cluster 10: CD8a, CD138, CD45RO, TIM3, PD1: matches dysfunctional CD8 signature in Xi et al, CD138 likely contamination
#Cluster 11: MKI67, matches standalone proliferating category in Windhager et al

if (length(broad_annotation) > 0) {
  spe$broad_celltype <- broad_annotation[as.character(spe$broad_clusters)]
  cat("Annotation applied.\n")
  print(table(spe$broad_celltype, useNA = "ifany"))
} else {
  cat("No annotation provided yet. Fill in broad_annotation above after inspecting plots.\n")
}

if (length(broad_annotation) > 0) {
  spe$broad_celltype <- broad_annotation[as.character(spe$broad_clusters)]
  cat("Annotation applied.\n")
  print(table(spe$broad_celltype, useNA = "ifany"))
} else {
  cat("No annotation provided yet. Fill in broad_annotation above after inspecting plots.\n")
}
```
---

## 15. UMAP Colored by Broad Cell Type

Run this chunk after filling in the annotation above.

```{r umap-celltype, fig.width=10, fig.height=7}
if ("broad_celltype" %in% colnames(colData(spe))) {
  spe_sub$broad_celltype <- spe$broad_celltype[match(colnames(spe_sub), colnames(spe))]

  p_celltype <- dittoDimPlot(spe_sub, var = "broad_celltype",
                             reduction.use = "UMAP_harmony",
                             size = 0.2, do.label = TRUE, labels.size = 3) +
    ggtitle("UMAP colored by Broad Cell Type (k=150)") +
    theme(legend.position = "right")
  print(p_celltype)
} else {
  cat("Run chunk 14 first to apply annotations.\n")
}
```

---

## 16. Save Annotated SPE

```{r save}
out_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/clustering/k200_clusters/harmony_200_cluster_annot.rds"

if ("broad_celltype" %in% colnames(colData(spe))) {
  saveRDS(spe, out_path)
  cat("Saved annotated SPE to:", out_path, "\n")
} else {
  cat("Annotation not yet applied. Fill in chunk 14 before saving.\n")}
```

