---
title: 'Patient 14 Dot Plot: Cluster Marker Expression'
author: ''
date: "06-18-2026"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 14, fig.height = 8)

set.seed(42)
```

## Overview

Dot plot showing mean marker expression per Louvain cluster, alongside
UMAP colored by cluster. The goal is to characterize which markers
define each cluster -- the grouping of the UMAP based off markers (groups
of the cluster and their marker expression).

Dot color = mean arcsinh-transformed expression per cluster
Dot size = fraction of cells in the cluster expressing above the
per-marker global median

This is exploratory annotation prior to formal Celda-based cell
phenotyping.

------------------------------------------------------------------------

## 1. Load Libraries

```{r libraries}
library(imcRtools)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(scater)
library(scran)
library(bluster)
library(patchwork)
library(RColorBrewer)
```

------------------------------------------------------------------------

## 2. Read in Steinbock Data

```{r read-data}
steinbock_path <- "/restricted/projectnb/camplab/home/tshah20/tumor_IMC/patient14_steinbock"

spe <- read_steinbock(steinbock_path)
spe
```

```{r process-spe}
# set informative column names: image_cellID
colnames(spe) <- paste0(spe$sample_id, "_", spe$ObjectNumber)

# extract patient and ROI from sample_id
spe$patient_id <- str_extract(spe$sample_id, "^[^_]+")
spe$ROI        <- spe$sample_id

# arcsinh transform (cofactor = 1, standard for IMC)
assay(spe, "exprs") <- asinh(counts(spe) / 1)

# flag biological channels (exclude DNA and membrane channels used for segmentation)
rowData(spe)$use_channel <- !grepl("DNA191|DNA193|Membrane", rownames(spe))

cat("Cells loaded:", ncol(spe), "\n")
cat("ROIs:", n_distinct(spe$sample_id), "\n")
cat("Channels:", nrow(spe), "\n")

image_path <- "/restricted/projectnb/camplab/home/tshah20/tumor_IMC/images"
```

------------------------------------------------------------------------

## 3. Subset and Run UMAP/tSNE

UMAP/tSNE projects cells to 2D -- looking to see if the ROIs mix well.
Marker expression is overlaid to verify that the embeddings separate
biologically meaningful populations aka the differences between patient
expressions.

```{r dimred, fig.height=12, fig.width=14}
set.seed(220225)
cell_sub <- sample(seq_len(ncol(spe)), min(15000 * n_distinct(spe$sample_id), ncol(spe)))
spe_sub  <- spe[, cell_sub]
spe_sub <- runUMAP(spe_sub, subset_row = rowData(spe_sub)$use_channel, exprs_values = "exprs")
spe_sub <- runTSNE(spe_sub, subset_row = rowData(spe_sub)$use_channel, exprs_values = "exprs")

# define ROI colors
roi_colors <- setNames(
  colorRampPalette(brewer.pal(4, "Set1"))(n_distinct(spe$sample_id)),
  unique(spe$sample_id)
)

p1 <- dittoDimPlot(spe_sub, var = "sample_id", reduction.use = "UMAP", size = 0.2) +
  scale_color_manual(values = roi_colors) +
  ggtitle("ROI on UMAP")

p2 <- dittoDimPlot(spe_sub, var = "sample_id", reduction.use = "TSNE", size = 0.2) +
  scale_color_manual(values = roi_colors) +
  ggtitle("ROI on tSNE")

(p1 + p2)

ggsave("patient14_ROI_UMAP_tSNE.pdf",
       plot   = (p1 + p2),
       width  = 14,
       height = 12,
       path   = image_path)
```

```{r dimred-markers, fig.height=12, fig.width=14}
p3 <- dittoDimPlot(spe_sub, var = "EpCAM", reduction.use = "UMAP",
                   assay = "exprs", size = 0.2) +
  scale_color_viridis(name = "EpCAM") +
  ggtitle("EpCAM on UMAP")

p4 <- dittoDimPlot(spe_sub, var = "SMA", reduction.use = "UMAP",
                   assay = "exprs", size = 0.2) +
  scale_color_viridis(name = "SMA") +
  ggtitle("SMA on UMAP")

p5 <- dittoDimPlot(spe_sub, var = "CD45RO", reduction.use = "UMAP",
                   assay = "exprs", size = 0.2) +
  scale_color_viridis(name = "CD45RO") +
  ggtitle("CD45RO on UMAP")

p6 <- dittoDimPlot(spe_sub, var = "CD68", reduction.use = "UMAP",
                   assay = "exprs", size = 0.2) +
  scale_color_viridis(name = "CD68") +
  ggtitle("CD68 on UMAP")

(p3 + p4) / (p5 + p6)

ggsave("patient14_marker_UMAP.pdf",
       plot   = (p3 + p4) / (p5 + p6),
       width  = 14,
       height = 12,
       path   = image_path)
```

------------------------------------------------------------------------

## 4. Louvain Clustering

Graph-based Louvain clustering on the full spe (not the UMAP subset) at
k = 20. Cluster labels are stored in spe$nn_clusters.

```{r clustering}
set.seed(220225)

mat <- t(assay(spe, "exprs")[rowData(spe)$use_channel, ])

g <- buildSNNGraph(t(mat), k = 20, type = "jaccard",
                   BNPARAM = BiocNeighbors::KmknnParam())
clusters <- igraph::cluster_louvain(g)$membership

spe$nn_clusters <- factor(clusters)

cat("Number of clusters:", nlevels(spe$nn_clusters), "\n")
print(table(spe$nn_clusters))
```

------------------------------------------------------------------------

## 5. UMAP Colored by Cluster

Overlay cluster labels on the existing spe_sub UMAP. Cluster labels are
propagated to the subset via cell name matching.

```{r umap-clusters, fig.width=10, fig.height=7}
# propagate cluster labels to spe_sub by matching cell names
spe_sub$nn_clusters <- spe$nn_clusters[match(colnames(spe_sub), colnames(spe))]

p_umap_clusters <- dittoDimPlot(spe_sub,
             var = "nn_clusters",
             reduction.use = "UMAP",
             size = 0.3,
             do.label = TRUE,
             labels.size = 3) +
  ggtitle("Patient 14: Louvain Clusters on UMAP") +
  theme(legend.position = "right")

print(p_umap_clusters)

ggsave("patient14_cluster_UMAP.pdf",
       plot   = p_umap_clusters,
       width  = 10,
       height = 7,
       path   = image_path)
```

------------------------------------------------------------------------

## 6. Dot Plot: Marker Expression per Cluster

Each dot encodes mean arcsinh expression (color) and fraction of cells
expressing above the per-marker global median (size).

```{r dotplot, fig.width=16, fig.height=8}
use_markers <- rownames(spe)[rowData(spe)$use_channel]
expr_mat    <- assay(spe, "exprs")[use_markers, ]
cluster_ids <- spe$nn_clusters

# mean expression per cluster
mean_expr <- sapply(levels(cluster_ids), function(cl) {
  idx <- which(cluster_ids == cl)
  rowMeans(expr_mat[, idx, drop = FALSE])
})

# fraction expressing above per-marker global median
global_medians <- apply(expr_mat, 1, median)
frac_expr <- sapply(levels(cluster_ids), function(cl) {
  idx <- which(cluster_ids == cl)
  rowMeans(expr_mat[, idx, drop = FALSE] > global_medians)
})

# reshape to long format
mean_df <- as.data.frame(mean_expr) |>
  rownames_to_column("marker") |>
  pivot_longer(-marker, names_to = "cluster", values_to = "mean_exprs")

frac_df <- as.data.frame(frac_expr) |>
  rownames_to_column("marker") |>
  pivot_longer(-marker, names_to = "cluster", values_to = "frac_expressing")

plot_df <- left_join(mean_df, frac_df, by = c("marker", "cluster"))
plot_df$cluster <- factor(plot_df$cluster, levels = levels(cluster_ids))

# order markers by hierarchical clustering for readability
marker_order <- rownames(mean_expr)[hclust(dist(mean_expr))$order]
plot_df$marker <- factor(plot_df$marker, levels = marker_order)

p_dot <- ggplot(plot_df, aes(x = cluster, y = marker)) +
  geom_point(aes(size = frac_expressing, color = mean_exprs)) +
  scale_color_viridis(option = "magma", name = "Mean\narcsinh expr") +
  scale_size_continuous(range = c(0.5, 6), name = "Fraction\nexpressing") +
  labs(
    title = "Patient 14: Marker Expression per Louvain Cluster",
    x = "Cluster",
    y = "Marker"
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x  = element_text(angle = 0, hjust = 0.5),
    axis.text.y  = element_text(size = 9),
    panel.grid.major = element_line(color = "grey90"),
    legend.position  = "right"
  )

print(p_dot)
```

------------------------------------------------------------------------

## 7. Save Dot Plot

```{r save-dotplot}
ggsave("patient14_dotplot_clusters.pdf",
       plot   = p_dot,
       width  = 16,
       height = 8,
       path   = image_path)

cat("Saved to:", file.path(image_path, "patient14_dotplot_clusters.pdf"), "\n")
```

------------------------------------------------------------------------

## Notes

- Clustering is exploratory (graph-based Louvain). Formal cell
  phenotyping will use Celda's celda_C two-pass approach per Xi et al.
- GATA3 (Eu151) is absent from the SPE until steinbock measure
  intensities is rerun; it is automatically excluded via the use_channel
  flag.
- TTF1orCD16 appears as TTForCD16 in the SPE due to a naming
  inconsistency in the raw TXT files; both names refer to the same
  channel.
