---
title: "Tumor IMC - Batch Correction and Dimensionality Reduction"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

# Setup
```{r setup, message=FALSE, warning=FALSE}
.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(imcRtools)
library(SpatialExperiment)
library(SingleCellExperiment)
library(ggplot2)
library(dplyr)
library(stringr)
library(harmony)
library(scater)
library(dittoSeq)
library(viridis)
library(cowplot)

set.seed(42)
```

# Define Paths
```{r paths}
base_dir <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/steinbock/"

steinbock_dirs <- list(
  Patient10   = file.path(base_dir, "p10_results"),
  Patient11   = file.path(base_dir, "p11_results"),
  Patient12   = file.path(base_dir, "p12_results"),
  Patient13   = file.path(base_dir, "p13_results"),
  Patient14   = file.path(base_dir, "p14_results"),
  Patient15   = file.path(base_dir, "p15_results"),
  Patient16   = file.path(base_dir, "p16_results"),
  Patient17   = file.path(base_dir, "p17_results"),
  Patient18   = file.path(base_dir, "p18_results"),
  Patient19   = file.path(base_dir, "p19_results"),
  Patient20_425 = file.path(base_dir, "p20_4_25_results"),
  Patient20_519 = file.path(base_dir, "p20_5_19_results")
)
```

# Load All Patients
```{r load-all}
spe_list <- lapply(names(steinbock_dirs), function(patient) {
  cat("Loading:", patient, "\n")
  spe <- read_steinbock(steinbock_dirs[[patient]])
  spe$patient_id  <- patient
  spe
})

names(spe_list) <- names(steinbock_dirs)

# Check cells loaded per patient
for (p in names(spe_list)) {
  cat(p, ":", ncol(spe_list[[p]]), "cells\n")
}
```

# Add Metadata
```{r metadata}
# Add acquisition metadata for Patient 20
spe_list$Patient20_425$acquisition <- "4-25"
spe_list$Patient20_519$acquisition <- "5-19"

# Add NA acquisition for all other patients
for (p in names(spe_list)[!names(spe_list) %in% c("Patient20_425", "Patient20_519")]) {
  spe_list[[p]]$acquisition <- NA
}
```

# Combine All Patients
```{r combine}
spe_combined <- do.call(cbind, spe_list)

cat("Total cells before filtering:", ncol(spe_combined), "\n")
cat("Total markers:", nrow(spe_combined), "\n")

# Check cells per patient
table(spe_combined$patient_id)
```

# Quality Control and Filtering
```{r qc-filter}
# Filter by cell area
spe_combined <- spe_combined[, spe_combined$area >= 20 & spe_combined$area < 500]

cat("Total cells after filtering:", ncol(spe_combined), "\n")
```

# Apply asinh Transformation
```{r asinh}
assay(spe_combined, "asin") <- asinh(counts(spe_combined) / 1)
```

# Select Markers for Batch Correction
```{r select-markers}
# Exclude structural/QC markers
markers_exclude <- c("DNA191", "DNA193", "Membrane")
markers_use <- rownames(spe_combined)[!rownames(spe_combined) %in% markers_exclude]

cat("Markers used for batch correction:", length(markers_use), "\n")
print(markers_use)
```

# Dimensionality Reduction BEFORE Batch Correction
```{r pca-before}
# Run PCA on asinh-transformed data
spe_combined <- runPCA(spe_combined,
                       assay.type = "asin",
                       subset_row = markers_use,
                       ncomponents = 50,
                       name = "PCA_uncorrected")

cat("PCA complete\n")
```

```{r umap-before}
# Run UMAP on uncorrected PCA
spe_combined <- runUMAP(spe_combined,
                        dimred = "PCA_uncorrected",
                        name = "UMAP_uncorrected")

cat("UMAP complete\n")
```

## Visualize UMAP Before Batch Correction
```{r plot-umap-before, fig.width=12, fig.height=5}
# Color by patient
p1 <- plotReducedDim(spe_combined,
                     dimred = "UMAP_uncorrected",
                     colour_by = "patient_id",
                     point_size = 0.3,
                     point_alpha = 0.3) +
  labs(title = "UMAP Before Batch Correction",
       subtitle = "Colored by patient") +
  theme_bw() +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 1)))

# Color by sample_id
p2 <- plotReducedDim(spe_combined,
                     dimred = "UMAP_uncorrected",
                     colour_by = "sample_id",
                     point_size = 0.3,
                     point_alpha = 0.3) +
  labs(title = "UMAP Before Batch Correction",
       subtitle = "Colored by sample_id (ROI)") +
  theme_bw() +
  theme(legend.position = "none")

plot_grid(p1, p2, ncol = 2)
```

# Run Harmony Batch Correction
```{r harmony}
# Extract PCA embeddings
pca_mat <- reducedDim(spe_combined, "PCA_uncorrected")

# Run Harmony using sample_id as batch variable
harmony_embeddings <- harmony::RunHarmony(
  pca_mat,
  meta_data = as.data.frame(colData(spe_combined)),
  vars_use  = "sample_id",
  verbose   = TRUE
)

# Store corrected embeddings
reducedDim(spe_combined, "Harmony") <- harmony_embeddings

cat("Harmony batch correction complete\n")
```

# Dimensionality Reduction AFTER Batch Correction
```{r umap-after}
# Run UMAP on Harmony-corrected embeddings
spe_combined <- runUMAP(spe_combined,
                        dimred = "Harmony",
                        name = "UMAP_harmony")

cat("UMAP on Harmony embeddings complete\n")
```

## Visualize UMAP After Batch Correction
```{r plot-umap-after, fig.width=12, fig.height=5}
# Color by patient
p3 <- plotReducedDim(spe_combined,
                     dimred = "UMAP_harmony",
                     colour_by = "patient_id",
                     point_size = 0.3,
                     point_alpha = 0.3) +
  labs(title = "UMAP After Harmony Batch Correction",
       subtitle = "Colored by patient") +
  theme_bw() +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 1)))

# Color by sample_id
p4 <- plotReducedDim(spe_combined,
                     dimred = "UMAP_harmony",
                     colour_by = "sample_id",
                     point_size = 0.3,
                     point_alpha = 0.3) +
  labs(title = "UMAP After Harmony Batch Correction",
       subtitle = "Colored by sample_id (ROI)") +
  theme_bw() +
  theme(legend.position = "none")

plot_grid(p3, p4, ncol = 2)
```

# Before vs After Comparison
```{r compare, fig.width=12, fig.height=10}
plot_grid(p1, p2, p3, p4, ncol = 2)
```

# Assess Batch Correction Quality
```{r assess}
# Check if patients mix well after correction
# Good batch correction = patients mix in UMAP
# Bad batch correction = patients still cluster separately

# Plot each patient separately to check mixing
plotReducedDim(spe_combined,
               dimred = "UMAP_harmony",
               colour_by = "patient_id",
               point_size = 0.3,
               point_alpha = 0.5) +
  facet_wrap(~ spe_combined$patient_id) +
  labs(title = "UMAP After Harmony - Per Patient") +
  theme_bw() +
  theme(legend.position = "none")
```
Cells should overlap across patients if batch correction worked

# Save Combined SPE Object
```{r save}
output_path <- file.path(base_dir, "../analysis/spe_combined_harmony.rds")
saveRDS(spe_combined, output_path)
cat("Saved combined SPE object to:", output_path, "\n")
```

# Summary
```{r summary}
cat("=== Batch Correction Summary ===\n")
cat("Total cells:", ncol(spe_combined), "\n")
cat("Total markers:", nrow(spe_combined), "\n")
cat("Markers used for correction:", length(markers_use), "\n")
cat("Batch variable: sample_id\n")
cat("Method: Harmony\n")
cat("\nCells per patient:\n")
print(table(spe_combined$patient_id))
cat("\nCells per ROI (sample_id):\n")
print(table(spe_combined$sample_id))
```
