---
title: 'Patient 14 Post-Steinbock Visualization'
author: ''
date: "05-27-2026"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 14, fig.height = 8)
set.seed(42)
```

## Overview

This notebook performs segmentation quality control by overlaying
Mesmer-generated cell mask outlines onto pseudo-colored multichannel IMC
images. Since Xi et al. uses Mesmer with a cytoplasmic channel set
(CD45RA, CD45RO, HLA-DRB + nuclear channels), cell masks expand beyond
the nucleus to capture full cell bodies — so mask outlines should
enclose the full cell, not just the nucleus.

We inspect three things:

1.  Visual overlay of mask outlines on composite channel images
2.  Distribution of cell areas per ROI (to catch
    over/under-segmentation)
3.  Fraction of image area covered by cells per ROI

------------------------------------------------------------------------

## 1. Load Libraries

```{r libraries}
library(imcRtools)
library(cytomapper)
library(tidyverse)
library(dittoSeq)

dir.create("images", showWarnings = FALSE)
```

------------------------------------------------------------------------

## 2. Read in Steinbock Data

```{r read-data}
steinbock_path <- "/restricted/projectnb/camplab/home/tshah20/tumor_IMC/patient14_steinbock"

spe <- read_steinbock(steinbock_path)
```

```{r}
spe
```
```{r}
# set informative column names: image_cellID
colnames(spe) <- paste0(spe$sample_id, "_", spe$ObjectNumber)

# extract patient and ROI from sample_id
spe$patient_id <- str_extract(spe$sample_id, "^[^_]+")
spe$ROI        <- spe$sample_id

# arcsinh transform (cofactor = 1, standard for IMC)
assay(spe, "exprs") <- asinh(counts(spe) / 1)

# flag biological channels (exclude DNA and membrane channels used for segmentation)
rowData(spe)$use_channel <- !grepl("DNA191|DNA193|Membrane|Pt196", rownames(spe))

cat("Cells loaded:", ncol(spe), "\n")
cat("ROIs:", n_distinct(spe$sample_id), "\n")
cat("Channels:", nrow(spe), "\n")
```

```{r}
spe
```

------------------------------------------------------------------------

## 3. Load Multichannel Images and Segmentation Masks

```{r load-images}
images <- loadImages(file.path(steinbock_path, "img/"))
masks  <- loadImages(file.path(steinbock_path, "masks_deepcell/"), as.is = TRUE)

# assign channel names from SPE
channelNames(images) <- rownames(spe)

# link images and masks to SPE via sample_id
mcols(images)$sample_id <- names(images)
mcols(masks)$sample_id  <- names(masks)

cat("Images loaded:", length(images), "\n")
cat("Masks loaded:", length(masks), "\n")
```

------------------------------------------------------------------------

## 4. Normalize Images for Display

Pixel intensities are normalized per image to [0, 1], then clipped at
0.2 to bring out dim signals. Adjust `inputRange` upper bound if markers
appear oversaturated or invisible.

```{r normalize}
images_norm <- cytomapper::normalize(images, separateImages = TRUE)
images_norm <- cytomapper::normalize(images_norm, inputRange = c(0, 0.2))
```

------------------------------------------------------------------------

## 5. Overlay Cell Masks on Composite Images

We sample 3 ROIs and overlay mask outlines colored by key cell-type
markers. Look for: outlines that cleanly enclose individual cells
without obvious merging (one outline around two nuclei) or fragmentation
(one nucleus split into multiple outlines).

```{r plot-masks, fig.height=10}
img_ids <- sample(seq_len(length(images_norm)), 3)

plotPixels(
  images_norm[img_ids],
  mask          = masks[img_ids],
  img_id        = "sample_id",
  missing_colour = "white",
  colour_by = c("SMA", "CD4", "CD8a", "CD19", "DNA191"),
  colour = list(
    SMA    = c("black", "yellow"),
    CD4    = c("black", "green"),
    CD8a   = c("black", "red"),
    CD19   = c("black", "cyan"),
    DNA191 = c("black", "blue")
  ),
  image_title = list(position = "topleft", cex = 1),
  legend      = list(colour_by.title.cex = 0.8, colour_by.labels.cex = 0.8)
)
```

> **Note:** Adjust the `colour_by` markers to match your panel. DNA191
> (Ir191) anchors nuclei in blue; the remaining markers highlight major
> cell types. Zoom in on the rendered image to assess individual cell
> boundaries.

------------------------------------------------------------------------

## 6. Cell Area Distribution per ROI

Small cells (area \< 5 pixels) are likely segmentation artifacts —
partial cell slices or noise. The distribution below helps you choose a
sensible area cutoff.

```{r cell-area, fig.height=6}
dittoPlot(spe,
          var      = "area",
          group.by = "sample_id",
          plots    = "boxplot") +
  ylab("Cell area (pixels)") +
  xlab("") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Cell area distribution per ROI")
```

```{r filter-cells}
# filter out cells with area below threshold
area_threshold <- 5
spe <- spe[, spe$area >= area_threshold]

cat("Cells after area filter (>=", area_threshold, "px):", ncol(spe), "\n")
```

------------------------------------------------------------------------

## 7. Fraction of Image Area Covered by Cells

High cell coverage (\~70–90%) is expected for dense tissue sections.
Sparse coverage may indicate tissue loss or large acellular regions
(e.g., vessels, necrosis).

```{r cell-coverage, fig.height=5}
cell_density <- colData(spe) %>%
  as.data.frame() %>%
  group_by(sample_id) %>%
  summarise(
    cell_area  = sum(area),
    no_pixels  = mean(width_px) * mean(height_px),
    covered_area = cell_area / no_pixels,
    .groups = "drop"
  )

ggplot(cell_density, aes(x = sample_id, y = covered_area)) +
  geom_point(size = 3, color = "steelblue") +
  geom_hline(yintercept = 0.7, linetype = "dashed", color = "grey50") +
  ylim(0, 1) +
  ylab("Fraction of image area covered by cells") +
  xlab("") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Image area covered by cells per ROI")
```

------------------------------------------------------------------------

## Session Info

```{r session-info}
sessionInfo()
```
