---
title: "Raw IMC Data QC: Pixel Intensity Distributions"
author: ""
date: "05-27-2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 14, fig.height = 8)
set.seed(42)
```

## Overview

This notebook visualizes pixel-level ion count distributions from raw
IMC TXT files — **before any steinbock preprocessing**. Each TXT file is
one ROI; rows are pixels, columns are already labeled with both protein
name and metal tag (e.g. `SMA(Pr141Di)`), so no external panel file is
needed.

------------------------------------------------------------------------

## 1. Load Libraries

```{r libraries}
library(tidyverse)
library(ggridges)

dir.create("images", showWarnings = FALSE)
```

------------------------------------------------------------------------

## 2. Read All TXT Files and Combine

TXT files live in the raw data directory, one per ROI. Column names
already contain the protein name and metal tag (e.g. `SMA(Pr141Di)`), so
no renaming is needed. We drop acquisition metadata and non-biological
channels, apply arcsinh transformation (cofactor = 1), and stack
everything into a single long-format tibble.

```{r read-txt}
cofactor <- 1

txt_files <- list.files(
  path       = "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/Data/Imaging/IMC/Tumor/PardeeTumorAblations/Patient_14T",
  pattern    = "\\.txt$",
  recursive  = TRUE,
  full.names = TRUE
)

# acquisition metadata and non-biological channels to drop
exclude_cols <- c("Start_push", "End_push", "Pushes_duration",
                  "X", "Y", "Z",
                  "DNA191(Ir191Di)", "DNA193(Ir193Di)", "Membrane(Pt196Di)")

raw_pixels <- map_dfr(txt_files, function(f) {
  df <- read_tsv(f, show_col_types = FALSE)

  # downsample to 50k pixels per ROI before any heavy operations
  df <- df %>% slice_sample(n = 50000)

  # drop metadata and non-biological channels
  df <- df %>% select(-any_of(exclude_cols))

  # arcsinh transform all remaining numeric columns
  df <- df %>% mutate(across(where(is.numeric), ~asinh(. / cofactor)))

  # tag with ROI label and patient ID parsed from filename
  fname      <- tools::file_path_sans_ext(basename(f))
  df$roi     <- fname
  df$patient <- str_extract(fname, "^[^_]+")  # everything before first underscore
  df
})

cat("Total pixels loaded:", nrow(raw_pixels), "\n")
cat("Patients found:", n_distinct(raw_pixels$patient), "\n")
cat("ROIs found:", n_distinct(raw_pixels$roi), "\n")
```

------------------------------------------------------------------------

## 3. Pivot to Long Format

```{r pivot}
plot_long <- raw_pixels %>%
  pivot_longer(
    cols      = -c(roi, patient),
    names_to  = "marker",
    values_to = "intensity"
  )

single_roi <- unique(plot_long$roi)[1]
```

------------------------------------------------------------------------

## 4. Histograms: All Channels, One ROI

A quick per-channel histogram faceted view for a single ROI to get an
overview of the intensity landscape.

```{r histogram-single-roi, fig.height=12}
single_roi <- unique(plot_long$roi)[1]

plot_long %>%
  filter(roi == single_roi) %>%
  ggplot(aes(x = intensity)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white", linewidth = 0.2) +
  facet_wrap(~marker, scales = "free_y", ncol = 6) +
  labs(
    title    = paste0("Pixel intensity distributions (arcsinh) — ", single_roi),
    subtitle = "Each panel = one protein marker; x-axis = arcsinh(count / 1)",
    x        = "arcsinh intensity",
    y        = "Pixel count"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    strip.text  = element_text(face = "bold", size = 8),
    axis.text.x = element_text(size = 7)
  )
```

------------------------------------------------------------------------

## 5. Density Plots: Selected Markers Across All ROIs

Ridge plots let you compare intensity distributions across ROIs for a
few markers of interest, which is useful for spotting between-ROI batch
effects or acquisition issues.

```{r ridgeplot, fig.height=20}
p <- plot_long %>%
  ggplot(aes(x = intensity, y = roi, fill = patient)) +
  geom_density_ridges(alpha = 0.6, scale = 1.2, rel_min_height = 0.01) +
  facet_wrap(~marker, scales = "free_x", ncol = 4) +
  labs(
    title    = "Per-ROI pixel intensity distributions (all markers)",
    subtitle = "arcsinh(count / 1); each ridge = one ROI, color = patient",
    x        = "arcsinh intensity",
    y        = NULL,
    fill     = "Patient"
  ) +
  theme_ridges(font_size = 9) +
  theme(strip.text = element_text(face = "bold", size = 7))

p
ggsave("images/02_ridgeplot_all_markers.png", plot = p, width = 14, height = 40, dpi = 150)
```

------------------------------------------------------------------------

## 6. Density Overlay: All Channels for One ROI

An alternative view: all markers overlaid in a single density plot to
compare the signal spread across channels. Useful for spotting channels
with no signal or unusually high background.

```{r density-overlay, fig.height=5}
plot_long %>%
  filter(roi == single_roi) %>%
  ggplot(aes(x = intensity, color = marker)) +
  geom_density(linewidth = 0.5, alpha = 0.8) +
  labs(
    title    = paste0("All-channel density overlay — ", single_roi),
    subtitle = "Channels with no biological signal will hug x = 0",
    x        = "arcsinh intensity",
    y        = "Density"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right",
        legend.text     = element_text(size = 7),
        legend.key.size = unit(0.4, "cm"))
```

------------------------------------------------------------------------

## 7. Summary Table: Per-Channel Median and % Zero Pixels

A numeric summary to flag channels with very low signal (high % zero
pixels) or unexpected medians — useful for early QC before committing to
preprocessing.

```{r summary-table}
plot_long %>%
  group_by(marker) %>%
  summarise(
    median_intensity = round(median(intensity, na.rm = TRUE), 3),
    pct_zero         = round(mean(intensity == 0, na.rm = TRUE) * 100, 1),
    max_intensity    = round(max(intensity, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_zero)) %>%
  knitr::kable(caption = "Per-channel pixel intensity summary (arcsinh-transformed)")
```
