---
title: '`r paste("Cell Segmentation Mask Visualization —", params$patient_id)`'
author: 'Tazein Shah'
date: "06-10-2026"
output:
  pdf_document:
    toc: true
params:
  patient_id:
  steinbock_path:
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 14, fig.height = 10)
```

## Overview

Cell segmentation mask overlays for **`r params$patient_id`** across all ROIs.
Channel combinations are matched to annotated cell types per Xi et al. (2026).

1. Nuclear anchor (DNA191)
2. Fibroblast / CAF (SMA)
3. CD4+ T Cell (CD4)
4. CD8+ T Cell (CD8a)
5. B Cell (CD19)
6. PMN-MDSC (CD15)
7. Macrophage (CD14 + CD86)
8. M-MDSC (CD14 + CD163)
9. DC / mregDC (LAMP3)

---

## 1. Load Libraries

```{r libraries}
library(imcRtools)
library(cytomapper)
library(tidyverse)
```

---

## 2. Load Images, Masks, and Panel

```{r load-data}
base_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/steinbock"

images <- loadImages(file.path(params$steinbock_path, "img/"))
masks  <- loadImages(file.path(params$steinbock_path, "masks_deepcell/"), as.is = TRUE)

panel         <- read.csv(file.path(base_path, "panel.csv"))
kept_channels <- panel$name[panel$keep == 1]

channelNames(images) <- kept_channels

mcols(images)$sample_id <- names(images)
mcols(masks)$sample_id  <- names(masks)

cat("Images loaded:", length(images), "\n")
cat("Masks loaded:", length(masks), "\n")
```

---

## 3. Normalize Images for Display

```{r normalize}
images_norm <- cytomapper::normalize(images, separateImages = TRUE)
images_norm <- cytomapper::normalize(images_norm, inputRange = c(0, 0.2))

img_ids <- seq_len(length(images_norm))
```

---

## 4. Helper

```{r helper}
plot_mask <- function(colour_by, colour) {
  plotPixels(
    images_norm[img_ids],
    mask           = masks[img_ids],
    img_id         = "sample_id",
    missing_colour = "white",
    colour_by      = colour_by,
    colour         = colour,
    image_title    = list(position = "topleft", cex = 1),
    legend         = list(colour_by.title.cex = 0.8, colour_by.labels.cex = 0.8)
  )
}
```

---

## 5. Nuclear Anchor

```{r nuclear, fig.height=12}
plot_mask(
  colour_by = "DNA191",
  colour    = list(DNA191 = c("black", "blue"))
)
```

---

## 6. Fibroblast / CAF

SMA (alpha-smooth muscle actin) marks fibroblasts and cancer-associated fibroblasts
in the tumor stroma.

```{r fibroblast, fig.height=12}
plot_mask(
  colour_by = c("SMA", "DNA191"),
  colour    = list(SMA    = c("black", "magenta"),
                   DNA191 = c("black", "blue"))
)
```

---

## 7. CD4+ T Cells

```{r cd4-tcell, fig.height=12}
plot_mask(
  colour_by = c("CD4", "DNA191"),
  colour    = list(CD4    = c("black", "orange"),
                   DNA191 = c("black", "blue"))
)
```

---

## 8. CD8+ T Cells

CD8+ T cells are the primary effector population in anti-tumor immunity.

```{r cd8-tcell, fig.height=12}
plot_mask(
  colour_by = c("CD8a", "DNA191"),
  colour    = list(CD8a   = c("black", "orange"),
                   DNA191 = c("black", "blue"))
)
```

---

## 9. B Cells

```{r bcell, fig.height=12}
plot_mask(
  colour_by = c("CD19", "DNA191"),
  colour    = list(CD19   = c("black", "yellow"),
                   DNA191 = c("black", "blue"))
)
```

---

## 10. PMN-MDSCs

```{r pmn-mdsc, fig.height=12}
plot_mask(
  colour_by = c("CD15", "DNA191"),
  colour    = list(CD15   = c("black", "yellow"),
                   DNA191 = c("black", "blue"))
)
```

---

## 11. Macrophages

CD14 marks monocytes/macrophages broadly; CD86 distinguishes activated
macrophages from M-MDSCs.

```{r macrophage, fig.height=12}
plot_mask(
  colour_by = c("CD14", "CD86", "DNA191"),
  colour    = list(CD14   = c("black", "yellow"),
                   CD86   = c("black", "magenta"),
                   DNA191 = c("black", "blue"))
)
```

---

## 12. M-MDSCs

CD14 + CD163 marks monocytic MDSCs; CD163 additionally marks M2-polarized
immunosuppressive macrophages.

```{r m-mdsc, fig.height=12}
plot_mask(
  colour_by = c("CD14", "CD163", "DNA191"),
  colour    = list(CD14   = c("black", "yellow"),
                   CD163  = c("black", "orange"),
                   DNA191 = c("black", "blue"))
)
```

---

## 13. DC / mregDC

LAMP3 is the canonical marker for mature regulatory dendritic cells (mregDCs).

```{r dc-mregdc, fig.height=12}
plot_mask(
  colour_by = c("LAMP3", "DNA191"),
  colour    = list(LAMP3  = c("black", "cyan"),
                   DNA191 = c("black", "blue"))
)
```

---

## Session Info

```{r session-info}
sessionInfo()
```