---
title: 'All Patients: Broad Cell Type Clustering: Harmony (no asinh) (k=150)'
author: "Tazein Shah"
date: "07-14-2026"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
subtitle: k=150 Clusters (h4)
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)
set.seed(42)
```

## 1. Load Libraries

```{r libraries}
library(imcRtools)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(scater)
library(scran)
library(bluster)
library(patchwork)
library(RColorBrewer)
```
---

## 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

clustering_path  <- file.path(analysis_path, "clustering")
spe_path         <- file.path(clustering_path, "h4_k150_clusters", "harmony_k150_clusters.rds")
findmarkers_path <- file.path(clustering_path, "h4_k150_clusters", "harmony_k150_cluster_findmarkers.csv")
out_path         <- file.path(clustering_path, "h4_k150_clusters", "harmony_k150_cluster_annot.rds")
```

---

## 3. Load Clustered SPE

```{r}
spe <- readRDS(spe_path)
```

---

## 4. Marker Sets

```{r markers}
# core markers for broad annotation and UMAP overlays
core_markers <- c("SMA", "CD4", "CD8a", "CD19", "CD15",
                  "CD14", "CD86", "LAMP3", "CD163", "EpCAM")
core_markers <- core_markers[core_markers %in% rownames(spe)]
cat("Core markers found:", paste(core_markers, collapse = ", "), "\n")
cat("Missing:", paste(setdiff(
  c("SMA", "CD4", "CD8a", "CD19", "CD15", "CD14", "CD86", "LAMP3", "CD163", "EpCAM"),
  rownames(spe)), collapse = ", "), "\n")

# full panel, used to investigate clusters that don't separate on core markers
full_markers <- c(
  "SMA", "CD4", "CD8a", "CD19", "CD15", "CD14", "CD86", "LAMP3",
  "Ki67", "MKI67", "CD138", "PD1", "PDL1", "FoxP3", "CD68", "CD163",
  "TIM3", "CD141", "CD11c", "CD56", "CD117", "EpCAM", "CD38",
  "CD45RO", "LAG3", "CD1c", "GATA3", "TTF1orCD16"
)
full_markers <- full_markers[full_markers %in% rownames(spe)]
cat("Full panel markers found:", length(full_markers), "\n")
cat("Missing:", paste(setdiff(
  c("SMA", "CD4", "CD8a", "CD19", "CD15", "CD14", "CD86", "LAMP3",
    "Ki67", "MKI67", "CD138", "PD1", "PDL1", "FoxP3", "CD68", "CD163",
    "TIM3", "CD141", "CD11c", "CD56", "CD117", "EpCAM", "CD38",
    "CD45RO", "LAG3", "CD1c", "GATA3", "TTF1orCD16"),
  rownames(spe)), collapse = ", "), "\n")
```

---

## 5. Preprocessing: 99th Percentile Cap + Z-score

```{r helpers}
cap_99 <- function(mat) {
  t(apply(mat, 1, function(x) {
    q99 <- quantile(x, 0.99)
    pmin(x, q99)
  }))
}

zscore_rows <- function(mat) {
  t(scale(t(mat)))
}
```

```{r preprocess}
mat_0100 <- counts(spe) |> cap_99() |> zscore_rows()
assay(spe, "scaled_0100") <- mat_0100

cat("Assays in SPE:", paste(assayNames(spe), collapse = ", "), "\n")
```

---

## 6. UMAP on Harmony Embeddings

`UMAP_harmony` already exists in this object from the batch correction
step, so it is reused here instead of recomputed.

```{r umap}
if ("UMAP_harmony" %in% reducedDimNames(spe)) {
  cat("Reusing existing UMAP_harmony from batch correction step.\n")
} else {
  set.seed(220225)
  spe <- runUMAP(spe, dimred = "Harmony", name = "UMAP_harmony")
  cat("UMAP computed.\n")
}
```

---

## 7. Subset for Visualization

20,000 cells per ROI cap to keep plots manageable.

```{r subset}
set.seed(220225)
cells_per_roi  <- split(seq_len(ncol(spe)), spe$sample_id)
sampled_cells  <- unlist(lapply(cells_per_roi, function(idx) {
  sample(idx, min(20000, length(idx)))
}))
spe_sub <- spe[, sampled_cells]
cat("Cells in subset:", ncol(spe_sub), "\n")
```

---

## 8. UMAP Colored by Broad Cluster

```{r umap-clusters, fig.width=10, fig.height=7}
p_clusters <- dittoDimPlot(spe_sub, var = "broad_clusters",
                           reduction.use = "UMAP_harmony",
                           size = 0.2, do.label = TRUE, labels.size = 3) +
  ggtitle("UMAP colored by Broad Louvain Cluster (k=150)") +
  theme(legend.position = "right")
print(p_clusters)
```

---

## 9. UMAP Per Core Marker

Used to visually match clusters to known marker expression before
annotating.

```{r umap-markers, fig.width=8, fig.height=6}
markers_to_plot <- union(core_markers, "FoxP3")
markers_to_plot <- markers_to_plot[markers_to_plot %in% rownames(spe)]

for (marker in markers_to_plot) {
  p <- dittoDimPlot(spe_sub, var = marker,
                    reduction.use = "UMAP_harmony",
                    assay = "scaled_0100", size = 0.2) +
    scale_color_viridis(name = marker) +
    ggtitle(paste("UMAP:", marker))
  print(p)
}
```

---

## 10. Dot Plot: Core Marker Expression per Cluster

The primary tool for annotating clusters. Each row is a marker, each
column is a cluster. Look for which clusters show high expression of
the core lineage markers.

```{r make-dotplot}
make_dotplot <- function(spe, assay_name, cluster_col, marker_set, title) {
  expr_mat   <- assay(spe, assay_name)[marker_set, ]
  cluster_id <- spe[[cluster_col]]

  mean_expr <- sapply(levels(cluster_id), function(cl) {
    idx <- which(cluster_id == cl)
    rowMeans(expr_mat[, idx, drop = FALSE])
  })

  global_med <- apply(expr_mat, 1, median)
  frac_expr  <- sapply(levels(cluster_id), function(cl) {
    idx <- which(cluster_id == cl)
    rowMeans(expr_mat[, idx, drop = FALSE] > global_med)
  })

  mean_df <- as.data.frame(mean_expr) |>
    rownames_to_column("marker") |>
    pivot_longer(-marker, names_to = "cluster", values_to = "mean_exprs")

  frac_df <- as.data.frame(frac_expr) |>
    rownames_to_column("marker") |>
    pivot_longer(-marker, names_to = "cluster", values_to = "frac_expressing")

  plot_df <- left_join(mean_df, frac_df, by = c("marker", "cluster"))
  plot_df$cluster <- factor(plot_df$cluster, levels = levels(cluster_id))

  marker_order <- rownames(mean_expr)[hclust(dist(mean_expr))$order]
  plot_df$marker <- factor(plot_df$marker, levels = marker_order)

  ggplot(plot_df, aes(x = cluster, y = marker)) +
    geom_point(aes(size = frac_expressing, color = mean_exprs)) +
    scale_color_viridis(option = "magma", name = "Mean\nZ-score expr") +
    scale_size_continuous(range = c(0.5, 6), name = "Fraction\nexpressing") +
    labs(title = title, x = "Cluster", y = "Marker") +
    theme_bw(base_size = 12) +
    theme(
      axis.text.x      = element_text(angle = 0, hjust = 0.5),
      axis.text.y      = element_text(size = 9),
      panel.grid.major = element_line(color = "grey90"),
      legend.position  = "right"
    )
}
```

```{r dotplot-core, fig.width=16, fig.height=6}
p_dot_core <- make_dotplot(spe, "scaled_0100", "broad_clusters",
                           core_markers,
                           "Harmony (no asinh, k=150): Core Marker Expression per Broad Louvain Cluster")
print(p_dot_core)
```

---

## 11. Full Panel Dot Plot for Low Signal Clusters

Several clusters may not show a clear standout marker on the 10 core
markers above, this can happen for macrophage/myeloid populations
where markers like CD14, CD86, CD163 are broadly and continuously
expressed rather than sharply spiking in one cluster, or because the
cluster is actually separated by markers outside the core panel
(e.g. PD1, Ki67, FoxP3, CD45RO), since this Harmony run was computed
on raw counts across the full panel, not just the 10 core markers.
Borderline or low quality segmented cells are also possible. Use this
expanded dot plot to investigate clusters that look flat above before
calling them "Unknown."

```{r dotplot-full, fig.width=18, fig.height=10}
p_dot_full <- make_dotplot(spe, "scaled_0100", "broad_clusters",
                           full_markers,
                           "Harmony (no asinh, k=150): Full Panel Marker Expression per Broad Louvain Cluster")
print(p_dot_full)
```

---

## 12. Statistical Marker Discovery (findMarkers)

For clusters with no obvious standout marker on either dot plot,
`findMarkers()` ranks markers by how distinctly they separate that
cluster from all others, accounting for variance rather than just
mean difference. This can surface subtle but consistent shifts that
don't show up clearly in a Z-score dot plot averaged across thousands
of cells.

```{r findmarkers}
markers_de <- findMarkers(spe,
                          groups     = spe$broad_clusters,
                          assay.type = "scaled_0100",
                          subset.row = full_markers)

# save all clusters to CSV
all_markers <- lapply(levels(spe$broad_clusters), function(cl) {
  df <- as.data.frame(markers_de[[cl]])
  data.frame(
    cluster = cl,
    marker  = rownames(df),
    Top     = df$Top,
    FDR     = df$FDR,
    logFC   = df$summary.logFC
  )
})

all_markers_df <- do.call(rbind, all_markers)

dir.create(dirname(findmarkers_path), recursive = TRUE, showWarnings = FALSE)
write.csv(all_markers_df, findmarkers_path, row.names = FALSE)

# print top 5 positive markers per cluster
df <- read.csv(findmarkers_path)

# summary table: top positive marker per cluster at a glance
top1_summary <- do.call(rbind, lapply(levels(spe$broad_clusters), function(cl) {
  sub <- df[df$cluster == cl & df$logFC > 1.0, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  if (nrow(sub) == 0) {
    data.frame(cluster = cl, top_markers = "none above logFC 1.0")
  } else {
    data.frame(cluster = cl,
               top_markers = paste(head(sub$marker, 5), collapse = ", "))
  }
}))

cat("Top positive markers (logFC > 1.0) per cluster:\n")
print(knitr::kable(top1_summary, col.names = c("Cluster", "Top Positive Markers")))

cat("\n\nDetailed tables:\n")

# detailed table per cluster, positive logFC only
for (cl in levels(spe$broad_clusters)) {
  sub <- df[df$cluster == cl & df$Top <= 5 & df$logFC > 1.0, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  if (nrow(sub) == 0) {
    cat("Cluster", cl, ": no markers with logFC > 1.0 in top 5\n\n")
  } else {
    print(knitr::kable(sub[, c("marker", "Top", "FDR", "logFC")],
                       caption = paste("Cluster", cl),
                       digits  = 3))
  }
}
```
---

## 13. Trying to Find "Unknown" Clusters

Cluster 2 was the one with no markers above logFC 1.0, so the
threshold is lowered here to look for anything subtler.

```{r}
for (cl in c("2")) {
  sub <- df[df$cluster == cl, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  print(knitr::kable(sub[, c("marker", "Top", "FDR", "logFC")],
                     caption = paste("Cluster", cl, "(all markers)"),
                     digits  = 3))
}
```
---

## 14. Diagnostic: Area and Patient Composition for Ambiguous Clusters

Clusters 2, 3, 10, and 11 are the ones flagged as clusters of concern
from the top positive marker table: cluster 2 had no marker above
logFC 1.0, cluster 3 mixed a DC signature (LAMP3) with epithelial and
Treg markers (EpCAM, FoxP3), and clusters 10 and 11 both mixed
lymphoid/plasma markers with the granulocyte marker CD15. Cell area
and patient composition are checked here to build evidence before
deciding whether any of these are doublets, low quality segmentation,
or real populations.

```{r diagnostic-patient, fig.width=12, fig.height=5}
ambiguous_clusters <- c("2")

if (length(ambiguous_clusters) > 0) {
  # patient composition per ambiguous cluster
  comp_df <- as.data.frame(
    table(cluster   = as.character(spe$broad_clusters),
          patient   = spe$patient_id)
  )
  comp_df <- comp_df[comp_df$cluster %in% ambiguous_clusters & comp_df$Freq > 0, ]

  p_comp <- ggplot(comp_df, aes(x = patient, y = Freq, fill = patient)) +
    geom_col() +
    facet_wrap(~ cluster, scales = "free_y", nrow = 2) +
    labs(title = "Patient Composition for Ambiguous Clusters",
         x = "Patient", y = "Cell Count") +
    theme_bw() +
    theme(axis.text.x  = element_text(angle = 45, hjust = 1),
          legend.position = "none")
  print(p_comp)
} else {
  cat("Fill in ambiguous_clusters above after inspecting the k=150 dot plots.\n")
}
```

---

## 15. Manual Broad Annotation

Filled in below based on the top positive marker table (`top1_summary`
from chunk 12) plus the diagnostics in chunk 14. Clusters 2, 10, and
11 are left as Unknown pending the area and patient composition
results; cluster 3 is tentatively called DC / mregDC but flagged as a
possible doublet population between DCs, tumor cells, and Tregs.

```{r annotate, echo=TRUE}
broad_annotation <- c(
  "1"  = "Fibroblast",
  "2"  = "Low Stain",
  "3"  = "DC / mregDC",
  "4"  = "Epithelial",
  "5"  = "B Cell",
  "6"  = "CD4+ T Cell",
  "7"  = "Macrophage",
  "8"  = "Tumor Epithelial (TTF1+)",
  "9"  = "Possible Doublet (Treg / CD138)",
  "10" = "Dysfunctional CD8+ T Cell",
  "11" = "Proliferating"
)

if (length(broad_annotation) > 0) {
  spe$broad_celltype <- broad_annotation[as.character(spe$broad_clusters)]
  cat("Annotation applied.\n")
  print(table(spe$broad_celltype, useNA = "ifany"))
} else {
  cat("No annotation provided yet. Fill in broad_annotation above after inspecting plots.\n")
}
```
---

## 16. UMAP Colored by Broad Cell Type

Run this chunk after filling in the annotation above.

```{r umap-celltype, fig.width=10, fig.height=7}
if ("broad_celltype" %in% colnames(colData(spe))) {
  spe_sub$broad_celltype <- spe$broad_celltype[match(colnames(spe_sub), colnames(spe))]

  p_celltype <- dittoDimPlot(spe_sub, var = "broad_celltype",
                             reduction.use = "UMAP_harmony",
                             size = 0.2, do.label = TRUE, labels.size = 3) +
    ggtitle("UMAP colored by Broad Cell Type (k=150)") +
    theme(legend.position = "right")
  print(p_celltype)
} else {
  cat("Run chunk 15 first to apply annotations.\n")
}
```

---

## 17. Save Annotated SPE

```{r save}
if ("broad_celltype" %in% colnames(colData(spe))) {
  dir.create(dirname(out_path), recursive = TRUE, showWarnings = FALSE)
  saveRDS(spe, out_path)
  cat("Saved annotated SPE to:", out_path, "\n")
} else {
  cat("Annotation not yet applied. Fill in chunk 15 before saving.\n")}
```

---
