---
title: "Niche Annotation: Composition Summary for Significant Niches"
author: "Tazein Shah"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 8)

suppressPackageStartupMessages({
  library(SpatialExperiment)
  library(tidyverse)
  library(pheatmap)
})
```

## Overview

This notebook summarizes the average neighborhood composition
(`spe$aggregatedNeighbors`) for niches found significant in the
`sccomp` N-status and Stage models from the niche generation
notebook, to support manual niche annotation following the same
convention used for cluster annotation.

---

## 1. Paths

Reuses `niche_path` from the niche generation notebook (unchanged).

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"
niche_path    <- file.path(analysis_path, "spatial_analysis", "niche_k50_thresh20")

niche_spe_path      <- file.path(niche_path, "spe_niche.rds")
sccomp_nstatus_path <- file.path(niche_path, "sccomp_nstatus.rds")
sccomp_stage_path   <- file.path(niche_path, "sccomp_stage.rds")

composition_out_path   <- file.path(niche_path, "niche_avg_composition.csv")
heatmap_out_path       <- file.path(niche_path, "niche_composition_heatmap.pdf")
top_celltypes_out_path <- file.path(niche_path, "niche_top_celltypes.csv")
```

---

## 2. Load objects

```{r load}
spe <- readRDS(niche_spe_path)
sccomp_niche_nstatus <- readRDS(sccomp_nstatus_path)
sccomp_niche_stage   <- readRDS(sccomp_stage_path)
```

---

## 2b. Diagnose cell type naming

Confirmed: the raw names are already clean at the source, no mangling
here. The actual bug turned out to be downstream -- `column_to_rownames()`
(used to build matrices for the dot plot and heatmap) internally calls
`as.data.frame()` with `check.names = TRUE` and no way to override it,
silently re-mangling clean names every time it's called. Fixed in
Sections 4b and 5 by building those matrices manually with `as.matrix()`
instead, which doesn't touch names.

```{r diagnose-names}
cat("Column names of spe$aggregatedNeighbors:\n")
print(colnames(spe$aggregatedNeighbors))

cat("\nLevels of spe$final_celltype:\n")
print(levels(spe$final_celltype))
```

---

## 3. Identify significant niches

Significant niches (FDR < 0.05) from each model, unioned.

```{r significant-niches}
sig_nstatus <- sccomp_niche_nstatus |>
  filter(factor == "N_status", c_FDR < 0.05) |>
  pull(Niche) |>
  unique()

sig_stage <- sccomp_niche_stage |>
  filter(factor == "Stage", c_FDR < 0.05) |>
  pull(Niche) |>
  unique()

sig_niches <- union(sig_nstatus, sig_stage)

cat("Significant niches (N-status):", length(sig_nstatus), "\n")
cat("Significant niches (Stage):", length(sig_stage), "\n")
cat("Significant niches (union):", length(sig_niches), "\n")
print(sort(as.numeric(as.character(sig_niches))))
```

---

## 3b. Significant niches: effect size and FDR

Combined table of every niche x contrast pair that cleared FDR < 0.05
in either model, for reference alongside the composition/annotation
work below.

```{r significant-niches-table}
sig_nstatus_table <- sccomp_niche_nstatus |>
  filter(factor == "N_status", c_FDR < 0.05) |>
  select(Niche, factor, parameter, c_effect, c_FDR)

sig_stage_table <- sccomp_niche_stage |>
  filter(factor == "Stage", c_FDR < 0.05) |>
  select(Niche, factor, parameter, c_effect, c_FDR)

sig_niches_table <- bind_rows(sig_nstatus_table, sig_stage_table) |>
  arrange(Niche, c_FDR)

sig_niches_table |>
  as.data.frame() |>
  knitr::kable(digits = 3, caption = "Significant niches: effect size and FDR (N-status and Stage models)")
```

---

## 4. Average neighborhood composition per niche

All 50 niches, saved for reference.

```{r composition}
# Confirmed via the diagnostic chunk above: spe$aggregatedNeighbors' column
# names and levels(spe$final_celltype) are already clean (no mangling at the
# source). check.names = FALSE just guards against as.data.frame() re-
# mangling them at this conversion step.
neighbor_mat <- as.data.frame(spe$aggregatedNeighbors, check.names = FALSE)
neighbor_mat$Niche <- spe$Niche

niche_composition <- neighbor_mat |>
  group_by(Niche) |>
  summarise(across(everything(), mean), .groups = "drop")

write_csv(niche_composition, composition_out_path)
cat("Saved full niche composition table to:", composition_out_path, "\n")

# Re-read the CSV rather than continuing to use the in-memory object.
# read_csv() takes the header row literally (no make.names-style checking,
# unlike as.data.frame()/column_to_rownames()), so every chunk below now
# works from a guaranteed-clean, disk-verified copy of the names.
niche_composition <- read_csv(composition_out_path, show_col_types = FALSE)
```

---

## 4b. Detailed composition dot plot (all niches)

Dot size and color both encode mean fraction; niches and cell types
are hierarchically clustered so similar composition profiles sit next
to each other, giving a more detailed view across all 50 niches than
a single heatmap panel.

```{r composition-dotplot, fig.width=14, fig.height=10}
# column_to_rownames() internally converts via as.data.frame() with
# check.names = TRUE (no way to override), which re-mangles the clean tibble
# column names. Build the matrix manually with as.matrix() instead, which
# doesn't touch names at all.
niche_mat_full <- niche_composition |> select(-Niche) |> as.matrix()
rownames(niche_mat_full) <- niche_composition$Niche

niche_order    <- rownames(niche_mat_full)[hclust(dist(niche_mat_full))$order]
celltype_order <- colnames(niche_mat_full)[hclust(dist(t(niche_mat_full)))$order]

niche_long <- niche_composition |>
  pivot_longer(-Niche, names_to = "cell_type", values_to = "mean_fraction") |>
  mutate(
    Niche = factor(Niche, levels = niche_order),
    cell_type = factor(cell_type, levels = celltype_order)
  )

ggplot(niche_long, aes(x = Niche, y = cell_type, size = mean_fraction, color = mean_fraction)) +
  geom_point() +
  scale_color_viridis_c() +
  scale_size_continuous(range = c(0.5, 6)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7),
        axis.text.y = element_text(size = 8)) +
  labs(x = "Niche", y = "Cell type", color = "Mean fraction", size = "Mean fraction",
       title = "Average neighborhood composition per niche (all 50 niches)")
```

---

## 5. Heatmap (significant niches only)

```{r heatmap}
suppressPackageStartupMessages({
  library(RColorBrewer)
})

niche_composition_sig <- niche_composition |>
  filter(Niche %in% sig_niches)

# Drop cell types with near-zero variance across these niches. Column
# z-scoring (scale = "column") divides by each column's SD -- a flat/near-zero
# column has a tiny SD, so a single stray value blows up into an extreme
# z-score. That's what was producing the noisy, extreme-red heatmap.
range_cutoff_sig <- 0.05

celltype_range_sig <- niche_composition_sig |>
  select(-Niche) |>
  summarise(across(everything(), ~ max(.x) - min(.x))) |>
  pivot_longer(everything(), names_to = "cell_type", values_to = "range")

keep_celltypes_sig <- celltype_range_sig |>
  filter(range > range_cutoff_sig) |>
  pull(cell_type)

cat("Keeping", length(keep_celltypes_sig), "of", ncol(niche_composition_sig) - 1,
    "cell types (range >", range_cutoff_sig, ")\n")

niche_composition_sig_filtered <- niche_composition_sig |>
  select(Niche, all_of(keep_celltypes_sig))

niche_mat_sig <- niche_composition_sig_filtered |> select(-Niche) |> as.matrix()
rownames(niche_mat_sig) <- niche_composition_sig_filtered$Niche

# Pre-scale (column z-score) ourselves so we can force symmetric, well-
# contrasted breaks around zero rather than relying on pheatmap's default
# range, which produced the washed-out/inconsistent coloring above.
niche_mat_scaled <- scale(niche_mat_sig)
max_abs <- max(abs(niche_mat_scaled), na.rm = TRUE)
breaks <- seq(-max_abs, max_abs, length.out = 101)
heat_colors <- colorRampPalette(rev(brewer.pal(11, "RdBu")))(100)

pdf(heatmap_out_path, width = 14, height = 10)
pheatmap(niche_mat_scaled,
         scale = "none",
         color = heat_colors,
         breaks = breaks,
         fontsize_row = 9,
         fontsize_col = 9,
         border_color = NA,
         angle_col = 45,
         main = "Average neighborhood composition: significant niches only")
dev.off()
cat("Saved heatmap to:", heatmap_out_path, "\n")

pheatmap(niche_mat_scaled,
         scale = "none",
         color = heat_colors,
         breaks = breaks,
         fontsize_row = 9,
         fontsize_col = 9,
         border_color = NA,
         angle_col = 45,
         main = "Average neighborhood composition: significant niches only")
```

---

## 6. Top dominant cell types per significant niche

For writing the usual `#Niche N: cell_type1, cell_type2: reasoning`
annotation comments.

```{r top-celltypes}
top_n <- 5

niche_top_celltypes <- niche_composition_sig |>
  pivot_longer(-Niche, names_to = "cell_type", values_to = "mean_fraction") |>
  group_by(Niche) |>
  slice_max(mean_fraction, n = top_n) |>
  arrange(Niche, desc(mean_fraction))

write_csv(niche_top_celltypes, top_celltypes_out_path)
cat("Saved top-", top_n, " cell types per significant niche to:", top_celltypes_out_path, "\n")

niche_top_celltypes |>
  as.data.frame() |>
  knitr::kable(digits = 3, caption = "Top dominant cell types per significant niche")
```
