---
title: "Niche Generation and Differential Abundance"
author: "Tazein Shah"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)

suppressPackageStartupMessages({
  library(imcRtools)
  library(tidyverse)
  library(SingleCellExperiment)
  library(SpatialExperiment)
  library(sccomp)
  library(ggrepel)
})

set.seed(12345)
```

## Overview

This notebook generates cellular neighborhoods (niches) on the fully
annotated, spatial-graph-equipped dataset, and tests niche composition
for association with N-status and AJCC stage. Structure follows the
preprint's original niche generation script as closely as possible.

Two deliberate deviations from that script are noted inline below
(both flagged where they occur, and summarized in the Notes section
at the end):

1. **Spatial graph threshold.** This notebook reuses an existing
   `expansion_interaction_graph` built upstream at `threshold = 20`.
   The original preprint script used `threshold = 50`. This
   discrepancy between the manuscript text/figure legend (20) and the
   actual code (50) has not yet been resolved, so this notebook
   proceeds with the already-built 20-threshold graph rather than
   rebuilding at 50.
2. Everything else -- `aggregateNeighbors`, `kmeans`, and the
   table/volcano/boxplot outputs -- follows the original script's
   structure directly. The differential abundance step itself uses
   `sccomp_estimate() |> sccomp_test()` rather than the original
   script's `sccomp_glm()`, since that function does not exist in the
   installed package version (see Notes).

---

## 1. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

spatial_graph_path <- file.path(analysis_path, "spatial_analysis", "spe_spatial_graph.rds")

niche_path <- file.path(analysis_path, "spatial_analysis", "niche_k50_thresh20")
dir.create(niche_path, recursive = TRUE, showWarnings = FALSE)

niche_spe_out_path      <- file.path(niche_path, "spe_niche.rds")
sccomp_nstatus_out_path <- file.path(niche_path, "sccomp_nstatus.rds")
sccomp_stage_out_path   <- file.path(niche_path, "sccomp_stage.rds")
```

---

## 2. Load spatial graph object

Loads the object that already has `final_celltype` populated (merged
broad annotations + all subclustering results, low-quality clusters
excluded) and the `expansion_interaction_graph` colPair already built
at `threshold = 20`.

```{r load}
spe <- readRDS(spatial_graph_path)

cat("Cells:", ncol(spe), "\n")
cat("Cell types:", nlevels(spe$final_celltype), "\n")
print(table(spe$final_celltype))
```

---

## 3. Aggregate neighbor composition

```{r aggregate-neighbors}
# Reusing the existing graph (threshold = 20) rather than rebuilding it.
# See Overview note above re: 20 vs. 50 discrepancy.
spe <- aggregateNeighbors(spe,
                          colPairName = "expansion_interaction_graph",
                          aggregate_by = "metadata",
                          count_by = "final_celltype")
```

---

## 4. K-means niche clustering

```{r kmeans-niches}
kmeansClust <- 50

cn_1 <- kmeans(spe$aggregatedNeighbors, centers = kmeansClust, iter.max = 10000)
spe$Niche <- as.factor(cn_1$cluster)

table(spe$Niche)
```

---

## 5. Patient-level covariates

```{r covariates}
spe$stage_group <- sapply(spe$patient_id, function(x) {
  if (x %in% c("Patient12", "Patient15")) return("IIB")
  else if (x == "Patient13") return("IIIA")
  else if (x == "Patient19") return("IB")
  else if (x == "Patient11" || grepl("^Patient20", x)) return("IA3")
  else if (x %in% c("Patient14", "Patient16", "Patient17", "Patient18")) return("IA2")
  else if (x == "Patient10") return("Benign")
  else return(NA)
})

spe$stage_group <- factor(spe$stage_group,
                          levels = c("Benign", "IA2", "IA3", "IB", "IIIA", "IIB"))

# Three-tier disease state grouping (Benign / Lower / Higher)
spe$disease_state <- sapply(spe$patient_id, function(x) {
  if (x == "Patient10") return("Benign")
  else if (x %in% c("Patient11", "Patient14", "Patient16", "Patient17", "Patient18") || grepl("^Patient20", x)) return("Lower")
  else if (x %in% c("Patient12", "Patient13", "Patient15", "Patient19")) return("Higher")
  else return(NA)
})

spe$disease_state <- factor(spe$disease_state, levels = c("Benign", "Lower", "Higher"))

# N-status: only IIB (P12, P15) is N1; everything else is N0
# NOTE: flagging for confirmation -- does IIIA (Patient13) carry N1 status?
# AJCC IIIA often implies nodal involvement, but this cohort's original
# stage list only assigned N1 possibility to IIB. Left as-is pending confirmation.
spe$N_status <- factor(ifelse(spe$stage_group == "IIB", "N1", "N0"),
                       levels = c("N0", "N1"))

table(spe$patient_id, spe$N_status)
table(spe$patient_id, spe$stage_group)
table(spe$patient_id, spe$disease_state)
```

```{r save-niche-spe}
saveRDS(spe, niche_spe_out_path)
cat("Saved object with niches to:", niche_spe_out_path, "\n")
```

---

## 6. Differential niche abundance (sccomp)

`.sample = Patient`, since N-status and stage are constant within
patient (not within ROI) -- mirroring how the original script's `LN`
is the invariant unit for its `TumorStage`/`HighN1` covariates.

```{r sccomp-prep}
sceIMC <- SingleCellExperiment(list(counts = assay(spe, "counts")))
sceIMC$Patient   <- factor(spe$patient_id)
sceIMC$Sample    <- factor(spe$sample_id)
sceIMC$Niche     <- factor(spe$Niche)
sceIMC$N_status  <- factor(spe$N_status, levels = c("N0", "N1"))
sceIMC$Stage     <- factor(spe$stage_group,
                           levels = c("Benign", "IA2", "IA3", "IB", "IIIA", "IIB"))
```

### N-status model (parallel to `HighN1` in the original script)

```{r sccomp-nstatus}
if (file.exists(sccomp_nstatus_out_path)) {
  cat("Found existing output, loading:", sccomp_nstatus_out_path, "\n")
  sccomp_niche_nstatus <- readRDS(sccomp_nstatus_out_path)
} else {
  sccomp_niche_nstatus <-
    sceIMC |>
    sccomp_estimate(
      formula_composition = ~ N_status + (1 | Patient),
      .sample = Sample,
      .cell_group = Niche,
      bimodal_mean_variability_association = FALSE,
      variational_inference = FALSE,
      output_directory = file.path(niche_path, "sccomp_draws_nstatus"),
      cores = 8
    ) |>
    sccomp_test()

  saveRDS(sccomp_niche_nstatus, sccomp_nstatus_out_path)
}
```

### Stage model (parallel to `TumorStage` in the original script)

```{r sccomp-stage}
if (file.exists(sccomp_stage_out_path)) {
  cat("Found existing output, loading:", sccomp_stage_out_path, "\n")
  sccomp_niche_stage <- readRDS(sccomp_stage_out_path)
} else {
  sccomp_niche_stage <-
    sceIMC |>
    sccomp_estimate(
      formula_composition = ~ Stage + (1 | Patient),
      .sample = Sample,
      .cell_group = Niche,
      bimodal_mean_variability_association = FALSE,
      variational_inference = FALSE,
      output_directory = file.path(niche_path, "sccomp_draws_stage"),
      cores = 8
    ) |>
    sccomp_test()

  saveRDS(sccomp_niche_stage, sccomp_stage_out_path)
}
```

---

## 7. Results table

```{r results-table}
sccomp_niche_nstatus <- readRDS(sccomp_nstatus_out_path)
sccomp_niche_nstatus_sub <- sccomp_niche_nstatus[which(sccomp_niche_nstatus$factor == "N_status"), ]

sccomp_niche_nstatus_sub |>
  filter(c_FDR < 0.1) |>
  arrange(c_FDR) |>
  as.data.frame() |>
  select(-starts_with("v_")) |>
  knitr::kable(digits = 3, caption = "Niche composition association with N-status (FDR < 0.1)")
```

---

## 8. Volcano plot

```{r volcano}
plot_df <- sccomp_niche_nstatus_sub |>
  as.data.frame() |>
  filter(!is.nan(c_FDR), c_FDR < 0.1) |>
  mutate(
    LogFDR = -log(c_FDR),
    Significant = ifelse(c_FDR < 0.05, "FDR < 0.05", "Not significant")
  )

ggplot(plot_df, aes(x = c_effect, y = LogFDR, color = Significant)) +
  geom_point(size = 2) +
  geom_text_repel(
    aes(label = Niche), size = 3, show.legend = FALSE
  ) +
  scale_color_manual(values = c("FDR < 0.05" = "firebrick", "Not significant" = "grey70")) +
  theme_classic() +
  labs(x = "Effect size", y = "-log(FDR)",
       title = "Niche composition vs. N-status")
```

---

## 9. Results table (Stage)

`factor == "Stage"` covers all 5 non-reference contrasts (each vs.
`Benign`) at once, distinguished by `parameter`
(`StageIA2`, `StageIA3`, `StageIB`, `StageIIIA`, `StageIIB`) --
250 rows total (50 niches x 5 contrasts).

```{r results-table-stage}
sccomp_niche_stage <- readRDS(sccomp_stage_out_path)
sccomp_niche_stage_sub <- sccomp_niche_stage[which(sccomp_niche_stage$factor == "Stage"), ]

sccomp_niche_stage_sub |>
  filter(c_FDR < 0.1) |>
  arrange(c_FDR) |>
  as.data.frame() |>
  select(-starts_with("v_")) |>
  knitr::kable(digits = 3, caption = "Niche composition association with Stage (vs. Benign, FDR < 0.1)")
```

---

## 10. Volcano plot (Stage) {.tabset}

Faceted by contrast, since each stage level vs. Benign is a separate
comparison rather than a single binary effect.

```{r volcano-stage, results = "asis", fig.width = 7, fig.height = 5.5}
plot_df_stage <- sccomp_niche_stage_sub |>
  as.data.frame() |>
  filter(!is.nan(c_FDR), c_FDR < 0.1) |>
  mutate(
    LogFDR = -log(c_FDR),
    Significant = ifelse(c_FDR < 0.05, "FDR < 0.05", "Not significant")
  )

for (contrast in unique(plot_df_stage$parameter)) {
  cat("\n\n### ", contrast, "\n\n")
  df_sub <- plot_df_stage[plot_df_stage$parameter == contrast, ]
  print(
    ggplot(df_sub, aes(x = c_effect, y = LogFDR, color = Significant)) +
      geom_point(size = 2) +
      geom_text_repel(aes(label = Niche), size = 3, show.legend = FALSE) +
      scale_color_manual(values = c("FDR < 0.05" = "firebrick", "Not significant" = "grey70")) +
      theme_classic() +
      labs(x = "Effect size", y = "-log(FDR)",
           title = paste0("Niche composition: ", contrast, " vs. Benign"))
  )
  cat("\n\n")
}
```

---

## 11. Boxplots per niche {.tabset}

```{r boxplots, results = "asis", fig.width = 5, fig.height = 5.5}
# Only plot niches significant (FDR < 0.1) in either the N-status or
# Stage model, rather than all 50.
sig_niches_nstatus <- sccomp_niche_nstatus_sub |>
  filter(c_FDR < 0.1) |>
  pull(Niche) |>
  as.character()

sig_niches_stage <- sccomp_niche_stage_sub |>
  filter(c_FDR < 0.1) |>
  pull(Niche) |>
  as.character()

sig_niches <- union(sig_niches_nstatus, sig_niches_stage)

cat("Niches significant (FDR < 0.1) in N-status model:", length(sig_niches_nstatus), "\n")
cat("Niches significant (FDR < 0.1) in Stage model:", length(sig_niches_stage), "\n")
cat("Total unique niches plotted below:", length(sig_niches), "\n")

niche_prop <- table(spe$sample_id, spe$Niche) |>
  prop.table(margin = 1) |>
  as.data.frame()
colnames(niche_prop) <- c("Sample", "Niche", "Value")

niche_prop$Patient <- spe$patient_id[match(niche_prop$Sample, spe$sample_id)]
niche_prop$Stage <- spe$stage_group[match(niche_prop$Patient, spe$patient_id)]

for (n in sig_niches) {
  cat("\n\n### Niche", n, "\n\n")
  df_sub <- niche_prop[niche_prop$Niche == n, ]
  df_sub$Value <- df_sub$Value * 100
  print(
    ggplot(df_sub, aes(x = Stage, y = Value)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(aes(color = Patient), width = 0.1, size = 1.5) +
      theme_classic() +
      ylab("% within sample") +
      ggtitle(paste0("Niche ", n))
  )
  cat("\n\n")
}
```

---

## Notes

- **Graph threshold (20 vs. 50):** this notebook builds on an
  upstream object whose spatial graph was constructed at
  `threshold = 20`, matching the manuscript text/figure legend. The
  original preprint's niche generation script itself used
  `threshold = 50`. That inconsistency between text and code is
  unresolved upstream; this notebook did not rebuild the graph and so
  inherits the 20-threshold version. Easy to regenerate with a
  different threshold later without touching anything downstream of
  the graph-building step.
- **`sccomp_glm()` vs. `sccomp_estimate()`/`sccomp_test()`:** the
  original script calls `sccomp_glm()` directly, but this installed
  package version does not have that function at all -- confirmed via
  a direct "could not find function" error. This notebook uses
  `sccomp_estimate() |> sccomp_test()` instead. Output object shape
  (`factor`, `c_effect`, `c_FDR` columns) is the same either way, so
  nothing downstream (results table, volcano, boxplots) needed to
  change.
- **`variational_inference = FALSE`:** the default variational Bayes
  fitting failed to converge after 5 attempts. Forcing full HMC
  sampling instead resolved this -- confirmed working (6 chains, no
  divergences, ~145 seconds total).
- **`output_directory`:** set explicitly to a subfolder under
  `niche_path`, since the default working-directory-relative path
  `sccomp` tries to use was not writable in this environment.
- `bimodal_mean_variability_association = FALSE` used for both models
  for consistency with the cell-type-level `sccomp` convention already
  established in this pipeline.
- `.sample = Patient` chosen because N-status and stage are constant
  within patient (not within ROI); confirmed via the
  `table(spe$patient_id, spe$N_status)` cross-tab above.
