---
title: "Spatial Visualization: Cell Centroids, Graphs, and Cell-Cell Interactions"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

Builds the spatial graph from the finalized annotated dataset, visualizes
cell centroids + graph edges per ROI, and runs formal cell-cell interaction
testing -- addressing the to-do item "visualize the cells' centroids and
cell-cell interactions as spatial graphs."

**IMPORTANT:** this loads `spe_final_celltype.rds`. As of writing, the final
annotation is still being finalized (a few label naming/marker-mismatch
questions are still open) -- re-run this notebook once that file is locked,
since anything built here on an interim version will need to be redone
against the final labels before being treated as a result.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# 1. Load Libraries

```{r libraries}
.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(imcRtools)
library(tidyverse)
library(SingleCellExperiment)
library(SpatialExperiment)
library(scater)
library(RColorBrewer)
```

# 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

final_celltype_path <- file.path(analysis_path, "final_cell_annotation", "spe_final_celltype.rds")

spatial_analysis_path <- file.path(analysis_path, "spatial_analysis")
dir.create(spatial_analysis_path, recursive = TRUE, showWarnings = FALSE)

spatial_graph_out_path <- file.path(spatial_analysis_path, "spe_spatial_graph.rds")
interactions_out_path  <- file.path(spatial_analysis_path, "cell_interactions.csv")
```

# 3. Load Finalized Annotated Object

```{r load-spe}
spe <- readRDS(final_celltype_path)

cat("Cells loaded:", ncol(spe), "\n")
cat("Cell types:", nlevels(spe$final_celltype), "\n")
cat("Samples/ROIs:", length(unique(spe$sample_id)), "\n")
```

# 4. Build Spatial Graph

Using `type = "expansion"` and `threshold = 20`, matching the reference
publication pipeline's graph-building approach (threshold value per team's
current working decision -- pending confirmation from Campbell given the
20 vs. 50 discrepancy between the manuscript text and code).

```{r build-spatial-graph}
spe <- buildSpatialGraph(
  spe,
  img_id    = "sample_id",
  type      = "expansion",
  threshold = 20
)

colPairNames(spe)
```

```{r graph-diagnostics}
graph_name <- colPairNames(spe)[1]

neighbor_counts <- imcRtools::countNeighbors(spe, colPairName = graph_name)
cat("Mean neighbors per cell:", mean(neighbor_counts, na.rm = TRUE), "\n")
cat("Median neighbors per cell:", median(neighbor_counts, na.rm = TRUE), "\n")
```

```{r save-spatial-graph}
saveRDS(spe, spatial_graph_out_path)
cat("Saved spatial graph object to:", spatial_graph_out_path, "\n")
```

# 5. Visualize Cell Centroids + Graph Edges

Spot-check individual ROIs to confirm the graph looks reasonable (respects
ROI boundaries, edges connect physically nearby cells) before running any
interaction statistics on it.

```{r plot-spatial-single, fig.width=10, fig.height=8}
one_sample_id <- unique(spe$sample_id)[1]
one_sample <- spe[, spe$sample_id == one_sample_id]

n_celltypes <- nlevels(spe$final_celltype)
celltype_colors <- colorRampPalette(brewer.pal(12, "Paired"))(n_celltypes)
names(celltype_colors) <- levels(spe$final_celltype)

plotSpatial(
  one_sample,
  node_color_by = "final_celltype",
  img_id        = "sample_id",
  draw_edges    = TRUE,
  colPairName   = graph_name,
  node_size_fix = 0.5
) +
  scale_color_manual(values = celltype_colors) +
  labs(title = paste("Spatial graph:", one_sample_id)) +
  theme(legend.text = element_text(size = 6),
        legend.key.size = unit(0.3, "cm"))
```

```{r plot-spatial-multiple, fig.width=16, fig.height=12}
# Spot-check a few more ROIs across different patients/disease states
sample_ids_to_check <- unique(spe$sample_id)[1:min(4, length(unique(spe$sample_id)))]

for (sid in sample_ids_to_check) {
  one_sample <- spe[, spe$sample_id == sid]
  p <- plotSpatial(
    one_sample,
    node_color_by = "final_celltype",
    img_id        = "sample_id",
    draw_edges    = TRUE,
    colPairName   = graph_name,
    node_size_fix = 0.5
  ) +
    scale_color_manual(values = celltype_colors) +
    labs(title = paste("Spatial graph:", sid)) +
    theme(legend.position = "none")
  print(p)
}
```

# 6. Cell-Cell Interaction Testing

Formal test of which cell type pairs are spatially enriched or depleted
relative to chance, using `imcRtools::testInteractions()` (permutation-based
neighborhood enrichment/depletion, conceptually similar to but distinct from
the preprint's custom neighbor-composition function --
see `countInteractionsPerCell()` in `Publication_Code_IMC_Analysis.Rmd` if
exact methodological consistency with the reference pipeline is required;
confirm with Campbell/Yusuke which approach the team wants).

```{r test-interactions}
set.seed(220225)
interactions <- testInteractions(
  spe,
  group_by    = "sample_id",
  label       = "final_celltype",
  colPairName = graph_name,
  method      = "classic"
)

interactions_df <- as.data.frame(interactions)
write.csv(interactions_df, interactions_out_path, row.names = FALSE)
cat("Saved interaction results to:", interactions_out_path, "\n")

head(interactions_df)
```

# 7. Visualize Interaction Results

```{r interaction-heatmap, fig.width=12, fig.height=10}
ggplot(interactions_df, aes(x = from_label, y = to_label, fill = strength)) +
  geom_tile() +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, name = "Interaction\nstrength") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6),
        axis.text.y = element_text(size = 6)) +
  labs(x = NULL, y = NULL,
       title = "Cell-Cell Interaction Enrichment/Depletion",
       subtitle = "Red = enriched (more neighbors than expected), Blue = depleted")
```

```{r interaction-significant, fig.width=12, fig.height=8}
# Focus on statistically significant interactions only, if a p-value/FDR
# column is present in the output (column name may vary by imcRtools version)
if ("p.value" %in% colnames(interactions_df)) {
  sig_interactions <- interactions_df |>
    filter(p.value < 0.05) |>
    arrange(p.value)

  cat("Number of significant interactions (p < 0.05):", nrow(sig_interactions), "\n")

  sig_interactions |>
    head(30) |>
    knitr::kable(digits = 3, caption = "Top 30 significant cell-cell interactions")
} else {
  cat("No p-value column found -- check colnames(interactions_df) for the actual column name.\n")
  colnames(interactions_df)
}
```

# 8. Save Final Spatial Graph + Interaction Object

```{r save-final}
saveRDS(spe, spatial_graph_out_path)
cat("Spatial graph object saved to:", spatial_graph_out_path, "\n")
cat("Interaction results saved to:", interactions_out_path, "\n")
```

# Notes

- **Re-run against the finalized `spe_final_celltype.rds`** once the open
  label questions (HLADRB-high naming, CD138/CD39 broad non-specificity,
  T Cell/DC Admixture CD11c mismatch, etc.) are resolved -- everything in
  this notebook is built directly on top of `final_celltype`, so any label
  changes propagate here too.
- `threshold = 20` used for `buildSpatialGraph()` per current team decision;
  flag to Campbell if this should be reconciled with the `threshold = 50`
  discrepancy noted between the reference manuscript text and its code.
- `testInteractions(method = "classic")` was used here as a standard
  permutation-based approach. This differs methodologically from the
  reference pipeline's custom `countInteractionsPerCell()` function (which
  computes neighbor composition ratios rather than enrichment/depletion
  significance) -- confirm with Campbell/Yusuke which approach the team
  wants for consistency with the published methodology, especially if this
  analysis is meant to be directly compared to the preprint's Figure 4G/H.
