---
title: "Final Annotation Violin Plots by Cell Type Category"
author: "Tazein Shah"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 1. Load Libraries

```{r libraries}
library(SpatialExperiment)
library(tidyverse)
```

## 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

spe_path <- file.path(analysis_path, "final_cell_annotation", "spe_final_celltype.rds")
```

## 3. Load Final Annotated SPE

```{r load-data}
spe <- readRDS(spe_path)

print(table(spe$final_celltype))
```

## 3b. Check Assay Scales

`scaled_0100` is z-scored despite the name (confirmed: range -2.23 to
6.08, mean ~0.097, negative values present) -- see assay inspection
below.

```{r check-assays}
print(assayNames(spe))

for (a in assayNames(spe)) {
  vals <- assay(spe, a)
  cat("\nAssay:", a, "\n")
  cat("  range:", paste(round(range(vals), 2), collapse = " to "), "\n")
  cat("  mean:", round(mean(vals), 3), "\n")
  cat("  has negative values:", any(vals < 0), "\n")
}
```

## 4. Group Final Cell Types into Broad Categories

```{r categories}
celltype_category <- c(
  "B Cell (CD19+)"                                       = "B Cell/Plasma Cell",
  "B Cell, Weak (CD19+)"                                 = "B Cell/Plasma Cell",
  "B Cell/Plasma Cell, HLADRB-high (CD19+CD138+HLADRB+)" = "B Cell/Plasma Cell",
  "Plasma Cell, Memory-like (Weak) (CD45RO+)"            = "B Cell/Plasma Cell",
  "Plasma Cell, TIM3+ (CD138+TIM3+CD45RO-)"              = "B Cell/Plasma Cell",
  "Plasmablast, Strongly Activated (CD38++)"             = "B Cell/Plasma Cell",
  "Plasmablast/Activated B Cell (CD38+)"                 = "B Cell/Plasma Cell",
  "Transitional B Cell/Plasmablast (CD19+CD138+HLADRB+)" = "B Cell/Plasma Cell",

  "CD4+ T Cell"                                          = "T Cell",
  "CD8+ T Cell"                                          = "T Cell",
  "Memory T Cell"                                        = "T Cell",
  "Memory T Cell (CD45RO+)"                              = "T Cell",
  "Memory/Activated T Cell (CD4++)"                      = "T Cell",
  "T Cell (TIM3+)"                                       = "T Cell",
  "T Cell, Exhaustion-leaning (PD1++)"                   = "T Cell",
  "Treg"                                                 = "T Cell",
  "Treg (FoxP3+)"                                        = "T Cell",
  "T Cell/DC Admixture (CD11c+)"                         = "T Cell",

  "Macrophage (CD68+)"                                   = "Myeloid",
  "Macrophage, Antigen-Presenting (HLADRB+)"             = "Myeloid",
  "Macrophage, M2-like (CD163+)"                         = "Myeloid",
  "Macrophage, Mixed Signal (PDL1+)"                     = "Myeloid",
  "Macrophage, Proliferating (HLADRB+)"                  = "Myeloid",
  "Macrophage, Proliferating (MKI67+)"                   = "Myeloid",
  "Monocyte (Generic)"                                   = "Myeloid",
  "Monocyte (PD1+)"                                      = "Myeloid",
  "Monocyte-Macrophage Transitional (CD14+)"             = "Myeloid",
  "Monocyte-Macrophage, Unresolved"                      = "Myeloid",
  "Monocyte, Unresolved"                                 = "Myeloid",
  "cDC1"                                                 = "Myeloid",
  "DC/mregDC"                                            = "Myeloid",
  "Dendritic Cell (CD11c+)"                              = "Myeloid",
  "PMN-MDSC"                                             = "Myeloid",

  "Epithelial (EpCAM+)"                                  = "Epithelial/Stroma",
  "Fibroblast/Stroma"                                    = "Epithelial/Stroma",
  "Tumor Epithelial (Neuroendocrine)"                    = "Epithelial/Stroma",
  "Tumor Epithelial (TTF1++)"                            = "Epithelial/Stroma",

  "Mast Cell"                                            = "Other",
  "Proliferating Cells"                                  = "Other"
)

spe$celltype_category <- celltype_category[as.character(spe$final_celltype)]
```

Flag anything that didn't map:

```{r check-unmapped}
unmapped <- setdiff(unique(as.character(spe$final_celltype)), names(celltype_category))
if (length(unmapped) > 0) {
  cat("Unmapped final_celltype labels -- add these to celltype_category:\n")
  print(unmapped)
} else {
  cat("All final_celltype labels mapped to a category.\n")
}
```

## 5. Cell Counts by Category and Final Cell Type

```{r counts-category}
cat("Cell count by broad category:\n")
print(table(spe$celltype_category, useNA = "ifany"))
```

## 6. Panel Markers (Excluding DNA/Membrane)

```{r markers}
exclude_channels <- c("DNA191", "DNA193", "Membrane")
panel_markers <- setdiff(rownames(spe), exclude_channels)

cat("Markers used for violin plots:", paste(panel_markers, collapse = ", "), "\n")
```

## 7. Violin Plot Function (One Plot Per Marker)

```{r make-violinplot}
make_violinplot_single_marker <- function(spe, assay_name, cluster_col, marker, title) {
  expr_vec   <- assay(spe, assay_name)[marker, ]
  cluster_id <- factor(spe[[cluster_col]])

  plot_df <- data.frame(cluster = cluster_id, expr = expr_vec)

  ggplot(plot_df, aes(x = cluster, y = expr, fill = cluster)) +
    geom_violin(scale = "width", trim = TRUE, linewidth = 0.2) +
    stat_summary(fun = median, geom = "point", size = 0.8, color = "black") +
    labs(title = title, x = "Cell Type", y = "Z-score expression") +
    theme_bw(base_size = 11) +
    theme(
      axis.text.x       = element_text(angle = 45, hjust = 1, size = 9),
      panel.grid.major.x = element_blank(),
      legend.position    = "none"
    )
}
```

## 8. Violin Plots per Category (One Plot per Marker)

### B Cell/Plasma Cell

```{r violin-bcell, fig.width=10, fig.height=5, results='asis'}
spe_bcell_group <- spe[, spe$celltype_category == "B Cell/Plasma Cell"]
spe_bcell_group$final_celltype <- droplevels(factor(spe_bcell_group$final_celltype))

for (m in panel_markers) {
  p <- make_violinplot_single_marker(spe_bcell_group, "scaled_0100", "final_celltype", m,
                                     paste("B Cell/Plasma Cell:", m))
  print(p)
}
```

### B Cell/Plasma Cell: Admixture Signal Shape Check

```{r make-density-plot}
make_density_plot <- function(spe, assay_name, cluster_col, cluster_label, marker, title) {
  expr_vec   <- assay(spe, assay_name)[marker, ]
  cluster_id <- spe[[cluster_col]]

  idx <- which(cluster_id == cluster_label)
  plot_df <- data.frame(expr = expr_vec[idx])

  ggplot(plot_df, aes(x = expr)) +
    geom_histogram(aes(y = after_stat(density)), bins = 60, fill = "steelblue", alpha = 0.6) +
    geom_density(color = "firebrick", linewidth = 0.8) +
    labs(title = title, x = "Z-score expression", y = "Density") +
    theme_bw(base_size = 11)
}
```

```{r density-tim3-cluster, fig.width=7, fig.height=4}
tim3_label <- "Plasma Cell, TIM3+ (CD138+TIM3+CD45RO-)"

p_tbet_density <- make_density_plot(spe_bcell_group, "scaled_0100", "final_celltype",
                                    tim3_label, "TBet",
                                    "TIM3+ Cluster: TBet Density")
print(p_tbet_density)

p_cd56_density <- make_density_plot(spe_bcell_group, "scaled_0100", "final_celltype",
                                    tim3_label, "CD56",
                                    "TIM3+ Cluster: CD56 Density")
print(p_cd56_density)

p_cd45ra_density <- make_density_plot(spe_bcell_group, "scaled_0100", "final_celltype",
                                      tim3_label, "CD45RA",
                                      "TIM3+ Cluster: CD45RA Density")
print(p_cd45ra_density)
```

```{r density-activated-cluster, fig.width=7, fig.height=4}
activated_label <- "Plasmablast, Strongly Activated (CD38++)"

p_cd68_density <- make_density_plot(spe_bcell_group, "scaled_0100", "final_celltype",
                                    activated_label, "CD68",
                                    "Strongly Activated Cluster: CD68 Density")
print(p_cd68_density)

p_cd14_density <- make_density_plot(spe_bcell_group, "scaled_0100", "final_celltype",
                                    activated_label, "CD14",
                                    "Strongly Activated Cluster: CD14 Density")
print(p_cd14_density)
```

### T Cell

```{r violin-tcell, fig.width=10, fig.height=5, results='asis'}
spe_tcell_group <- spe[, spe$celltype_category == "T Cell"]
spe_tcell_group$final_celltype <- droplevels(factor(spe_tcell_group$final_celltype))

for (m in panel_markers) {
  p <- make_violinplot_single_marker(spe_tcell_group, "scaled_0100", "final_celltype", m,
                                     paste("T Cell:", m))
  print(p)
}
```

### T Cell: Memory T Cell Comparison and CD138 Cross-Category Check

```{r heatmap-memory-tcell-comparison, fig.width=6, fig.height=8}
memory_labels <- c("Memory T Cell", "Memory T Cell (CD45RO+)")

spe_memory_compare <- spe_tcell_group[, spe_tcell_group$final_celltype %in% memory_labels]
spe_memory_compare$final_celltype <- droplevels(factor(spe_memory_compare$final_celltype))

mean_expr_memory <- sapply(levels(spe_memory_compare$final_celltype), function(cl) {
  idx <- which(spe_memory_compare$final_celltype == cl)
  rowMeans(assay(spe_memory_compare, "scaled_0100")[panel_markers, idx, drop = FALSE])
})

marker_order_memory <- panel_markers[hclust(dist(mean_expr_memory))$order]

heatmap_memory_df <- as.data.frame(mean_expr_memory) |>
  rownames_to_column("marker") |>
  pivot_longer(-marker, names_to = "cell_type", values_to = "mean_expr")

heatmap_memory_df$marker <- factor(heatmap_memory_df$marker, levels = marker_order_memory)

p_heatmap_memory <- ggplot(heatmap_memory_df, aes(x = cell_type, y = marker, fill = mean_expr)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(mean_expr, 2)), size = 3) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, name = "Mean\nZ-score") +
  labs(title = "Memory T Cell vs. Memory T Cell (CD45RO+): Full Panel Comparison",
       x = "Cell Type", y = "Marker") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())
print(p_heatmap_memory)
```

```{r cd138-cross-category-check, fig.width=10, fig.height=5}
cd138_by_category <- data.frame(
  category = spe$celltype_category,
  cd138    = assay(spe, "scaled_0100")["CD138", ]
)

p_cd138_category <- ggplot(cd138_by_category, aes(x = category, y = cd138, fill = category)) +
  geom_violin(scale = "width", trim = TRUE, linewidth = 0.2) +
  stat_summary(fun = median, geom = "point", size = 0.8, color = "black") +
  labs(title = "CD138 by Broad Category (Cross-Category Spillover Check)",
       x = "Category", y = "Z-score expression") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major.x = element_blank(),
        legend.position = "none")
print(p_cd138_category)
```

### Myeloid

```{r violin-myeloid, fig.width=10, fig.height=5, results='asis'}
spe_myeloid_group <- spe[, spe$celltype_category == "Myeloid"]
spe_myeloid_group$final_celltype <- droplevels(factor(spe_myeloid_group$final_celltype))

for (m in panel_markers) {
  p <- make_violinplot_single_marker(spe_myeloid_group, "scaled_0100", "final_celltype", m,
                                     paste("Myeloid:", m))
  print(p)
}
```

### Myeloid: Macrophage, Proliferating (MKI67+) Diagnostic

Same diagnostic used previously for the k150 "Unknown" cluster artifact
-- total counts, cell area, and patient/ROI origin -- applied here to
check whether the broad off-lineage marker elevation in this cluster
reflects a technical artifact, spatial spillover, or real biology
layered with background noise.

```{r prolif-setup}
prolif_label <- "Macrophage, Proliferating (MKI67+)"

idx_prolif <- which(spe_myeloid_group$final_celltype == prolif_label)

cat("Number of cells in", prolif_label, ":", length(idx_prolif), "\n")
```

```{r prolif-total-counts, fig.width=7, fig.height=5}
total_counts_all     <- colSums(assay(spe_myeloid_group, "counts"))
total_counts_prolif  <- total_counts_all[idx_prolif]
total_counts_other   <- total_counts_all[-idx_prolif]

cat("Total counts -- Proliferating cluster:\n")
print(summary(total_counts_prolif))

cat("\nTotal counts -- all other myeloid cells:\n")
print(summary(total_counts_other))

count_compare_df <- data.frame(
  total_counts = c(total_counts_prolif, total_counts_other),
  group = c(rep("Proliferating (MKI67+)", length(total_counts_prolif)),
            rep("Other Myeloid", length(total_counts_other)))
)

p_counts_compare <- ggplot(count_compare_df, aes(x = group, y = total_counts, fill = group)) +
  geom_violin(scale = "width", trim = TRUE) +
  stat_summary(fun = median, geom = "point", size = 1, color = "black") +
  labs(title = "Total Counts: Proliferating Cluster vs. Other Myeloid",
       x = "", y = "Total Counts per Cell") +
  theme_bw(base_size = 11) +
  theme(legend.position = "none")
print(p_counts_compare)
```

```{r prolif-check-coldata-columns}
# Check available colData columns to confirm the correct area/ROI column names below
print(colnames(colData(spe_myeloid_group)))
```

```{r prolif-cell-area, fig.width=7, fig.height=5}
area_col <- "area"  # <-- update if colData uses a different column name

if (area_col %in% colnames(colData(spe_myeloid_group))) {
  cell_area_all    <- colData(spe_myeloid_group)[[area_col]]
  cell_area_prolif <- cell_area_all[idx_prolif]
  cell_area_other  <- cell_area_all[-idx_prolif]

  cat("Cell area -- Proliferating cluster:\n")
  print(summary(cell_area_prolif))

  cat("\nCell area -- all other myeloid cells:\n")
  print(summary(cell_area_other))

  area_compare_df <- data.frame(
    area  = c(cell_area_prolif, cell_area_other),
    group = c(rep("Proliferating (MKI67+)", length(cell_area_prolif)),
              rep("Other Myeloid", length(cell_area_other)))
  )

  p_area_compare <- ggplot(area_compare_df, aes(x = group, y = area, fill = group)) +
    geom_violin(scale = "width", trim = TRUE) +
    stat_summary(fun = median, geom = "point", size = 1, color = "black") +
    labs(title = "Cell Area: Proliferating Cluster vs. Other Myeloid",
         x = "", y = "Cell Area") +
    theme_bw(base_size = 11) +
    theme(legend.position = "none")
  print(p_area_compare)
} else {
  cat("Column '", area_col, "' not found in colData -- check colnames(colData(spe)) output above and update area_col.\n")
}
```

```{r prolif-patient-origin}
patient_counts_prolif <- table(spe_myeloid_group$patient_id[idx_prolif])
patient_totals        <- table(spe_myeloid_group$patient_id)

prolif_origin <- data.frame(
  patient              = names(patient_counts_prolif),
  prolif_cell_count    = as.integer(patient_counts_prolif),
  pct_of_prolif_cluster = round(100 * as.integer(patient_counts_prolif) / sum(patient_counts_prolif), 1),
  pct_of_patient_total  = round(100 * as.integer(patient_counts_prolif) /
                                 as.integer(patient_totals[names(patient_counts_prolif)]), 1)
)
prolif_origin <- prolif_origin[order(-prolif_origin$pct_of_prolif_cluster), ]

cat("Patient origin of Macrophage, Proliferating (MKI67+) cluster:\n")
print(prolif_origin)
```

```{r prolif-roi-origin}
roi_col <- "sample_id"

if (roi_col %in% colnames(colData(spe_myeloid_group))) {
  roi_counts_prolif <- table(colData(spe_myeloid_group)[[roi_col]][idx_prolif])
  roi_totals        <- table(colData(spe_myeloid_group)[[roi_col]])

  prolif_roi_origin <- data.frame(
    roi                  = names(roi_counts_prolif),
    prolif_cell_count    = as.integer(roi_counts_prolif),
    pct_of_prolif_cluster = round(100 * as.integer(roi_counts_prolif) / sum(roi_counts_prolif), 1),
    pct_of_roi_total     = round(100 * as.integer(roi_counts_prolif) /
                                  as.integer(roi_totals[names(roi_counts_prolif)]), 1)
  )
  prolif_roi_origin <- prolif_roi_origin[order(-prolif_roi_origin$pct_of_prolif_cluster), ]

  cat("ROI origin of Macrophage, Proliferating (MKI67+) cluster:\n")
  print(prolif_roi_origin)
} else {
  cat("Column '", roi_col, "' not found in colData -- check colnames(colData(spe)) output above and update roi_col.\n")
}
```

```{r prolif-density-check, fig.width=7, fig.height=4}
p_rorgt_density <- make_density_plot(spe_myeloid_group, "scaled_0100", "final_celltype",
                                     prolif_label, "RORgT",
                                     "Proliferating Cluster: RORgT Density")
print(p_rorgt_density)

p_ccr7_density <- make_density_plot(spe_myeloid_group, "scaled_0100", "final_celltype",
                                    prolif_label, "CCR7",
                                    "Proliferating Cluster: CCR7 Density")
print(p_ccr7_density)

p_cd127_density <- make_density_plot(spe_myeloid_group, "scaled_0100", "final_celltype",
                                     prolif_label, "CD127",
                                     "Proliferating Cluster: CD127 Density")
print(p_cd127_density)
```

### Epithelial/Stroma

```{r violin-epi-stroma, fig.width=10, fig.height=5, results='asis'}
spe_epi_group <- spe[, spe$celltype_category == "Epithelial/Stroma"]
spe_epi_group$final_celltype <- droplevels(factor(spe_epi_group$final_celltype))

for (m in panel_markers) {
  p <- make_violinplot_single_marker(spe_epi_group, "scaled_0100", "final_celltype", m,
                                     paste("Epithelial/Stroma:", m))
  print(p)
}
```

```{r}
## Optional Check: Tumor Epithelial (TTF1++) Patient/ROI Distribution

tumor_ttf1_label <- "Tumor Epithelial (TTF1++)"

idx_tumor_ttf1 <- which(spe_epi_group$final_celltype == tumor_ttf1_label)

patient_counts_ttf1 <- table(spe_epi_group$patient_id[idx_tumor_ttf1])
patient_totals_epi  <- table(spe_epi_group$patient_id)

ttf1_origin <- data.frame(
  patient               = names(patient_counts_ttf1),
  ttf1_cell_count       = as.integer(patient_counts_ttf1),
  pct_of_ttf1_cluster   = round(100 * as.integer(patient_counts_ttf1) / sum(patient_counts_ttf1), 1),
  pct_of_patient_total  = round(100 * as.integer(patient_counts_ttf1) /
                                 as.integer(patient_totals_epi[names(patient_counts_ttf1)]), 1)
)
ttf1_origin <- ttf1_origin[order(-ttf1_origin$pct_of_ttf1_cluster), ]

cat("Patient origin of Tumor Epithelial (TTF1++) cluster:\n")
print(ttf1_origin)
```

### Other

```{r violin-other, fig.width=10, fig.height=5, results='asis'}
spe_other_group <- spe[, spe$celltype_category == "Other"]
spe_other_group$final_celltype <- droplevels(factor(spe_other_group$final_celltype))

for (m in panel_markers) {
  p <- make_violinplot_single_marker(spe_other_group, "scaled_0100", "final_celltype", m,
                                     paste("Other:", m))
  print(p)
}
```
