---
title: 'T Cell Subclustering: Combined Cluster 15 (Cleaned) + Clusters 5, 6, 7 (Round 2)'
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

This notebook combines cluster 15 (CD4+ T cell/Treg population) with the other 
broad T cell clusters(5 = Treg, 6 = CD8+ T Cell, 7 = Memory T Cell) for a joint 
subclustering pass, using k=20 Louvain.

Cluster 15 contributes only its already-cleaned cells (excluding
subclusters 2, 4, 7 (low quality/Patient 20 driven) and 8 (confirmed
Patient20 CD19 technical artifact), rather than the raw broad cluster,
since those issues are already diagnosed and there's no reason to
reintroduce them. 

This is a round-2 pass following PI review of the initial subclustering
(`tcell_subclustering.Rmd`): adds a CD45RA violin panel, remakes the
marker heatmap with the full marker panel labeled by annotated subcluster
name rather than numeric ID, drops the intensity-confound check, and
updates subcluster naming per PI feedback (resting -> naive;
CD45RO-elevated -> memory rather than activated).

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# 1. Load Libraries

```{r libraries}
.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(imcRtools)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(scater)
library(scran)
library(bluster)
library(patchwork)
library(RColorBrewer)
library(ComplexHeatmap)
library(circlize)
```

# 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

spe_path <- file.path(analysis_path, "final_cell_annotation", "final_broad_cluster", "h5_cluster_annot.rds")

cluster15_clean_path <- file.path(analysis_path, "final_cell_annotation", "subclustering",
                                   "cluster15", "cluster15_subcluster_clean.rds")

tcell_clustering_path <- file.path(analysis_path, "final_cell_annotation", "subclustering", "tcells")
out_path <- file.path(tcell_clustering_path, "tcell_subclustering.rds")
findmarkers_path <- file.path(tcell_clustering_path, "tcell_subclustering_findmarkers.csv")
```

# 3. Load Data and Assemble Combined Cell Set

```{r load-and-assemble}
spe <- readRDS(spe_path)
cluster15_clean <- readRDS(cluster15_clean_path)

# Cluster 15's already-cleaned cells
cluster15_clean_ids <- colnames(cluster15_clean)
cat("Cluster 15 (cleaned) cells:", length(cluster15_clean_ids), "\n")

# Full, unsubclustered cells from the other broad T cell clusters
tcell_broad_clusters <- c("5", "6", "7")  # Treg, CD8+ T Cell, Memory T Cell
other_tcell_ids <- colnames(spe)[as.character(spe$broad_clusters) %in% tcell_broad_clusters]
cat("Cells from broad clusters 5/6/7:", length(other_tcell_ids), "\n")

for (cl in tcell_broad_clusters) {
  n <- sum(as.character(spe$broad_clusters) == cl)
  cat("  Cluster", cl, ":", n, "cells\n")
}

combined_ids <- union(cluster15_clean_ids, other_tcell_ids)
cat("\nTotal combined T cell subset:", length(combined_ids), "\n")

spe_tcell <- spe[, combined_ids]
```

```{r patient-cell-counts, fig.width=10, fig.height=6}
patient_counts <- as.data.frame(table(patient_id = spe_tcell$patient_id))

p_patient_comp <- ggplot(patient_counts, aes(x = patient_id, y = Freq, fill = patient_id)) +
  geom_col() +
  labs(title = "T Cell Subset: Cell Count by Patient",
       x = "Patient", y = "Cell Count") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
print(p_patient_comp)

cat("T cell subset composition by patient:\n")
print(patient_counts[order(-patient_counts$Freq), ])
```

# 4. Recluster: Louvain on Existing Harmony Embedding (k = 20)

```{r recluster}
if (file.exists(out_path)) {
  cat("Output already exists, loading instead of reclustering:", out_path, "\n")
  spe_tcell <- readRDS(out_path)
} else {
  harmony_tcell <- reducedDim(spe_tcell, "Harmony")

  set.seed(220225)
  snn_graph <- buildSNNGraph(spe_tcell, use.dimred = "Harmony", k = 20,
                              BPPARAM = BiocParallel::SerialParam())
  clusters <- igraph::cluster_louvain(snn_graph)$membership
  spe_tcell$tcell_subclusters <- factor(clusters)
}

cat("Number of subclusters:", nlevels(spe_tcell$tcell_subclusters), "\n")
print(table(spe_tcell$tcell_subclusters))
```

# 5. UMAP on the Subset

```{r umap}
if ("UMAP_sub" %in% reducedDimNames(spe_tcell)) {
  cat("UMAP_sub already exists, skipping recomputation.\n")
} else {
  set.seed(220225)
  spe_tcell <- runUMAP(spe_tcell, dimred = "Harmony", name = "UMAP_sub")
}

p_sub_original <- dittoDimPlot(spe_tcell, var = "broad_celltype",
                                reduction.use = "UMAP_sub",
                                size = 0.4) +
  ggtitle("T Cell Subset: UMAP colored by Original Broad Label") +
  theme(legend.position = "right")
print(p_sub_original)
```

```{r umap-subclusters, fig.width=10, fig.height=8}
p_sub_clusters <- dittoDimPlot(spe_tcell, var = "tcell_subclusters",
                                reduction.use = "UMAP_sub",
                                size = 0.4, do.label = TRUE, labels.size = 4) +
  ggtitle("T Cell Subset: UMAP colored by Subcluster") +
  theme(legend.position = "right")
print(p_sub_clusters)
```

# 6. Marker Sets for T Cell Subtyping

```{r markers}
lineage_markers <- c("CD4", "CD8a", "FoxP3")
state_markers <- c("CD45RA", "CD45RO", "MKI67", "PD1", "PDL1", "TIM3", "HLADRB", "CD127", "TIGIT", "CTLA4")

# Used to catch contamination/spillover
admixture_markers <- c("CD19", "CD138", "CD11c", "CD14", "CD1c", "CD68")

tcell_markers <- c(lineage_markers, state_markers, admixture_markers)
tcell_markers <- tcell_markers[tcell_markers %in% rownames(spe_tcell)]
cat("Markers found:", paste(tcell_markers, collapse = ", "), "\n")
```

# 7. Manual Subcluster Annotation

Naive: T cell that has not yet encountered its cognate antigen.
Characterized by low levels of activation, proliferation, and CD45RO
(memory) markers.

Activated: T cell that has recognized its antigen or received other
activating signals and is actively engaged -- proliferating, producing
cytokines, and upregulating markers that support its immune function.

Memory: T cell that has previously encountered antigen and persists in a
poised, CD45RO+ state that allows faster reactivation upon re-exposure --
distinct from actively engaged (Activated) or naive (CD45RO-) cells.

```{r annotate}
tcell_subannotation <- c(
  "1" = "CD4+ Memory T Cell, Exhaustion-leaning", # CD45RO+, PD1+/TIGIT+, mild CTLA4 and LAG3 positive (exhausion signature)
  "2" = "CD4+ Memory T Cell, Partial-exhausion",  # CD45RO+, CD4+, TIM3/CD138 co-elevation attributed to intensity confound
  "3" = "Treg, Memory",  # FoxP3+, CD4 and CD45RO co-elevated
  "4" = "Memory T Cell (Unresolved)",  # CD45RO+, negative for both CD4 and CD8a which is weird
  "5" = "CD8+ Memory T Cell",  # CD45RO and CD138 also elevated
  "6" = "Treg, Naive",  # FoxP3+, everything else strongly negative
  "7" = "CD8+ Naive T Cell" # CD8a sole positive marker, everything else negative
)

if (!any(tcell_subannotation == "PLACEHOLDER") &&
    length(tcell_subannotation) == nlevels(spe_tcell$tcell_subclusters)) {
  spe_tcell$sub_celltype <- tcell_subannotation[as.character(spe_tcell$tcell_subclusters)]
  spe_tcell$sub_celltype <- factor(spe_tcell$sub_celltype, levels = unique(tcell_subannotation))
  cat("Subcluster annotation applied.\n")
  print(table(spe_tcell$sub_celltype, useNA = "ifany"))
} else {
  cat("Fill in tcell_subannotation above before proceeding",
      "-- check length matches nlevels(spe_tcell$tcell_subclusters).\n")
}
```

```{r umap-celltype, fig.width=10, fig.height=8}
if ("sub_celltype" %in% colnames(colData(spe_tcell))) {
  p_sub_celltype <- dittoDimPlot(spe_tcell, var = "sub_celltype",
                                  reduction.use = "UMAP_sub",
                                  size = 0.4, do.label = TRUE, labels.size = 3.5) +
    ggtitle("T Cell Subset: UMAP colored by Subcluster Cell Type") +
    theme(legend.position = "right")
  print(p_sub_celltype)
} else {
  cat("Run the annotate chunk first.\n")
}
```

# 8. Violin Plots per Subcluster

```{r violin-fn, fig.width=14, fig.height=6}
make_violinplot <- function(spe, assay_name, cluster_col, marker_set, title, ncol = 3) {
  expr_mat  <- assay(spe, assay_name)[marker_set, , drop = FALSE]
  cluster_id <- spe[[cluster_col]]

  mean_expr <- sapply(levels(cluster_id), function(cl) {
    idx <- which(cluster_id == cl)
    rowMeans(expr_mat[, idx, drop = FALSE])
  })

  if (is.matrix(mean_expr) && nrow(mean_expr) > 1) {
    marker_order <- rownames(mean_expr)[hclust(dist(mean_expr))$order]
  } else {
    marker_order <- marker_set
  }

  plot_df <- as.data.frame(t(expr_mat)) |>
    mutate(cluster = cluster_id) |>
    pivot_longer(-cluster, names_to = "marker", values_to = "expr")

  plot_df$marker  <- factor(plot_df$marker, levels = marker_order)
  plot_df$cluster <- factor(plot_df$cluster, levels = levels(cluster_id))

  ggplot(plot_df, aes(x = cluster, y = expr, fill = cluster)) +
    geom_violin(scale = "width", trim = TRUE, linewidth = 0.2) +
    stat_summary(fun = median, geom = "point", size = 0.6, color = "black") +
    facet_wrap(~ marker, scales = "free_y", ncol = ncol) +
    labs(title = title, x = "Subcluster", y = "Z-score expression") +
    theme_bw(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 0, hjust = 0.5, size = 9),
      strip.text = element_text(face = "bold"),
      panel.grid.major.x = element_blank(),
      legend.position = "none"
    )
}

p_violin_lineage <- make_violinplot(spe_tcell, "scaled_0100", "tcell_subclusters",
                                     lineage_markers,
                                     "T Cell Subset: Lineage Markers per Subcluster",
                                     ncol = 3)
print(p_violin_lineage)
```

```{r violin-state, fig.width=12, fig.height=10}
p_violin_state <- make_violinplot(spe_tcell, "scaled_0100", "tcell_subclusters",
                                   state_markers,
                                   "T Cell Subset: State/Checkpoint Markers per Subcluster",
                                   ncol = 3)
print(p_violin_state)
```

```{r violin-admixture, fig.width=12, fig.height=8}
p_violin_admixture <- make_violinplot(spe_tcell, "scaled_0100", "tcell_subclusters",
                                       admixture_markers,
                                       "T Cell Subset: Admixture/Spillover Markers per Subcluster",
                                       ncol = 3)
print(p_violin_admixture)
```

# 9. Heatmap: All Markers, Labeled by Subcluster Name

Remade from the first-pass heatmap to (1) use the full marker panel rather
than just the curated lineage/state/admixture subset, and (2) label
columns by annotated subcluster name (Section 7) instead of numeric ID.

```{r heatmap-all-named, fig.width=12, fig.height=14}
all_panel_markers <- rownames(spe_tcell)

celltype_levels <- levels(spe_tcell$sub_celltype)

mean_expr_all <- sapply(celltype_levels, function(ct) {
  idx <- which(spe_tcell$sub_celltype == ct)
  rowMeans(assay(spe_tcell, "scaled_0100")[all_panel_markers, idx, drop = FALSE])
})

marker_order <- all_panel_markers[hclust(dist(mean_expr_all))$order]

celltype_sizes <- table(spe_tcell$sub_celltype)
celltype_labels <- paste0(celltype_levels, "\n(n=", celltype_sizes[celltype_levels], ")")
names(celltype_labels) <- celltype_levels

heatmap_df <- as.data.frame(mean_expr_all) |>
  rownames_to_column("marker") |>
  pivot_longer(-marker, names_to = "sub_celltype", values_to = "mean_expr")

heatmap_df$marker       <- factor(heatmap_df$marker, levels = marker_order)
heatmap_df$sub_celltype <- factor(heatmap_df$sub_celltype, levels = celltype_levels)

p_heatmap_named <- ggplot(heatmap_df, aes(x = sub_celltype, y = marker, fill = mean_expr)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(mean_expr, 2)), size = 2.5) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, name = "Mean\nZ-score") +
  scale_x_discrete(labels = celltype_labels) +
  labs(title = "Mean Marker Expression: All Markers, Annotated T Cell Subclusters",
       x = "Subcluster (cell count)", y = "Marker") +
  theme_bw(base_size = 10) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10))
print(p_heatmap_named)
```

# 10. Statistical Marker Discovery (findMarkers)

```{r findmarkers}
markers_de <- findMarkers(spe_tcell,
                           groups = spe_tcell$tcell_subclusters,
                           assay.type = "scaled_0100",
                           subset.row = tcell_markers)

all_markers_df <- do.call(rbind, lapply(levels(spe_tcell$tcell_subclusters), function(cl) {
  df <- as.data.frame(markers_de[[cl]])
  data.frame(
    subcluster = cl,
    marker     = rownames(df),
    Top        = df$Top,
    FDR        = df$FDR,
    logFC      = df$summary.logFC
  )
}))

#dir.create(dirname(findmarkers_path), recursive = TRUE, showWarnings = FALSE)
#write.csv(all_markers_df, findmarkers_path, row.names = FALSE)

top1_summary <- do.call(rbind, lapply(levels(spe_tcell$tcell_subclusters), function(cl) {
  sub <- all_markers_df[all_markers_df$subcluster == cl & all_markers_df$logFC > 1.0, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  if (nrow(sub) == 0) {
    data.frame(subcluster = cl, top_markers = "none above logFC 1.0")
  } else {
    data.frame(subcluster = cl, top_markers = paste(head(sub$marker, 5), collapse = ", "))
  }
}))

cat("Top positive markers (logFC > 1.0) per subcluster:\n")
knitr::kable(top1_summary, col.names = c("Subcluster", "Top Positive Markers"))
```

```{r findmarkers-per-subcluster-tables}
cat("Detailed tables:\n\n")
for (cl in levels(spe_tcell$tcell_subclusters)) {
  sub <- all_markers_df[all_markers_df$subcluster == cl & all_markers_df$Top <= 5, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  if (nrow(sub) == 0) {
    cat("Subcluster", cl, ": no markers in top 5\n\n")
  } else {
    print(knitr::kable(sub[, c("marker", "Top", "FDR", "logFC")],
                       caption = paste("Subcluster", cl),
                       digits  = 3, row.names = FALSE))
  }
}
```

# 11. Quality/Origin Checks

Following the same diagnostic pattern used for cluster 15 and the B cell
subclustering -- check total counts, cell area, and patient composition for
any subcluster that looks flat/low-signal in the heatmap above (update
`check_clusters` once you've looked at chunks 8-9).

```{r quality-checks}
check_clusters <- c("4")  # large, flat CD45RO-only signature -- same profile as prior artifacts

spe_tcell$total_counts <- colSums(assay(spe_tcell, "counts"))

quality_check <- colData(spe_tcell) |>
  as.data.frame() |>
  mutate(group = ifelse(tcell_subclusters %in% check_clusters,
                         "Flagged subcluster(s)", "Other subclusters")) |>
  group_by(group) |>
  summarise(
    n = n(),
    mean_area = mean(area, na.rm = TRUE),
    median_area = median(area, na.rm = TRUE),
    mean_total_counts = mean(total_counts, na.rm = TRUE),
    median_total_counts = median(total_counts, na.rm = TRUE)
  )
print(quality_check)
```
Mean area is nearly identical and total counts are only modestly lower (about 10% less). So this looks like a genuine (lower signal) resting memory T cell population, not a staining artifact. 

## Subcluster 4: Lineage Ambiguity Check

```{r lineage-ambiguity}

# CD4/CD8a/FoxP3 are all near-zero for subcluster 4 despite elevated CD45RO,
# with CD1c (admixture marker) also elevated -- checking whether this is a
# genuine double-negative memory population or myeloid contamination/artifact
sub4_mask <- spe_tcell$tcell_subclusters == "4"

lineage_check_markers <- c("CD4", "CD8a", "FoxP3", "CCR7", "CD1c", "CD11c", "CD68")
lineage_check_markers <- lineage_check_markers[lineage_check_markers %in% rownames(spe_tcell)]

sub4_expr <- rowMeans(assay(spe_tcell, "scaled_0100")[lineage_check_markers, sub4_mask, drop = FALSE])
cat("Subcluster 4 -- mean expression (lineage + admixture markers):\n")
print(round(sub4_expr, 2))

# Same area/total-counts comparison used for the earlier flagged cluster
sub4_quality <- colData(spe_tcell) |>
  as.data.frame() |>
  mutate(group = ifelse(tcell_subclusters == "4", "Subcluster 4", "Other subclusters")) |>
  group_by(group) |>
  summarise(
    n = n(),
    mean_area = mean(area, na.rm = TRUE),
    median_area = median(area, na.rm = TRUE),
    mean_total_counts = mean(total_counts, na.rm = TRUE),
    median_total_counts = median(total_counts, na.rm = TRUE)
  )
cat("\nSubcluster 4 vs. rest -- area/total counts:\n")
print(sub4_quality)

```
Subcluster 4 is likely a genuine biological population, not an artifact. It's negative for both CD4 and CD8a, which is an uncommon category of T cells.

## Patient Composition heatmap

```{r patient-composition, fig.width=10, fig.height=6}
comp_patient <- table(subcluster = spe_tcell$tcell_subclusters,
                       patient    = spe_tcell$patient_id)
comp_patient_props <- prop.table(comp_patient, margin = 1)

col_fun_patient <- colorRamp2(c(0, 0.5, 1), rev(brewer.pal(3, "RdBu")))

ht_patient <- Heatmap(
  comp_patient_props,
  name = "Proportion",
  col = col_fun_patient,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  column_title = "Patient",
  column_title_side = "bottom",
  row_title = "Subcluster",
  column_names_rot = 45,
  cell_fun = function(j, i, x, y, width, height, fill) {
    val <- comp_patient_props[i, j]
    text_col <- if (val < 0.2 || val > 0.8) "white" else "black"
    grid.text(sprintf("%.2f", val), x, y, gp = gpar(col = text_col, fontsize = 7))
  }
)
draw(ht_patient, column_title = "T Cell Subcluster Composition by Patient (row-normalized %)")
```

# 12. Save Subclustered/Annotated Object

```{r save}
dir.create(dirname(out_path), recursive = TRUE, showWarnings = FALSE)
saveRDS(spe_tcell, out_path)
cat("Saved T cell subcluster object to:", out_path, "\n")
```

# Notes

- Combined input: cluster 15's already-cleaned cells (excludes subclusters
  2, 4, 7 -- low quality/artifact -- and 8 -- Patient20 CD19 artifact) +
  full cell sets from broad clusters 5 (Treg), 6 (CD8+ T Cell), 7 (Memory
  T Cell). Cluster 10 (Memory T/Plasma Cell) intentionally excluded --
  already covered by the B cell/plasma subclustering.
- k=20 used for Louvain clustering per team convention (matching the B
  cell/plasma subclustering approach), rather than k=10 used in the
  standalone cluster 15 pass.
- Round-2 changes from PI feedback: added CD45RA to the state/checkpoint
  marker set (Section 6, flows into the violin panel in Section 8);
  remade the marker heatmap (Section 9) with the full marker panel labeled
  by annotated subcluster name rather than numeric ID; dropped the
  intensity-confound correlation check; relabeled subclusters 6/7 from
  "Resting" to "Naive" and subclusters 2/3/5 from "Activated" to "Memory"
  given their CD45RO elevation (Section 7).
- Annotation (Section 7) was moved earlier in this revision, right after
  clustering, so the heatmap and everything downstream can reference
  subcluster names -- in the original first-pass notebook it came after
  the heatmap/findMarkers/quality checks, since those were needed to
  decide the annotation in the first place.
- The CD45RO-elevated -> "Memory" relabeling was applied to subclusters
  2, 3, and 5 based on their existing comments; double check this matches
  intent, since it wasn't given as an explicit 1:1 mapping.
