---
title: "Cell Type Composition Association with Disease State (sccomp)"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

Runs `sccomp` on the finalized cell type annotation to test whether cell
type proportions are associated with disease stage, producing two
comparisons:

1. Lower & Higher vs. Benign (Benign as reference)
2. Higher vs. Lower directly (Lower as reference)

Both are shown as ranked forest plots (effect size + credible interval),
skipping the volcano plot this round since the forest plots proved more
directly interpretable last time.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# 1. Load Libraries

```{r libraries}
.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(imcRtools)
library(tidyverse)
library(SingleCellExperiment)
library(SpatialExperiment)
library(scater)
library(cmdstanr)
library(sccomp)
library(tidytext)
library(kableExtra)
```

# 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

final_celltype_path <- file.path(analysis_path, "final_cell_annotation", "r3_annotation",
                                  "spe_final_celltype_r3.rds")

sccomp_out_dir <- file.path(analysis_path, "celltype_prop_analysis")

sccomp_vs_benign_path   <- file.path(sccomp_out_dir, "sccomp_celltype_vs_benign.rds")
sccomp_higher_v_lower_path <- file.path(sccomp_out_dir, "sccomp_celltype_higher_v_lower.rds")
```

# 3. Load Finalized Annotated Object

```{r load-spe}
spe <- readRDS(final_celltype_path)

cat("Cells:", ncol(spe), "\n")
cat("Cell types:", nlevels(spe$final_celltype), "\n")
```

## 3a. Sanity Check on Updated Annotation

Given the T cell subclustering update, worth confirming the labels look
right before running sccomp: no leftover numbered prefixes, new T cell
subtypes present, no stray low-quality labels slipped through.

```{r sanity-check-annotation}
levels(spe$final_celltype)

cat("\nAny bare-number labels remaining:",
    any(grepl("^[0-9]+$", levels(spe$final_celltype))), "\n")
```

# 4. Build sccomp Input Object (Benign as Reference)

```{r sccomp-prep}
sceIMC <- SingleCellExperiment(list(counts = assay(spe, "counts")))
sceIMC$Patient      <- factor(spe$patient_id)
sceIMC$Sample       <- factor(spe$sample_id)
sceIMC$DiseaseState <- factor(spe$disease_state, levels = c("Benign", "Lower", "Higher"))
sceIMC$CellType     <- factor(spe$final_celltype)

cat("Patients:", n_distinct(sceIMC$Patient), "\n")
cat("Disease states:", paste(levels(sceIMC$DiseaseState), collapse = ", "), "\n")
cat("Cell types:", n_distinct(sceIMC$CellType), "\n")
```

# 5. Run sccomp: Lower & Higher vs. Benign

```{r sccomp-run-vs-benign}
if (file.exists(sccomp_vs_benign_path)) {
  sccomp_vs_benign_res <- readRDS(sccomp_vs_benign_path)
} else {
  sccomp_vs_benign_res <- sceIMC |>
    sccomp_estimate(
      formula_composition = ~ DiseaseState + (1 | Patient),
      .sample     = Sample,
      .cell_group = CellType,
      verbose = FALSE
    ) |>
    sccomp_test()

  saveRDS(sccomp_vs_benign_res, sccomp_vs_benign_path)
}

unique(sccomp_vs_benign_res$parameter)
```

# 6. Run sccomp: Higher vs. Lower (Direct Contrast)

```{r sccomp-run-higher-v-lower}
if (file.exists(sccomp_higher_v_lower_path)) {
  sccomp_higher_v_lower_res <- readRDS(sccomp_higher_v_lower_path)
} else {
  sceIMC_hvl <- sceIMC
  sceIMC_hvl$DiseaseState <- factor(spe$disease_state, levels = c("Lower", "Higher", "Benign"))

  sccomp_higher_v_lower_res <- sceIMC_hvl |>
    sccomp_estimate(
      formula_composition = ~ DiseaseState + (1 | Patient),
      .sample     = Sample,
      .cell_group = CellType,
      verbose = FALSE
    ) |>
    sccomp_test()

  saveRDS(sccomp_higher_v_lower_res, sccomp_higher_v_lower_path)
}

unique(sccomp_higher_v_lower_res$parameter)
```

# 7. Forest Plot: Lower & Higher vs. Benign

Faceted by contrast, sorted by effect size within each facet. FDR<0.05
only; CI-crosses-zero cases are excluded rather than just flagged, given
the earlier finding that FDR alone can pass while the credible interval
still spans zero.

```{r forest-vs-benign, fig.width=10, fig.height=12}
effect_df_benign <- sccomp_vs_benign_res |>
  as.data.frame() |>
  filter(parameter %in% c("DiseaseStateLower", "DiseaseStateHigher")) |>
  filter(c_FDR < 0.05) |>
  filter(!(c_lower < 0 & c_upper > 0)) |>   # drop cases where CI still crosses zero
  mutate(CellType_ordered = reorder_within(CellType, c_effect, parameter))

ggplot(effect_df_benign, aes(x = c_effect, y = CellType_ordered)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = c_lower, xmax = c_upper), height = 0, color = "grey40") +
  geom_point(aes(color = c_effect > 0), size = 2, show.legend = FALSE) +
  scale_color_manual(values = c("TRUE" = "firebrick", "FALSE" = "steelblue")) +
  scale_y_reordered() +
  facet_wrap(~ parameter, scales = "free_y") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 7)) +
  labs(
    x = "Effect size (logit fold-change)",
    y = NULL,
    title = "Cell type composition: Lower & Higher vs. Benign",
    subtitle = "FDR < 0.05, CI excludes zero; red = increases, blue = decreases"
  )
```

```{r save-forest-vs-benign, eval=FALSE}
ggsave(file.path(sccomp_out_dir, "celltype_comp_vs_benign.png"),
       width = 10, height = 12, dpi = 300)
```

# 8. Forest Plot: Higher vs. Lower (Direct Contrast)

```{r forest-higher-v-lower, fig.width=8, fig.height=12}
effect_df_hvl <- sccomp_higher_v_lower_res |>
  as.data.frame() |>
  filter(parameter == "DiseaseStateHigher") |>   # Higher vs. Lower, since Lower is the reference
  filter(c_FDR < 0.05) |>
  filter(!(c_lower < 0 & c_upper > 0)) |>
  mutate(CellType = factor(CellType, levels = CellType[order(c_effect)]))

ggplot(effect_df_hvl, aes(x = c_effect, y = CellType)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = c_lower, xmax = c_upper), height = 0, color = "grey40") +
  geom_point(aes(color = c_effect > 0), size = 2, show.legend = FALSE) +
  scale_color_manual(values = c("TRUE" = "firebrick", "FALSE" = "steelblue")) +
  theme_classic() +
  theme(axis.text.y = element_text(size = 9)) +
  labs(
    x = "Effect size (logit fold-change)",
    y = NULL,
    title = "Cell type composition: Higher vs. Lower disease stage",
    subtitle = "FDR < 0.05, CI excludes zero; red = increases, blue = decreases"
  )
```

```{r save-forest-higher-v-lower, eval=FALSE}
ggsave(file.path(sccomp_out_dir, "celltype_comp_higher_vs_lower.png"),
       width = 8, height = 12, dpi = 300)
```

# 9. Stacked Bar Chart: Cell Type Proportions by Disease State

Complementary view to the forest plots above -- shows actual proportions
per sample rather than effect sizes, grouped by disease state.

## Compute per-sample proportions

```{r per-sample-props}
selected_types <- c(
  "Epithelial", "Tumor Epithelial", "Proliferating Cells", "CD8+ T Cell",
  "CD4+ T Cell", "Treg", "Dendritic Cell", "Macrophage", "Monocyte",
  "B Cell/Plasmablast", "Fibroblast", "Mast Cell"
)
display_levels <- c(selected_types, "Other")

spe$CellTypeLabel_display <- case_when(
  spe$final_celltype == "Epithelial (EpCAM+)" ~ "Epithelial",
  str_detect(spe$final_celltype, "Tumor Epithelial") ~ "Tumor Epithelial",
  spe$final_celltype == "Proliferating Cells" ~ "Proliferating Cells",
  str_detect(spe$final_celltype, "^CD8") ~ "CD8+ T Cell",
  str_detect(spe$final_celltype, "^CD4") ~ "CD4+ T Cell",
  str_detect(spe$final_celltype, "^Treg") ~ "Treg",
  str_detect(spe$final_celltype, "^cDC|DC/mregDC|Dendritic Cell") ~ "Dendritic Cell",
  str_detect(spe$final_celltype, "^Macrophage") ~ "Macrophage",
  str_detect(spe$final_celltype, "^Monocyte") ~ "Monocyte",
  str_detect(spe$final_celltype, "B Cell|Plasma") ~ "B Cell/Plasmablast",
  spe$final_celltype == "Fibroblast/Stroma" ~ "Fibroblast",
  spe$final_celltype == "Mast Cell" ~ "Mast Cell",
  TRUE ~ "Other"
)
spe$CellTypeLabel_display <- factor(spe$CellTypeLabel_display, levels = display_levels)

celltype_props <- as.data.frame(
  table(sample_id = spe$sample_id, CellType = spe$CellTypeLabel_display)
) |>
  group_by(sample_id) |>
  mutate(Proportion = Freq / sum(Freq)) |>
  ungroup()

sample_disease_state <- unique(data.frame(sample_id = spe$sample_id,
                                           disease_state = spe$disease_state))
celltype_props <- celltype_props |> left_join(sample_disease_state, by = "sample_id")

celltype_props <- celltype_props |>
  mutate(patient_num = str_extract(sample_id, "(?<=Patient_?)\\d+"),
         roi_suffix = str_extract(sample_id, "(?<=ROI_\\d{3}_)[0-9A-Za-z]+"),
         sample_short = paste0("P", patient_num, "-", roi_suffix))

celltype_cols <- c(
  "Epithelial"          = "lightpink",
  "Tumor Epithelial"     = "darkorange",
  "Proliferating Cells"  = "deeppink",
  "CD8+ T Cell"          = "forestgreen",
  "CD4+ T Cell"          = "yellowgreen",
  "Treg"                 = "purple",
  "Dendritic Cell"       = "lightblue",
  "Macrophage"           = "dodgerblue",
  "Monocyte"             = "navy",
  "B Cell/Plasmablast"   = "gold",
  "Fibroblast"           = "firebrick",
  "Mast Cell"            = "violet",
  "Other"                = "gray85"
)

```

```{r stacked-bar-chart, fig.width=12, fig.height=4}
ggplot(celltype_props, aes(x = reorder(sample_short, as.numeric(factor(disease_state))),
                            y = Proportion, fill = CellType)) +
  geom_col() +
  facet_grid(~ disease_state, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = celltype_cols) +
  theme_bw() +
  theme(axis.text.x = element_blank()) +
  labs(x = "Sample", y = "Proportion", fill = "Cell Type",
       title = "Cell Type Composition by Sample, Grouped by Disease State")

```


