---
title: "Poster Figures: Cell Type Composition (Higher vs. Lower Stage)"
author: "Vaidehi Gupta"
output:
 html_document:
     toc: true
     toc_float: true
     code_folding: hide
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(SpatialExperiment)
    library(sccomp)
    library(dplyr)
    library(ggplot2)
    library(forcats)
})

set.seed(12345)
```

# Step 0. Paths and Load Data

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

final_celltype_path <- file.path(analysis_path, "final_cell_annotation",
                                  "r3_annotation", "spe_final_celltype_r3.rds")

celltype_prop_dir <- file.path(analysis_path, "celltype_prop_analysis")

sccomp_higher_v_lower_path <- file.path(celltype_prop_dir, "sccomp_celltype_higher_v_lower.rds")
```

```{r load-data}
spe <- readRDS(final_celltype_path)

sceIMC <- SingleCellExperiment(list(counts = assay(spe, "counts")))
sceIMC$Patient      <- factor(spe$patient_id)
sceIMC$Sample       <- factor(spe$sample_id)
sceIMC$DiseaseState <- factor(spe$disease_state, levels = c("Benign", "Lower", "Higher"))
sceIMC$CellType     <- factor(spe$final_celltype)

cat("Cells:", ncol(sceIMC), "\n")
cat("Cell types:", nlevels(sceIMC$CellType), "\n")
```

# Step 1. Load Existing sccomp Result

Reusing the already-run Higher vs. Lower result rather than re-fitting.

```{r load-sccomp-result}
stopifnot(file.exists(sccomp_higher_v_lower_path))
sccompRes <- readRDS(sccomp_higher_v_lower_path)

unique(sccompRes$parameter)
```

# Step 2. Tidy sccomp Output for Plotting

```{r tidy-sccomp}
sceIMCHvL <- sceIMC[, sceIMC$DiseaseState %in% c("Lower", "Higher")]
sceIMCHvL$DiseaseState <- factor(sceIMCHvL$DiseaseState, levels = c("Lower", "Higher"))

sccompResSub <- sccompRes |>
  as.data.frame() |>
  filter(parameter == "DiseaseStateHigher")   # Higher vs. Lower, since Lower is the reference

sccompResSub$Significant <- ifelse(sccompResSub$c_FDR < 0.05, "FDR < 0.05", "Not significant")
sccompResSub$Direction <- ifelse(sccompResSub$c_effect > 0, "Higher stage", "Lower stage")

# Cell counts per cell type, for optional point/bar sizing
numCells <- as.data.frame(table(sceIMCHvL$CellType))
colnames(numCells) <- c("CellType", "n_cells")
sccompResSub <- merge(sccompResSub, numCells, by = "CellType")
```

---

# Figure Option 1: Curated Diverging Bar Chart (Poster Main Figure)

This is the poster-friendly replacement for the dense dot-and-whisker forest
plot. Two things make it poster-legible: (1) only significant, story-relevant
cell types are shown -- edit `keepCellTypes` below to your final curated list --
and (2) solid bars read better than thin dot+CI lines at a distance.

```{r diverging-bar, fig.width=10, fig.height=6}
# Note: pick the cell types that carry your actual narrative.

keepCellTypes <- c(
  "Proliferating Cells",
  "Macrophage, M2-like (CD163+)",
  "Monocyte-Macrophage Transitional (CD14+)",
  "Treg, Memory",
  "PMN-MDSC",
  "CD4+ Memory T Cell, Exhaustion-leaning",
  "CD8+ Memory T Cell",
  "Plasmablast/Activated B Cell (CD38+)",
  "B Cell (CD19+)",
  "DC/mregDC (Epi Adj)",
  "Mast Cell",
  "Tumor Epithelial (TTF1++)"
)

still_present <- setdiff(keepCellTypes, levels(sceIMC$CellType))
if (length(still_present) > 0) {
  cat("WARNING -- these keepCellTypes entries weren't found in current annotation:\n")
  print(still_present)
}

plotDF <- sccompResSub[sccompResSub$CellType %in% keepCellTypes, ]
plotDF <- plotDF[plotDF$c_FDR < 0.05, ]  # keep only significant, since these are your curated "story" cell types

# Display-only rename -- TTF1 is a normal lung differentiation marker, not
# tumor-specific, per PI feedback. Real annotation label unchanged for now.
plotDF$CellType <- recode(plotDF$CellType,
                           "Tumor Epithelial (TTF1++)" = "Lung Epithelial (TTF1+)")

# Order bars by effect size for a clean diverging look
plotDF$CellType <- fct_reorder(plotDF$CellType, plotDF$c_effect)

divergingBarPlot <- ggplot(plotDF, aes(x = CellType, y = c_effect, fill = Direction)) +
  geom_col(width = 0.7) +
  geom_errorbar(aes(ymin = c_lower, ymax = c_upper), width = 0.2, linewidth = 0.5) +
  coord_flip() +
  scale_fill_manual(values = c("Higher stage" = "firebrick3", "Lower stage" = "steelblue3")) +
  scale_y_continuous(breaks = seq(-1.5, 1, by = 0.5)) +
  labs(
    x = NULL,
    y = "Effect size (logit fold-change)",
    fill = NULL,
    title = "Cell Type Proportion Differences Between Higher and Lower Disease Stage"
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(size = 10),
    axis.title.x = element_text(size = 12),
    plot.title = element_text(size = 14, hjust = 0),
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12),
    legend.position = "top"
  )

divergingBarPlot

#ggsave(file.path(celltype_prop_dir, "Fig_DivergingBar_HigherVsLower.pdf"),
       #divergingBarPlot, width = 10, height = 7)
```

```{r}
sccompResSub |>
  filter(c_FDR < 0.05, !(c_lower < 0 & c_upper > 0), Direction == "Lower stage") |>
  arrange(c_effect) |>
  select(CellType, c_effect, c_lower, c_upper, c_FDR)

```
## Full bar plot 

```{r fig.width=10, fig.height=10}
plotDF_full <- sccompResSub[sccompResSub$c_FDR < 0.05 & !(sccompResSub$c_lower < 0 & sccompResSub$c_upper > 0), ]

plotDF_full$CellType <- fct_reorder(plotDF_full$CellType, plotDF_full$c_effect)

n_full <- nrow(plotDF_full)
cat("Cell types passing FDR < 0.05 and CI excludes zero:", n_full, "\n")

divergingBarPlotFull <- ggplot(plotDF_full, aes(x = CellType, y = c_effect, fill = Direction)) +
  geom_col(width = 0.7) +
  geom_errorbar(aes(ymin = c_lower, ymax = c_upper), width = 0.2, linewidth = 0.5) +
  coord_flip() +
  scale_fill_manual(values = c("Higher stage" = "firebrick3", "Lower stage" = "steelblue3")) +
  labs(
    x = NULL,
    y = "Effect size (logit fold-change)",
    fill = NULL,
    title = "Cell type proportion differences: Higher vs. Lower disease stage (all significant cell types)"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.y = element_text(size = 9, face = "bold"),
    axis.text.x = element_text(size = 9),
    axis.title.x = element_text(size = 11),
    plot.title = element_text(size = 12, hjust = 0),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 11),
    legend.position = "top"
  )

divergingBarPlotFull

#ggsave(file.path(sccomp_out_dir, "Fig_DivergingBar_HigherVsLower_Full.pdf"),
       #divergingBarPlotFull, width = 9, height = max(6, n_full * 0.3))

```

---

# Figure Option 2: Small-Multiples Boxplots per Cell Type

One clean panel per curated cell type, showing individual samples (colored by
patient) plus a boxplot summary, faceted by cell type.

```{r small-multiples, fig.width=10, fig.height=12}
# Build a long-format proportion table: one row per sample x cell type
propTable <- table(sceIMC$Sample, sceIMC$CellType) |> prop.table(margin = 1)
propDF <- as.data.frame(propTable)
colnames(propDF) <- c("Sample", "CellType", "Proportion")
propDF$Proportion <- propDF$Proportion * 100

# Attach patient + disease state metadata back onto each sample
sampleMeta <- unique(data.frame(
  Sample       = sceIMC$Sample,
  Patient      = sceIMC$Patient,
  DiseaseState = sceIMC$DiseaseState
))
propDF <- merge(propDF, sampleMeta, by = "Sample")

# Keep only Lower/Higher for this comparison (drop Benign, or keep it -- your call)
propDF <- propDF[propDF$DiseaseState %in% c("Lower", "Higher"), ]
propDF$DiseaseState <- factor(propDF$DiseaseState, levels = c("Lower", "Higher"))

propDFSub <- propDF[propDF$CellType %in% keepCellTypes, ]
propDFSub$CellType <- factor(propDFSub$CellType, levels = keepCellTypes)

# Restrict to the same curated cell types as Figure 1, for a consistent story
smallMultiplesPlot <- ggplot(propDFSub, aes(x = DiseaseState, y = Proportion)) +
  geom_boxplot(outlier.shape = NA, width = 0.5) +
  geom_jitter(aes(color = Patient), width = 0.12, size = 1.8) +
  facet_wrap(~ CellType, scales = "free_y", ncol = 3,
             labeller = labeller(CellType = label_wrap_gen(20))) +
  labs(x = NULL, y = "% of cells in sample") +
  theme_classic(base_size = 14) +
  theme(
    strip.text = element_text(size = 10, face = "bold"),
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 11),
    axis.title.y = element_text(size = 14),
    legend.position = "right"
  )

smallMultiplesPlot

#ggsave(file.path(celltype_prop_dir, "Fig_SmallMultiplesBoxplots_HigherVsLower.pdf"),
       #smallMultiplesPlot, width = 14, height = 10)
```

---

```{r, fig.width=10, fig.height=8}
# Order panels to match the forest plot's effect-size ranking
forest_order <- plotDF |> arrange(c_effect) |> pull(CellType) |> as.character()

# Trim to a smaller, poster-appropriate subset -- adjust this list as you like
poster_subset <- c(
  "Proliferating Cells", "Macrophage, M2-like (CD163+)",
  "Treg, Memory", "CD4+ Memory T Cell, Exhaustion-leaning",
  "DC/mregDC (Epi Adj)", "Mast Cell"
)

propDFPoster <- propDFSub |>
  filter(CellType %in% poster_subset) |>
  mutate(CellType = factor(CellType, levels = intersect(forest_order, poster_subset)))

smallMultiplesPosterPlot <- ggplot(propDFPoster, aes(x = DiseaseState, y = Proportion)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, fill = "gray95") +
  geom_jitter(aes(color = DiseaseState), width = 0.12, size = 1.6, alpha = 0.7) +
  scale_color_manual(values = c("Lower" = "steelblue3", "Higher" = "firebrick3")) +
  facet_wrap(~ CellType, scales = "free_y", ncol = 3,
             labeller = labeller(CellType = label_wrap_gen(20))) +
  labs(x = NULL, y = "% of cells in sample") +
  theme_classic(base_size = 14) +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 11),
    axis.title.y = element_text(size = 14),
    legend.position = "none"
  )

smallMultiplesPosterPlot

```
```{r, fig.width=10, fig.height=14}
# Build proportion data for every significant cell type, not just the curated subset
forest_order_full <- plotDF_full |> arrange(c_effect) |> pull(CellType) |> as.character()

propDFAll <- propDF[propDF$CellType %in% forest_order_full, ]
propDFAll$CellType <- factor(propDFAll$CellType, levels = forest_order_full)

n_panels <- nlevels(propDFAll$CellType)
cat("Cell types to plot:", n_panels, "\n")

smallMultiplesAllPlot <- ggplot(propDFAll, aes(x = DiseaseState, y = Proportion)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, fill = "gray95") +
  geom_jitter(aes(color = DiseaseState), width = 0.12, size = 1.4, alpha = 0.7) +
  scale_color_manual(values = c("Lower" = "steelblue3", "Higher" = "firebrick3")) +
  facet_wrap(~ CellType, scales = "free_y", ncol = 4,
             labeller = labeller(CellType = label_wrap_gen(20))) +
  labs(x = NULL, y = "% of cells in sample") +
  theme_classic(base_size = 12) +
  theme(
    strip.text = element_text(size = 9, face = "bold"),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 9),
    axis.title.y = element_text(size = 12),
    legend.position = "none"
  )

smallMultiplesAllPlot

```


# Notes

- Both figures pull from the **same curated `keepCellTypes` list** -- edit that
  one vector and both figures update consistently. Verify it against
  `levels(spe$final_celltype)` before finalizing -- several entries were
  updated here to match the post-relabeling names (naive/memory/exhaustion),
  but other drift may exist that wasn't caught.
- Figure 1 (diverging bar) is best as your **main poster figure**: it directly
  shows the statistical result (effect size + CI) your talk already leans on.
- Figure 2 (small multiples) is a strong **secondary/supporting figure**: it
  reassures a reader that the effect isn't being driven by one outlier patient,
  since every dot (sample) is visible and colored by patient.
- If poster space is tight, Figure 1 alone is sufficient; Figure 2 is the first
  thing to cut, not the reverse.
- Assumes `sccomp_celltype_r3.Rmd` follows the same object-building convention
  as the original `sccomp_celltype.Rmd` (`Patient`/`Sample`/`DiseaseState`/
  `CellType` columns on a `SingleCellExperiment`) -- if that's since changed,
  the Step 0 chunk above will need matching updates.

```{r session-info}
sessionInfo()
```
