---
title: "Subclustering: Cluster 15 (T Cell/Plasma Cell, Mixed)"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)
set.seed(42)
```

## Overview

Subclustering cluster 15 (currently labeled "T Cell/Plasma Cell, Mixed
(CD4+/CD138+)") from the broad k=20 annotation. Unlike the myeloid
subclustering (which combined two broad clusters that looked related),
this is a single broad cluster that showed markers from two different
lineages at once. The goal here is to test whether it splits into a
real CD4+ T cell population and a real plasma cell population, or
whether it holds together as one population with some background/
spillover on one of the two marker sets.

## 1. Load Libraries

```{r libraries}
# Only run this if you start a new RStudio session
#.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(imcRtools)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(scater)
library(scran)
library(bluster)
library(patchwork)
library(RColorBrewer)
library(ComplexHeatmap)
library(circlize)
```

## 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

spe_path <- file.path(analysis_path, "final_cell_annotation", "h5_cluster_annot.rds")
out_path <- file.path(analysis_path, "final_cell_annotation", "subclustering",
                      "cluster15_subcluster.rds")
findmarkers_path <- file.path(analysis_path, "final_cell_annotation", "subclustering",
                              "cluster15_subcluster_findmarkers.csv")
```

## 3. Load Annotated Data and Subset

```{r load-and-subset}
spe <- readRDS(spe_path)

spe_15 <- spe[, spe$broad_clusters == "15"]

cat("Cells in cluster 15 subset:", ncol(spe_15), "\n")
```

## 4. Re-cluster the Subset on the Existing Harmony Embedding

Starting with k = 10, same as the myeloid subclustering, to encourage
finer splits than the broad k = 20 pass.

```{r subcluster}
harmony_sub <- reducedDim(spe_15, "Harmony")
cat("Harmony embedding dimensions (subset):", dim(harmony_sub), "\n")

set.seed(220225)
g_sub     <- buildSNNGraph(t(harmony_sub), k = 10, type = "rank")
clust_sub <- igraph::cluster_louvain(g_sub)
spe_15$sub_clusters <- factor(clust_sub$membership)

cat("Number of subclusters:", nlevels(spe_15$sub_clusters), "\n")
print(table(spe_15$sub_clusters))
```

## 5. UMAP on the Subset

```{r umap-sub}
set.seed(220225)
spe_15 <- runUMAP(spe_15, dimred = "Harmony", name = "UMAP_sub")
```

```{r plot-umap-sub-clusters, fig.width=10, fig.height=7}
p_sub_clusters <- dittoDimPlot(spe_15, var = "sub_clusters",
                               reduction.use = "UMAP_sub",
                               size = 0.4, do.label = TRUE, labels.size = 4) +
  ggtitle("Cluster 15 Subset: UMAP colored by Subcluster") +
  theme(legend.position = "right")
print(p_sub_clusters)
```

## 6. Marker Sets for T Cell / Plasma Cell Disambiguation

T/checkpoint markers, plasma/B markers, and a few unrelated-lineage
markers included specifically to catch any surprise contamination
(similar to what flagged cluster 13's FoxP3/CD15 mix at the broad
level).

```{r markers}
disambiguation_markers <- c(
  "CD4", "CD8a", "CD45RO", "FoxP3", "TIM3", "PD1",
  "CD138", "CD38", "CD19",
  "CD14", "CD15", "CD117", "CD11c"
)
disambiguation_markers <- disambiguation_markers[disambiguation_markers %in% rownames(spe_15)]
cat("Markers found:", paste(disambiguation_markers, collapse = ", "), "\n")
```

## 7. Violin Plots per Subcluster

```{r make-violinplot}
make_violinplot <- function(spe, assay_name, cluster_col, marker_set, title, ncol = 3) {
  expr_mat   <- assay(spe, assay_name)[marker_set, , drop = FALSE]
  cluster_id <- spe[[cluster_col]]

  mean_expr <- sapply(levels(cluster_id), function(cl) {
    idx <- which(cluster_id == cl)
    rowMeans(expr_mat[, idx, drop = FALSE])
  })

  if (is.matrix(mean_expr) && nrow(mean_expr) > 1) {
    marker_order <- rownames(mean_expr)[hclust(dist(mean_expr))$order]
  } else {
    marker_order <- marker_set
  }

  plot_df <- as.data.frame(t(expr_mat)) |>
    mutate(cluster = cluster_id) |>
    pivot_longer(-cluster, names_to = "marker", values_to = "expr")

  plot_df$marker  <- factor(plot_df$marker, levels = marker_order)
  plot_df$cluster <- factor(plot_df$cluster, levels = levels(cluster_id))

  ggplot(plot_df, aes(x = cluster, y = expr, fill = cluster)) +
    geom_violin(scale = "width", trim = TRUE, linewidth = 0.2) +
    stat_summary(fun = median, geom = "point", size = 0.6, color = "black") +
    facet_wrap(~ marker, scales = "free_y", ncol = ncol) +
    labs(title = title, x = "Subcluster", y = "Z-score expression") +
    theme_bw(base_size = 11) +
    theme(
      axis.text.x      = element_text(angle = 0, hjust = 0.5, size = 9),
      strip.text        = element_text(face = "bold"),
      panel.grid.major.x = element_blank(),
      legend.position   = "none"
    )
}
```

```{r violin-cluster15, fig.width=16, fig.height=12}
p_violin_15 <- make_violinplot(spe_15, "scaled_0100", "sub_clusters",
                               disambiguation_markers,
                               "Cluster 15 Subset: Marker Expression per Subcluster",
                               ncol = 3)
print(p_violin_15)
```

## 8. Statistical Marker Discovery (findMarkers) on Subclusters

```{r findmarkers}
markers_de <- findMarkers(spe_15,
                          groups     = spe_15$sub_clusters,
                          assay.type = "scaled_0100",
                          subset.row = disambiguation_markers)

all_markers <- lapply(levels(spe_15$sub_clusters), function(cl) {
  df <- as.data.frame(markers_de[[cl]])
  data.frame(
    subcluster = cl,
    marker     = rownames(df),
    Top        = df$Top,
    FDR        = df$FDR,
    logFC      = df$summary.logFC
  )
})

all_markers_df <- do.call(rbind, all_markers)
dir.create(dirname(findmarkers_path), recursive = TRUE, showWarnings = FALSE)
write.csv(all_markers_df, findmarkers_path, row.names = FALSE)

top1_summary <- do.call(rbind, lapply(levels(spe_15$sub_clusters), function(cl) {
  sub <- all_markers_df[all_markers_df$subcluster == cl & all_markers_df$logFC > 1.0, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  if (nrow(sub) == 0) {
    data.frame(subcluster = cl, top_markers = "none above logFC 1.0")
  } else {
    data.frame(subcluster = cl,
               top_markers = paste(head(sub$marker, 5), collapse = ", "))
  }
}))

cat("Top positive markers (logFC > 1.0) per subcluster:\n")
print(knitr::kable(top1_summary, col.names = c("Subcluster", "Top Positive Markers")))
```

Everywhere CD4 (T cell marker) goes up, CD138 (B cell marker) goes up right alongside it. Doesn't really prove that these are two distinct cell populations.

If there were real plasma cells embedded in this cluster, we'd expect to see at least one subcluster to show CD38 (the plasma marker) light up along with CD138, but that doesn't happen. This could be evidence that CD138's signal here is background/spillover, not plasma cell contamination.

Cluster 15 most likely is a CD4 T cell population with many distinct activation/differentiation states, with CD138 riding along as noise.

```{r logfc-wide-table-formatted}
logfc_wide <- all_markers_df |>
  select(subcluster, marker, logFC) |>
  mutate(logFC = round(logFC, 2)) |>
  pivot_wider(names_from = subcluster, values_from = logFC)

knitr::kable(logfc_wide, caption = "logFC per marker, per subcluster (Cluster 15 subset)")
```
```{r findmarkers-per-subcluster-tables}
cat("Detailed tables:\n\n")

for (cl in levels(spe_15$sub_clusters)) {
  sub <- all_markers_df[all_markers_df$subcluster == cl & all_markers_df$Top <= 5, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  if (nrow(sub) == 0) {
    cat("Subcluster", cl, ": no markers in top 5\n\n")
  } else {
    print(knitr::kable(sub[, c("marker", "Top", "FDR", "logFC")],
                       caption = paste("Subcluster", cl),
                       digits  = 3))
  }
}
```

Subclusters 2 and 4: no markers clear threshold, very isolated on the right of the UMAP, doesn't blend in with the main cluster mass.

Subcluster 3: CD4+ T cell, but is elevated for CD11c, which is a dendritic/myeloid marker

Subcluster 4: low quality noise most likely

Subcluster 5: TIM3 elevated, likely Exhausted T cell

Subcluster 6: has a sharp, isolated FoxP3 signal and nothing else. Looks like a real Treg population.

Subcluster 8: strong PD1 signal (confirmed by LogFC value and violin plot). Looks like a real, distinct exhausted/checkpoint-high CD4 population

Subcluster 9: elevated in CD45RO and CD4, likely Memory T Cell

## Subcluster 2 and 4 Quality Check

```{r subcluster2-4-quality-check}
spe_15$total_counts <- colSums(assay(spe_15, "counts"))

quality_check <- colData(spe_15) |>
  as.data.frame() |>
  mutate(group = ifelse(sub_clusters %in% c("2", "4"), "Subcluster 2/4", "Other subclusters")) |>
  group_by(group) |>
  summarise(
    n = n(),
    mean_area = mean(area, na.rm = TRUE),
    median_area = median(area, na.rm = TRUE),
    mean_total_counts = mean(total_counts, na.rm = TRUE),
    median_total_counts = median(total_counts, na.rm = TRUE)
  )

print(quality_check)
```
Area is nearly identical, so this isn't a segmentation-size issue. 
However, the total counts are much lower. This is evidence that these cells didn't stain well.

## CD138 Spillover Check

```{r cd138-spillover-check}
check_markers <- c("CD138", "EpCAM", "CD4", "CD45RO", "CD19", "CD38")
check_markers <- check_markers[check_markers %in% rownames(spe_15)]

expr_df <- as.data.frame(t(assay(spe_15, "counts")[check_markers, ]))
cor_matrix <- cor(expr_df, method = "spearman")

cat("Correlation of CD138 with panel markers, within cluster 15:\n")
print(round(sort(cor_matrix["CD138", ], decreasing = TRUE), 3))
```

CD138 has a noticeable correlation with CD4 (0.59) and CD45RO (0.51) (they tend to be high in the same cells), which means that the CD138 signal in this cluster might be real, low-level expression on activated/memory CD4+ T cells, rather than plasma cell contamination.

## CD138 vs. Total Counts (Checking for Generalized Intensity Effect)

```{r cd138-total-counts-check}
spe_15$total_counts <- colSums(assay(spe_15, "counts"))
cor(assay(spe_15, "counts")["CD138", ], spe_15$total_counts, method = "spearman")
```

This points to the possibility of a generalized intensity effect. Cells with more total signal overall show elevated CD138 along with the other markers, simply because those markers are also picking up more signal in high-intensity cells.

## Subcluster 5 Checkpoint Marker Correlations (Checking for Intensity Confound in TIM3)

```{r subcluster5-checkpoint-check}
checkpoint_markers <- c("TIM3", "PD1", "LAG3", "CTLA4", "CD138")
checkpoint_markers <- checkpoint_markers[checkpoint_markers %in% rownames(spe_15)]

expr_df <- as.data.frame(t(assay(spe_15, "counts")[checkpoint_markers, ]))
expr_df$total_counts <- spe_15$total_counts

cor(expr_df, method = "spearman")["TIM3", ]
```

TIM3 correlates more strongly with total_counts (0.57) than with any of the checkpoint/exhaustion markers. This further supports a generalized intensity effect that we are seeing with CD138, rather than a real exhaustion signature.

## Correlate PD1 against the other checkpoint markers in Subcluster 8

```{r subcluster8-checkpoint-check}
checkpoint_markers <- c("PD1", "TIM3", "LAG3", "CTLA4", "CD138")
checkpoint_markers <- checkpoint_markers[checkpoint_markers %in% rownames(spe_15)]

sub8_cells <- spe_15$sub_clusters == "8"

expr_df <- as.data.frame(t(assay(spe_15, "counts")[checkpoint_markers, sub8_cells]))
expr_df$total_counts <- spe_15$total_counts[sub8_cells]

cor(expr_df, method = "spearman")["PD1", ]
```

Within subcluster 8, PD1 and TIM3 (the two main exhausion markers) don't track together at all. Total_counts drops to 0.21, which means that within the cells that actually make up subcluster 8, PD1 is not strongly explained by overall intensity.

## PD1 check (across all subclusters)

```{r pd1-checkpoint-check-full-cluster}
checkpoint_markers <- c("PD1", "TIM3", "LAG3", "CTLA4", "CD138")
checkpoint_markers <- checkpoint_markers[checkpoint_markers %in% rownames(spe_15)]

expr_df <- as.data.frame(t(assay(spe_15, "counts")[checkpoint_markers, ]))
expr_df$total_counts <- spe_15$total_counts

cor(expr_df, method = "spearman")["PD1", ]
```

At the whole-cluster level, PD1 shows the same intensity-tracking signature found for CD138 and TIM3. This may be driven by cross-subcluster differences (subclusters with generally higher total signal also happen to have somewhat higher PD1 on average), not by a within-population intensity confound.

## 9. Composition Check: Patient Composition of Each Subcluster

Cluster 15 is a single broad cluster (not a merge of two, like the
myeloid subclustering), so there's no "original broad cluster" split
to check here, only whether any subcluster is patient-driven rather
than a real shared population.

```{r composition-patient-heatmap, fig.width=12, fig.height=6}
comp_patient <- table(subcluster = spe_15$sub_clusters,
                      patient = spe_15$patient_id)
comp_patient_props <- prop.table(comp_patient, margin = 1)

col_fun_patient <- colorRamp2(c(0, 0.5, 1), rev(brewer.pal(3, "RdBu")))

ht_patient <- Heatmap(
  comp_patient_props,
  name              = "Proportion",
  col               = col_fun_patient,
  cluster_rows      = FALSE,
  cluster_columns   = FALSE,
  column_title      = "Patient",
  column_title_side = "bottom",
  row_title         = "Subcluster",
  column_names_rot  = 45,
  cell_fun = function(j, i, x, y, width, height, fill) {
    val <- comp_patient_props[i, j]
    text_col <- if (val < 0.2 || val > 0.8) "white" else "black"
    grid.text(sprintf("%.2f", val), x, y, gp = gpar(col = text_col, fontsize = 7))
  }
)

draw(ht_patient, column_title = "Subcluster composition by patient (row-normalized %)")
```

Subcluster 7 definitely looks patient driven, with Patient20_519 making up 47% and Patient20_425 making up 22%, which means 69% of this subcluster is from one patient. It is also sitting as an isolated island in the UMAP, so this is probably a technical/segmentation artifact.

## 10. Manual Subcluster Annotation

```{r annotate-sub, echo=TRUE}
cluster15_subannotation <- c(
  "1"  = "1. Memory/Activated T Cell (CD4++)",  # CD4++, CD45RO+, FoxP3, PD1
  "2"  = "2. Unresolved/Low Quality", # no marker above logFC 1.0
  "3"  = "3. CD4+ T Cell", # CD4, CD138, CD11c (could be contamination), CD45RO+
  "4"  = "4. Unresolved/Low Quality",  # no marker above logFC 1.0
  "5"  = "5. T Cell (TIM3+)",  # Has TIM3 as its top marker
  "6"  = "6. Treg (FoxP3+)", # Has FoxP3 as its top marker
  "7"  = "7. Unresolved",  # no marker above logFC 1.0, patient-driven, likely technical artifact
  "8"  = "8. T Cell, Exhausion-leaning (PD1++)", # Highly elevated in PD1
  "9"  = "9. Memory T Cell (CD45RO+)" # CD45RO+ and CD4+
)

if (length(cluster15_subannotation) > 0) {
  spe_15$sub_celltype <- cluster15_subannotation[as.character(spe_15$sub_clusters)]
  spe_15$sub_celltype <- factor(spe_15$sub_celltype, levels = unique(cluster15_subannotation))
  cat("Subcluster annotation applied.\n")
  print(table(spe_15$sub_celltype, useNA = "ifany"))
} else {
  cat("No subcluster annotation yet. Fill in cluster15_subannotation above.\n")
}
```

```{r umap-sub-celltype, fig.width=10, fig.height=7}
if ("sub_celltype" %in% colnames(colData(spe_15))) {

  n_types <- nlevels(spe_15$sub_celltype)
  subtype_cols <- setNames(
    colorRampPalette(brewer.pal(12, "Paired"))(n_types),
    levels(spe_15$sub_celltype)
  )

  subtype_cols["T Cell, Exhausion-leaning (PD1++)"] <- "#4B0082" 
  subtype_cols["T Cell (TIM3+)"] <- "#009DC4"
  subtype_cols["Memory/Activated T Cell (CD4++)"] <- "#C30F16"

  p_sub_celltype <- dittoDimPlot(spe_15, var = "sub_celltype",
                                 reduction.use = "UMAP_sub",
                                 size = 0.4, do.label = TRUE, labels.size = 3.5) +
    scale_color_manual(values = subtype_cols) +
    ggtitle("Cluster 15 Subset: UMAP colored by Subcluster Cell Type") +
    theme(legend.position = "right")
  print(p_sub_celltype)

} else {
  cat("Run the annotate-sub chunk first to apply subcluster annotations.\n")
}
```

## 11. Save Subclustered/Annotated Object

```{r save}
dir.create(dirname(out_path), recursive = TRUE, showWarnings = FALSE)
saveRDS(spe_15, out_path)
cat("Saved cluster 15 subcluster object to:", out_path, "\n")
```

## Notes

- **Cluster 15 isn't really T Cell/Plasma Cell (Mixed).** CD138 never separates from CD4, and CD38/CD19 stay flat throughout, so there's no real plasma cell population hiding in here. It's a heterogeneous CD4+ T Cell/Treg population with some technical noise (from subclusters 2, 4 and 7).

- **CD138's presence turned out to be a generalized intensity effect, not biology or spillover.** It correlates more strongly with total counts than with CD4 or CD45R0, and doesn't correlate with EpCAM or SMA, ruling out segmentation spillover from neighboring cells. Cells with more total signal overall simply read higher on CD138 along with everything else.

- **This same intensity effect explains subcluster 5's TIM3 signal.** TIM3
  correlates more with total_counts (0.57) than with PD1, LAG3, or CTLA4
  (0.20-0.30) -- if this were a real, coordinated exhaustion phenotype, TIM3
  should track with the other checkpoint markers more tightly than with
  generic intensity. Labeled as "CD4+ T Cell (TIM3+, unconfirmed)" rather
  than "Exhausted T Cell" until corroborated by additional checkpoint
  markers.

- **Subclusters 2 and 4 are low-quality, not a real cell state.** Every
  marker checked (including CD4, CD45RO, FoxP3, PD1, TIM3) is significantly
  *lower* than the rest of cluster 15, and total counts are ~38% lower with
  no difference in cell area -- consistent with poorly-stained cells, the
  same phenomenon behind the original broad "unknown" cluster, just
  resurfacing here at finer resolution. Labeled Unresolved/Low Quality.

- **Subcluster 7 is likely a Patient20-specific technical artifact.** No
  marker clears threshold, and Patient20_425 + Patient20_519 together make
  up ~69% of this subcluster -- similar pattern to broad cluster 17's
  Patient12 concentration. 

- **Subcluster 3's CD11c signal is still an open flag.** CD11c is a
  myeloid/DC marker, not expected alongside CD4/CD138/CD45RO/FoxP3.

