---
title: "DC/mregDC Subclustering — Run5 Cluster 4 (DC/mregDC vs. AT2 Epithelial Check)"
author: "Tazein Shah"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(42)
```

## Overview

Extract the DC/mregDC cluster (cluster 4) from the annotated run5 SPE
(`h5_cluster_annot.rds`) and recluster on the existing Harmony
embedding (Louvain, k = 20). The panel has no surfactant-protein
marker, so AT2 identity can't be confirmed directly -- the question
here is resolved indirectly: do subclusters show a clean DC/mregDC
signature (CD11c/HLADRB/LAMP3/CCR7/CD141/CD1c) with epithelial markers
riding along as spillover, or does a subset instead show a
predominantly epithelial signature (EpCAM/CD117/TTF1orCD16/GATA3) with
DC markers as the weaker, spillover-level signal -- which would point
to misclustered AT2-like epithelial cells rather than true DC/mregDC.

Broad clusters 9 (Tumor Epithelial), 12 (Epithelial), and 19 (Tumor
Epithelial Neuroendocrine) are pulled in as visual reference groups
for the violin plots and heatmap only -- they are not reclustered and
never merged into the saved `spe_dc` object. The UMAP (Section 11)
plots the DC/mregDC subset only.

---

## 1. Load Libraries

```{r libraries}
library(imcRtools)
library(scran)
library(bluster)
library(BiocParallel)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(scater)
library(patchwork)
library(RColorBrewer)
library(pheatmap)
```

---

## 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

spe_path <- file.path(analysis_path, "final_cell_annotation", "final_broad_cluster", "h5_cluster_annot.rds")

dc_clustering_path <- file.path(analysis_path, "final_cell_annotation", "subclustering", "dc_mregdc")

out_path <- file.path(dc_clustering_path, "dc_mregdc_subclustering.rds")
```

---

## 3. Load Annotated SPE

```{r load-data}
spe <- readRDS(spe_path)

print(table(spe$broad_clusters))
```

---

## 4. Extract DC/mregDC Cluster

```{r subset}
dc_cluster <- "4"

spe_dc <- spe[, as.character(spe$broad_clusters) %in% dc_cluster]

cat("Cells in DC/mregDC subset:", ncol(spe_dc), "\n")
```

---

## 5. Recluster: Louvain on Existing Harmony Embedding (k = 20)

```{r recluster}
snn_graph <- buildSNNGraph(spe_dc, use.dimred = "Harmony", k = 20,
                            BPPARAM = SerialParam())

clusters <- igraph::cluster_louvain(snn_graph)$membership

spe_dc$dc_subclusters <- factor(clusters)

cat("Number of subclusters:", nlevels(spe_dc$dc_subclusters), "\n")
print(table(spe_dc$dc_subclusters))
```

---

## 6. UMAP and Distribution on the Subset

```{r umap-sub}
set.seed(220225)
spe_dc <- runUMAP(spe_dc, dimred = "Harmony", name = "UMAP_sub")
```

```{r plot-patient-composition, fig.width=10, fig.height=5}
patient_counts <- as.data.frame(table(patient_id = spe_dc$patient_id))
patient_counts$pct <- patient_counts$Freq / sum(patient_counts$Freq) * 100
patient_counts$pct_label <- paste0(round(patient_counts$pct, 1), "%")

p_patient_comp <- ggplot(patient_counts, aes(x = patient_id, y = Freq, fill = patient_id)) +
  geom_col() +
  geom_text(aes(label = pct_label), vjust = -0.5, size = 3.5) +
  labs(title = "DC/mregDC Subset: Cell Count by Patient",
       x = "Patient", y = "Cell Count") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))  # headroom so labels don't clip
print(p_patient_comp)

cat("DC/mregDC subset composition by patient:\n")
print(patient_counts[order(-patient_counts$Freq), c("patient_id", "Freq", "pct")])
```

---

## 6.5. Reference Epithelial Clusters for Visualization (No Reclustering)

```{r epi-reference-extract}
# Broad clusters pulled in purely as visual reference groups -- not
# reclustered, not merged into spe_dc, just plotted alongside it.
epi_ref_clusters <- c("9", "12", "19")

epi_ref_labels <- c(
  "9"  = "Tumor Epithelial (C9)",
  "12" = "Epithelial (C12)",
  "19" = "Tumor Epithelial NE (C19)"
)

spe_epi_ref <- spe[, as.character(spe$broad_clusters) %in% epi_ref_clusters]
spe_epi_ref$vis_group <- epi_ref_labels[as.character(spe_epi_ref$broad_clusters)]

cat("Reference epithelial cells pulled in for visualization:\n")
print(table(spe_epi_ref$vis_group))
```

---

## 7. Marker Sets: DC/mregDC vs. Epithelial (AT2-like)

```{r markers}
# true DC/mregDC lineage markers
dc_markers <- c("CD11c", "HLADRB", "CD141", "CD1c", "LAMP3", "CCR7")

# epithelial markers -- if a subcluster is dominated by these instead,
# that's the AT2-like/misclustered-epithelial signal, not DC/mregDC
epithelial_markers <- c("EpCAM", "CD117", "TTF1orCD16", "GATA3")

# activation / functional state markers, useful either way
state_markers <- c("PD1", "PDL1", "TIM3", "CD86", "MKI67", "CD45RO")

# other-lineage spillover flags, to rule out non-DC/non-epithelial admixture
admixture_markers <- c("CD68", "CD14", "CD19")

# everything else in the panel, plotted too so nothing is left out
other_markers <- setdiff(rownames(spe_dc), c(dc_markers, epithelial_markers,
                                              state_markers, admixture_markers,
                                              "DNA191", "DNA193", "Membrane"))

dc_all_markers <- c(dc_markers, epithelial_markers, state_markers, admixture_markers, other_markers)

cat("Markers found:", paste(dc_all_markers, collapse = ", "), "\n")
```

---

## 8. Violin Plots: DC/mregDC Subclusters vs. Reference Epithelial Clusters

```{r make-violinplot}
make_violinplot_df <- function(expr_mat, group_vec, group_levels, marker_set, title, ncol = 3) {
  expr_mat <- expr_mat[marker_set, , drop = FALSE]

  mean_expr <- sapply(group_levels, function(g) {
    idx <- which(group_vec == g)
    rowMeans(expr_mat[, idx, drop = FALSE])
  })

  if (is.matrix(mean_expr) && nrow(mean_expr) > 1) {
    marker_order <- rownames(mean_expr)[hclust(dist(mean_expr))$order]
  } else {
    marker_order <- marker_set
  }

  plot_df <- as.data.frame(t(expr_mat)) |>
    mutate(group = factor(group_vec, levels = group_levels)) |>
    pivot_longer(-group, names_to = "marker", values_to = "expr")

  plot_df$marker <- factor(plot_df$marker, levels = marker_order)

  ggplot(plot_df, aes(x = group, y = expr, fill = group)) +
    geom_violin(scale = "width", trim = TRUE, linewidth = 0.2) +
    stat_summary(fun = median, geom = "point", size = 0.6, color = "black") +
    facet_wrap(~ marker, scales = "free_y", ncol = ncol) +
    labs(title = title, x = "Group", y = "Z-score expression") +
    theme_bw(base_size = 11) +
    theme(
      axis.text.x       = element_text(angle = 45, hjust = 1, size = 8),
      strip.text        = element_text(face = "bold"),
      panel.grid.major.x = element_blank(),
      legend.position   = "none"
    )
}
```

```{r combine-for-viz}
# DC subclusters keep their own labels; reference epithelial cells keep
# their broad-cluster labels. Nothing here feeds back into spe_dc.
dc_group_vec  <- paste0("DC Sub ", as.character(spe_dc$dc_subclusters))
epi_group_vec <- spe_epi_ref$vis_group

vis_group_levels <- c(paste0("DC Sub ", levels(spe_dc$dc_subclusters)),
                       unname(epi_ref_labels))  # strip names so they aren't used as identifiers downstream

vis_group_vec <- c(dc_group_vec, epi_group_vec)

expr_vis_0100 <- cbind(
  assay(spe_dc, "scaled_0100")[dc_all_markers, , drop = FALSE],
  assay(spe_epi_ref, "scaled_0100")[dc_all_markers, , drop = FALSE]
)
```

```{r violin-dc-global, fig.width=10, fig.height=5}
p_violin_dc_global <- make_violinplot_df(expr_vis_0100, vis_group_vec, vis_group_levels,
                                         dc_markers,
                                         "DC/mregDC Subclusters vs. Reference Epithelial Clusters: DC Lineage Markers",
                                         ncol = 3)
print(p_violin_dc_global)
```

```{r violin-epithelial-global, fig.width=10, fig.height=5}
p_violin_epi_global <- make_violinplot_df(expr_vis_0100, vis_group_vec, vis_group_levels,
                                          epithelial_markers,
                                          "DC/mregDC Subclusters vs. Reference Epithelial Clusters: Epithelial Markers",
                                          ncol = 3)
print(p_violin_epi_global)
```

```{r violin-state-global, fig.width=12, fig.height=8}
p_violin_state_global <- make_violinplot_df(expr_vis_0100, vis_group_vec, vis_group_levels,
                                            state_markers,
                                            "DC/mregDC Subclusters vs. Reference Epithelial Clusters: State Markers",
                                            ncol = 3)
print(p_violin_state_global)
```

```{r violin-admixture-global, fig.width=10, fig.height=5}
p_violin_admixture_global <- make_violinplot_df(expr_vis_0100, vis_group_vec, vis_group_levels,
                                                admixture_markers,
                                                "DC/mregDC Subclusters vs. Reference Epithelial Clusters: Other-Lineage Admixture Markers",
                                                ncol = 3)
print(p_violin_admixture_global)
```

```{r violin-other-global, fig.width=14, fig.height=16}
p_violin_other_global <- make_violinplot_df(expr_vis_0100, vis_group_vec, vis_group_levels,
                                            other_markers,
                                            "DC/mregDC Subclusters vs. Reference Epithelial Clusters: All Other Panel Markers",
                                            ncol = 4)
print(p_violin_other_global)
```

---

## 9. Heatmap: DC/mregDC Subclusters vs. Reference Epithelial Clusters (All Markers)

```{r dc-vs-epi-heatmap-matrix}
mean_expr_mat <- sapply(vis_group_levels, function(g) {
  idx <- which(vis_group_vec == g)
  rowMeans(expr_vis_0100[, idx, drop = FALSE])
})

# keep DC / Epithelial / State / Admixture / Other markers grouped as
# blocks rather than clustering rows freely, so the marker-class split
# stays visually clear
mean_expr_mat <- mean_expr_mat[dc_all_markers, , drop = FALSE]

#print(round(mean_expr_mat, 2))
```

```{r dc-vs-epi-heatmap-plot, fig.width=11, fig.height=14}
n_dc_markers    <- sum(rownames(mean_expr_mat) %in% dc_markers)
n_epi_markers   <- sum(rownames(mean_expr_mat) %in% epithelial_markers)
n_state_markers <- sum(rownames(mean_expr_mat) %in% state_markers)
n_admix_markers <- sum(rownames(mean_expr_mat) %in% admixture_markers)
gap_positions   <- cumsum(c(n_dc_markers, n_epi_markers, n_state_markers, n_admix_markers))

p_heatmap <- pheatmap::pheatmap(
  mean_expr_mat,
  scale = "row",  # rescale each marker across groups so LAMP3's
                  # magnitude doesn't compress every other marker's range
  color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
  cluster_rows = FALSE,
  cluster_cols = FALSE,  # keep DC subclusters left, reference epithelial groups right
  gaps_row = gap_positions,  # breaks between DC | Epithelial | State | Admixture | Other blocks
  gaps_col = nlevels(spe_dc$dc_subclusters),  # break between DC subclusters and reference groups
  display_numbers = TRUE,
  number_format = "%.2f",
  number_color = "black",
  fontsize_number = 7,
  main = "DC/mregDC Subclusters vs. Reference Epithelial Clusters: Mean Marker Expression (All Markers)",
  fontsize = 10,
  fontsize_row = 8,
  fontsize_col = 9,
  angle_col = 45,  # reference group labels are full text now, need the angle
  border_color = NA,
  silent = TRUE
)

grid::grid.newpage()
grid::grid.draw(p_heatmap$gtable)
grid::grid.text("Row Z-score", x = 0.94, y = 0.98,
                 gp = grid::gpar(fontsize = 9, fontface = "bold"))
```

A DC subcluster that lights up red/blue consistently across the DC
block and stays flat across the Epithelial block supports true
DC/mregDC. A DC subcluster with the pattern flipped -- flat or
negative across DC markers, strongly positive across
EpCAM/CD117/TTF1orCD16/GATA3 (resembling the reference C9/C12/C19
columns) -- is the AT2-like/misclustered-epithelial signal.

---

## 10. Manual Subcluster Annotation

```{r annotate-sub, echo=TRUE}
dc_subannotation <- c(
  "5" = "DC/mregDC",
  "3" = "DC/mregDC (Epi Adj)",
  "1" = "DC/mregDC Weak",
  "2" = "DC/mregDC Weak",
  "4" = "DC/mregDC (LAMP3-)"
)

if (length(dc_subannotation) > 0 && !"PLACEHOLDER" %in% dc_subannotation) {
  spe_dc$sub_celltype <- dc_subannotation[as.character(spe_dc$dc_subclusters)]
  cat("Subcluster annotation applied.\n")
  print(table(spe_dc$sub_celltype, useNA = "ifany"))
} else {
  cat("No subcluster annotation yet. Fill in dc_subannotation above using the\n")
  cat("violin plots (Section 8) and the DC vs. Epithelial heatmap (Section 9).\n")
}
```

---

## 11. UMAP Colored by Subcluster Cell Type

```{r umap-sub-celltype, fig.width=10, fig.height=5}
if ("sub_celltype" %in% colnames(colData(spe_dc))) {

  spe_dc$sub_celltype <- factor(spe_dc$sub_celltype, levels = unique(dc_subannotation))

  p_sub_celltype <- dittoDimPlot(spe_dc, var = "sub_celltype",
                                 reduction.use = "UMAP_sub",
                                 size = 1.2) +
    ggtitle("DC/mregDC Subset: UMAP colored by Subcluster Cell Type") +
    theme(legend.position = "right")
  print(p_sub_celltype)

} else {
  cat("Run the annotate-sub chunk first to apply annotations.\n")
}
```

---

## 12. Cell Type Composition by Patient

```{r celltype-patient-composition, fig.width=14, fig.height=8}
if ("sub_celltype" %in% colnames(colData(spe_dc))) {

  celltype_patient_counts <- as.data.frame(
    table(patient_id = spe_dc$patient_id,
          cell_type  = spe_dc$sub_celltype)
  )

  p_celltype_patient <- ggplot(celltype_patient_counts, aes(x = patient_id, y = Freq, fill = patient_id)) +
    geom_col() +
    facet_wrap(~ cell_type, scales = "free", ncol = 2,
               labeller = labeller(cell_type = label_wrap_gen(width = 35))) +
    labs(title = "DC/mregDC Subset: Cell Type Composition by Patient",
         x = "Patient", y = "Cell Count") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
          strip.text = element_text(face = "bold", size = 8),
          legend.position = "none")

  print(p_celltype_patient)

  celltype_patient_plot_path <- file.path(dc_clustering_path, "celltype_composition_by_patient.png")
  ggsave(celltype_patient_plot_path, p_celltype_patient, width = 14, height = 8, dpi = 300)
  cat("Saved full-size plot to:", celltype_patient_plot_path, "\n")

} else {
  cat("Run the annotate-sub chunk first to apply annotations.\n")
}
```

---

## 13. Save Subclustered/Annotated Object

```{r save}
dir.create(dirname(out_path), recursive = TRUE, showWarnings = FALSE)
saveRDS(spe_dc, out_path)
cat("Saved DC/mregDC subcluster object to:", out_path, "\n")
```

Note: only `spe_dc` is saved here -- `spe_epi_ref` (the reference
epithelial cells used in Sections 8–9) is a plotting-only object and
is never written to disk or merged in.