---
title: "DC/mregDC Subclustering — Run5 Cluster 4 (DC/mregDC vs. AT2 Epithelial Check)"
author: "Tazein Shah"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(42)
```

## Overview

Extract the DC/mregDC cluster (cluster 4) from the annotated run5 SPE
(`h5_cluster_annot.rds`) and recluster on the existing Harmony
embedding (Louvain, k = 20). The panel has no surfactant-protein
marker, so AT2 identity can't be confirmed directly -- the question
here is resolved indirectly: do subclusters show a clean DC/mregDC
signature (CD11c/HLADRB/LAMP3/CCR7/CD141/CD1c) with epithelial markers
riding along as spillover, or does a subset instead show a
predominantly epithelial signature (EpCAM/CD117/TTF1orCD16) with DC
markers as the weaker, spillover-level signal -- which would point to
misclustered AT2-like epithelial cells rather than true DC/mregDC.

---

## 1. Load Libraries

```{r libraries}
library(imcRtools)
library(scran)
library(bluster)
library(BiocParallel)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(scater)
library(patchwork)
library(RColorBrewer)
library(pheatmap)
```

---

## 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

spe_path <- file.path(analysis_path, "final_cell_annotation", "final_broad_cluster", "h5_cluster_annot.rds")

dc_clustering_path <- file.path(analysis_path, "final_cell_annotation", "subclustering", "dc_mregdc")

out_path <- file.path(dc_clustering_path, "dc_mregdc_subclustering.rds")
```

---

## 3. Load Annotated SPE

```{r load-data}
spe <- readRDS(spe_path)

print(table(spe$broad_clusters))
```

---

## 4. Extract DC/mregDC Cluster

```{r subset}
dc_cluster <- "4"

spe_dc <- spe[, as.character(spe$broad_clusters) %in% dc_cluster]

cat("Cells in DC/mregDC subset:", ncol(spe_dc), "\n")
```

---

## 5. Recluster: Louvain on Existing Harmony Embedding (k = 20)

```{r recluster}
snn_graph <- buildSNNGraph(spe_dc, use.dimred = "Harmony", k = 20,
                            BPPARAM = SerialParam())

clusters <- igraph::cluster_louvain(snn_graph)$membership

spe_dc$dc_subclusters <- factor(clusters)

cat("Number of subclusters:", nlevels(spe_dc$dc_subclusters), "\n")
print(table(spe_dc$dc_subclusters))
```

---

## 6. UMAP and Distribution on the Subset

```{r umap-sub}
set.seed(220225)
spe_dc <- runUMAP(spe_dc, dimred = "Harmony", name = "UMAP_sub")
```

```{r plot-patient-composition, fig.width=10, fig.height=5}
patient_counts <- as.data.frame(table(patient_id = spe_dc$patient_id))

p_patient_comp <- ggplot(patient_counts, aes(x = patient_id, y = Freq, fill = patient_id)) +
  geom_col() +
  labs(title = "DC/mregDC Subset: Cell Count by Patient",
       x = "Patient", y = "Cell Count") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
print(p_patient_comp)

cat("DC/mregDC subset composition by patient:\n")
print(patient_counts[order(-patient_counts$Freq), ])
```

---

## 7. Marker Sets: DC/mregDC vs. Epithelial (AT2-like)

```{r markers}
# true DC/mregDC lineage markers
dc_markers <- c("CD11c", "HLADRB", "CD141", "CD1c", "LAMP3", "CCR7")

# epithelial markers -- if a subcluster is dominated by these instead,
# that's the AT2-like/misclustered-epithelial signal, not DC/mregDC
epithelial_markers <- c("EpCAM", "CD117", "TTF1orCD16")

# activation / functional state markers, useful either way
state_markers <- c("PD1", "PDL1", "TIM3", "CD86", "MKI67", "CD45RO")

# other-lineage spillover flags, to rule out non-DC/non-epithelial admixture
admixture_markers <- c("CD68", "CD14", "CD19")

dc_all_markers <- c(dc_markers, epithelial_markers, state_markers, admixture_markers)
dc_all_markers <- dc_all_markers[dc_all_markers %in% rownames(spe_dc)]

cat("Markers found:", paste(dc_all_markers, collapse = ", "), "\n")
```

---

## 8. Violin Plots per Subcluster

```{r make-violinplot}
make_violinplot <- function(spe, assay_name, cluster_col, marker_set, title, ncol = 3) {
  expr_mat   <- assay(spe, assay_name)[marker_set, , drop = FALSE]
  cluster_id <- spe[[cluster_col]]

  mean_expr <- sapply(levels(cluster_id), function(cl) {
    idx <- which(cluster_id == cl)
    rowMeans(expr_mat[, idx, drop = FALSE])
  })

  if (is.matrix(mean_expr) && nrow(mean_expr) > 1) {
    marker_order <- rownames(mean_expr)[hclust(dist(mean_expr))$order]
  } else {
    marker_order <- marker_set
  }

  plot_df <- as.data.frame(t(expr_mat)) |>
    mutate(cluster = cluster_id) |>
    pivot_longer(-cluster, names_to = "marker", values_to = "expr")

  plot_df$marker  <- factor(plot_df$marker, levels = marker_order)
  plot_df$cluster <- factor(plot_df$cluster, levels = levels(cluster_id))

  ggplot(plot_df, aes(x = cluster, y = expr, fill = cluster)) +
    geom_violin(scale = "width", trim = TRUE, linewidth = 0.2) +
    stat_summary(fun = median, geom = "point", size = 0.6, color = "black") +
    facet_wrap(~ marker, scales = "free_y", ncol = ncol) +
    labs(title = title, x = "Subcluster", y = "Z-score expression") +
    theme_bw(base_size = 11) +
    theme(
      axis.text.x      = element_text(angle = 0, hjust = 0.5, size = 9),
      strip.text        = element_text(face = "bold"),
      panel.grid.major.x = element_blank(),
      legend.position   = "none"
    )
}
```

```{r violin-dc-global, fig.width=10, fig.height=5}
p_violin_dc_global <- make_violinplot(spe_dc, "scaled_0100", "dc_subclusters",
                                      dc_markers,
                                      "DC/mregDC Subset: DC Lineage Markers per Subcluster (Global Z-score)",
                                      ncol = 3)
print(p_violin_dc_global)
```

```{r violin-epithelial-global, fig.width=10, fig.height=5}
p_violin_epi_global <- make_violinplot(spe_dc, "scaled_0100", "dc_subclusters",
                                       epithelial_markers,
                                       "DC/mregDC Subset: Epithelial (AT2-like) Markers per Subcluster (Global Z-score)",
                                       ncol = 3)
print(p_violin_epi_global)
```

```{r violin-state-global, fig.width=12, fig.height=8}
p_violin_state_global <- make_violinplot(spe_dc, "scaled_0100", "dc_subclusters",
                                         state_markers,
                                         "DC/mregDC Subset: State Markers per Subcluster (Global Z-score)",
                                         ncol = 3)
print(p_violin_state_global)
```

```{r violin-admixture-global, fig.width=10, fig.height=5}
p_violin_admixture_global <- make_violinplot(spe_dc, "scaled_0100", "dc_subclusters",
                                             admixture_markers,
                                             "DC/mregDC Subset: Other-Lineage Admixture Markers per Subcluster (Global Z-score)",
                                             ncol = 3)
print(p_violin_admixture_global)
```

Subset-relative renormalization: since `scaled_0100` was capped and
Z-scored over the full run5 dataset, this recomputes the same
cap+Z-score recipe from raw counts using only this DC/mregDC subset's
own distribution -- one normalization layer, just a different
denominator.

```{r renorm-subset}
cap_99 <- function(mat) {
  t(apply(mat, 1, function(x) {
    q99 <- quantile(x, 0.99)
    pmin(x, q99)
  }))
}

zscore_rows <- function(mat) {
  t(scale(t(mat)))
}

mat_sub_0100 <- counts(spe_dc) |> cap_99() |> zscore_rows()
assay(spe_dc, "scaled_0100_sub") <- mat_sub_0100
```

```{r violin-dc-subset, fig.width=10, fig.height=5}
p_violin_dc_sub <- make_violinplot(spe_dc, "scaled_0100_sub", "dc_subclusters",
                                   dc_markers,
                                   "DC/mregDC Subset: DC Lineage Markers per Subcluster (Subset-Renormalized Z-score)",
                                   ncol = 3)
print(p_violin_dc_sub)
```

```{r violin-epithelial-subset, fig.width=10, fig.height=5}
p_violin_epi_sub <- make_violinplot(spe_dc, "scaled_0100_sub", "dc_subclusters",
                                    epithelial_markers,
                                    "DC/mregDC Subset: Epithelial (AT2-like) Markers per Subcluster (Subset-Renormalized Z-score)",
                                    ncol = 3)
print(p_violin_epi_sub)
```

```{r violin-state-subset, fig.width=12, fig.height=8}
p_violin_state_sub <- make_violinplot(spe_dc, "scaled_0100_sub", "dc_subclusters",
                                      state_markers,
                                      "DC/mregDC Subset: State Markers per Subcluster (Subset-Renormalized Z-score)",
                                      ncol = 3)
print(p_violin_state_sub)
```

```{r violin-admixture-subset, fig.width=10, fig.height=5}
p_violin_admixture_sub <- make_violinplot(spe_dc, "scaled_0100_sub", "dc_subclusters",
                                          admixture_markers,
                                          "DC/mregDC Subset: Other-Lineage Admixture Markers per Subcluster (Subset-Renormalized Z-score)",
                                          ncol = 3)
print(p_violin_admixture_sub)
```

---

## 9. Heatmap: DC vs. Epithelial Signal per Subcluster

```{r dc-vs-epi-heatmap-matrix}
heatmap_markers <- c(dc_markers, epithelial_markers)
heatmap_markers <- heatmap_markers[heatmap_markers %in% rownames(spe_dc)]

expr_scaled <- assay(spe_dc, "scaled_0100")[heatmap_markers, , drop = FALSE]
cluster_id  <- spe_dc$dc_subclusters

mean_expr_mat <- sapply(levels(cluster_id), function(cl) {
  idx <- which(cluster_id == cl)
  rowMeans(expr_scaled[, idx, drop = FALSE])
})

# keep DC markers and epithelial markers grouped as blocks rather than
# clustering rows freely, so the DC-vs-epithelial split stays visually clear
row_order <- heatmap_markers
mean_expr_mat <- mean_expr_mat[row_order, , drop = FALSE]

print(round(mean_expr_mat, 2))
```

```{r dc-vs-epi-heatmap-plot, fig.width=8, fig.height=6}
n_dc_markers <- sum(rownames(mean_expr_mat) %in% dc_markers)

p_heatmap <- pheatmap(
  mean_expr_mat,
  scale = "row",  # rescale each marker across subclusters so LAMP3's
                  # magnitude doesn't compress every other marker's range
  color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  gaps_row = n_dc_markers,  # visual break between DC block (top) and Epithelial block (bottom)
  display_numbers = TRUE,
  number_format = "%.2f",
  number_color = "black",
  fontsize_number = 8,
  main = "DC/mregDC Subset: Mean DC vs. Epithelial Marker Expression per Subcluster",
  fontsize = 10,
  fontsize_row = 9,
  fontsize_col = 9,
  border_color = NA,
  silent = TRUE
)

grid::grid.newpage()
grid::grid.draw(p_heatmap$gtable)
grid::grid.text("Row Z-score", x = 0.92, y = 0.85,
                 gp = grid::gpar(fontsize = 9, fontface = "bold"))
```

A subcluster that lights up red/blue consistently across the DC block
and stays flat across the Epithelial block supports true DC/mregDC. A
subcluster with the pattern flipped -- flat or negative across DC
markers, strongly positive across EpCAM/CD117/TTF1orCD16 -- is the
AT2-like/misclustered-epithelial signal.

---

## 10. Manual Subcluster Annotation

```{r annotate-sub, echo=TRUE}
dc_subannotation <- c(
  "5" = "DC/mregDC",
  "3" = "DC/mregDC (Epi Adj)",
  "1" = "DC/mregDC Weak",
  "2" = "DC/mregDC Weak",
  "4" = "DC/mregDC (LAMP3-)"
)

if (length(dc_subannotation) > 0 && !"PLACEHOLDER" %in% dc_subannotation) {
  spe_dc$sub_celltype <- dc_subannotation[as.character(spe_dc$dc_subclusters)]
  cat("Subcluster annotation applied.\n")
  print(table(spe_dc$sub_celltype, useNA = "ifany"))
} else {
  cat("No subcluster annotation yet. Fill in dc_subannotation above using the\n")
  cat("violin plots (Section 8) and the DC vs. Epithelial heatmap (Section 9).\n")
}
```

---

## 11. UMAP Colored by Subcluster Cell Type

```{r umap-sub-celltype, fig.width=10, fig.height=5}
if ("sub_celltype" %in% colnames(colData(spe_dc))) {

  spe_dc$sub_celltype <- factor(spe_dc$sub_celltype, levels = unique(dc_subannotation))

  p_sub_celltype <- dittoDimPlot(spe_dc, var = "sub_celltype",
                                 reduction.use = "UMAP_sub",
                                 size = 0.4) +
    ggtitle("DC/mregDC Subset: UMAP colored by Subcluster Cell Type") +
    theme(legend.position = "right")
  print(p_sub_celltype)

} else {
  cat("Run the annotate-sub chunk first to apply annotations.\n")
}
```

---

## 12. Cell Type Composition by Patient

```{r celltype-patient-composition, fig.width=14, fig.height=8}
if ("sub_celltype" %in% colnames(colData(spe_dc))) {

  celltype_patient_counts <- as.data.frame(
    table(patient_id = spe_dc$patient_id,
          cell_type  = spe_dc$sub_celltype)
  )

  p_celltype_patient <- ggplot(celltype_patient_counts, aes(x = patient_id, y = Freq, fill = patient_id)) +
    geom_col() +
    facet_wrap(~ cell_type, scales = "free", ncol = 2,
               labeller = labeller(cell_type = label_wrap_gen(width = 35))) +
    labs(title = "DC/mregDC Subset: Cell Type Composition by Patient",
         x = "Patient", y = "Cell Count") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
          strip.text = element_text(face = "bold", size = 8),
          legend.position = "none")

  print(p_celltype_patient)

  celltype_patient_plot_path <- file.path(dc_clustering_path, "celltype_composition_by_patient.png")
  ggsave(celltype_patient_plot_path, p_celltype_patient, width = 14, height = 8, dpi = 300)
  cat("Saved full-size plot to:", celltype_patient_plot_path, "\n")

} else {
  cat("Run the annotate-sub chunk first to apply annotations.\n")
}
```

---

## 13. Save Subclustered/Annotated Object

```{r save}
dir.create(dirname(out_path), recursive = TRUE, showWarnings = FALSE)
saveRDS(spe_dc, out_path)
cat("Saved DC/mregDC subcluster object to:", out_path, "\n")
```
