---
title: "Final Cell Type Annotation Merge"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

Final merge of all subclustering results (myeloid, cluster 15, B cell/plasma)
into a single annotated `spe` object. Strips tracking-number prefixes from
labels, excludes confirmed low-quality/artifact clusters, and validates the
final annotation with a heatmap and violin plots before saving.

This does NOT re-derive any of the subclustering QC itself (myeloid,
cluster 15, B cell/plasma) — those were already done separately and are
documented in their own notebooks. This notebook only combines the finalized
outputs.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# 1. Load Libraries

```{r libraries}
.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(tidyverse)
library(SingleCellExperiment)
library(SpatialExperiment)
library(scater)
```

# 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

spe_path <- file.path(analysis_path, "final_cell_annotation", "h5_cluster_annot.rds")

subclustering_path <- file.path(analysis_path, "final_cell_annotation", "subclustering")

myeloid_path <- file.path(subclustering_path, "myeloid_cells", "myeloid_subcluster.rds")

cluster15_path <- file.path(subclustering_path, "cluster15", "cluster15_subcluster_clean.rds")
bcell_path     <- file.path(subclustering_path, "bcells", "bcell_subclustering_wo15.rds")

final_out_path <- file.path(analysis_path, "final_cell_annotation", "spe_final_celltype.rds")
```

# 3. Load Broad-Annotated SPE and Subclustering Results

```{r load-intermediates}
spe <- readRDS(spe_path)

myeloid_obj   <- readRDS(myeloid_path)
cluster15_obj <- readRDS(cluster15_path)
bcell_obj     <- readRDS(bcell_path)

cat("Broad-annotated cells:", ncol(spe), "\n")
cat("Myeloid subset cells:", ncol(myeloid_obj), "\n")
cat("Cluster 15 (clean) subset cells:", ncol(cluster15_obj), "\n")
cat("B cell/plasma (w/o cluster 15) subset cells:", ncol(bcell_obj), "\n")
```

# 4. Final Merge

Cluster 15 was already excluded from the B cell/plasma subclustering input
(rerun without cluster 15, confirmed stable without needing removal/
reclustering), and cluster 15's own low-quality/artifact subclusters were
separately removed and reclustered — so there is no overlap between these
three subclustering results, and no origin-based filtering is needed at
merge time (unlike the earlier version of this pipeline).

```{r final-merge}
spe$final_celltype <- as.character(spe$broad_celltype)

# Myeloid subcluster labels
spe$final_celltype[match(colnames(myeloid_obj), colnames(spe))] <- as.character(myeloid_obj$sub_celltype)

# Cluster 15 (post-removal, annotated) subcluster labels
spe$final_celltype[match(colnames(cluster15_obj), colnames(spe))] <- as.character(cluster15_obj$sub_celltype_clean)

# Cells from broad cluster 15 that were excluded during cluster 15 subclustering
# (subclusters 2, 4, 7 -- low quality / Patient20 artifact) never got a
# sub_celltype_clean label, since they aren't in cluster15_obj at all.
# Flag them explicitly so they get caught by the exclusion step below,
# rather than silently keeping their stale pre-subclustering broad label.
cluster15_removed_cells <- setdiff(
  colnames(spe)[spe$broad_clusters == "15"],
  colnames(cluster15_obj)
)
spe$final_celltype[colnames(spe) %in% cluster15_removed_cells] <- "Unresolved/Low Quality"

# B cell/plasma (rerun without cluster 15) subcluster labels
spe$final_celltype[match(colnames(bcell_obj), colnames(spe))] <- as.character(bcell_obj$sub_celltype)

spe$final_celltype <- factor(spe$final_celltype)

cat("Cell types after merge:", nlevels(spe$final_celltype), "\n")
cat("NAs:", sum(is.na(spe$final_celltype)), "\n")
```


# 5. Strip Tracking-Number Prefixes from Labels

The leading numbers (e.g. `"10. Memory T/Plasma Cell (CD45RO+)"`) were only
ever meant as an internal bookkeeping aid during annotation, but they
interfere with downstream niche annotation, so they're removed here at the
source before saving the final object.

```{r strip-label-numbers}
spe$final_celltype <- factor(gsub("^\\d+\\.\\s*", "", as.character(spe$final_celltype)))

# Sanity check: confirm no numbers remain, no unintended matches
levels(spe$final_celltype)
```

```{r}
compare_markers <- c("FoxP3", "CD4", "CD25", "CD45RO", "TIM3", "PD1", "CD138")  # adjust to your panel
compare_markers <- compare_markers[compare_markers %in% rownames(spe)]

mean_expr_compare <- sapply(c("Treg", "Treg (FoxP3+)"), function(ct) {
  idx <- which(spe$final_celltype == ct)
  rowMeans(assay(spe, "scaled_0100")[compare_markers, idx, drop = FALSE])
})

round(mean_expr_compare, 2)
```
Note: Treg and Treg (FoxP3+) are distinct populations, not duplicates — confirmed via marker comparison (CD4/CD45RO/TIM3/PD1/CD138 all elevated in the latter, flat in the former).

```{r}
compare_markers_memory <- c("CD45RO", "CD4", "TIM3", "PD1", "CD138", "FoxP3")
compare_markers_memory <- compare_markers_memory[compare_markers_memory %in% rownames(spe)]

mean_expr_memory_compare <- sapply(c("Memory T Cell", "Memory T Cell (CD45RO+)"), function(ct) {
  idx <- which(spe$final_celltype == ct)
  rowMeans(assay(spe, "scaled_0100")[compare_markers_memory, idx, drop = FALSE])
})

round(mean_expr_memory_compare, 2)
```

Same story with Memory T Cell and Memory T Cell (CD45RO+).

# 6. Exclude Confirmed Low-Quality/Artifact Clusters

- **Broad cluster 17**: confirmed low quality at the broad level.
- **"Unresolved/Technical Artifact (Patient20)"**: confirmed technical
  artifact from the cluster 15 rerun — CD19+ signal concentrated in
  Patient20_425/519 (51% of cells, ~4x higher CD19 than other patients),
  not explained by generalized intensity. Bleeds into some
  originally-labeled Memory T Cell (CD45RO+) cells as well; excluded here
  rather than treated as a real cell type.
- **"Unresolved/Low Quality"**: generic low-signal category from earlier
  subclustering rounds (flat/negative across every marker).

Any tiny (< 3 cell) leftover categories from prior merge-boundary conflicts
should no longer occur, since cluster 15 was excluded from the B cell/plasma
subclustering input from the start this time — but the table in chunk 7
below is worth a visual scan to confirm nothing unexpectedly small slipped
through regardless.

```{r exclude-low-quality}
n_before <- ncol(spe)

spe <- spe[, spe$broad_clusters != "17"]

low_quality_labels <- c(
  "Unresolved/Low Quality",
  "Unresolved/Technical Artifact (Patient20)"
)
spe <- spe[, !(spe$final_celltype %in% low_quality_labels)]

n_after <- ncol(spe)

cat("Cells before exclusion:", n_before, "\n")
cat("Cells after excluding low-quality/artifact clusters:", n_after, "\n")
cat("Cells removed:", n_before - n_after,
    sprintf("(%.1f%% of dataset)\n", 100 * (n_before - n_after) / n_before))

spe$final_celltype <- droplevels(spe$final_celltype)
```

# 7. Final Cell Type Counts

```{r final-celltype-table}
spe$final_celltype |>
  table(useNA = "ifany") |>
  as.data.frame() |>
  setNames(c("CellType", "Count")) |>
  arrange(desc(Count)) |>
  knitr::kable(caption = "Final cell type counts after merge and QC exclusion")
```

# 8. Validation: Heatmap of Marker Expression by Final Cell Type

Confirm each final cell type's name-implied marker(s) are actually elevated
in that group relative to others, before treating this object as the shared
source of truth.

```{r validate-heatmap, fig.width=14, fig.height=10}
all_markers <- rownames(spe)

celltype_levels <- levels(spe$final_celltype)

mean_expr <- sapply(celltype_levels, function(ct) {
  idx <- which(spe$final_celltype == ct)
  rowMeans(assay(spe, "scaled_0100")[all_markers, idx, drop = FALSE])
})

marker_order <- all_markers[hclust(dist(mean_expr))$order]

celltype_sizes <- table(spe$final_celltype)
celltype_labels <- paste0(celltype_levels, "\n(n=", celltype_sizes[celltype_levels], ")")
names(celltype_labels) <- celltype_levels

heatmap_df <- as.data.frame(mean_expr) |>
  rownames_to_column("marker") |>
  pivot_longer(-marker, names_to = "celltype", values_to = "mean_expr")

heatmap_df$marker   <- factor(heatmap_df$marker, levels = marker_order)
heatmap_df$celltype <- factor(heatmap_df$celltype, levels = celltype_levels)

ggplot(heatmap_df, aes(x = celltype, y = marker, fill = mean_expr)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, name = "Mean\nZ-score") +
  scale_x_discrete(labels = celltype_labels) +
  theme_bw(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        panel.grid = element_blank()) +
  labs(x = "Final Cell Type (n)", y = "Marker",
       title = "Final Annotation Validation: Mean Marker Expression by Cell Type")
```
### Check EpCAM
```{r}
mean_expr["EpCAM", c("Epithelial (EpCAM+)", "Tumor Epithelial (TTF1++)", "Tumor Epithelial (Neuroendocrine)")]
```

### Check Plasma Cell, HLADRB-high
```{r}
mean_expr[c("HLADRB", "CD19", "CD138"), "B Cell/Plasma Cell, HLADRB-high (CD19+CD138+HLADRB+)"]
```

CD19 and CD138 are more elevated than HLADRB, which is not consistent with this cell type label. Might be worth renaming to Transitional B Cell/Plasmablast, HLADRB+ (CD19+CD138+)

```{r auto-flag-label-mismatches}
# For each cell type, find its top 3 markers by mean expression, and check
# whether the marker(s) named in the label actually show up among them
top_markers_per_celltype <- apply(mean_expr, 2, function(col) {
  names(sort(col, decreasing = TRUE))[1:3]
})

top_markers_df <- as.data.frame(t(top_markers_per_celltype)) |>
  rownames_to_column("CellType") |>
  setNames(c("CellType", "Top1", "Top2", "Top3"))

knitr::kable(top_markers_df, caption = "Top 3 markers per final cell type (by mean Z-score)")
```

# 9. Validation: Violin Plots (Targeted Marker Check)

Focused check on a few key markers, worth adjusting to whichever cell
types/markers the heatmap above flags as ambiguous.

```{r validate-violin-1, fig.width=12, fig.height=8}
library(RColorBrewer)

markers_check_1 <- c("FoxP3", "CD19", "CD138", "EpCAM", "CD117", "SMA")
markers_check_1 <- markers_check_1[markers_check_1 %in% rownames(spe)]

plot_df_1 <- as.data.frame(t(assay(spe, "scaled_0100")[markers_check_1, ])) |>
  mutate(CellType = spe$final_celltype) |>
  pivot_longer(-CellType, names_to = "marker", values_to = "expression")

n_celltypes <- length(unique(plot_df_1$CellType))
celltype_colors <- colorRampPalette(brewer.pal(12, "Paired"))(n_celltypes)

ggplot(plot_df_1, aes(x = CellType, y = expression, fill = CellType)) +
  geom_violin(scale = "width", trim = TRUE) +
  scale_fill_manual(values = celltype_colors) +
  facet_wrap(~ marker, scales = "free_y", ncol = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
        legend.position = "none") +
  labs(x = NULL, y = "Z-score expression",
       title = "Marker Expression Validation (1 of 2)")
```

```{r validate-violin-2, fig.width=12, fig.height=9}
markers_check_2 <- c("HLADRB", "CD38", "CD8a", "CD11c", "CD45RA", "CD39", "TIM3")
markers_check_2 <- markers_check_2[markers_check_2 %in% rownames(spe)]

plot_df_2 <- as.data.frame(t(assay(spe, "scaled_0100")[markers_check_2, ])) |>
  mutate(CellType = spe$final_celltype) |>
  pivot_longer(-CellType, names_to = "marker", values_to = "expression")

ggplot(plot_df_2, aes(x = CellType, y = expression, fill = CellType)) +
  geom_violin(scale = "width", trim = TRUE) +
  scale_fill_manual(values = celltype_colors) +
  facet_wrap(~ marker, scales = "free_y", ncol = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
        legend.position = "none") +
  labs(x = NULL, y = "Z-score expression",
       title = "Marker Expression Validation (2 of 2)")
```

# 10. Validation: Final UMAP

```{r final-umap, fig.width=12, fig.height=8}
library(dittoSeq)
library(RColorBrewer)

# Reuse existing UMAP if already computed; only recompute if missing
if (!("UMAP" %in% reducedDimNames(spe))) {
  set.seed(220225)
  spe <- runUMAP(spe, dimred = "Harmony", name = "UMAP")}

n_celltypes <- nlevels(spe$final_celltype)
celltype_colors <- colorRampPalette(brewer.pal(12, "Paired"))(n_celltypes)

p_final_umap <- dittoDimPlot(spe, var = "final_celltype",
                              reduction.use = "UMAP",
                              size = 0.3) +
  scale_color_manual(values = celltype_colors) +
  ggtitle("Final Cell Type Annotation: UMAP") +
  theme(legend.position = "right",
        legend.text = element_text(size = 7),
        legend.key.size = unit(0.4, "cm"))

print(p_final_umap)

# Save full-size to file
umap_out_path <- file.path(analysis_path, "final_cell_annotation", "final_celltype_umap.png")
ggsave(umap_out_path, p_final_umap, width = 14, height = 10, dpi = 300)
cat("Saved full-size UMAP to:", umap_out_path, "\n")
```

# 11. Save Finalized Object

```{r save-final}
saveRDS(spe, final_out_path)
cat("Saved final annotated object to:", final_out_path, "\n")
```

# Handoff Notes

- `spe_final_celltype.rds` has unnumbered cell type labels — any downstream
  code (sccomp exclusion lists, spatial graph exclusion lists, niche
  annotation) that references old numbered label strings needs to be
  updated to the new unnumbered versions.
- Excluded from this object: broad cluster 17, "Unresolved/Low Quality",
  and "Unresolved/Technical Artifact (Patient20)". No small leftover
  categories are expected this time, since cluster 15 no longer overlaps
  with the B cell/plasma subclustering input — confirmed by the table in
  chunk 7.
- Flag for Campbell: Patient20_425/519 showed a recurring
  patient-specific CD19 technical artifact across both the cluster 15 and
  B cell/plasma subclustering pipelines — worth a dedicated sample-level QC
  pass on those two patients specifically.
