---
title: 'All Patients: Broad Cell Type Clustering — Harmony (no asinh)'
author: "Tazein Shah"
date: "06-29-2026"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
subtitle: 20 Clusters
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)
set.seed(42)
```

## Overview

Broad cell type clustering on the harmonized combined SPE
(`spe_combined_harmony_r3.rds`). This Harmony run was computed on raw
counts directly (PCA on `counts`, no asinh, no cap/Z-score) before
Harmony. Clustering runs on the Harmony corrected embeddings
(analogous to Xi et al.'s PCA on top 2,000 variable features). The
cap + Z-score assay used here is for visualization only and is
computed fresh from raw counts; it was not part of how Harmony was
run. After inspecting the dot plot and UMAP, broad cell type labels
are assigned manually via a lookup table.

---

## 1. Load Libraries

```{r libraries}
library(imcRtools)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(scater)
library(scran)
library(bluster)
library(patchwork)
library(RColorBrewer)
```

---

## 2. Load Harmonized Data

```{r load-data}
spe_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/batch_correction/spe_combined_harmony_r3.rds"

spe <- readRDS(spe_path)

cat("Cells:", ncol(spe), "\n")
cat("Patients:", n_distinct(spe$patient_id), "\n")
cat("ROIs:", n_distinct(spe$sample_id), "\n")
cat("Channels:", nrow(spe), "\n")
cat("Reduced dims available:", paste(reducedDimNames(spe), collapse = ", "), "\n")
cat("Assays available:", paste(assayNames(spe), collapse = ", "), "\n")
```

---

## 3. Marker Sets

```{r markers}
# core markers for broad annotation and UMAP overlays
core_markers <- c("SMA", "CD4", "CD8a", "CD19", "CD15",
                  "CD14", "CD86", "LAMP3", "CD163", "EpCAM")
core_markers <- core_markers[core_markers %in% rownames(spe)]
cat("Core markers found:", paste(core_markers, collapse = ", "), "\n")
cat("Missing:", paste(setdiff(
  c("SMA", "CD4", "CD8a", "CD19", "CD15", "CD14", "CD86", "LAMP3", "CD163", "EpCAM"),
  rownames(spe)), collapse = ", "), "\n")

# full panel, used to investigate clusters that don't separate on core markers
full_markers <- c(
  "SMA", "CD4", "CD8a", "CD19", "CD15", "CD14", "CD86", "LAMP3",
  "Ki67", "MKI67", "CD138", "PD1", "PDL1", "FoxP3", "CD68", "CD163",
  "TIM3", "CD141", "CD11c", "CD56", "CD117", "EpCAM", "CD38",
  "CD45RO", "LAG3", "CD1c", "GATA3", "TTF1orCD16"
)
full_markers <- full_markers[full_markers %in% rownames(spe)]
cat("Full panel markers found:", length(full_markers), "\n")
cat("Missing:", paste(setdiff(
  c("SMA", "CD4", "CD8a", "CD19", "CD15", "CD14", "CD86", "LAMP3",
    "Ki67", "MKI67", "CD138", "PD1", "PDL1", "FoxP3", "CD68", "CD163",
    "TIM3", "CD141", "CD11c", "CD56", "CD117", "EpCAM", "CD38",
    "CD45RO", "LAG3", "CD1c", "GATA3", "TTF1orCD16"),
  rownames(spe)), collapse = ", "), "\n")
```

---

## 4. Preprocessing: 99th Percentile Cap + Z-score

```{r helpers}
cap_99 <- function(mat) {
  t(apply(mat, 1, function(x) {
    q99 <- quantile(x, 0.99)
    pmin(x, q99)
  }))
}

zscore_rows <- function(mat) {
  t(scale(t(mat)))
}
```

```{r preprocess}
mat_0100 <- counts(spe) |> cap_99() |> zscore_rows()
assay(spe, "scaled_0100") <- mat_0100

cat("Assays in SPE:", paste(assayNames(spe), collapse = ", "), "\n")
```

---

## 5. Louvain Clustering on Harmony Embeddings

Clustering runs on the Harmony corrected embeddings, which serve the
same role as PCA on top variable features in the Xi et al. approach.
Resolution 1.0 as per the preprint's first pass. The clustered SPE is
saved immediately after so clustering does not need to rerun on every
render.

```{r cluster}
intermediate_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/clustering/harmony_20_clustered_intermediate.rds"

if (file.exists(intermediate_path)) {
  cat("Loading existing clustered SPE from intermediate file.\n")
  spe <- readRDS(intermediate_path)
  cat("Number of broad clusters:", nlevels(spe$broad_clusters), "\n")
  print(table(spe$broad_clusters))
} else {
  set.seed(220225)

  harmony_mat <- reducedDim(spe, "Harmony")
  cat("Harmony embedding dimensions:", dim(harmony_mat), "\n")

  g     <- buildSNNGraph(t(harmony_mat), k = 20, type = "rank")
  clust <- igraph::cluster_louvain(g)
  spe$broad_clusters <- factor(clust$membership)

  cat("Number of broad clusters:", nlevels(spe$broad_clusters), "\n")
  print(table(spe$broad_clusters))

  saveRDS(spe, intermediate_path)
  cat("Saved clustered intermediate SPE to:", intermediate_path, "\n")
}
```

---

## 6. UMAP on Harmony Embeddings

`UMAP_harmony` already exists in this object from the batch correction
step, so it is reused here instead of recomputed.

```{r umap}
if ("UMAP_harmony" %in% reducedDimNames(spe)) {
  cat("Reusing existing UMAP_harmony from batch correction step.\n")
} else {
  set.seed(220225)
  spe <- runUMAP(spe, dimred = "Harmony", name = "UMAP_harmony")
  cat("UMAP computed.\n")
}
```

---

## 7. Subset for Visualization

20,000 cells per ROI cap to keep plots manageable.

```{r subset}
set.seed(220225)
cells_per_roi  <- split(seq_len(ncol(spe)), spe$sample_id)
sampled_cells  <- unlist(lapply(cells_per_roi, function(idx) {
  sample(idx, min(20000, length(idx)))
}))
spe_sub <- spe[, sampled_cells]
cat("Cells in subset:", ncol(spe_sub), "\n")
```

---

## 8. UMAP Colored by Patient

```{r umap-patient, fig.width=10, fig.height=7}
n_patients    <- n_distinct(spe$patient_id)
patient_cols  <- setNames(
  colorRampPalette(brewer.pal(9, "Set1"))(n_patients),
  unique(spe$patient_id)
)

p_patient <- dittoDimPlot(spe_sub, var = "patient_id",
                          reduction.use = "UMAP_harmony", size = 0.2) +
  scale_color_manual(values = patient_cols) +
  ggtitle("UMAP colored by Patient ID (Harmony, no asinh)")
print(p_patient)
```

---

## 9. UMAP Colored by Disease State

Quick check, not the main focus. Useful to confirm Harmony correction
isn't accidentally erasing real biological signal tied to disease
state.

```{r umap-disease, fig.width=10, fig.height=7}
p_disease <- dittoDimPlot(spe_sub, var = "disease_state",
                          reduction.use = "UMAP_harmony", size = 0.2) +
  ggtitle("UMAP colored by Disease State (Harmony, no asinh)")
print(p_disease)
```

---

## 10. UMAP Colored by Broad Cluster

```{r umap-clusters, fig.width=10, fig.height=7}
p_clusters <- dittoDimPlot(spe_sub, var = "broad_clusters",
                           reduction.use = "UMAP_harmony",
                           size = 0.2, do.label = TRUE, labels.size = 3) +
  ggtitle("UMAP colored by Broad Louvain Cluster") +
  theme(legend.position = "right")
print(p_clusters)
```

---

## 11. UMAP Per Core Marker

Used to visually match clusters to known marker expression before
annotating.

```{r umap-markers, fig.width=8, fig.height=6}
for (marker in core_markers) {
  p <- dittoDimPlot(spe_sub, var = marker,
                    reduction.use = "UMAP_harmony",
                    assay = "scaled_0100", size = 0.2) +
    scale_color_viridis(name = marker) +
    ggtitle(paste("UMAP:", marker))
  print(p)
}
```

---

## 12. Dot Plot: Core Marker Expression per Cluster

The primary tool for annotating clusters. Each row is a marker, each
column is a cluster. Look for which clusters show high expression of
the core lineage markers.

```{r make-dotplot}
make_dotplot <- function(spe, assay_name, cluster_col, marker_set, title) {
  expr_mat   <- assay(spe, assay_name)[marker_set, ]
  cluster_id <- spe[[cluster_col]]

  mean_expr <- sapply(levels(cluster_id), function(cl) {
    idx <- which(cluster_id == cl)
    rowMeans(expr_mat[, idx, drop = FALSE])
  })

  global_med <- apply(expr_mat, 1, median)
  frac_expr  <- sapply(levels(cluster_id), function(cl) {
    idx <- which(cluster_id == cl)
    rowMeans(expr_mat[, idx, drop = FALSE] > global_med)
  })

  mean_df <- as.data.frame(mean_expr) |>
    rownames_to_column("marker") |>
    pivot_longer(-marker, names_to = "cluster", values_to = "mean_exprs")

  frac_df <- as.data.frame(frac_expr) |>
    rownames_to_column("marker") |>
    pivot_longer(-marker, names_to = "cluster", values_to = "frac_expressing")

  plot_df <- left_join(mean_df, frac_df, by = c("marker", "cluster"))
  plot_df$cluster <- factor(plot_df$cluster, levels = levels(cluster_id))

  marker_order <- rownames(mean_expr)[hclust(dist(mean_expr))$order]
  plot_df$marker <- factor(plot_df$marker, levels = marker_order)

  ggplot(plot_df, aes(x = cluster, y = marker)) +
    geom_point(aes(size = frac_expressing, color = mean_exprs)) +
    scale_color_viridis(option = "magma", name = "Mean\nZ-score expr") +
    scale_size_continuous(range = c(0.5, 6), name = "Fraction\nexpressing") +
    labs(title = title, x = "Cluster", y = "Marker") +
    theme_bw(base_size = 12) +
    theme(
      axis.text.x      = element_text(angle = 0, hjust = 0.5),
      axis.text.y      = element_text(size = 9),
      panel.grid.major = element_line(color = "grey90"),
      legend.position  = "right"
    )
}
```

```{r dotplot-core, fig.width=16, fig.height=6}
p_dot_core <- make_dotplot(spe, "scaled_0100", "broad_clusters",
                           core_markers,
                           "Harmony (no asinh): Core Marker Expression per Broad Louvain Cluster")
print(p_dot_core)
```

---

## 13. Full Panel Dot Plot for Low-Signal Clusters

Several clusters may not show a clear standout marker on the 10 core
markers above, this can happen for macrophage/myeloid populations
where markers like CD14, CD86, CD163 are broadly and continuously
expressed rather than sharply spiking in one cluster, or because the
cluster is actually separated by markers outside the core panel
(e.g. PD1, Ki67, FoxP3, CD45RO), since this Harmony run was computed
on raw counts across the full panel, not just the 10 core markers.
Borderline or low quality segmented cells are also possible. Use this
expanded dot plot to investigate clusters that look flat above before
calling them "Unknown."

```{r dotplot-full, fig.width=18, fig.height=10}
p_dot_full <- make_dotplot(spe, "scaled_0100", "broad_clusters",
                           full_markers,
                           "Harmony (no asinh): Full Panel Marker Expression per Broad Louvain Cluster")
print(p_dot_full)
```

---

## 14. Statistical Marker Discovery (findMarkers)

For clusters with no obvious standout marker on either dot plot,
`findMarkers()` ranks markers by how distinctly they separate that
cluster from all others, accounting for variance rather than just
mean difference. This can surface subtle but consistent shifts that
don't show up clearly in a Z-score dot plot averaged across thousands
of cells.

```{r findmarkers}
markers_de <- findMarkers(spe,
                          groups     = spe$broad_clusters,
                          assay.type = "scaled_0100",
                          subset.row = full_markers)

# save all clusters to CSV
all_markers <- lapply(levels(spe$broad_clusters), function(cl) {
  df <- as.data.frame(markers_de[[cl]])
  data.frame(
    cluster = cl,
    marker  = rownames(df),
    Top     = df$Top,
    FDR     = df$FDR,
    logFC   = df$summary.logFC
  )
})

all_markers_df <- do.call(rbind, all_markers)

write.csv(all_markers_df,
          "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/clustering/harmony_20_cluster_findmarkers.csv",
          row.names = FALSE)

# print top 5 positive markers per cluster
df <- read.csv("/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/clustering/harmony_20_cluster_findmarkers.csv")

# summary table: top positive marker per cluster at a glance
top1_summary <- do.call(rbind, lapply(levels(spe$broad_clusters), function(cl) {
  sub <- df[df$cluster == cl & df$logFC > 1.0, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  if (nrow(sub) == 0) {
    data.frame(cluster = cl, top_markers = "none above logFC 1.0")
  } else {
    data.frame(cluster = cl,
               top_markers = paste(head(sub$marker, 5), collapse = ", "))
  }
}))

cat("Top positive markers (logFC > 1.0) per cluster:\n")
print(knitr::kable(top1_summary, col.names = c("Cluster", "Top Positive Markers")))

cat("\n\nDetailed tables:\n")

# detailed table per cluster, positive logFC only
for (cl in levels(spe$broad_clusters)) {
  sub <- df[df$cluster == cl & df$Top <= 5 & df$logFC > 1.0, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  if (nrow(sub) == 0) {
    cat("Cluster", cl, ": no markers with logFC > 1.0 in top 5\n\n")
  } else {
    print(knitr::kable(sub[, c("marker", "Top", "FDR", "logFC")],
                       caption = paste("Cluster", cl),
                       digits  = 3))
  }
}
```
--- 

## 14b. Trying to Find "Unknown" Clusters
```{r}
for (cl in c("1", "10", "14", "16", "17", "20")) {
  sub <- df[df$cluster == cl & df$logFC > 0.5, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  print(knitr::kable(sub[, c("marker", "Top", "FDR", "logFC")],
                     caption = paste("Cluster", cl, "(logFC > 0.5)"),
                     digits  = 3))
}
```
---

## 15. Diagnostic: Area and Patient Composition for Ambiguous Clusters

Clusters 10 and 17 are candidates for removal before subclustering.
Cluster 10 shows high expression across completely unrelated lineage
markers (doublet signature). Cluster 17 shows no signal above logFC
0.5 across the full panel. Cell area and patient composition are
checked here to build evidence before removing them.

```{r diagnostic-unknown, fig.width=10, fig.height=5}
ambiguous_clusters <- c("1", "10", "14", "16", "17", "20")

# cell area boxplot
df_area <- data.frame(
  area    = spe$area,
  cluster = as.character(spe$broad_clusters)
)

# reference: median area across all annotated clusters
df_area_ref <- df_area[!df_area$cluster %in% ambiguous_clusters, ]
ref_median   <- median(df_area_ref$area)

p_area <- ggplot(df_area[df_area$cluster %in% ambiguous_clusters, ],
                 aes(x = cluster, y = area, fill = cluster)) +
  geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.3) +
  geom_hline(yintercept = ref_median, linetype = "dashed", color = "grey40") +
  annotate("text", x = 0.6, y = ref_median + 5,
           label = "median (annotated clusters)", size = 3, color = "grey40") +
  labs(title = "Cell Area Distribution for Ambiguous Clusters",
       x = "Cluster", y = "Cell Area (pixels)") +
  theme_bw() +
  theme(legend.position = "none")
print(p_area)
```

```{r diagnostic-patient, fig.width=12, fig.height=5}
# patient composition per ambiguous cluster
comp_df <- as.data.frame(
  table(cluster   = as.character(spe$broad_clusters),
        patient   = spe$patient_id)
)
comp_df <- comp_df[comp_df$cluster %in% ambiguous_clusters & comp_df$Freq > 0, ]

p_comp <- ggplot(comp_df, aes(x = patient, y = Freq, fill = patient)) +
  geom_col() +
  facet_wrap(~ cluster, scales = "free_y", nrow = 2) +
  labs(title = "Patient Composition for Ambiguous Clusters",
       x = "Patient", y = "Cell Count") +
  theme_bw() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        legend.position = "none")
print(p_comp)
```

Interpretation:
- Cluster dominated by 1-2 patients → likely batch artifact, candidate for removal
- Cluster with abnormally small area → likely debris or segmentation artifact
- Cluster with abnormally large area → likely merged/oversegmented cells (doublets)

## 15. Manual Broad Annotation

After inspecting the dot plots and `findMarkers()` output above, fill
in the lookup table below. Map each cluster number to one of:

- Tumor / Epithelial (EpCAM)
- Fibroblast / Stroma (SMA)
- CD4+ T Cell (CD4)
- CD8+ T Cell (CD8a)
- B Cell (CD19)
- Macrophage (CD14, CD86)
- M-MDSC (CD14, CD163)
- PMN-MDSC / Neutrophil (CD15)
- DC / mregDC (LAMP3)
- Unknown (low marker expression, revisit in subclustering)

```{r annotate}
broad_annotation <- c(
  "1"  = "cDC1",
  "2"  = "Fibroblast / Stroma",
  "3"  = "Plasmablast / Activated B Cell",
  "4"  = "DC / mregDC",
  "5"  = "PMN-MDSC",
  "6"  = "CD8+ T Cell",
  "7"  = "CD4+ T Cell",
  "8"  = "Memory T Cell",
  "9"  = "Macrophage",
  "10" = "Transitional Myeloid / Unresolved",
  "11" = "Treg",
  "12" = "Mast Cell",
  "13" = "Epithelial",
  "14" = "M2 Macrophage",
  "15" = "Plasma Cell",
  "16" = "Epithelial",
  "17" = "Unresolved Stromal / Other",
  "18" = "Tumor Epithelial",
  "19" = "Proliferating Cells",
  "20" = "Exhausted T Cell"
)

if (length(broad_annotation) > 0) {
  spe$broad_celltype <- broad_annotation[as.character(spe$broad_clusters)]
  cat("Annotation applied.\n")
  print(table(spe$broad_celltype, useNA = "ifany"))
} else {
  cat("No annotation provided yet. Fill in broad_annotation above after inspecting plots.\n")
}


if (length(broad_annotation) > 0) {
  spe$broad_celltype <- broad_annotation[as.character(spe$broad_clusters)]
  cat("Annotation applied.\n")
  print(table(spe$broad_celltype))
} else {
  cat("No annotation provided yet. Fill in broad_annotation above after inspecting plots.\n")
}
```
---

## 16. UMAP Colored by Broad Cell Type

Run this chunk after filling in the annotation above.

```{r umap-celltype, fig.width=10, fig.height=7}
if ("broad_celltype" %in% colnames(colData(spe))) {
  spe_sub$broad_celltype <- spe$broad_celltype[match(colnames(spe_sub), colnames(spe))]

  p_celltype <- dittoDimPlot(spe_sub, var = "broad_celltype",
                             reduction.use = "UMAP_harmony",
                             size = 0.2, do.label = TRUE, labels.size = 3) +
    ggtitle("UMAP colored by Broad Cell Type") +
    theme(legend.position = "right")
  print(p_celltype)
} else {
  cat("Run chunk 15 first to apply annotations.\n")
}
```

---

## 17. Save Annotated SPE

```{r save}
out_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/clustering/k20_clusters/harmony_20_cluster_annot.rds"

if ("broad_celltype" %in% colnames(colData(spe))) {
  saveRDS(spe, out_path)
  cat("Saved annotated SPE to:", out_path, "\n")
} else {
  cat("Annotation not yet applied. Fill in chunk 15 before saving.\n")}
```

---

## Notes

- Input `spe_combined_harmony_r3.rds` has Harmony embeddings computed
  via PCA on raw `counts`, no asinh and no cap/Z-score; the cap +
  Z-score assay used in this notebook is for visualization only and
  was not part of how Harmony was run
- Because raw counts can have a long right tail, this PCA/Harmony run
  is more exposed to outlier-driven distortion than the asinh or
  cap+Z-score versions; worth keeping in mind when comparing UMAPs
  and cluster quality across the three conditions
- Preprocessing: counts already 0-100 filtered; 99th percentile cap + Z-score applied and stored as `scaled_0100` assay
- Clustering uses Harmony corrected embeddings (`reducedDim(spe, "Harmony")`),
  which play the same role as PCA on top 2,000 variable features in Xi et al.
- Dot plot and UMAP marker overlays use `scaled_0100` assay
- `igraph::cluster_louvain` does not accept a resolution parameter; if the
  number of clusters is too coarse or too fine, switch to
  `igraph::cluster_leiden(g, resolution_parameter = 1.0)` from the `leiden`
  package
- Doublet-like clusters (hybrid marker expression across lineages) should be
  assigned "Unknown" and flagged for removal before subclustering
- Note: rerunning this notebook may shuffle Louvain cluster numbers; if you
  need to preserve a finalized annotation across reruns, save the clustered
  SPE before annotating and load it separately
