---
title: 'All Patients: Violin Plots per Cluster — Harmony (no asinh)'
author: "Vaidehi Gupta"
date: "07-06-2026"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
subtitle: 20 Clusters — Full Panel Marker Distributions
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 14, fig.height = 6)
set.seed(42)
```

## Overview

The dot plot from `cell_pheno_20.Rmd` shows the *mean* Z-score
expression per marker per cluster, which can hide whether a marker is
uniformly expressed across a cluster or only in a subset of cells
within it. That distinction matters because two markers can both show
an elevated mean in the same cluster without ever being expressed in
the same cell (e.g. the FoxP3 / LAMP3 concern raised at lab meeting).

This notebook builds one violin plot per cluster: markers on the
x-axis, and the full distribution of per-cell Z-score expression
(all cells belonging to that cluster) on the y-axis. Reading the
shape of each marker's violin (not just its mean) tells you:

- **Tall, narrow, high violin** -> marker cleanly defines this
  cluster's identity (most cells agree)
- **Squashed, low violin near zero** -> marker is not really "on" in
  this cluster; likely background/noise or an irrelevant marker
- **Two-humped (bimodal) violin** -> the cluster is probably a mix of
  two subpopulations, one positive and one negative for that marker,
  rather than one cell type genuinely co-expressing both

Three clusters are specifically flagged going in, based on the dot
plot review:

- **Cluster 4** ("DC / mregDC") -- LAMP3 is very high, but CD117,
  MKI67, PD1, and EpCAM are also co-elevated in the same column. That
  combination doesn't correspond to one coherent lineage, so this is
  a candidate for a doublet/hybrid population.
- **Cluster 9** ("Macrophage") -- broad, overlapping elevation across
  many myeloid markers (CD11c, TIM3, CD163, CD14, CD68, CD86, LAG3,
  PDL1, EpCAM, CD1c, TTF1orCD16, MKI67, LAMP3). Myeloid populations
  often show this continuous pattern, so the violin shapes here help
  tell "real transitional myeloid state" apart from "multiple merged
  populations."
- **Cluster 17** ("Unresolved Stromal / Other") -- flat in the dot
  plot, no marker rises above baseline, and this matches the 10-cluster
  `findMarkers()` result (no marker above logFC 0.5). Confirming the
  violin shapes are flat here supports excluding this cluster before
  subclustering.

---

## 1. Load Libraries

```{r libraries}
.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(imcRtools)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(scater)
library(scran)
library(patchwork)
library(RColorBrewer)
```

---

## 2. Load Annotated Clustered Data

```{r load-data}
# harmony_20_cluster.rds is the fully annotated output of cell_pheno_20.Rmd

spe_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/clustering/k20_clusters/harmony_20_cluster.rds"

spe <- readRDS(spe_path)

cat("Cells:", ncol(spe), "\n")
cat("Patients:", n_distinct(spe$patient_id), "\n")
cat("Number of clusters:", nlevels(spe$broad_clusters), "\n")
cat("Assays available:", paste(assayNames(spe), collapse = ", "), "\n")
cat("Has broad_celltype annotation:", "broad_celltype" %in% colnames(colData(spe)), "\n")
```

---

## 3. Marker Panel

Using the full marker panel, same list used for the dot plot / findMarkers in cell_pheno_20.Rmd
Keeping this identical means the violin marker order below is directly comparable to the dot plot.

```{r markers}

full_markers <- c(
  "SMA", "CD4", "CD8a", "CD19", "CD15", "CD14", "CD86", "LAMP3",
  "Ki67", "MKI67", "CD138", "PD1", "PDL1", "FoxP3", "CD68", "CD163",
  "TIM3", "CD141", "CD11c", "CD56", "CD117", "EpCAM", "CD38",
  "CD45RO", "LAG3", "CD1c", "GATA3", "TTF1orCD16"
)
full_markers <- full_markers[full_markers %in% rownames(spe)]

cat("Full panel markers found:", length(full_markers), "\n")
cat("Missing:", paste(setdiff(
  c("SMA", "CD4", "CD8a", "CD19", "CD15", "CD14", "CD86", "LAMP3",
    "Ki67", "MKI67", "CD138", "PD1", "PDL1", "FoxP3", "CD68", "CD163",
    "TIM3", "CD141", "CD11c", "CD56", "CD117", "EpCAM", "CD38",
    "CD45RO", "LAG3", "CD1c", "GATA3", "TTF1orCD16"),
  rownames(spe)), collapse = ", "), "\n")
```

---

## 4. Cluster -> Cell Type Lookup and Flagged Clusters

Pull the cluster -> cell type label mapping straight out of the annotated SPE, so plot titles stay in sync with cell_pheno_20.Rmd without needing to retype the lookup table here.

```{r lookup}

cluster_levels <- levels(spe$broad_clusters)

celltype_lookup <- setNames(
  spe$broad_celltype[match(cluster_levels, as.character(spe$broad_clusters))],
  cluster_levels
)

print(celltype_lookup)

# Clusters flagged from the dot plot review, with a short reason each.
# These get a highlighted subtitle on their violin plot so they're easy to spot when flipping through the PDF

flagged_clusters <- list(
  "4"  = "FLAGGED: LAMP3 co-elevated with CD117/MKI67/PD1/EpCAM -- check for doublet/hybrid",
  "9"  = "FLAGGED: broad overlapping myeloid marker elevation -- check merged vs. transitional state",
  "17" = "FLAGGED: no marker above baseline in dot plot -- candidate for exclusion"
)
```

---

## 5. Output Directory for Saved Plots

Individual PNGs and one combined multi-page PDF get saved here.

```{r out-dir}

out_dir <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/clustering/k20_clusters/violin_plots"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
cat("Saving plots to:", out_dir, "\n")
```

---

## 6. Violin Plot Function

This builds one violin plot for a single cluster: x-axis = marker, y-axis = per-cell Z-score expression (scaled_0100 assay) for every cell in that cluster. 

Markers are ordered by their mean expression within the cluster (highest first) so the "defining" markers for that cluster show up on the left.

Each violin is filled by that marker's own mean expression (viridis scale) so it visually matches the color logic already used in the dot plot. Bright yellow means a marker is genuinely driving this cluster's identity, dark purple means it's near background.

```{r violin-function}
make_cluster_violin <- function(spe, assay_name, cluster_col, cluster_id,
                                marker_set, celltype_label = NULL,
                                note = NULL) {

  idx      <- which(spe[[cluster_col]] == cluster_id)
  expr_mat <- assay(spe, assay_name)[marker_set, idx, drop = FALSE]

  # per-marker mean expression within this cluster -- drives both the
  # x-axis ordering and the violin fill color
  marker_means <- rowMeans(expr_mat)
  marker_order <- names(sort(marker_means, decreasing = TRUE))

  plot_df <- as.data.frame(t(expr_mat)) |>
    pivot_longer(everything(), names_to = "marker", values_to = "zscore")
  plot_df$marker <- factor(plot_df$marker, levels = marker_order)

  # attach each cell's row to its marker's cluster-level mean, so the
  # fill color is constant within a violin but varies across violins
  plot_df$marker_mean <- marker_means[as.character(plot_df$marker)]

  title_text <- paste0(
    "Cluster ", cluster_id,
    if (!is.null(celltype_label) && !is.na(celltype_label)) paste0(" \u2014 ", celltype_label) else ""
  )
  subtitle_text <- if (!is.null(note)) note else paste0("n = ", length(idx), " cells")
  subtitle_color <- if (!is.null(note)) "firebrick" else "grey30"

  ggplot(plot_df, aes(x = marker, y = zscore, fill = marker_mean)) +
    geom_violin(color = "grey40", scale = "width", trim = TRUE) +
    geom_boxplot(width = 0.12, outlier.size = 0.3, outlier.alpha = 0.3, fill = "white") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    scale_fill_viridis_c(option = "viridis", name = "Mean\nZ-score") +
    labs(title = title_text, subtitle = subtitle_text,
        x = "Marker", y = "Z-score expression (scaled_0100)") +
    theme_bw(base_size = 11) +
    theme(
      axis.text.x    = element_text(angle = 45, hjust = 1),
      plot.subtitle  = element_text(color = subtitle_color, face = "bold")
    )
}
```

---

## 7. Generate and Save a Violin Plot for Every Cluster

Loop over every cluster: print inline for this HTML/PDF render. Save an individual PNG per cluster, and collect all plots into a list for the combined multi-page PDF in the next chunk.

```{r violin-loop, fig.width=14, fig.height=6}

violin_plots <- list()

for (cl in cluster_levels) {

  celltype <- celltype_lookup[[cl]]
  note     <- flagged_clusters[[cl]]  # NULL if this cluster isn't flagged

  p <- make_cluster_violin(
    spe             = spe,
    assay_name      = "scaled_0100",
    cluster_col     = "broad_clusters",
    cluster_id      = cl,
    marker_set      = full_markers,
    celltype_label  = celltype,
    note            = note
  )

  print(p)
  violin_plots[[cl]] <- p

  # individual PNG, one file per cluster
  ggsave(
    filename = file.path(out_dir, paste0("violin_cluster_", cl, ".png")),
    plot     = p, width = 14, height = 6, dpi = 300
  )
}

cat("Saved", length(violin_plots), "individual violin plots to:", out_dir, "\n")
```

---

## 8. Combined Multi-Page PDF (All 20 Clusters)

```{r combined-pdf}

pdf_path <- file.path(out_dir, "violin_plots_all_clusters.pdf")

pdf(pdf_path, width = 14, height = 6)
for (cl in cluster_levels) {
  print(violin_plots[[cl]])
}
dev.off()

cat("Saved combined PDF to:", pdf_path, "\n")
```

---

## 9. Quick Reference: Flagged Clusters Side-by-Side

```{r flagged-side-by-side, fig.width=14, fig.height=16}

flagged_ids <- names(flagged_clusters)
flagged_ids <- flagged_ids[flagged_ids %in% cluster_levels]

if (length(flagged_ids) > 0) {
  wrap_plots(violin_plots[flagged_ids], ncol = 1)
} else {
  cat("No flagged clusters found in this cluster set.\n")
}
```

---

## Notes

- Uses the `scaled_0100` assay (99th percentile cap + Z-score) that is
  already attached to `harmony_20_cluster.rds` from `cell_pheno_20.Rmd`
  -- no re-computation of this assay happens here, so the values match
  the dot plot exactly.
- Marker order is determined *per cluster* (highest mean first), not
  the global hclust order used in the dot plot, so the left-most
  markers in each panel are always that cluster's own top markers.
- A bimodal violin for a marker that also had a high dot-plot mean is
  the direct signature of a doublet/hybrid population -- that cell
  type is not genuinely co-expressing both markers, it's two cell
  types blended into one cluster.
- If a flagged cluster's violins turn out clean (all unimodal, no
  double humps), that argues the co-elevated markers are a real,
  if unusual, biological signature and not a clustering artifact --
  worth a second look with `findMarkers()` and spatial context (QuPath)
  before relabeling.
- Rerunning `cell_pheno_20.Rmd` may reshuffle cluster numbers; if that
  happens, re-run this notebook against the newly saved
  `harmony_20_cluster.rds` rather than reusing old cluster numbers.
