---
title: "Cell Type Distribution Across Patients and ROIs"
author: "Tazein Shah"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    theme: cosmo
    code_folding: show
    toc: true
    toc_float: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(SpatialExperiment)
library(ComplexHeatmap)
library(circlize)
library(knitr)
library(kableExtra)

print_kable <- function(df, ...) {
  kable(df, ...) |>
    kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8)
}
```

# 1. Load Data

```{r load-data}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/final_cell_annotation/r3_annotation"
spe_path <- file.path(analysis_path, "spe_final_celltype_r3.rds")
figure_path <- file.path(analysis_path, "figures")

spe <- readRDS(spe_path)

plot_df <- data.frame(
  celltype       = as.character(spe$final_celltype),
  broad_celltype = as.character(spe$broad_celltype),
  patient_id     = as.character(as.vector(spe$patient_id)),
  sample_id      = as.character(as.vector(spe$sample_id))
)
```

# 2. ROIs and Total Cells per Patient

Baseline sanity check before any composition comparison: how many
ROIs and how many total cells each patient contributes. A patient
with far more ROIs or cells than the rest is likely to dominate
downstream annotation-level percentages, so this context is worth
having in mind while reading Sections 5–8.

```{r patient-summary}
patient_summary <- plot_df |>
  dplyr::group_by(patient_id) |>
  dplyr::summarise(
    n_rois  = dplyr::n_distinct(sample_id),
    n_cells = dplyr::n(),
    .groups = "drop"
  ) |>
  dplyr::arrange(patient_id)
```

```{r patient-roi-combined, fig.width=12, fig.height=8}
library(gridExtra)
library(patchwork)

p_roi_count <- ggplot(patient_summary, aes(x = patient_id, y = n_rois, fill = patient_id)) +
  geom_col(show.legend = FALSE, width = 0.6) +
  geom_text(aes(label = paste0("ROIs: ", n_rois)),
            vjust = -0.3, size = 3) +
  scale_x_discrete(expand = expansion(add = 0.8)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "Patient", y = "Number of ROIs",
       title = "ROIs and Total Cells per Patient") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

table_theme <- ttheme_minimal(
  core = list(fg_params = list(fontsize = 10)),
  colhead = list(fg_params = list(fontsize = 11, fontface = "bold"))
)

cell_count_tbl <- patient_summary |>
  dplyr::select(patient_id, n_cells) |>
  dplyr::rename(Patient = patient_id, `Total Cells` = n_cells) |>
  dplyr::mutate(`Total Cells` = scales::comma(`Total Cells`))

t_cell_count <- tableGrob(cell_count_tbl, rows = NULL, theme = table_theme)

p_roi_count + wrap_elements(t_cell_count) + plot_layout(widths = c(3, 1))
```

# 3. Group Final Cell Types into Broad Categories

```{r categories}
celltype_category <- c(
  "B Cell (CD19+)"                                       = "B Cell/Plasma Cell",
  "B Cell, Weak (CD19+)"                                 = "B Cell/Plasma Cell",
  "B Cell/Plasma Cell, HLADRB-high (CD19+CD138+HLADRB+)" = "B Cell/Plasma Cell",
  "Plasma Cell, Memory-like (Weak) (CD45RO+)"            = "B Cell/Plasma Cell",
  "Plasma Cell, TIM3+ (CD138+TIM3+CD45RO-)"              = "B Cell/Plasma Cell",
  "Plasmablast, Strongly Activated (CD38++)"             = "B Cell/Plasma Cell",
  "Plasmablast/Activated B Cell (CD38+)"                 = "B Cell/Plasma Cell",
  "Transitional B Cell/Plasmablast (CD19+CD138+HLADRB+)" = "B Cell/Plasma Cell",
  "CD4+ T Cell, Activated"                               = "T Cell",
  "CD4+ T Cell, Exhaustion-leaning"                      = "T Cell",
  "CD8+ T Cell, Activated"                                = "T Cell",
  "CD8+ T Cell, Resting"                                  = "T Cell",
  "Memory T Cell"                                         = "T Cell",
  "Treg, Activated"                                       = "T Cell",
  "Treg, Resting"                                         = "T Cell",
  "cDC1"                                                  = "Myeloid",
  "DC/mregDC"                                             = "Myeloid",
  "DC/mregDC (Epi Adj)"                                   = "Myeloid",
  "DC/mregDC (LAMP3-)"                                    = "Myeloid",
  "DC/mregDC Weak"                                         = "Myeloid",
  "Dendritic Cell (CD11c+)"                               = "Myeloid",
  "Macrophage (CD68+)"                                    = "Myeloid",
  "Macrophage, Antigen-Presenting (HLADRB+)"              = "Myeloid",
  "Macrophage, M2-like (CD163+)"                          = "Myeloid",
  "Macrophage, Mixed Signal (PDL1+)"                      = "Myeloid",
  "Macrophage, Proliferating (HLADRB+)"                   = "Myeloid",
  "Macrophage, Proliferating (MKI67+)"                    = "Myeloid",
  "Monocyte (Generic)"                                    = "Myeloid",
  "Monocyte (PD1+)"                                       = "Myeloid",
  "Monocyte-Macrophage Transitional (CD14+)"              = "Myeloid",
  "Monocyte-Macrophage, Unresolved"                       = "Myeloid",
  "Monocyte, Unresolved"                                  = "Myeloid",
  "PMN-MDSC"                                               = "Myeloid",
  "Epithelial (EpCAM+)"                                   = "Epithelial/Stroma",
  "Fibroblast/Stroma"                                     = "Epithelial/Stroma",
  "Tumor Epithelial (Neuroendocrine)"                     = "Epithelial/Stroma",
  "Tumor Epithelial (TTF1++)"                             = "Epithelial/Stroma",
  "Mast Cell"                                             = "Other",
  "Proliferating Cells"                                   = "Other"
)

spe$celltype_category <- celltype_category[as.character(spe$final_celltype)]
plot_df$celltype_category <- celltype_category[plot_df$celltype]

# Flag any final_celltype not covered by the mapping above
unmapped <- unique(plot_df$celltype[is.na(plot_df$celltype_category)])
if (length(unmapped) > 0) {
  warning("Unmapped final_celltype values (will show as NA category): ",
          paste(unmapped, collapse = ", "))
}
```

# 4. Broad vs. Final Cell Type Consistency Check

Sanity check that `final_celltype` nests cleanly within `broad_celltype`
— i.e. each fine-grained annotation should trace back to overwhelmingly
one broad cluster, with only small "boundary spillover" from
neighboring clusters. A `final_celltype` split roughly evenly across
multiple unrelated broad parents would flag a genuinely ambiguous
cluster worth investigating.

```{r broad-vs-final-check, results='asis'}
broad_final_tbl <- plot_df |>
  dplyr::count(broad_celltype, celltype, name = "Freq") |>
  dplyr::filter(Freq > 0) |>
  dplyr::arrange(celltype)

print_kable(broad_final_tbl)

# Flag final_celltypes that split across multiple broad parents,
# and how dominant the top parent is
nesting_check <- broad_final_tbl |>
  dplyr::group_by(celltype) |>
  dplyr::mutate(
    n_broad_parents = dplyr::n(),
    total = sum(Freq),
    pct = Freq / total * 100
  ) |>
  dplyr::slice_max(pct, n = 1) |>
  dplyr::ungroup() |>
  dplyr::rename(top_broad_parent = broad_celltype, pct_top_parent = pct) |>
  dplyr::mutate(flag = dplyr::case_when(
    n_broad_parents == 1 ~ "clean 1:1 mapping",
    pct_top_parent > 90 ~ "ok: minor boundary spillover",
    pct_top_parent > 70 ~ "moderate split - review",
    TRUE ~ "HIGH RISK: ambiguous, no dominant parent"
  )) |>
  dplyr::arrange(pct_top_parent) |>
  dplyr::select(celltype, n_broad_parents, top_broad_parent, pct_top_parent, flag)

k <- kable(nesting_check, digits = 1) |>
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8)

char_cols <- which(sapply(nesting_check, is.character))
for (i in char_cols) {
  k <- k |> column_spec(i, width = "3cm")
}

k
```

# 5. Cell Type Composition by Patient

One bar chart per cell type, printed separately, with one bar per
patient labeled with % and cell count.

```{r prep-patient-counts}
patient_counts <- plot_df |>
  dplyr::count(patient_id, celltype, celltype_category, name = "n") |>
  dplyr::group_by(celltype) |>
  dplyr::mutate(prop = n / sum(n),
                pct = round(prop * 100, 1),
                label = paste0(pct, "% (", n, ")")) |>
  dplyr::ungroup()

categories <- sort(unique(patient_counts$celltype_category))

make_patient_barplot <- function(df_ct, ct_name) {
  ggplot(df_ct, aes(x = patient_id, y = pct, fill = patient_id)) +
    geom_col(show.legend = FALSE) +
    geom_text(aes(label = label), vjust = -0.3, size = 3) +
    labs(x = "Patient", y = "% of cell type's total cells",
         title = ct_name) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## 5.1 B Cell/Plasma Cell

```{r patient-bcell, fig.width=10, fig.height=5, results='asis'}
df_bcell <- patient_counts |> dplyr::filter(celltype_category == "B Cell/Plasma Cell")
celltypes_bcell <- sort(unique(df_bcell$celltype))

for (ct in celltypes_bcell) {
  cat("\n\n### ", ct, "\n\n")
  p <- make_patient_barplot(df_bcell |> dplyr::filter(celltype == ct), ct)
  print(p)
}
```

## 5.2 T Cell

```{r patient-tcell, fig.width=10, fig.height=5, results='asis'}
df_tcell <- patient_counts |> dplyr::filter(celltype_category == "T Cell")
celltypes_tcell <- sort(unique(df_tcell$celltype))

for (ct in celltypes_tcell) {
  cat("\n\n### ", ct, "\n\n")
  p <- make_patient_barplot(df_tcell |> dplyr::filter(celltype == ct), ct)
  print(p)
}
```

## 5.3 Myeloid

```{r patient-myeloid, fig.width=10, fig.height=5, results='asis'}
df_myeloid <- patient_counts |> dplyr::filter(celltype_category == "Myeloid")
celltypes_myeloid <- sort(unique(df_myeloid$celltype))

for (ct in celltypes_myeloid) {
  cat("\n\n### ", ct, "\n\n")
  p <- make_patient_barplot(df_myeloid |> dplyr::filter(celltype == ct), ct)
  print(p)
}
```

## 5.4 Epithelial/Stroma

```{r patient-epi-stroma, fig.width=10, fig.height=5, results='asis'}
df_epi <- patient_counts |> dplyr::filter(celltype_category == "Epithelial/Stroma")
celltypes_epi <- sort(unique(df_epi$celltype))

for (ct in celltypes_epi) {
  cat("\n\n### ", ct, "\n\n")
  p <- make_patient_barplot(df_epi |> dplyr::filter(celltype == ct), ct)
  print(p)
}
```

## 5.5 Other

```{r patient-other, fig.width=10, fig.height=5, results='asis'}
df_other <- patient_counts |> dplyr::filter(celltype_category == "Other")
celltypes_other <- sort(unique(df_other$celltype))

for (ct in celltypes_other) {
  cat("\n\n### ", ct, "\n\n")
  p <- make_patient_barplot(df_other |> dplyr::filter(celltype == ct), ct)
  print(p)
}
```

## 5.6 High-Concentration Annotations (Patient)

Flags every annotation/patient pair where a single patient accounts
for more than 25% of that annotation's cells — pulled directly from
the bar charts above. Unlike the top-patient-only summary later in
this document, this can list more than one patient per annotation if
several each cross the 25% threshold.

```{r high-patient-concentration-flags, results='asis'}
patient_flags <- plot_df |>
  dplyr::count(celltype, patient_id, name = "n") |>
  dplyr::group_by(celltype) |>
  dplyr::mutate(pct = round(n / sum(n) * 100, 1)) |>
  dplyr::ungroup() |>
  dplyr::filter(pct > 25) |>
  dplyr::arrange(dplyr::desc(pct)) |>
  dplyr::select(celltype, patient_id, pct)

print_kable(patient_flags, col.names = c("Annotation", "Patient", "% of Cells"), digits = 1)
```

# 6. Cell Type Composition by ROI

Same format at the ROI level — one bar chart per cell type, printed
separately, with one bar per ROI labeled with % and cell count.

```{r prep-roi-counts}
roi_counts <- plot_df |>
  dplyr::count(sample_id, celltype, celltype_category, name = "n") |>
  dplyr::group_by(celltype) |>
  dplyr::mutate(prop = n / sum(n),
                pct = round(prop * 100, 1),
                label = paste0(pct, "% (", n, ")")) |>
  dplyr::ungroup()

make_roi_barplot <- function(df_ct, ct_name) {
  ggplot(df_ct, aes(x = sample_id, y = pct, fill = sample_id)) +
    geom_col(show.legend = FALSE) +
    geom_text(aes(label = label), vjust = -0.3, size = 2.5) +
    labs(x = "ROI", y = "% of cell type's total cells",
         title = ct_name) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7))
}
```

## 6.1 B Cell/Plasma Cell

```{r roi-bcell, fig.width=12, fig.height=5, results='asis'}
df_bcell_roi <- roi_counts |> dplyr::filter(celltype_category == "B Cell/Plasma Cell")
celltypes_bcell_roi <- sort(unique(df_bcell_roi$celltype))

for (ct in celltypes_bcell_roi) {
  cat("\n\n### ", ct, "\n\n")
  p <- make_roi_barplot(df_bcell_roi |> dplyr::filter(celltype == ct), ct)
  print(p)
}
```

## 6.2 T Cell

```{r roi-tcell, fig.width=12, fig.height=5, results='asis'}
df_tcell_roi <- roi_counts |> dplyr::filter(celltype_category == "T Cell")
celltypes_tcell_roi <- sort(unique(df_tcell_roi$celltype))

for (ct in celltypes_tcell_roi) {
  cat("\n\n### ", ct, "\n\n")
  p <- make_roi_barplot(df_tcell_roi |> dplyr::filter(celltype == ct), ct)
  print(p)
}
```

## 6.3 Myeloid

```{r roi-myeloid, fig.width=12, fig.height=5, results='asis'}
df_myeloid_roi <- roi_counts |> dplyr::filter(celltype_category == "Myeloid")
celltypes_myeloid_roi <- sort(unique(df_myeloid_roi$celltype))

for (ct in celltypes_myeloid_roi) {
  cat("\n\n### ", ct, "\n\n")
  p <- make_roi_barplot(df_myeloid_roi |> dplyr::filter(celltype == ct), ct)
  print(p)
}
```

## 6.4 Epithelial/Stroma

```{r roi-epi-stroma, fig.width=12, fig.height=5, results='asis'}
df_epi_roi <- roi_counts |> dplyr::filter(celltype_category == "Epithelial/Stroma")
celltypes_epi_roi <- sort(unique(df_epi_roi$celltype))

for (ct in celltypes_epi_roi) {
  cat("\n\n### ", ct, "\n\n")
  p <- make_roi_barplot(df_epi_roi |> dplyr::filter(celltype == ct), ct)
  print(p)
}
```

## 6.5 Other

```{r roi-other, fig.width=12, fig.height=5, results='asis'}
df_other_roi <- roi_counts |> dplyr::filter(celltype_category == "Other")
celltypes_other_roi <- sort(unique(df_other_roi$celltype))

for (ct in celltypes_other_roi) {
  cat("\n\n### ", ct, "\n\n")
  p <- make_roi_barplot(df_other_roi |> dplyr::filter(celltype == ct), ct)
  print(p)
}
```

## 6.6 High-Concentration Annotations (ROI)

Flags every annotation/ROI pair where a single ROI accounts for more
than 7% of that annotation's cells — pulled directly from the bar
charts in Section 6. Since there are far more ROIs than patients,
7% is a lower bar than the 25% patient threshold, but still flags
ROIs punching above an even distribution.

```{r high-roi-concentration-flags, results='asis'}
roi_flags <- plot_df |>
  dplyr::count(celltype, sample_id, patient_id, name = "n") |>
  dplyr::group_by(celltype) |>
  dplyr::mutate(pct = round(n / sum(n) * 100, 1)) |>
  dplyr::ungroup() |>
  dplyr::filter(pct > 7) |>
  dplyr::arrange(dplyr::desc(pct)) |>
  dplyr::select(celltype, sample_id, patient_id, pct)

print_kable(roi_flags, col.names = c("Annotation", "ROI", "Patient", "% of Cells"), digits = 1)
```

## 6.7 ROIs Driving High-Concentration Annotations (Grouped by Annotation)

Same flagged ROI/annotation pairs as above, but grouped by annotation
so it's easy to see, for a given cell type, which ROI(s) are
inflating its composition.

```{r high-roi-concentration-grouped, results='asis'}
flagged_celltypes <- roi_flags |>
  dplyr::pull(celltype) |>
  unique() |>
  sort()

for (ct in flagged_celltypes) {

  cat("\n\n### ", ct, "\n\n")

  df_ct <- roi_flags |>
    dplyr::filter(celltype == ct) |>
    dplyr::arrange(dplyr::desc(pct)) |>
    dplyr::select(sample_id, patient_id, pct)

  print(print_kable(df_ct, col.names = c("ROI", "Patient", "% of Cells"), digits = 1, row.names = FALSE))
}
```
