---
title: "Final Annotation Violin Plots by Cell Type Category"
author: "Tazein Shah"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 1. Load Libraries

```{r libraries}
library(SpatialExperiment)
library(tidyverse)
```

## 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/final_cell_annotation"

spe_path <- file.path(analysis_path, "r3_annotation", "spe_final_celltype_r3.rds")
```

## 3. Load Final Annotated SPE

```{r load-data}
spe <- readRDS(spe_path)

print(table(spe$final_celltype))
```

## 4. Group Final Cell Types into Broad Categories

```{r categories}
celltype_category <- c(
  "B Cell (CD19+)"                                       = "B Cell/Plasma Cell",
  "B Cell, Weak (CD19+)"                                 = "B Cell/Plasma Cell",
  "B Cell/Plasma Cell, HLADRB-high (CD19+CD138+HLADRB+)" = "B Cell/Plasma Cell",
  "Plasma Cell, Memory-like (Weak) (CD45RO+)"            = "B Cell/Plasma Cell",
  "Plasma Cell, TIM3+ (CD138+TIM3+CD45RO-)"              = "B Cell/Plasma Cell",
  "Plasmablast, Strongly Activated (CD38++)"             = "B Cell/Plasma Cell",
  "Plasmablast/Activated B Cell (CD38+)"                 = "B Cell/Plasma Cell",
  "Transitional B Cell/Plasmablast (CD19+CD138+HLADRB+)" = "B Cell/Plasma Cell",

  "CD4+ Memory T Cell, Exhaustion-leaning"                = "T Cell",
  "CD4+ Memory T Cell, Partial-exhausion"                 = "T Cell",
  "CD8+ Memory T Cell"                                     = "T Cell",
  "CD8+ Naive T Cell"                                      = "T Cell",
  "Memory T Cell (Unresolved)"                             = "T Cell",
  "Treg, Memory"                                            = "T Cell",
  "Treg, Naive"                                             = "T Cell",

  "cDC1"                                                  = "Myeloid",
  "DC/mregDC"                                             = "Myeloid",
  "DC/mregDC (Epi Adj)"                                   = "Myeloid",
  "DC/mregDC (LAMP3-)"                                    = "Myeloid",
  "DC/mregDC Weak"                                         = "Myeloid",
  "Dendritic Cell (CD11c+)"                               = "Myeloid",
  "Macrophage (CD68+)"                                    = "Myeloid",
  "Macrophage, Antigen-Presenting (HLADRB+)"              = "Myeloid",
  "Macrophage, M2-like (CD163+)"                          = "Myeloid",
  "Macrophage, Mixed Signal (PDL1+)"                      = "Myeloid",
  "Macrophage, Proliferating (HLADRB+)"                   = "Myeloid",
  "Macrophage, Proliferating (MKI67+)"                    = "Myeloid",
  "Monocyte (Generic)"                                    = "Myeloid",
  "Monocyte (PD1+)"                                       = "Myeloid",
  "Monocyte-Macrophage Transitional (CD14+)"              = "Myeloid",
  "Monocyte-Macrophage, Unresolved"                       = "Myeloid",
  "Monocyte, Unresolved"                                  = "Myeloid",
  "PMN-MDSC"                                               = "Myeloid",

  "Epithelial (EpCAM+)"                                   = "Epithelial/Stroma",
  "Fibroblast/Stroma"                                     = "Epithelial/Stroma",
  "Tumor Epithelial (Neuroendocrine)"                     = "Epithelial/Stroma",
  "Tumor Epithelial (TTF1++)"                             = "Epithelial/Stroma",

  "Mast Cell"                                             = "Other",
  "Proliferating Cells"                                   = "Other"
)

spe$celltype_category <- celltype_category[as.character(spe$final_celltype)]
```

Flag anything that didn't map (hard stop, since NAs here silently break downstream subsetting):

```{r check-unmapped}
unmapped <- setdiff(unique(as.character(spe$final_celltype)), names(celltype_category))
if (length(unmapped) > 0) {
  stop("Unmapped final_celltype labels -- add these to celltype_category:\n",
       paste(unmapped, collapse = "\n"))
} else {
  cat("All final_celltype labels mapped to a category.\n")
}
```

## 5. Cell Counts by Category and Final Cell Type

```{r counts-category}
cat("Cell count by broad category:\n")
print(table(spe$celltype_category, useNA = "ifany"))
```

## 6. Panel Markers (Excluding DNA/Membrane)

```{r markers}
exclude_channels <- c("DNA191", "DNA193", "Membrane")
panel_markers <- setdiff(rownames(spe), exclude_channels)

cat("Markers used for violin plots:", paste(panel_markers, collapse = ", "), "\n")
```

## 7. Violin Plot Function (One Plot Per Marker)

```{r make-violinplot}
make_violinplot_single_marker <- function(spe, assay_name, cluster_col, marker, title) {
  expr_vec   <- assay(spe, assay_name)[marker, ]
  cluster_id <- factor(spe[[cluster_col]])

  plot_df <- data.frame(cluster = cluster_id, expr = expr_vec)

  ggplot(plot_df, aes(x = cluster, y = expr, fill = cluster)) +
    geom_violin(scale = "width", trim = TRUE, linewidth = 0.2) +
    stat_summary(fun = median, geom = "point", size = 0.8, color = "black") +
    labs(title = title, x = "Cell Type", y = "Z-score expression") +
    theme_bw(base_size = 11) +
    theme(
      axis.text.x       = element_text(angle = 45, hjust = 1, size = 9),
      panel.grid.major.x = element_blank(),
      legend.position    = "none"
    )
}
```

## 8. Violin Plots per Category (One Plot per Marker)

### B Cell/Plasma Cell

```{r violin-bcell, fig.width=10, fig.height=5, results='asis'}
spe_bcell_group <- spe[, spe$celltype_category == "B Cell/Plasma Cell"]
spe_bcell_group$final_celltype <- droplevels(factor(spe_bcell_group$final_celltype))

for (m in panel_markers) {
  p <- make_violinplot_single_marker(spe_bcell_group, "scaled_0100", "final_celltype", m,
                                     paste("B Cell/Plasma Cell:", m))
  print(p)
}
```

### T Cell

```{r violin-tcell, fig.width=10, fig.height=5, results='asis'}
spe_tcell_group <- spe[, spe$celltype_category == "T Cell"]
spe_tcell_group$final_celltype <- droplevels(factor(spe_tcell_group$final_celltype))

for (m in panel_markers) {
  p <- make_violinplot_single_marker(spe_tcell_group, "scaled_0100", "final_celltype", m,
                                     paste("T Cell:", m))
  print(p)
}
```

### Myeloid

```{r violin-myeloid, fig.width=10, fig.height=5, results='asis'}
spe_myeloid_group <- spe[, spe$celltype_category == "Myeloid"]
spe_myeloid_group$final_celltype <- droplevels(factor(spe_myeloid_group$final_celltype))

for (m in panel_markers) {
  p <- make_violinplot_single_marker(spe_myeloid_group, "scaled_0100", "final_celltype", m,
                                     paste("Myeloid:", m))
  print(p)
}
```

### Epithelial/Stroma

```{r violin-epi-stroma, fig.width=10, fig.height=5, results='asis'}
spe_epi_group <- spe[, spe$celltype_category == "Epithelial/Stroma"]
spe_epi_group$final_celltype <- droplevels(factor(spe_epi_group$final_celltype))

for (m in panel_markers) {
  p <- make_violinplot_single_marker(spe_epi_group, "scaled_0100", "final_celltype", m,
                                     paste("Epithelial/Stroma:", m))
  print(p)
}
```

### Other

```{r violin-other, fig.width=10, fig.height=5, results='asis'}
spe_other_group <- spe[, spe$celltype_category == "Other"]
spe_other_group$final_celltype <- droplevels(factor(spe_other_group$final_celltype))

for (m in panel_markers) {
  p <- make_violinplot_single_marker(spe_other_group, "scaled_0100", "final_celltype", m,
                                     paste("Other:", m))
  print(p)
}
```