---
title: "Niche Generation and Differential Abundance"
author: "Tazein Shah"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)

suppressPackageStartupMessages({
  library(imcRtools)
  library(tidyverse)
  library(SingleCellExperiment)
  library(SpatialExperiment)
  library(sccomp)
  library(ggrepel)
})

set.seed(12345)
```

## Overview

This notebook generates cellular neighborhoods (niches) on the fully
annotated, spatial-graph-equipped dataset, and tests niche composition
for association with N-status and AJCC stage. Structure follows the
preprint's original niche generation script as closely as possible.

Three deliberate deviations from that script are noted inline below
(all flagged where they occur, and summarized in the Notes section
at the end):

1. **Spatial graph threshold.** Reuses an existing
   `expansion_interaction_graph` built at `threshold = 20`. The
   original preprint script used `threshold = 50`; this text-vs-code
   discrepancy is unresolved upstream, so the notebook proceeds with
   the already-built 20-threshold graph.
2. `aggregateNeighbors`, `kmeans`, and the table/volcano/boxplot
   outputs follow the original script's structure directly, except
   `sccomp_estimate() |> sccomp_test()` is used in place of
   `sccomp_glm()`, which doesn't exist in the installed package version
   (see Notes).
3. **Patient10 / Myeloid Proliferating caveat.** Patient10 (the sole
   benign control) accounts for 44.7% of the "Macrophage, Proliferating
   (MKI67+, Patient10-Driven Artifact)" cluster -- already flagged as
   an artifact in its label from the annotation review. Niches leaning
   heavily on this cell type should be read with that in mind -- see
   Section 5b and Notes.

---

## 1. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

# Spatial graph object. final_celltype merge (post-inspection renames
# + Patient10 flag) is saved to a new file rather than overwriting the
# original -- see Section 1b.
spatial_graph_path <- file.path(analysis_path, "spatial_analysis", "spe_spatial_graph_post_inspect.rds")
spatial_graph_path_orig <- file.path(analysis_path, "spatial_analysis", "spe_spatial_graph.rds")

niche_path <- file.path(analysis_path, "spatial_analysis", "niche_k50_thresh20")
dir.create(niche_path, recursive = TRUE, showWarnings = FALSE)

niche_spe_out_path      <- file.path(niche_path, "spe_niche.rds")
sccomp_nstatus_out_path <- file.path(niche_path, "sccomp_nstatus.rds")
sccomp_stage_out_path   <- file.path(niche_path, "sccomp_stage.rds")
```

---

## 1b. Merge updated annotations into spatial graph object (one-time)

The spatial graph (`expansion_interaction_graph`) is purely geometric
-- built from cell coordinates/segmentation, not from `final_celltype`
-- so correcting annotations doesn't require rebuilding it. This chunk
instead merges the corrected `final_celltype` (post-inspection renames
+ Patient10 flag) into the existing graph-equipped object, matching on
cell ID rather than assuming row order, and saves the result to a new
file -- the original `spe_spatial_graph.rds` is left untouched. Only
needs to be run once; safe to skip on subsequent runs once
`spe_spatial_graph_post_inspect.rds` exists.

```{r merge-updated-annotations, eval=FALSE}
post_inspect_path <- file.path(analysis_path, "final_cell_annotation", "final_cell_annot_post_inspect.rds")

spe_graph <- readRDS(spatial_graph_path_orig)
spe_post_inspect <- readRDS(post_inspect_path)

# Match on cell IDs (colnames), not row order, in case the two objects
# aren't in the same cell order.
stopifnot(all(colnames(spe_graph) %in% colnames(spe_post_inspect)))

match_idx <- match(colnames(spe_graph), colnames(spe_post_inspect))

cat("Cells in spatial graph object:", ncol(spe_graph), "\n")
cat("Cells matched in post-inspect object:", sum(!is.na(match_idx)), "\n")

# Overwrite with corrected annotations
spe_graph$final_celltype <- spe_post_inspect$final_celltype[match_idx]

cat("Updated cell type levels:\n")
print(table(spe_graph$final_celltype))

saveRDS(spe_graph, spatial_graph_path)
cat("Saved (original left untouched):", spatial_graph_path, "\n")
```

---

## 2. Load spatial graph object

Loads the object that already has `final_celltype` populated (merged
broad annotations + all subclustering results, low-quality clusters
excluded, and the post-inspection renames applied) and the
`expansion_interaction_graph` colPair already built at `threshold = 20`.

```{r load}
spe <- readRDS(spatial_graph_path)

cat("Cells:", ncol(spe), "\n")
cat("Cell types:", nlevels(spe$final_celltype), "\n")
print(table(spe$final_celltype))
```

---

## 3. Aggregate neighbor composition

```{r aggregate-neighbors}
# Reusing the existing graph (threshold = 20) rather than rebuilding it.
# See Overview note above re: 20 vs. 50 discrepancy.
spe <- aggregateNeighbors(spe,
                          colPairName = "expansion_interaction_graph",
                          aggregate_by = "metadata",
                          count_by = "final_celltype")
```

---

## 4. K-means niche clustering

```{r kmeans-niches}
kmeansClust <- 50

cn_1 <- kmeans(spe$aggregatedNeighbors, centers = kmeansClust, iter.max = 10000)
spe$Niche <- as.factor(cn_1$cluster)

table(spe$Niche)
```

---

## 5. Patient-level covariates

```{r covariates}
spe$stage_group <- sapply(spe$patient_id, function(x) {
  if (x %in% c("Patient12", "Patient15")) return("IIB")
  else if (x == "Patient13") return("IIIA")
  else if (x == "Patient19") return("IB")
  else if (x == "Patient11" || grepl("^Patient20", x)) return("IA3")
  else if (x %in% c("Patient14", "Patient16", "Patient17", "Patient18")) return("IA2")
  else if (x == "Patient10") return("Benign")
  else return(NA)
})

spe$stage_group <- factor(spe$stage_group,
                          levels = c("Benign", "IA2", "IA3", "IB", "IIIA", "IIB"))

# Three-tier disease state grouping (Benign / Lower / Higher)
spe$disease_state <- sapply(spe$patient_id, function(x) {
  if (x == "Patient10") return("Benign")
  else if (x %in% c("Patient11", "Patient14", "Patient16", "Patient17", "Patient18") || grepl("^Patient20", x)) return("Lower")
  else if (x %in% c("Patient12", "Patient13", "Patient15", "Patient19")) return("Higher")
  else return(NA)
})

spe$disease_state <- factor(spe$disease_state, levels = c("Benign", "Lower", "Higher"))

# N-status: only IIB (P12, P15) is N1; everything else is N0
# NOTE: flagging for confirmation -- does IIIA (Patient13) carry N1 status?
# AJCC IIIA often implies nodal involvement, but this cohort's original
# stage list only assigned N1 possibility to IIB. Left as-is pending confirmation.
spe$N_status <- factor(ifelse(spe$stage_group == "IIB", "N1", "N0"),
                       levels = c("N0", "N1"))

table(spe$patient_id, spe$N_status)
table(spe$patient_id, spe$stage_group)
table(spe$patient_id, spe$disease_state)
```

### 5b. Caveat check: Patient10 contribution to Myeloid Proliferating cluster, per niche

Patient10 is the sole benign control and was found (in the annotation
review session) to account for 44.7% of the "Macrophage, Proliferating
(MKI67+, Patient10-Driven Artifact)" cluster cohort-wide -- the label
itself was updated during that review to flag it as a technical
artifact rather than real biology. Since that cluster feeds directly
into niche composition here, this chunk reports, per niche, what
fraction of that cluster's cells within the niche come from Patient10
-- any niche with a high value here should be read as a possible
Patient10/technical-artifact effect rather than a genuine stage or
N-status association.

```{r patient10-myeloid-prolif-check}
prolif_label <- "Macrophage, Proliferating (MKI67+, Patient10-Driven Artifact)"

prolif_idx <- which(spe$final_celltype == prolif_label)

niche_prolif_patient10 <- data.frame(
  Niche = spe$Niche[prolif_idx],
  Patient = spe$patient_id[prolif_idx]
) |>
  group_by(Niche) |>
  summarise(
    n_prolif_cells = n(),
    pct_patient10 = mean(Patient == "Patient10") * 100,
    .groups = "drop"
  ) |>
  arrange(desc(pct_patient10))

niche_prolif_patient10 |>
  knitr::kable(digits = 1, caption = "Per-niche Patient10 share of Myeloid Proliferating cells")
```

```{r save-niche-spe}
saveRDS(spe, niche_spe_out_path)
cat("Saved object with niches to:", niche_spe_out_path, "\n")
```

---

## 6. Differential niche abundance (sccomp)

`.sample = Patient`, since N-status and stage are constant within
patient (not within ROI) -- mirroring how the original script's `LN`
is the invariant unit for its `TumorStage`/`HighN1` covariates.

```{r sccomp-prep}
sceIMC <- SingleCellExperiment(list(counts = assay(spe, "counts")))
sceIMC$Patient   <- factor(spe$patient_id)
sceIMC$Sample    <- factor(spe$sample_id)
sceIMC$Niche     <- factor(spe$Niche)
sceIMC$N_status  <- factor(spe$N_status, levels = c("N0", "N1"))
sceIMC$Stage     <- factor(spe$stage_group,
                           levels = c("Benign", "IA2", "IA3", "IB", "IIIA", "IIB"))
```

### N-status model (parallel to `HighN1` in the original script)

```{r sccomp-nstatus}
if (file.exists(sccomp_nstatus_out_path)) {
  cat("Found existing output, loading:", sccomp_nstatus_out_path, "\n")
  sccomp_niche_nstatus <- readRDS(sccomp_nstatus_out_path)
} else {
  sccomp_niche_nstatus <-
    sceIMC |>
    sccomp_estimate(
      formula_composition = ~ N_status + (1 | Patient),
      .sample = Sample,
      .cell_group = Niche,
      bimodal_mean_variability_association = FALSE,
      variational_inference = FALSE,
      output_directory = file.path(niche_path, "sccomp_draws_nstatus"),
      cores = 8
    ) |>
    sccomp_test()

  saveRDS(sccomp_niche_nstatus, sccomp_nstatus_out_path)
}
```

### Stage model (parallel to `TumorStage` in the original script)

```{r sccomp-stage}
if (file.exists(sccomp_stage_out_path)) {
  cat("Found existing output, loading:", sccomp_stage_out_path, "\n")
  sccomp_niche_stage <- readRDS(sccomp_stage_out_path)
} else {
  sccomp_niche_stage <-
    sceIMC |>
    sccomp_estimate(
      formula_composition = ~ Stage + (1 | Patient),
      .sample = Sample,
      .cell_group = Niche,
      bimodal_mean_variability_association = FALSE,
      variational_inference = FALSE,
      output_directory = file.path(niche_path, "sccomp_draws_stage"),
      cores = 8
    ) |>
    sccomp_test()

  saveRDS(sccomp_niche_stage, sccomp_stage_out_path)
}
```

---

## 7. Results table

```{r results-table}
sccomp_niche_nstatus <- readRDS(sccomp_nstatus_out_path)
sccomp_niche_nstatus_sub <- sccomp_niche_nstatus[which(sccomp_niche_nstatus$factor == "N_status"), ]

# Join Patient10 share of the Myeloid Proliferating (artifact) cluster
# per niche, so any niche at risk of that confound is visible directly
# on the results table rather than requiring a manual cross-check
# against Section 5b. Niches with no proliferating cells at all (not
# present in niche_prolif_patient10) get 0 -- there's no Patient10 risk
# for this cluster if the cluster isn't present in that niche.
sccomp_niche_nstatus_sub <- sccomp_niche_nstatus_sub |>
  as.data.frame() |>
  mutate(Niche = as.character(Niche)) |>
  left_join(
    niche_prolif_patient10 |> mutate(Niche = as.character(Niche)),
    by = "Niche"
  ) |>
  mutate(
    n_prolif_cells = replace_na(n_prolif_cells, 0),
    pct_patient10 = replace_na(pct_patient10, 0)
  )

sccomp_niche_nstatus_sub |>
  filter(c_FDR < 0.1) |>
  arrange(c_FDR) |>
  select(Niche, parameter, c_effect, c_FDR, n_prolif_cells, pct_patient10) |>
  knitr::kable(digits = 3, caption = "Niche composition association with N-status (FDR < 0.1), with Patient10 share of the Myeloid Proliferating artifact cluster")
```

---

## 8. Volcano plot

```{r volcano}
plot_df <- sccomp_niche_nstatus_sub |>
  as.data.frame() |>
  filter(!is.nan(c_FDR), c_FDR < 0.1) |>
  mutate(
    LogFDR = -log(c_FDR),
    Significant = ifelse(c_FDR < 0.05, "FDR < 0.05", "Not significant")
  )

ggplot(plot_df, aes(x = c_effect, y = LogFDR, color = Significant)) +
  geom_point(size = 2) +
  geom_text_repel(
    aes(label = Niche), size = 3, show.legend = FALSE
  ) +
  scale_color_manual(values = c("FDR < 0.05" = "firebrick", "Not significant" = "grey70")) +
  theme_classic() +
  labs(x = "Effect size", y = "-log(FDR)",
       title = "Niche composition vs. N-status")
```

---

## 9. Results table (Stage)

`factor == "Stage"` covers all 5 non-reference contrasts (each vs.
`Benign`) at once, distinguished by `parameter`
(`StageIA2`, `StageIA3`, `StageIB`, `StageIIIA`, `StageIIB`) --
250 rows total (50 niches x 5 contrasts).

```{r results-table-stage}
sccomp_niche_stage <- readRDS(sccomp_stage_out_path)
sccomp_niche_stage_sub <- sccomp_niche_stage[which(sccomp_niche_stage$factor == "Stage"), ]

# Same Patient10-share join as the N-status table (Section 7).
sccomp_niche_stage_sub <- sccomp_niche_stage_sub |>
  as.data.frame() |>
  mutate(Niche = as.character(Niche)) |>
  left_join(
    niche_prolif_patient10 |> mutate(Niche = as.character(Niche)),
    by = "Niche"
  ) |>
  mutate(
    n_prolif_cells = replace_na(n_prolif_cells, 0),
    pct_patient10 = replace_na(pct_patient10, 0)
  )

sccomp_niche_stage_sub |>
  filter(c_FDR < 0.1) |>
  arrange(c_FDR) |>
  select(Niche, parameter, c_effect, c_FDR, n_prolif_cells, pct_patient10) |>
  knitr::kable(digits = 3, caption = "Niche composition association with Stage (vs. Benign, FDR < 0.1), with Patient10 share of the Myeloid Proliferating artifact cluster")
```

---

## 10. Volcano plot (Stage) {.tabset}

Faceted by contrast, since each stage level vs. Benign is a separate
comparison rather than a single binary effect.

```{r volcano-stage, results = "asis", fig.width = 7, fig.height = 5.5}
plot_df_stage <- sccomp_niche_stage_sub |>
  as.data.frame() |>
  filter(!is.nan(c_FDR), c_FDR < 0.1) |>
  mutate(
    LogFDR = -log(c_FDR),
    Significant = ifelse(c_FDR < 0.05, "FDR < 0.05", "Not significant")
  )

for (contrast in unique(plot_df_stage$parameter)) {
  cat("\n\n### ", contrast, "\n\n")
  df_sub <- plot_df_stage[plot_df_stage$parameter == contrast, ]
  print(
    ggplot(df_sub, aes(x = c_effect, y = LogFDR, color = Significant)) +
      geom_point(size = 2) +
      geom_text_repel(aes(label = Niche), size = 3, show.legend = FALSE) +
      scale_color_manual(values = c("FDR < 0.05" = "firebrick", "Not significant" = "grey70")) +
      theme_classic() +
      labs(x = "Effect size", y = "-log(FDR)",
           title = paste0("Niche composition: ", contrast, " vs. Benign"))
  )
  cat("\n\n")
}
```

---

## 11. Boxplots per niche {.tabset}

```{r boxplots, results = "asis", fig.width = 5, fig.height = 5.5}
# Only plot niches significant at FDR < 0.05 in either the N-status or
# Stage model -- matching the "FDR < 0.05" red points on the volcano
# plots (Sections 8 and 10), rather than the wider FDR < 0.1 threshold
# used just to populate the results tables.
sig_niches_nstatus <- sccomp_niche_nstatus_sub |>
  filter(c_FDR < 0.05) |>
  pull(Niche) |>
  as.character()

sig_niches_stage <- sccomp_niche_stage_sub |>
  filter(c_FDR < 0.05) |>
  pull(Niche) |>
  as.character()

sig_niches <- union(sig_niches_nstatus, sig_niches_stage)
sig_niches <- as.character(sort(as.numeric(sig_niches)))

cat("Niches significant (FDR < 0.05) in N-status model:", length(sig_niches_nstatus), "\n")
cat("Niches significant (FDR < 0.05) in Stage model:", length(sig_niches_stage), "\n")
cat("Total unique niches plotted below:", length(sig_niches), "\n")

niche_prop <- table(spe$sample_id, spe$Niche) |>
  prop.table(margin = 1) |>
  as.data.frame()
colnames(niche_prop) <- c("Sample", "Niche", "Value")

niche_prop$Patient <- spe$patient_id[match(niche_prop$Sample, spe$sample_id)]
niche_prop$Stage <- spe$stage_group[match(niche_prop$Patient, spe$patient_id)]

for (n in sig_niches) {
  cat("\n\n### Niche", n, "\n\n")
  df_sub <- niche_prop[niche_prop$Niche == n, ]
  df_sub$Value <- df_sub$Value * 100
  print(
    ggplot(df_sub, aes(x = Stage, y = Value)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(aes(color = Patient), width = 0.1, size = 1.5) +
      theme_classic() +
      ylab("% within sample") +
      ggtitle(paste0("Niche ", n))
  )
  cat("\n\n")
}
```

---

## Notes

- **Graph threshold (20 vs. 50):** this notebook builds on an
  upstream object whose spatial graph was constructed at
  `threshold = 20`, matching the manuscript text/figure legend. The
  original preprint's niche generation script itself used
  `threshold = 50`. That inconsistency between text and code is
  unresolved upstream; this notebook did not rebuild the graph and so
  inherits the 20-threshold version. Easy to regenerate with a
  different threshold later without touching anything downstream of
  the graph-building step.
- **`sccomp_glm()` vs. `sccomp_estimate()`/`sccomp_test()`:** the
  original script calls `sccomp_glm()` directly, but this installed
  package version does not have that function at all -- confirmed via
  a direct "could not find function" error. This notebook uses
  `sccomp_estimate() |> sccomp_test()` instead. Output object shape
  (`factor`, `c_effect`, `c_FDR` columns) is the same either way, so
  nothing downstream (results table, volcano, boxplots) needed to
  change.
- **`variational_inference = FALSE`:** the default variational Bayes
  fitting failed to converge after 5 attempts. Forcing full HMC
  sampling instead resolved this -- confirmed working (6 chains, no
  divergences, ~145 seconds total).
- **`output_directory`:** set explicitly to a subfolder under
  `niche_path`, since the default working-directory-relative path
  `sccomp` tries to use was not writable in this environment.
- `bimodal_mean_variability_association = FALSE` used for both models
  for consistency with the cell-type-level `sccomp` convention already
  established in this pipeline.
- `.sample = Patient` chosen because N-status and stage are constant
  within patient (not within ROI); confirmed via the
  `table(spe$patient_id, spe$N_status)` cross-tab above.
- **Patient10 / Myeloid Proliferating cluster:** the annotation review
  session found Patient10 (the sole Benign-stage patient) accounts for
  44.7% of the "Macrophage, Proliferating (MKI67+, Patient10-Driven
  Artifact)" cluster cohort-wide, with broad off-lineage marker
  elevation in that cluster. That renamed label reflects the review
  session's conclusion that this cluster is a technical artifact
  rather than a real population, not just a caveat. Since Patient10 is
  also the only Benign-stage sample, any niche whose composition is
  dominated by this cell type risks conflating a Patient10-specific
  technical effect with a genuine Benign-vs-cancer stage effect.
  Section 5b's per-niche breakdown flags which niches carry that risk;
  any significant niche flagged there in Sections 7-11 should be
  footnoted accordingly in a writeup rather than reported at face
  value. Whether to exclude Patient10 (or this cell type entirely) from
  the `sccomp` models is still an open decision from the annotation
  review session and has not been implemented here. Sections 7 and 9
  now join `pct_patient10` directly onto the results tables so this is
  visible without cross-checking Section 5b by eye; niches with no
  proliferating cells present get `pct_patient10 = 0`. Section 11's
  boxplots were also tightened to FDR < 0.05 (matching the volcano
  plots' red "Significant" points) rather than the wider FDR < 0.1 used
  for table inclusion, so borderline/non-significant niches aren't
  plotted alongside genuinely significant ones.
