---
title: "Cell Type Composition Association with Disease State (sccomp)"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

Runs `sccomp` on the finalized cell type annotation to test whether cell
type proportions are associated with disease stage, producing two
comparisons:

1. Lower & Higher vs. Benign (Benign as reference)
2. Higher vs. Lower directly (Lower as reference)

Both are shown as ranked forest plots (effect size + credible interval),
skipping the volcano plot this round since the forest plots proved more
directly interpretable last time.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# 1. Load Libraries

```{r libraries}
.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(imcRtools)
library(tidyverse)
library(SingleCellExperiment)
library(SpatialExperiment)
library(scater)
library(cmdstanr)
library(sccomp)
library(tidytext)
library(kableExtra)
```

# 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

# Update to whichever is the current, Campbell-approved final annotation file
final_celltype_path <- file.path(analysis_path, "final_cell_annotation",
                                  "final_cell_annotation_post_inspect.rds")

sccomp_out_dir <- file.path(analysis_path, "spatial_analysis")
dir.create(sccomp_out_dir, recursive = TRUE, showWarnings = FALSE)

sccomp_vs_benign_path   <- file.path(sccomp_out_dir, "sccomp_celltype_vs_benign.rds")
sccomp_higher_v_lower_path <- file.path(sccomp_out_dir, "sccomp_celltype_higher_v_lower.rds")
```

# 3. Load Finalized Annotated Object

```{r load-spe}
spe <- readRDS(final_celltype_path)

cat("Cells:", ncol(spe), "\n")
cat("Cell types:", nlevels(spe$final_celltype), "\n")
```

# 4. Build sccomp Input Object (Benign as Reference)

```{r sccomp-prep}
sceIMC <- SingleCellExperiment(list(counts = assay(spe, "counts")))
sceIMC$Patient      <- factor(spe$patient_id)
sceIMC$Sample       <- factor(spe$sample_id)
sceIMC$DiseaseState <- factor(spe$disease_state, levels = c("Benign", "Lower", "Higher"))
sceIMC$CellType     <- factor(spe$final_celltype)

cat("Patients:", n_distinct(sceIMC$Patient), "\n")
cat("Disease states:", paste(levels(sceIMC$DiseaseState), collapse = ", "), "\n")
cat("Cell types:", n_distinct(sceIMC$CellType), "\n")
```

# 5. Run sccomp: Lower & Higher vs. Benign

Loads from disk if already run, to avoid re-fitting unnecessarily.

```{r sccomp-run-vs-benign}
if (file.exists(sccomp_vs_benign_path)) {
  sccomp_vs_benign_res <- readRDS(sccomp_vs_benign_path)
} else {
  sccomp_vs_benign_res <- sceIMC |>
    sccomp_estimate(
      formula_composition = ~ DiseaseState + (1 | Patient),
      .sample     = Sample,
      .cell_group = CellType,
      verbose = FALSE
    ) |>
    sccomp_test()

  saveRDS(sccomp_vs_benign_res, sccomp_vs_benign_path)
}

unique(sccomp_vs_benign_res$parameter)
```

# 6. Run sccomp: Higher vs. Lower (Direct Contrast)

```{r sccomp-run-higher-v-lower}
if (file.exists(sccomp_higher_v_lower_path)) {
  sccomp_higher_v_lower_res <- readRDS(sccomp_higher_v_lower_path)
} else {
  sceIMC_hvl <- sceIMC
  sceIMC_hvl$DiseaseState <- factor(spe$disease_state, levels = c("Lower", "Higher", "Benign"))

  sccomp_higher_v_lower_res <- sceIMC_hvl |>
    sccomp_estimate(
      formula_composition = ~ DiseaseState + (1 | Patient),
      .sample     = Sample,
      .cell_group = CellType,
      verbose = FALSE
    ) |>
    sccomp_test()

  saveRDS(sccomp_higher_v_lower_res, sccomp_higher_v_lower_path)
}

unique(sccomp_higher_v_lower_res$parameter)
```

# 7. Forest Plot: Lower & Higher vs. Benign

Faceted by contrast, sorted by effect size within each facet. FDR<0.05
only; CI-crosses-zero cases are excluded rather than just flagged, given
the earlier finding that FDR alone can pass while the credible interval
still spans zero.

```{r forest-vs-benign, fig.width=16, fig.height=12}
effect_df_benign <- sccomp_vs_benign_res |>
  as.data.frame() |>
  filter(parameter %in% c("DiseaseStateLower", "DiseaseStateHigher")) |>
  filter(c_FDR < 0.05) |>
  filter(!(c_lower < 0 & c_upper > 0)) |>   # drop cases where CI still crosses zero
  mutate(CellType_ordered = reorder_within(CellType, c_effect, parameter))

ggplot(effect_df_benign, aes(x = c_effect, y = CellType_ordered)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = c_lower, xmax = c_upper), height = 0, color = "grey40") +
  geom_point(aes(color = c_effect > 0), size = 2, show.legend = FALSE) +
  scale_color_manual(values = c("TRUE" = "firebrick", "FALSE" = "steelblue")) +
  scale_y_reordered() +
  facet_wrap(~ parameter, scales = "free_y") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 11)) +
  labs(
    x = "Effect size (logit fold-change)",
    y = NULL,
    title = "Cell type composition: Lower & Higher vs. Benign",
    subtitle = "FDR < 0.05, CI excludes zero; red = increases, blue = decreases"
  )
```

```{r save-forest-vs-benign}
ggsave(file.path(analysis_path, "figures", "celltype_comp_vs_benign.png"),
       width = 10, height = 12, dpi = 300)
```

# 8. Forest Plot: Higher vs. Lower (Direct Contrast)

```{r forest-higher-v-lower, fig.width=8, fig.height=12}
effect_df_hvl <- sccomp_higher_v_lower_res |>
  as.data.frame() |>
  filter(parameter == "DiseaseStateHigher") |>   # Higher vs. Lower, since Lower is the reference
  filter(c_FDR < 0.05) |>
  filter(!(c_lower < 0 & c_upper > 0)) |>
  mutate(CellType = factor(CellType, levels = CellType[order(c_effect)]))

ggplot(effect_df_hvl, aes(x = c_effect, y = CellType)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = c_lower, xmax = c_upper), height = 0, color = "grey40") +
  geom_point(aes(color = c_effect > 0), size = 2, show.legend = FALSE) +
  scale_color_manual(values = c("TRUE" = "firebrick", "FALSE" = "steelblue")) +
  theme_classic() +
  theme(axis.text.y = element_text(size = 11)) +
  labs(
    x = "Effect size (logit fold-change)",
    y = NULL,
    title = "Cell type composition: Higher vs. Lower disease stage",
    subtitle = "FDR < 0.05, CI excludes zero; red = increases, blue = decreases"
  )
```

```{r save-forest-higher-v-lower}
ggsave(file.path(analysis_path, "figures", "celltype_comp_higher_vs_lower.png"),
       width = 8, height = 12, dpi = 300)
```

# 9. Results Table

```{r sccomp-results-table}
sccomp_results_df <- sccomp_vs_benign_res |>
  arrange(c_FDR) |>
  as.data.frame() |>
  select(CellType, parameter, c_effect, c_lower, c_upper, c_FDR)

write.csv(sccomp_results_df,
          file.path(sccomp_out_dir, "sccomp_celltype_results_vs_benign.csv"),
          row.names = FALSE)

sccomp_higher_v_lower_df <- sccomp_higher_v_lower_res |>
  arrange(c_FDR) |>
  as.data.frame() |>
  select(CellType, parameter, c_effect, c_lower, c_upper, c_FDR)

write.csv(sccomp_higher_v_lower_df,
          file.path(sccomp_out_dir, "sccomp_celltype_results_higher_v_lower.csv"),
          row.names = FALSE)

sccomp_results_df |>
  knitr::kable(digits = 3, caption = "Cell type composition: Lower & Higher vs. Benign",
               format = "latex", booktabs = TRUE, longtable = TRUE) |>
  kable_styling(latex_options = c("striped", "repeat_header", "hold_position"),
                font_size = 8)
```

# Notes

- Loaded from `final_cell_annotation_post_inspect.rds` -- update
  `final_celltype_path` once Campbell's feedback produces a further-updated
  file, and re-run both sccomp calls fresh (delete the cached `.rds` outputs
  in `spatial_analysis/` first, since the load-from-cache logic will
  otherwise silently reuse stale results).
- Both forest plots now exclude cases where `c_FDR < 0.05` but the credible
  interval still crosses zero (the cDC1/Lower issue found earlier) --
  double check this doesn't unexpectedly drop a cell type you expected to
  see, given the updated annotation labels from today's inspection.
- Volcano plot intentionally omitted this round -- code available in the
  original `spatial_setup_celltype.Rmd` (`eval=FALSE`'d there) if needed
  again for a specific request.
