---
title: "Spatial Visualization: Cell Centroids, Graphs, and Cell-Cell Interactions"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

Builds the spatial graph from the finalized annotated dataset, visualizes
cell centroids + graph edges per ROI, and runs formal cell-cell interaction
testing.

**IMPORTANT:** this loads `spe_final_celltype.rds`. As of writing, the final
annotation is still being finalized (a few label naming/marker-mismatch
questions are still open) -- re-run this notebook once that file is locked,
since anything built here on an interim version will need to be redone
against the final labels before being treated as a result.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# 1. Load Libraries

```{r libraries}
#.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(imcRtools)
library(tidyverse)
library(SingleCellExperiment)
library(SpatialExperiment)
library(scater)
library(RColorBrewer)
```

# 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

final_celltype_path <- file.path(analysis_path, "final_cell_annotation", "spe_final_celltype.rds")

spatial_analysis_path <- file.path(analysis_path, "spatial_analysis")

spatial_graph_out_path <- file.path(spatial_analysis_path, "spe_spatial_graph.rds")
interactions_out_path  <- file.path(spatial_analysis_path, "cell_interactions.csv")
```

# 3. Load Finalized Annotated Object

```{r load-spe}
spe <- readRDS(final_celltype_path)

cat("Cells loaded:", ncol(spe), "\n")
cat("Cell types:", nlevels(spe$final_celltype), "\n")
cat("Samples/ROIs:", length(unique(spe$sample_id)), "\n")
```

```{r}
levels(spe$final_celltype)
```

# 4. Build Spatial Graph

Using `type = "expansion"` and `threshold = 20`, matching the reference
publication pipeline's graph-building approach (threshold value per team's
current working decision -- pending confirmation from Campbell given the
20 vs. 50 discrepancy between the manuscript text and code).

```{r build-spatial-graph}
t0 <- Sys.time()

spe <- buildSpatialGraph(
  spe,
  img_id    = "sample_id",
  type      = "expansion",
  threshold = 20,
  name      = "expansion_interaction_graph"
)

t1 <- Sys.time()
cat("Spatial graph built on full dataset in",
    round(as.numeric(t1 - t0, units = "mins"), 2), "minutes\n")

colPairNames(spe)
graph_name <- "expansion_interaction_graph"
```

```{r graph-diagnostics}
graph_obj <- colPair(spe, graph_name)
cat("Total edges in graph:", length(graph_obj), "\n")
cat("Mean neighbors per cell:", round(length(graph_obj) / ncol(spe), 2), "\n")

# Per-cell neighbor counts, for median/distribution
neighbor_counts <- table(factor(from(graph_obj), levels = seq_len(ncol(spe))))
cat("Median neighbors per cell:", median(neighbor_counts), "\n")
```

On average, each cell has about 8.36 neighbors within the 20 um radius (roughly 8 cells physically close enough to be considered its spatial neighborhood).

```{r save-spatial-graph}
saveRDS(spe, spatial_graph_out_path)
cat("Saved spatial graph object to:", spatial_graph_out_path, "\n")
```

# 5. Visualize Cell Centroids + Graph Edges

Spot-check individual ROIs to confirm the graph looks reasonable (respects
ROI boundaries, edges connect physically nearby cells) before running any
interaction statistics on it.

```{r plot-spatial-single, fig.width=10, fig.height=8}
one_sample_id <- unique(spe$sample_id)[1]
one_sample <- spe[, spe$sample_id == one_sample_id]

n_celltypes <- nlevels(spe$final_celltype)
celltype_colors <- colorRampPalette(brewer.pal(12, "Paired"))(n_celltypes)
names(celltype_colors) <- levels(spe$final_celltype)

plotSpatial(
  one_sample,
  node_color_by = "final_celltype",
  img_id        = "sample_id",
  draw_edges    = TRUE,
  colPairName   = graph_name,
  node_size_fix = 0.8,
  edge_width_fix = 0.1,
  edge_color_fix = "grey70"
) +
  scale_color_manual(values = celltype_colors) +
  labs(title = paste("Cell centroids:", one_sample_id)) +
  theme(legend.text = element_text(size = 6),
        legend.key.size = unit(0.3, "cm"))
```


```{r plot-spatial-multiple, fig.width=16, fig.height=12}
# One ROI each from a Benign, Lower, and Higher disease-state patient
sample_ids_to_check <- c(
  "Patient_10T_ROI_001_1",  # Patient10 -- Benign
  "Patient_14T_ROI_001_1",  # Patient14 -- Lower
  "Patient_12T_ROI_001_1"   # Patient12 -- Higher
)

for (sid in sample_ids_to_check) {
  one_sample <- spe[, spe$sample_id == sid]
  disease_state <- unique(one_sample$disease_state)

  p <- plotSpatial(
    one_sample,
    node_color_by = "final_celltype",
    img_id        = "sample_id",
    draw_edges    = TRUE,
    colPairName   = graph_name,
    node_size_fix = 0.8,
    edge_width_fix = 0.1,
    edge_color_fix = "grey70"
  ) +
    scale_color_manual(values = celltype_colors) +
    labs(title = paste0("Spatial graph: ", sid, " (", disease_state, ")")) +
    theme(legend.position = "none")
  print(p)
}
```

Patient 12 (higher disease state) and Patient 14 (lower disease state) ROIs look denser than Patient 10 (benign). 

Could be because they have genuinely higher cell density (makes sense as they are tumor tissues) --> more cells per unit area mean than more pairs fall within 20 um of each other, making more edges. Could also be because these two ROIs simply contain more total cells than the benign one.

```{r}
#Checking cell counts for these ROIs
table(spe$sample_id)[c("Patient_10T_ROI_001_1", "Patient_14T_ROI_001_1", "Patient_12T_ROI_001_1")]

#Per ROI density (comparing cells, edges and mean neighbors)
per_roi_density <- sapply(c("Patient_10T_ROI_001_1", "Patient_14T_ROI_001_1", "Patient_12T_ROI_001_1"), function(sid) {
  n_cells <- sum(spe$sample_id == sid)
  n_edges <- length(colPair(spe[, spe$sample_id == sid], graph_name))
  c(n_cells = n_cells, n_edges = n_edges, mean_neighbors = n_edges / n_cells)
})
t(per_roi_density)
```
Mean neighbors roughly double from benign (4.4) -> lower (6.7) -> higher (8.3)

```{r plot-spatial-bt-only, fig.width=16, fig.height=12}
b_colors <- c(
  "B Cell (CD19+)"                                          = "#0000FF",  # sky blue
  "B Cell, Weak (CD19+)"                                    = "#7FB3D5",  # medium blue
  "B Cell/Plasma Cell, HLADRB-high (CD19+CD138+HLADRB+)"    = "#000080",  # navy
  "Plasmablast/Activated B Cell (CD38+)"                    = "#89CFEF",  # dark navy
  "Transitional B Cell/Plasmablast (CD19+CD138+HLADRB+)"    = "#6693F5"   # pale steel blue
)

t_colors <- c(
  "CD4+ T Cell"                                = "#01ff01",  # bright green
  "CD8+ T Cell"                                = "#008080",
  "Memory T Cell"                               = "#bfdd50",  # forest green
  "Memory/Activated T Cell (CD4++)"             = "#9bc022",  # mint
  "T Cell (TIM3+)"                              = "#07DD05",  # deep teal-green
  "T Cell, Exhaustion-leaning (PD1++)"          = "#9cff94",  # sea green
  "Treg"                                        = "#0e5913"  # dark forest
)

bt_colors <- c(b_colors, t_colors, "Other" = "grey90")
spe$celltype_bt_highlight <- ifelse(
  spe$final_celltype %in% c(b_labels, t_labels),
  as.character(spe$final_celltype),
  "Other"
)
spe$celltype_bt_highlight <- factor(spe$celltype_bt_highlight, levels = c(b_labels, t_labels, "Other"))

bt_colors <- c(b_colors, t_colors, "Other" = "grey90")

for (sid in sample_ids_to_check) {
  one_sample <- spe[, spe$sample_id == sid]
  disease_state <- unique(one_sample$disease_state)

  p <- plotSpatial(
    one_sample,
    node_color_by = "celltype_bt_highlight",
    img_id        = "sample_id",
    draw_edges    = TRUE,
    colPairName   = graph_name,
    node_size_fix = 2,
    edge_width_fix = 0.15,
    edge_color_fix = "grey70",
    nodes_first = FALSE   # draw edges first, so colored nodes sit on top, not buried under them
  ) +
    scale_color_manual(values = bt_colors, name = "Cell Type") +
    labs(title = paste0("B/T Cells: ", sid, " (", disease_state, ")")) +
    theme(legend.text = element_text(size = 10),
        legend.title = element_text(size = 12),
        legend.key.size = unit(0.6, "cm")) +
  guides(color = guide_legend(override.aes = list(size = 4)))
  print(p)
}
```

# 6. Cell-Cell Interaction Testing

Formal test of which cell type pairs are spatially enriched or depleted
relative to chance, using `imcRtools::testInteractions()` (permutation-based
neighborhood enrichment/depletion, conceptually similar to but distinct from
the preprint's custom neighbor-composition function --
see `countInteractionsPerCell()` in `Publication_Code_IMC_Analysis.Rmd` if
exact methodological consistency with the reference pipeline is required;
confirm with Campbell/Yusuke which approach the team wants).

```{r test-interactions}
set.seed(220225)
interactions <- testInteractions(
  spe,
  group_by    = "sample_id",
  label       = "final_celltype",
  colPairName = graph_name,
  method      = "classic"
)

interactions_df <- as.data.frame(interactions)
write.csv(interactions_df, interactions_out_path, row.names = FALSE)
cat("Saved interaction results to:", interactions_out_path, "\n")

head(interactions_df)
```
```{r}
cat("Interaction testing completed in", round(as.numeric(t1 - t0, units = "mins"), 2), "minutes\n")
```

# 7. Visualize Interaction Results

```{r interaction-heatmap, fig.width=12, fig.height=10}
ggplot(interactions_df, aes(x = from_label, y = to_label, fill = strength)) +
  geom_tile() +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, name = "Interaction\nstrength") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6),
        axis.text.y = element_text(size = 6)) +
  labs(x = NULL, y = NULL,
       title = "Cell-Cell Interaction Enrichment/Depletion",
       subtitle = "Red = enriched (more neighbors than expected), Blue = depleted")
```

```{r interaction-significant, fig.width=12, fig.height=8}
# Focus on statistically significant interactions only, if a p-value/FDR
# column is present in the output (column name may vary by imcRtools version)
if ("p.value" %in% colnames(interactions_df)) {
  sig_interactions <- interactions_df |>
    filter(p.value < 0.05) |>
    arrange(p.value)

  cat("Number of significant interactions (p < 0.05):", nrow(sig_interactions), "\n")

  sig_interactions |>
    head(30) |>
    knitr::kable(digits = 3, caption = "Top 30 significant cell-cell interactions")
} else {
  cat("No p-value column found -- check colnames(interactions_df) for the actual column name.\n")
  colnames(interactions_df)
}
```

# 8. Save Final Spatial Graph + Interaction Object

```{r save-final}
saveRDS(spe, spatial_graph_out_path)
cat("Spatial graph object saved to:", spatial_graph_out_path, "\n")
cat("Interaction results saved to:", interactions_out_path, "\n")
```

# Notes

- **Re-run against the finalized `spe_final_celltype.rds`** once the open
  label questions (HLADRB-high naming, CD138/CD39 broad non-specificity,
  T Cell/DC Admixture CD11c mismatch, etc.) are resolved -- everything in
  this notebook is built directly on top of `final_celltype`, so any label
  changes propagate here too.
- `threshold = 20` used for `buildSpatialGraph()` per current team decision;
  flag to Campbell if this should be reconciled with the `threshold = 50`
  discrepancy noted between the reference manuscript text and its code.
- `testInteractions(method = "classic")` was used here as a standard
  permutation-based approach. This differs methodologically from the
  reference pipeline's custom `countInteractionsPerCell()` function (which
  computes neighbor composition ratios rather than enrichment/depletion
  significance) -- confirm with Campbell/Yusuke which approach the team
  wants for consistency with the published methodology, especially if this
  analysis is meant to be directly compared to the preprint's Figure 4G/H.
