---
title: "Patient 12 Tumor: Raw Pixel Intensity Distribution"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output: html_document
---

# Load libraries
```{r}
library(ggplot2)
library(ggridges)
library(dplyr)
library(reshape2)
```

# See what files exist
```{r}
list.files('/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/Data/Imaging/IMC/Tumor/PardeeTumorAblations/Patient_12T')
```

# Load raw pixel data
```{r}
data_dir <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/Data/Imaging/IMC/Tumor/PardeeTumorAblations/Patient_12T"

txt_files <- list.files(data_dir, pattern = "\\.txt$", full.names = TRUE)
cat("Found", length(txt_files), "ROI files:\n")
cat(basename(txt_files), sep = "\n")

# Read one ROI first to inspect structure
roi1 <- read.delim(txt_files[1], check.names = FALSE)
cat("\nDimensions:", nrow(roi1), "pixels x", ncol(roi1), "columns\n")
cat("Column names:\n")
print(colnames(roi1))
```

# Take a quick look at the data
```{r}
head(roi1)
```

The first 6 columns are metadata and coordinates. Columns 7-47 are the 41 marker channels. Each row is one pixel.

# Load ROIs and sample a subset
```{r}
# Columns 1-6 are metadata, 7 onwards are markers
marker_cols <- colnames(roi1)[7:ncol(roi1)]

# Load all ROIs, sample 50k pixels each to keep it manageable
set.seed(42)
all_rois <- lapply(seq_along(txt_files), function(i) {
  df <- read.delim(txt_files[i], check.names = FALSE)
  df_markers <- df[, marker_cols]
  df_sample <- df_markers[sample(nrow(df_markers), 50000), ]
  df_sample$ROI <- paste0("ROI_", i)
  df_sample
})

pixels <- do.call(rbind, all_rois)
cat("Total pixels loaded:", nrow(pixels), "\n")
```

# Reshape to long format and apply asinh transformation
```{r}
# Reshape to long format
pixels_long <- melt(pixels, id.vars = "ROI",
                    variable.name = "marker",
                    value.name = "intensity")

# Apply asinh transformation once so plots and summary table are consistent
pixels_long <- pixels_long %>%
  mutate(intensity_asinh = asinh(intensity / 1))
```


# Histograms: Pixel Intensity Landscape (ROI 1)
```{r histograms, fig.width=14, fig.height=16}
roi1_data <- pixels_long %>%
  filter(ROI == "ROI_1")

ggplot(roi1_data, aes(x = intensity_asinh)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", size = 0.1, alpha = 0.8) +
  facet_wrap(~ marker, scales = "free", ncol = 5) +
  labs(
    title    = "Pixel Intensity Landscape: Patient 12 (ROI 1)",
    subtitle = "Histogram distribution of all 41 markers",
    x        = "arcsinh(Raw Intensity)",
    y        = "Pixel Count"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.text        = element_text(face = "bold", size = 9),
    panel.grid.minor  = element_blank(),
    axis.text.x       = element_text(size = 8),
    axis.text.y       = element_text(size = 8)
  )
```

# Density Ridges: Per-ROI Pixel Intensity Distributions (All Markers)
Each ridge represents one ROI. This makes it easy to compare distributions across ROIs and spot any ROI-level artifacts.

```{r ridge-plot, fig.width=14, fig.height=22}
ggplot(pixels_long, aes(x = intensity_asinh, y = ROI, fill = ROI)) +
  # scale < 1.0 guarantees ridges never physically overlap
  geom_density_ridges(alpha = 0.8, scale = 0.9, rel_min_height = 0.01, show.legend = FALSE) +
  facet_wrap(~ marker, scales = "free_x", ncol = 5) +
  labs(
    title    = "Patient 12 - Pixel Intensity Distributions per Marker",
    subtitle = "Arcsinh-Transformed, Cofactor = 1",
    x        = "arcsinh(Raw Intensity)",
    y        = "Region of Interest (ROI)"
  ) +
  theme_ridges(font_size = 10) +
  theme(
    strip.text      = element_text(face = "bold", size = 9),
    axis.text.y     = element_text(size = 8),
    axis.title      = element_text(size = 11, face = "bold"),
    axis.title.x    = element_text(hjust = 0.5),
    axis.title.y    = element_text(hjust = 0.5, angle = 90)
  )
```


# Summary Table: Marker Intensity and Sparsity
```{r summary-table}
marker_summary <- pixels_long %>%
  group_by(marker) %>%
  summarise(
    median_intensity = median(intensity_asinh),
    pct_zero         = mean(intensity == 0) * 100,  # zero check on raw counts
    max_intensity    = max(intensity_asinh),
    .groups          = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  arrange(desc(pct_zero))

print(marker_summary)

knitr::kable(marker_summary, caption = "Patient 12 Marker Intensity and Sparsity Summary (asinh-transformed)")
```


# Visualize one channel image (using the pixel coordinates)
```{r}
library(ggplot2)

# Read ROI 1
roi1 <- read.delim(txt_files[1], check.names = FALSE)

# Visualize CD19 as a spatial heatmap
ggplot(roi1, aes(x = X, y = Y, fill = asinh(`CD19(Nd142Di)` / 1))) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(title = "CD19 - ROI 1", fill = "asinh\nIntensity") +
  theme_void()

```

There are two very small bright spots in the lower left and upper right...potential artifacts?

# Visualize a selection of immune markers (ROI 1)
```{r}
# Downsample to every 3rd pixel in X and Y (~11% of pixels) for speed
set.seed(42)
roi1_sub <- roi1 %>%
  filter(X %% 3 == 0, Y %% 3 == 0)

cat("Downsampled pixels:", nrow(roi1_sub), "\n")

# Focused marker list
focused_markers <- c(
  "EpCAM(Nd144Di)",
  "SMA(Pr141Di)",
  "FoxP3(Dy162Di)",
  "CD68(Tb159Di)",
  "CD163(Sm147Di)",
  "CD1c(Sm154Di)",
  "TIM3(Nd143Di)",
  "PD1(Ho165Di)",
  "TTF1orCD16(Pt194Di)"
)

# Reshape to long format
roi1_focused <- roi1_sub %>%
  select(X, Y, all_of(focused_markers)) %>%
  melt(id.vars = c("X", "Y"),
       variable.name = "marker",
       value.name    = "intensity")

# Plot
p <- ggplot(roi1_focused, aes(x = X, y = Y, fill = asinh(intensity / 1))) +
  geom_tile() +
  scale_fill_viridis_c(
    option = "viridis",
    limits = c(0, quantile(roi1_focused$intensity_asinh, 0.99, na.rm = TRUE)),
    oob    = scales::squish  # values above the limit get clamped to max color
    ) +
  facet_wrap(~ marker, ncol = 3) +
  labs(
    title    = "Patient 12 ROI 1 - Selected Immune Markers",
    subtitle = "asinh transformed, downsampled every 3rd pixel",
    fill     = "asinh\nIntensity"
  ) +
  theme_void(base_size = 9) +
  theme(
    strip.text      = element_text(size = 8, face = "bold"),
    legend.position = "bottom"
  )

#ggsave("Patient12T_ROI1_focused_markers.pdf", p, width = 12, height = 14)
print(p)
```

Even after capping the color scale at the 99th percentile (to identify extreme outliers aka hot pixels), the plots still look the same. This probably means that the data doesn't have extreme hot pixels in these 9 markers (unlike CD19). Suggests these channels are relatively clean.

Takeaway: pixel-level visualization is most useful for structural markers like EpCAM and SMA and markers with broad expression. For rare immune populations like FoxP3, the signal is too sparse at the pixel. Steinbock is needed to segment cells first and them sum up the signal per cell, which will make those populations much more detectable.


