---
title: "Final Cell Type Annotation (R2) — Follow-up Diagnostic Checks"
author: "Tazein Shah"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 0. Setup

```{r libraries}
library(SpatialExperiment)
library(SummarizedExperiment)
library(tidyverse)
library(ggplot2)
```

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis/final_cell_annotation"
spe_path      <- file.path(analysis_path, "r2_annotation", "spe_final_celltype_r2.rds")
```

```{r load-data}
spe <- readRDS(spe_path)

if (!"exprs" %in% assayNames(spe)) {
  assay(spe, "exprs") <- asinh(assay(spe, "counts") / 1)
}

spe$patient_id <- as.character(as.vector(spe$patient_id))
spe$sample_id  <- as.character(as.vector(spe$sample_id))

print(table(spe$final_celltype, useNA = "ifany"))
```

```{r celltype-category}
celltype_category <- c(
  "B Cell (CD19+)"                                       = "B Cell/Plasma Cell",
  "B Cell, Weak (CD19+)"                                 = "B Cell/Plasma Cell",
  "B Cell/Plasma Cell, HLADRB-high (CD19+CD138+HLADRB+)" = "B Cell/Plasma Cell",
  "Plasma Cell, Memory-like (Weak) (CD45RO+)"            = "B Cell/Plasma Cell",
  "Plasma Cell, TIM3+ (CD138+TIM3+CD45RO-)"              = "B Cell/Plasma Cell",
  "Plasmablast, Strongly Activated (CD38++)"             = "B Cell/Plasma Cell",
  "Plasmablast/Activated B Cell (CD38+)"                 = "B Cell/Plasma Cell",
  "Transitional B Cell/Plasmablast (CD19+CD138+HLADRB+)" = "B Cell/Plasma Cell",

  "CD4+ T Cell, Activated"                               = "T Cell",
  "CD4+ T Cell, Exhaustion-leaning"                      = "T Cell",
  "CD8+ T Cell, Activated"                                = "T Cell",
  "CD8+ T Cell, Resting"                                  = "T Cell",
  "Memory T Cell"                                         = "T Cell",
  "Treg, Activated"                                       = "T Cell",
  "Treg, Resting"                                         = "T Cell",

  "cDC1"                                                  = "Myeloid",
  "DC/mregDC"                                             = "Myeloid",
  "DC/mregDC (Epi Adj)"                                   = "Myeloid",
  "DC/mregDC (LAMP3-)"                                    = "Myeloid",
  "DC/mregDC Weak"                                         = "Myeloid",
  "Dendritic Cell (CD11c+)"                               = "Myeloid",
  "Macrophage (CD68+)"                                    = "Myeloid",
  "Macrophage, Antigen-Presenting (HLADRB+)"              = "Myeloid",
  "Macrophage, M2-like (CD163+)"                          = "Myeloid",
  "Macrophage, Mixed Signal (PDL1+)"                      = "Myeloid",
  "Macrophage, Proliferating (HLADRB+)"                   = "Myeloid",
  "Macrophage, Proliferating (MKI67+)"                    = "Myeloid",
  "Monocyte (Generic)"                                    = "Myeloid",
  "Monocyte (PD1+)"                                       = "Myeloid",
  "Monocyte-Macrophage Transitional (CD14+)"              = "Myeloid",
  "Monocyte-Macrophage, Unresolved"                       = "Myeloid",
  "Monocyte, Unresolved"                                  = "Myeloid",
  "PMN-MDSC"                                               = "Myeloid",

  "Epithelial (EpCAM+)"                                   = "Epithelial/Stroma",
  "Fibroblast/Stroma"                                     = "Epithelial/Stroma",
  "Tumor Epithelial (Neuroendocrine)"                     = "Epithelial/Stroma",
  "Tumor Epithelial (TTF1++)"                             = "Epithelial/Stroma",

  "Mast Cell"                                             = "Other",
  "Proliferating Cells"                                   = "Other"
)

spe$celltype_category <- celltype_category[as.character(spe$final_celltype)]

unmapped <- setdiff(unique(as.character(spe$final_celltype)), names(celltype_category))
if (length(unmapped) > 0) {
  cat("Unmapped final_celltype labels -- add these to celltype_category:\n")
  print(unmapped)
} else {
  cat("All final_celltype labels mapped to a category.\n")
}

print(table(spe$celltype_category, useNA = "ifany"))
```

---

## 1. DC/mregDC (LAMP3-) — Residual LAMP3 Signal Check

The cluster is named LAMP3-negative but its LAMP3 median sits well
above the myeloid baseline. Check whether this is graded spillover
from the other DC/mregDC clusters, or a genuinely separable
LAMP3-intermediate population.

```{r lamp3neg-subset}
lamp3neg_label <- "DC/mregDC (LAMP3-)"

spe_myeloid_group <- spe[, spe$celltype_category == "Myeloid"]
spe_myeloid_group$final_celltype <- droplevels(factor(spe_myeloid_group$final_celltype))

idx_lamp3neg <- which(spe_myeloid_group$final_celltype == lamp3neg_label)
cat("Number of cells in", lamp3neg_label, ":", length(idx_lamp3neg), "\n")
```

```{r lamp3neg-marker-compare}
dc_labels <- c("DC/mregDC", "DC/mregDC (Epi Adj)", "DC/mregDC (LAMP3-)", "DC/mregDC Weak")

lamp3_means <- sapply(dc_labels, function(cl) {
  idx <- which(spe_myeloid_group$final_celltype == cl)
  mean(assay(spe_myeloid_group, "exprs")["LAMP3", idx])
})

cat("Mean LAMP3 (asinh) expression across DC/mregDC labels:\n")
print(lamp3_means)

cat("\nFor comparison, myeloid-wide LAMP3-negative baseline (cDC1, monocytes, macrophages):\n")
baseline_labels <- setdiff(unique(as.character(spe_myeloid_group$final_celltype)), dc_labels)
idx_baseline <- which(spe_myeloid_group$final_celltype %in% baseline_labels)
cat(mean(assay(spe_myeloid_group, "exprs")["LAMP3", idx_baseline]), "\n")
```

```{r lamp3neg-density}
make_density_plot <- function(spe, assay_name, cluster_col, cluster_label, marker, title) {
  expr_vec   <- assay(spe, assay_name)[marker, ]
  cluster_id <- spe[[cluster_col]]

  idx <- which(cluster_id == cluster_label)
  plot_df <- data.frame(expr = expr_vec[idx])

  ggplot(plot_df, aes(x = expr)) +
    geom_histogram(aes(y = after_stat(density)), bins = 60, fill = "steelblue", alpha = 0.6) +
    geom_density(color = "firebrick", linewidth = 0.8) +
    labs(title = title, x = "asinh expression", y = "Density") +
    theme_bw(base_size = 11)
}

p_lamp3neg_density <- make_density_plot(spe_myeloid_group, "exprs", "final_celltype",
                                        lamp3neg_label, "LAMP3",
                                        "DC/mregDC (LAMP3-): LAMP3 Density")
print(p_lamp3neg_density)
```

A single unimodal peak at an intermediate level (between true LAMP3-
myeloid and the other DC/mregDC clusters) argues for a graded/spillover
signal and supports a rename to something like "LAMP3-intermediate."
A bimodal shape with a genuine low-LAMP3 mode plus a smaller
higher-LAMP3 tail would instead suggest the negative-gating threshold
just needs adjusting, with most cells legitimately LAMP3-.

```{r lamp3neg-patient-composition}
cat("Patient composition of DC/mregDC (LAMP3-) cluster:\n")
print(sort(table(spe_myeloid_group$patient_id[idx_lamp3neg]), decreasing = TRUE))
```

---

## 2. Plasmablast, Strongly Activated (CD38++) — CD138 Check

Strongly activated plasmablasts are typically CD38+CD138+. This
cluster's CD138 median is negative, lower than several other B/plasma
clusters. Check whether that's a real phenotype or spillover from an
adjacent low-CD138 population.

```{r sa-cd138-subset}
sa_label <- "Plasmablast, Strongly Activated (CD38++)"

spe_bcell_group <- spe[, spe$celltype_category == "B Cell/Plasma Cell"]
spe_bcell_group$final_celltype <- droplevels(factor(spe_bcell_group$final_celltype))

idx_sa <- which(spe_bcell_group$final_celltype == sa_label)
cat("Number of cells in", sa_label, ":", length(idx_sa), "\n")
```

```{r sa-cd138-compare}
bcell_labels <- levels(spe_bcell_group$final_celltype)

cd138_means <- sapply(bcell_labels, function(cl) {
  idx <- which(spe_bcell_group$final_celltype == cl)
  mean(assay(spe_bcell_group, "exprs")["CD138", idx])
})

cat("Mean CD138 (asinh) expression across all B Cell/Plasma Cell labels:\n")
print(sort(cd138_means, decreasing = TRUE))
```

```{r sa-cd138-density}
p_sa_cd138_density <- make_density_plot(spe_bcell_group, "exprs", "final_celltype",
                                        sa_label, "CD138",
                                        "Plasmablast, Strongly Activated: CD138 Density")
print(p_sa_cd138_density)

p_sa_cd38_density <- make_density_plot(spe_bcell_group, "exprs", "final_celltype",
                                       sa_label, "CD38",
                                       "Plasmablast, Strongly Activated: CD38 Density")
print(p_sa_cd38_density)
```

A unimodal CD138-low peak alongside a clean unimodal CD38-high peak
would support a genuine CD38++CD138- phenotype (possibly an earlier
activation stage before CD138 upregulation). A bimodal CD138
distribution -- some cells clearly CD138+ mixed with a CD138- majority
-- would instead suggest an admixed population that could be split.

```{r sa-patient-composition}
cat("Patient composition of Plasmablast, Strongly Activated cluster:\n")
print(sort(table(spe_bcell_group$patient_id[idx_sa]), decreasing = TRUE))
```

---

## 3. Plasma Cell Clusters — Unexpected CD45RA Elevation

Plasma cells are terminally differentiated and typically CD45RA-low.
Two plasma-cell-labeled clusters show CD45RA near ceiling, comparable
to naive B cells. Check whether this reflects spillover from an
adjacent naive B cell population or true co-expression.

```{r cd45ra-subset}
cd45ra_check_labels <- c(
  "Plasma Cell, TIM3+ (CD138+TIM3+CD45RO-)",
  "B Cell/Plasma Cell, HLADRB-high (CD19+CD138+HLADRB+)"
)

cd45ra_means <- sapply(bcell_labels, function(cl) {
  idx <- which(spe_bcell_group$final_celltype == cl)
  mean(assay(spe_bcell_group, "exprs")["CD45RA", idx])
})

cat("Mean CD45RA (asinh) expression across all B Cell/Plasma Cell labels:\n")
print(sort(cd45ra_means, decreasing = TRUE))
```

```{r cd45ra-density}
for (lbl in cd45ra_check_labels) {
  p <- make_density_plot(spe_bcell_group, "exprs", "final_celltype",
                          lbl, "CD45RA",
                          paste(lbl, ": CD45RA Density"))
  print(p)
}
```

```{r cd45ra-cd138-crosscheck}
# If CD45RA-high cells in these clusters are also CD138-low, that's
# consistent with naive B cell contamination rather than a genuine
# CD45RA+ plasma cell state (plasma cells should be CD138+ almost by
# definition given how they were gated into this category).
for (lbl in cd45ra_check_labels) {
  idx <- which(spe_bcell_group$final_celltype == lbl)
  cd45ra_vals <- assay(spe_bcell_group, "exprs")["CD45RA", idx]
  cd138_vals  <- assay(spe_bcell_group, "exprs")["CD138", idx]

  cat(lbl, "\n")
  cat("  Correlation CD45RA vs CD138:", round(cor(cd45ra_vals, cd138_vals), 3), "\n")
  cat("  Proportion CD45RA-high (>2) AND CD138-low (<0):",
      round(mean(cd45ra_vals > 2 & cd138_vals < 0), 3), "\n\n")
}
```

A negative CD45RA-CD138 correlation, with a meaningful share of cells
CD45RA-high/CD138-low, points to admixture with a naive B cell
population rather than a real CD45RA+ plasma cell phenotype. A
positive or flat correlation, with CD45RA-high cells still CD138+,
would argue the co-expression is real and worth a note rather than a
split.

```{r cd45ra-patient-composition}
for (lbl in cd45ra_check_labels) {
  idx <- which(spe_bcell_group$final_celltype == lbl)
  cat("Patient composition of", lbl, ":\n")
  print(sort(table(spe_bcell_group$patient_id[idx]), decreasing = TRUE))
  cat("\n")
}
```

---

## 4. Monocyte, Unresolved — CD138 Check

CD138 is a plasma cell marker with no expected myeloid role, but this
"unresolved" monocyte cluster has the highest CD138 mean in the entire
Myeloid category. Check whether this reflects genuine cross-lineage
spillover consistent with an "unresolved" catch-all, or a specific
subpopulation worth pulling out.

```{r monounresolved-subset}
mono_unresolved_label <- "Monocyte, Unresolved"

idx_mono_unresolved <- which(spe_myeloid_group$final_celltype == mono_unresolved_label)
cat("Number of cells in", mono_unresolved_label, ":", length(idx_mono_unresolved), "\n")
```

```{r monounresolved-cd138-compare}
myeloid_labels <- levels(spe_myeloid_group$final_celltype)

cd138_means_myeloid <- sapply(myeloid_labels, function(cl) {
  idx <- which(spe_myeloid_group$final_celltype == cl)
  mean(assay(spe_myeloid_group, "exprs")["CD138", idx])
})

cat("Mean CD138 (asinh) expression across all Myeloid labels:\n")
print(sort(cd138_means_myeloid, decreasing = TRUE))
```

```{r monounresolved-cd138-density}
p_mono_cd138_density <- make_density_plot(spe_myeloid_group, "exprs", "final_celltype",
                                          mono_unresolved_label, "CD138",
                                          "Monocyte, Unresolved: CD138 Density")
print(p_mono_cd138_density)
```

A bimodal CD138 distribution -- most cells low, a distinct minority
high -- would suggest a genuine plasma-cell-admixed subpopulation
buried in the "unresolved" bucket that could potentially be pulled out
with targeted resubclustering. A unimodal, moderately-elevated peak
would just reflect diffuse cross-lineage noise consistent with an
honest "unresolved" label, not warranting further splitting.

```{r monounresolved-patient-composition}
cat("Patient composition of Monocyte, Unresolved cluster:\n")
print(sort(table(spe_myeloid_group$patient_id[idx_mono_unresolved]), decreasing = TRUE))
```

