---
title: "Epithelial Cell Subclustering — Run5 Extraction"
author: "Tazein Shah"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(42)
```

## Overview

Extract epithelial clusters (9, 12, 17, 19) from the annotated run5 SPE
(`h5_cluster_annot.rds`) and recluster on the existing Harmony
embedding (Louvain, k = 20).

---

## 1. Load Libraries

```{r libraries}
library(imcRtools)
library(scran)
library(bluster)
library(BiocParallel)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(scater)
library(patchwork)
library(RColorBrewer)
```

---

## 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

spe_path <- file.path(analysis_path, "final_cell_annotation", "h5_cluster_annot.rds")

epithelial_clustering_path <- file.path(analysis_path, "final_cell_annotation", "subclustering", "epithelial")

out_path <- file.path(epithelial_clustering_path, "epithelial_subclustering.rds")
```

---

## 3. Load Annotated SPE

```{r load-data}
spe <- readRDS(spe_path)

print(table(spe$broad_clusters))
```

---

## 4. Extract Epithelial Clusters

```{r subset}
epithelial_clusters <- c("9", "12", "17", "19")

spe_epithelial <- spe[, as.character(spe$broad_clusters) %in% epithelial_clusters]

cat("Cells in epithelial subset:", ncol(spe_epithelial), "\n")
cat("From cluster 9 (Tumor Epithelial, TTF1+):", sum(spe_epithelial$broad_clusters == "9"), "\n")
cat("From cluster 12 (Epithelial, low confidence):", sum(spe_epithelial$broad_clusters == "12"), "\n")
cat("From cluster 17 (Unknown/Low Quality):", sum(spe_epithelial$broad_clusters == "17"), "\n")
cat("From cluster 19 (Tumor Epithelial Neuroendocrine, CD56+EpCAM+):", sum(spe_epithelial$broad_clusters == "19"), "\n")
```

---

## 5. Recluster: Louvain on Existing Harmony Embedding (k = 20)

```{r recluster}
if (file.exists(out_path)) {
  cat("Output already exists, loading instead of reclustering:", out_path, "\n")
  spe_epithelial <- readRDS(out_path)
} else {
  snn_graph <- buildSNNGraph(spe_epithelial, use.dimred = "Harmony", k = 20,
                              BPPARAM = SerialParam())

  clusters <- igraph::cluster_louvain(snn_graph)$membership

  spe_epithelial$epithelial_subclusters <- factor(clusters)
}

cat("Number of subclusters:", nlevels(spe_epithelial$epithelial_subclusters), "\n")
print(table(spe_epithelial$epithelial_subclusters))
```

---

## 6. UMAP on the Subset

```{r umap-sub}
if ("UMAP_sub" %in% reducedDimNames(spe_epithelial)) {
  cat("UMAP_sub already exists, skipping recomputation.\n")
} else {
  set.seed(220225)
  spe_epithelial <- runUMAP(spe_epithelial, dimred = "Harmony", name = "UMAP_sub")
}
```

```{r plot-umap-sub-original, fig.width=10, fig.height=7}
p_sub_original <- dittoDimPlot(spe_epithelial, var = "broad_celltype",
                               reduction.use = "UMAP_sub",
                               size = 0.4) +
  ggtitle("Epithelial Subset: UMAP colored by Original Broad Label") +
  theme(legend.position = "right")
print(p_sub_original)
```

```{r plot-patient-composition, fig.width=10, fig.height=5}
patient_counts <- as.data.frame(table(patient_id = spe_epithelial$patient_id))

p_patient_comp <- ggplot(patient_counts, aes(x = patient_id, y = Freq, fill = patient_id)) +
  geom_col() +
  labs(title = "Epithelial Subset: Cell Count by Patient",
       x = "Patient", y = "Cell Count") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
print(p_patient_comp)

cat("Epithelial subset composition by patient:\n")
print(patient_counts[order(-patient_counts$Freq), ])
```

---

## 7. Marker Sets for Epithelial Subtyping

```{r markers}
# core epithelial/tumor lineage markers
# note: "TTF1" is measured on the channel named "TTF1orCD16" -- TTF1 is the
# actual marker on that channel, not CD16
lineage_markers <- c("EpCAM", "TTF1orCD16", "CD56")

# functional/state markers relevant to tumor epithelium
state_markers <- c("MKI67", "PDL1", "HLADRB")

# admixture / spillover flags (non-epithelial lineage markers included in
# the pixel-filtration mutual-exclusion group -- useful for catching
# doublets or spatial bleed from stromal/immune lineages)
admixture_markers <- c("CD68", "SMA", "CD117", "CD138")

lineage_markers   <- lineage_markers[lineage_markers %in% rownames(spe_epithelial)]
state_markers     <- state_markers[state_markers %in% rownames(spe_epithelial)]
admixture_markers <- admixture_markers[admixture_markers %in% rownames(spe_epithelial)]

epithelial_markers <- c(lineage_markers, state_markers, admixture_markers)

cat("Lineage markers found:", paste(lineage_markers, collapse = ", "), "\n")
cat("State markers found:", paste(state_markers, collapse = ", "), "\n")
cat("Admixture markers found:", paste(admixture_markers, collapse = ", "), "\n")
```

---

## 8. Violin Plots per Subcluster

```{r make-violinplot}
make_violinplot <- function(spe, assay_name, cluster_col, marker_set, title, ncol = 3) {
  expr_mat   <- assay(spe, assay_name)[marker_set, , drop = FALSE]
  cluster_id <- spe[[cluster_col]]

  mean_expr <- sapply(levels(cluster_id), function(cl) {
    idx <- which(cluster_id == cl)
    rowMeans(expr_mat[, idx, drop = FALSE])
  })

  if (is.matrix(mean_expr) && nrow(mean_expr) > 1) {
    marker_order <- rownames(mean_expr)[hclust(dist(mean_expr))$order]
  } else {
    marker_order <- marker_set
  }

  plot_df <- as.data.frame(t(expr_mat)) |>
    mutate(cluster = cluster_id) |>
    pivot_longer(-cluster, names_to = "marker", values_to = "expr")

  plot_df$marker  <- factor(plot_df$marker, levels = marker_order)
  plot_df$cluster <- factor(plot_df$cluster, levels = levels(cluster_id))

  ggplot(plot_df, aes(x = cluster, y = expr, fill = cluster)) +
    geom_violin(scale = "width", trim = TRUE, linewidth = 0.2) +
    stat_summary(fun = median, geom = "point", size = 0.6, color = "black") +
    facet_wrap(~ marker, scales = "free_y", ncol = ncol) +
    labs(title = title, x = "Subcluster", y = "Z-score expression") +
    theme_bw(base_size = 11) +
    theme(
      axis.text.x      = element_text(angle = 0, hjust = 0.5, size = 9),
      strip.text        = element_text(face = "bold"),
      panel.grid.major.x = element_blank(),
      legend.position   = "none"
    )
}
```

```{r violin-epithelial-global-lineage, fig.width=10, fig.height=5}
p_violin_lineage_global <- make_violinplot(spe_epithelial, "scaled_0100", "epithelial_subclusters",
                                           lineage_markers,
                                           "Epithelial Subset: Lineage Markers per Subcluster (Global Z-score)",
                                           ncol = 3)
print(p_violin_lineage_global)
```

```{r violin-epithelial-global-state, fig.width=12, fig.height=8}
p_violin_state_global <- make_violinplot(spe_epithelial, "scaled_0100", "epithelial_subclusters",
                                         state_markers,
                                         "Epithelial Subset: State Markers per Subcluster (Global Z-score)",
                                         ncol = 3)
print(p_violin_state_global)
```

```{r violin-epithelial-global-admixture, fig.width=10, fig.height=5}
p_violin_admixture_global <- make_violinplot(spe_epithelial, "scaled_0100", "epithelial_subclusters",
                                             admixture_markers,
                                             "Epithelial Subset: Admixture Markers per Subcluster (Global Z-score)",
                                             ncol = 3)
print(p_violin_admixture_global)
```

Subset-relative renormalization: since `scaled_0100` was capped and Z-scored
over the full run5 dataset, this recomputes the same cap+Z-score recipe
from raw counts using only this epithelial subset's own distribution --
one normalization layer, just a different denominator.

```{r renorm-subset}
cap_99 <- function(mat) {
  t(apply(mat, 1, function(x) {
    q99 <- quantile(x, 0.99)
    pmin(x, q99)
  }))
}

zscore_rows <- function(mat) {
  t(scale(t(mat)))
}

mat_sub_0100 <- counts(spe_epithelial) |> cap_99() |> zscore_rows()
assay(spe_epithelial, "scaled_0100_sub") <- mat_sub_0100
```

```{r violin-epithelial-subset-lineage, fig.width=10, fig.height=5}
p_violin_lineage_sub <- make_violinplot(spe_epithelial, "scaled_0100_sub", "epithelial_subclusters",
                                        lineage_markers,
                                        "Epithelial Subset: Lineage Markers per Subcluster (Subset-Renormalized Z-score)",
                                        ncol = 3)
print(p_violin_lineage_sub)
```

```{r violin-epithelial-subset-state, fig.width=12, fig.height=8}
p_violin_state_sub <- make_violinplot(spe_epithelial, "scaled_0100_sub", "epithelial_subclusters",
                                      state_markers,
                                      "Epithelial Subset: State Markers per Subcluster (Subset-Renormalized Z-score)",
                                      ncol = 3)
print(p_violin_state_sub)
```

```{r violin-epithelial-subset-admixture, fig.width=10, fig.height=5}
p_violin_admixture_sub <- make_violinplot(spe_epithelial, "scaled_0100_sub", "epithelial_subclusters",
                                          admixture_markers,
                                          "Epithelial Subset: Admixture Markers per Subcluster (Subset-Renormalized Z-score)",
                                          ncol = 3)
print(p_violin_admixture_sub)
```

---

## 9. Heatmap: All-Subcluster Expression Overview

Findings from this check: clusters 3 and 6 both showed a selective
positive signal isolated to EpCAM and TTF1orCD16 (the only positive
values in either row), distinguishing them as weak-but-real tumor
epithelial populations rather than flat/ambiguous clusters. Clusters
5, 8, and 9 showed no such selective signal -- negative or flat across
essentially the entire marker panel, confirming Unknown/Low Quality.
Cluster 1 retained the highest CD117 and highest HLADRB of any
subcluster alongside its EpCAM/TTF1orCD16 signal, consistent with the
immune-admixture flag raised from the violin plots.

```{r unknown-heatmap, fig.width=10, fig.height=6}
all_clusters <- levels(spe_epithelial$epithelial_subclusters)

mean_expr_all <- sapply(all_clusters, function(cl) {
  idx <- which(spe_epithelial$epithelial_subclusters == cl)
  rowMeans(assay(spe_epithelial, "scaled_0100")[epithelial_markers, idx, drop = FALSE])
})

marker_order <- epithelial_markers[hclust(dist(mean_expr_all))$order]

cluster_sizes <- table(spe_epithelial$epithelial_subclusters)
cluster_labels <- paste0(all_clusters, "\n(n=", cluster_sizes[all_clusters], ")")
names(cluster_labels) <- all_clusters

heatmap_df <- as.data.frame(mean_expr_all) |>
  rownames_to_column("marker") |>
  pivot_longer(-marker, names_to = "subcluster", values_to = "mean_expr")

heatmap_df$marker     <- factor(heatmap_df$marker, levels = marker_order)
heatmap_df$subcluster <- factor(heatmap_df$subcluster, levels = all_clusters)
heatmap_df$is_ambiguous <- heatmap_df$subcluster %in% c("3", "5", "6", "8", "9")

p_unknown_heatmap <- ggplot(heatmap_df, aes(x = subcluster, y = marker, fill = mean_expr)) +
  geom_tile(color = "white") +
  geom_tile(data = subset(heatmap_df, is_ambiguous), fill = NA, color = "black", linewidth = 0.8) +
  geom_text(aes(label = round(mean_expr, 2)), size = 3) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, name = "Mean\nZ-score") +
  scale_x_discrete(labels = cluster_labels) +
  labs(title = "Mean Marker Expression: All Subclusters (Ambiguous/Low Quality Outlined)",
       x = "Subcluster (cell count)", y = "Marker") +
  theme_bw(base_size = 11) +
  theme(panel.grid = element_blank())
print(p_unknown_heatmap)
```

---

## 10. Manual Subcluster Annotation

```{r annotate-sub, echo=TRUE}
epithelial_subannotation <- c(
  "1"  = "Tumor Epithelial (TTF1+EpCAM+)",  # TTF1orCD16 and EpCAM both elevated; also highest CD117 and highest HLADRB of any subcluster -- worth an origin check before treating as clean tumor epithelium
  "2"  = "Tumor Epithelial (TTF1+EpCAM+)",  # TTF1orCD16 highest of all subclusters (1.9); EpCAM (1.34) also higher than cluster 1's EpCAM (1.08), which carries the same dual label -- PDL1 mildly elevated
  "3"  = "Tumor Epithelial, Weak (TTF1+EpCAM+)",  # Heatmap shows EpCAM (0.32) and TTF1orCD16 (0.48) as the only two positive values in the row, all other markers negative -- selective epithelial signal, not flat; distinguishes it from truly flat clusters 5/8/9
  "4"  = "Tumor Epithelial (EpCAM+TTF1+)",  # EpCAM highest of all subclusters (2.25); TTF1orCD16 (0.5) is also higher than cluster 3's TTF1orCD16 (0.48), which carries a TTF1+ tag -- should be tagged here too
  "5"  = "Unknown/Low Quality",  # Flat to negative across nearly every marker in the heatmap, including EpCAM/TTF1orCD16 -- no selective lineage signal
  "6"  = "Tumor Epithelial, Weak (TTF1+EpCAM+)",  # Same pattern as cluster 3: EpCAM (0.53) and TTF1orCD16 (0.21) the only positive values, rest negative
  "7"  = "Tumor Epithelial (TTF1+EpCAM+)",  # TTF1orCD16 (1.29) elevated; EpCAM (0.71) is also higher than clusters 3 (0.32) and 6 (0.53), both of which carry EpCAM+ tags -- should be tagged here too
  "8"  = "Unknown/Low Quality",  # Flat to negative across lineage, state, and admixture markers alike
  "9"  = "Unknown/Low Quality",  # Most uniformly negative subcluster across the panel, including lowest CD138
  "10" = "Tumor Epithelial Neuroendocrine (CD56+EpCAM+)",  # CD56 far above every other subcluster (3.16); EpCAM (0.53) is real and positive, same magnitude as cluster 6's tagged EpCAM -- matches the CD56+EpCAM+ signature already on record for this cluster from broad clustering
  "11" = "Proliferative, Lineage-Low (MKI67+)"  # MKI67 highest of all subclusters, PDL1 elevated, but EpCAM/TTF1orCD16/CD56 all low -- real signal (not flat like 5/8/9), but lacks epithelial lineage markers
)

if (length(epithelial_subannotation) > 0) {
  spe_epithelial$sub_celltype <- epithelial_subannotation[as.character(spe_epithelial$epithelial_subclusters)]
  cat("Subcluster annotation applied.\n")
  print(table(spe_epithelial$sub_celltype, useNA = "ifany"))
} else {
  cat("No subcluster annotation yet. Fill in epithelial_subannotation above.\n")
}
```

---

## 11. UMAP Colored by Subcluster Cell Type (Pre-Removal)

```{r umap-sub-celltype, fig.width=10, fig.height=5}
spe_epithelial$sub_celltype <- factor(spe_epithelial$sub_celltype, levels = unique(epithelial_subannotation))

p_sub_celltype <- dittoDimPlot(spe_epithelial, var = "sub_celltype",
                               reduction.use = "UMAP_sub",
                               size = 0.4) +
  ggtitle("Epithelial Subset: UMAP colored by Subcluster Cell Type (Pre-Removal)") +
  theme(legend.position = "right")
print(p_sub_celltype)
```

---

## 12. Remove Unknown/Low Quality Clusters

Clusters 5, 8, and 9 were flat to negative across essentially the
entire marker panel in the heatmap above -- no selective lineage
signal, unlike clusters 3 and 6.

```{r remove-unknown}
spe_epithelial_clean <- spe_epithelial[, !(spe_epithelial$epithelial_subclusters %in% c("5", "8", "9"))]

cat("Cells before removal:", ncol(spe_epithelial), "\n")
cat("Cells after removing Unknown/Low Quality (clusters 5, 8, 9):", ncol(spe_epithelial_clean), "\n")
```

---

## 13. Remove Proliferative Cluster

Cluster 11 (Proliferative, Lineage-Low) had the highest MKI67 of any
subcluster but lacked EpCAM/TTF1orCD16/CD56 -- a real signal, but not
an epithelial lineage call, so it's removed alongside the
Unknown/Low Quality clusters before reclustering.

```{r remove-proliferative}
cells_before_prolif_removal <- ncol(spe_epithelial_clean)

spe_epithelial_clean <- spe_epithelial_clean[, spe_epithelial_clean$epithelial_subclusters != "11"]

cat("Cells before removing Proliferative cluster (11):", cells_before_prolif_removal, "\n")
cat("Cells after removing Proliferative cluster (11):", ncol(spe_epithelial_clean), "\n")
```

---

## 14. Recluster After Removal

Recluster on the existing Harmony embedding (same k = 20 as the
original subclustering) to check whether the remaining subclusters
hold up once the Unknown/Low Quality and Proliferative cells are gone.

```{r recluster-after-removal}
if ("epithelial_subclusters_clean" %in% colnames(colData(spe_epithelial_clean))) {
  cat("epithelial_subclusters_clean already exists, skipping recomputation.\n")
} else {
  set.seed(220225)
  snn_graph_clean <- buildSNNGraph(spe_epithelial_clean, use.dimred = "Harmony", k = 20,
                                    BPPARAM = SerialParam())
  clusters_clean <- igraph::cluster_louvain(snn_graph_clean)$membership
  spe_epithelial_clean$epithelial_subclusters_clean <- factor(clusters_clean)
}

cat("Number of subclusters after removal:", nlevels(spe_epithelial_clean$epithelial_subclusters_clean), "\n")
print(table(spe_epithelial_clean$epithelial_subclusters_clean))
```

---

## 15. Violin Plots on Post-Removal Subclusters

```{r violin-clean, fig.width=14, fig.height=12}
p_violin_clean <- make_violinplot(spe_epithelial_clean, "scaled_0100", "epithelial_subclusters_clean",
                                  epithelial_markers,
                                  "Epithelial Subset (Unknown/Proliferative-Removed): Marker Expression per Subcluster",
                                  ncol = 3)
print(p_violin_clean)
```

---

## 16. Heatmap: All Post-Removal Subclusters

Same approach as the pre-removal heatmap (chunk 9) -- mean Z-score
expression across the full marker panel for all post-removal
subclusters side by side.

```{r heatmap-clean, fig.width=8, fig.height=5}
clean_clusters <- levels(spe_epithelial_clean$epithelial_subclusters_clean)

mean_expr_clean <- sapply(clean_clusters, function(cl) {
  idx <- which(spe_epithelial_clean$epithelial_subclusters_clean == cl)
  rowMeans(assay(spe_epithelial_clean, "scaled_0100")[epithelial_markers, idx, drop = FALSE])
})

marker_order_clean <- epithelial_markers[hclust(dist(mean_expr_clean))$order]

cluster_sizes_clean <- table(spe_epithelial_clean$epithelial_subclusters_clean)
cluster_labels_clean <- paste0(clean_clusters, "\n(n=", cluster_sizes_clean[clean_clusters], ")")
names(cluster_labels_clean) <- clean_clusters

heatmap_clean_df <- as.data.frame(mean_expr_clean) |>
  rownames_to_column("marker") |>
  pivot_longer(-marker, names_to = "subcluster", values_to = "mean_expr")

heatmap_clean_df$marker     <- factor(heatmap_clean_df$marker, levels = marker_order_clean)
heatmap_clean_df$subcluster <- factor(heatmap_clean_df$subcluster, levels = clean_clusters)

p_heatmap_clean <- ggplot(heatmap_clean_df, aes(x = subcluster, y = marker, fill = mean_expr)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(mean_expr, 2)), size = 3) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, name = "Mean\nZ-score") +
  scale_x_discrete(labels = cluster_labels_clean) +
  labs(title = "Mean Marker Expression: All Post-Removal Subclusters",
       x = "Subcluster (cell count)", y = "Marker") +
  theme_bw(base_size = 11) +
  theme(panel.grid = element_blank())
print(p_heatmap_clean)
```

---

## 17. Manual Annotation of Post-Removal Subclusters

```{r annotate-sub-clean, echo=TRUE}
epithelial_subannotation_clean <- c(
  "1"  = "Tumor Epithelial (TTF1+EpCAM+)",  # TTF1orCD16=1.28, EpCAM=1.36; also highest CD117 (1.45), highest HLADRB (0.91), and highest MKI67 (0.62) of any post-removal subcluster -- same immune-admixture flag as pre-removal cluster 1
  "2"  = "Tumor Epithelial (TTF1+EpCAM+)",  # TTF1orCD16=1.54, EpCAM=0.4 -- both positive, rest of panel near/below zero
  "3"  = "Tumor Epithelial (TTF1+EpCAM+)",  # TTF1orCD16 highest of all post-removal subclusters (1.94), EpCAM=1.48; PDL1 (0.44) and CD56 (0.35) mildly elevated but well below cluster 1's admixture-level signal
  "4"  = "Tumor Epithelial (EpCAM+)",  # EpCAM=1.86, second highest overall; TTF1orCD16 near zero (-0.05) -- unlike the pre-removal solution, this cluster carries no TTF1 signal
  "5"  = "Tumor Epithelial, Weak (TTF1+EpCAM+)",  # TTF1orCD16=0.21, EpCAM=0.38 -- modest signal; CD117 (0.62) second-highest after cluster 1, worth a secondary admixture check
  "6"  = "Tumor Epithelial (EpCAM+TTF1+)",  # EpCAM highest of all post-removal subclusters (2.41), TTF1orCD16=0.85
  "7"  = "Tumor Epithelial, Weak (TTF1+)",  # TTF1orCD16=0.46 modest positive; EpCAM=0.18 below the threshold used elsewhere for a dual tag -- rest of panel negative
  "8"  = "Tumor Epithelial, Weak (TTF1+EpCAM+)",  # TTF1orCD16=0.22, EpCAM=0.53 -- modest dual signal, rest of panel negative
  "9"  = "Tumor Epithelial (TTF1+EpCAM+)",  # TTF1orCD16=1.16, EpCAM=0.96; CD56 (0.41) mildly elevated but well below cluster 10's dominant neuroendocrine signal
  "10" = "Tumor Epithelial Neuroendocrine (CD56+EpCAM+)"  # CD56 dominant (3.16), EpCAM=0.53; SMA (0.02) is the only non-negative SMA value in the panel, same minor stromal-admixture note as pre-removal
)

if (length(epithelial_subannotation_clean) > 0) {
  spe_epithelial_clean$sub_celltype_clean <- epithelial_subannotation_clean[as.character(spe_epithelial_clean$epithelial_subclusters_clean)]
  cat("Post-removal subcluster annotation applied.\n")
  print(table(spe_epithelial_clean$sub_celltype_clean, useNA = "ifany"))
} else {
  cat("No post-removal annotation yet. Fill in epithelial_subannotation_clean above.\n")
}
```

---

## 18. UMAP Colored by Subcluster Cell Type (Post-Removal)

```{r umap-sub-celltype-clean, fig.width=10, fig.height=5}
spe_epithelial_clean$sub_celltype_clean <- factor(spe_epithelial_clean$sub_celltype_clean,
                                                   levels = unique(epithelial_subannotation_clean))

p_sub_celltype_clean <- dittoDimPlot(spe_epithelial_clean, var = "sub_celltype_clean",
                                     reduction.use = "UMAP_sub",
                                     size = 0.4) +
  ggtitle("Epithelial Subset (Unknown/Proliferative-Removed): UMAP colored by Subcluster Cell Type (Post-Removal)") +
  theme(legend.position = "right")
print(p_sub_celltype_clean)
```

---

## 19. Neuroendocrine Cluster: Spatial Island Composition Check

The UMAP above shows the neuroendocrine cluster (CD56+EpCAM+) split
into two disconnected islands rather than one contiguous region. This
checks whether that split is driven by patient/ROI composition (i.e.
two distinct populations) or just an embedding artifact (i.e. one
population, split apart by the layout).

```{r neuroendocrine-island-check, fig.width=10, fig.height=5}
ne_label <- "Tumor Epithelial Neuroendocrine (CD56+EpCAM+)"
ne_mask  <- spe_epithelial_clean$sub_celltype_clean == ne_label

ne_umap <- reducedDim(spe_epithelial_clean, "UMAP_sub")[ne_mask, ]

# threshold based on the visual gap in the UMAP between the main cloud
# (around UMAP2 ~ 0) and the isolated island (around UMAP2 ~ -7)
ne_island <- ifelse(ne_umap[, 2] < -4, "Isolated Island", "Main Cloud")

ne_patient_ids <- as.character(spe_epithelial_clean$patient_id[ne_mask])

ne_composition <- as.data.frame(table(island = ne_island, patient_id = ne_patient_ids))
ne_composition <- ne_composition[ne_composition$Freq > 0, ]

cat("Neuroendocrine cluster cell counts by island:\n")
print(table(ne_island))

cat("\nPatient composition by island:\n")
print(ne_composition[order(ne_composition$island, -ne_composition$Freq), ])

p_ne_island <- ggplot(ne_composition, aes(x = patient_id, y = Freq, fill = patient_id)) +
  geom_col() +
  facet_wrap(~ island, scales = "free_x") +
  labs(title = "Neuroendocrine Cluster: Patient Composition by Spatial Island",
       x = "Patient", y = "Cell Count") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(face = "bold"),
        legend.position = "none")
print(p_ne_island)
```
