---
title: "Tumor IMC - Full Cohort Cytomapper Catalog (All Cells)"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: cosmo
    code_folding: hide
---

Cytomapper "All cells" overlays for every ROI in the cohort, organized as
Patient tabs. Colored by the same curated 13-category display grouping
used throughout the poster figures.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = FALSE)
```

```{r libraries}
#.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

suppressPackageStartupMessages({
    library(imcRtools)
    library(cytomapper)
    library(SpatialExperiment)
    library(tidyverse)
    library(scales)
})

set.seed(12345)
```

```{r paths}
analysis_path  <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"
steinbock_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/steinbock"

final_celltype_path <- file.path(analysis_path, "final_cell_annotation",
                                  "r3_annotation", "spe_final_celltype_r3.rds")

# TEST_MODE: set TRUE to run on just the first 2 patients, to sanity-check
# paths/labels/rendering before committing to the full cohort.
TEST_MODE <- FALSE
```

```{r load-spe}
spe <- readRDS(final_celltype_path)
cat("Cells:", ncol(spe), "\n")
cat("Patients:", n_distinct(spe$patient_id), "\n")
cat("ROIs:", n_distinct(spe$sample_id), "\n")
```

# Build ROI -> file lookup

```{r build-file-lookup}
result_dirs <- list.dirs(steinbock_path, recursive = FALSE)
result_dirs <- result_dirs[grepl("^p[0-9]+(_[0-9]+_[0-9]+)?_results$", basename(result_dirs))]

cat("Steinbock result folders found:\n")
print(basename(result_dirs))

img_files  <- list.files(file.path(result_dirs, "img"), pattern = "\\.tiff?$",
                          full.names = TRUE, recursive = FALSE)
mask_files <- list.files(file.path(result_dirs, "masks_deepcell"), pattern = "\\.tiff?$",
                          full.names = TRUE, recursive = FALSE)

names(img_files)  <- tools::file_path_sans_ext(basename(img_files))
names(mask_files) <- tools::file_path_sans_ext(basename(mask_files))

missing_img  <- setdiff(unique(spe$sample_id), names(img_files))
missing_mask <- setdiff(unique(spe$sample_id), names(mask_files))

if (length(missing_img) > 0) {
  cat("WARNING -- sample_ids with no matching image file:\n")
  print(missing_img)
}
if (length(missing_mask) > 0) {
  cat("WARNING -- sample_ids with no matching mask file:\n")
  print(missing_mask)
}
if (length(missing_img) == 0 && length(missing_mask) == 0) {
  cat("All sample_ids matched to an image and mask file.\n")
}
```

# Build curated display grouping

```{r derive-display-celltype}
spe$CellTypeLabel_display <- case_when(
  spe$final_celltype == "Epithelial (EpCAM+)" ~ "Epithelial",
  str_detect(spe$final_celltype, "Tumor Epithelial") ~ "Lung Epithelial",
  spe$final_celltype == "Proliferating Cells" ~ "Proliferating Cells",
  str_detect(spe$final_celltype, "^CD8") ~ "CD8+ T Cell",
  str_detect(spe$final_celltype, "^CD4") ~ "CD4+ T Cell",
  str_detect(spe$final_celltype, "^Treg") ~ "Treg",
  str_detect(spe$final_celltype, "^cDC|DC/mregDC|Dendritic Cell") ~ "Dendritic Cell",
  spe$final_celltype == "Macrophage, Proliferating (MKI67+)" ~ "Other",  # Patient10-driven artifact
  str_detect(spe$final_celltype, "^Macrophage") ~ "Macrophage",
  str_detect(spe$final_celltype, "^Monocyte") ~ "Monocyte",
  str_detect(spe$final_celltype, "B Cell|Plasma") ~ "B Cell/Plasmablast",
  spe$final_celltype == "Fibroblast/Stroma" ~ "Fibroblast",
  spe$final_celltype == "Mast Cell" ~ "Mast Cell",
  TRUE ~ "Other"
)

display_levels <- c("Epithelial", "Lung Epithelial", "Proliferating Cells",
                     "CD8+ T Cell", "CD4+ T Cell", "Treg", "Dendritic Cell",
                     "Macrophage", "Monocyte", "B Cell/Plasmablast",
                     "Fibroblast", "Mast Cell", "Other")
spe$CellTypeLabel_display <- factor(spe$CellTypeLabel_display, levels = display_levels)

display_cols <- c(
  "Epithelial"          = "red",
  "Lung Epithelial"     = "darkorange",
  "Proliferating Cells"  = "yellow",
  "CD8+ T Cell"          = "forestgreen",
  "CD4+ T Cell"          = "yellowgreen",
  "Treg"                 = "purple",
  "Dendritic Cell"       = "lightblue",
  "Macrophage"           = "dodgerblue",
  "Monocyte"             = "navy",
  "B Cell/Plasmablast"   = "pink",
  "Fibroblast"           = "violet",
  "Mast Cell"            = "deeppink",
  "Other"                = "gray85"
)

table(spe$CellTypeLabel_display)
```

# Cohort catalog {.tabset}

```{r roi-list}
patients <- sort(unique(spe$patient_id))
if (TEST_MODE) {
  patients <- head(patients, 2)
  cat("TEST_MODE active -- rendering only:", paste(patients, collapse = ", "), "\n")
}
```

```{r render-catalog, results="asis", fig.width=14, fig.height=10}
for (pt in patients) {

  cat("\n\n## ", pt, " {.tabset}\n\n")

  pt_rois <- sort(unique(spe$sample_id[spe$patient_id == pt]))

  for (roi in pt_rois) {

    if (!(roi %in% names(img_files)) || !(roi %in% names(mask_files))) {
      cat("\n\n### ", roi, " {-}\n\n")
      cat("Skipped -- no matching image/mask file found (see lookup warnings above).\n\n")
      next
    }

    cat("\n\n### ", roi, " {-}\n\n")

    cur_image <- loadImages(img_files[roi])
    cur_mask  <- loadImages(mask_files[roi], as.is = TRUE)
    names(cur_image) <- roi
    names(cur_mask)  <- roi
    mcols(cur_image)$sample_id <- roi
    mcols(cur_mask)$sample_id  <- roi

    tryCatch({
      plotCells(
        mask = cur_mask,
        object = spe,
        cell_id = "ObjectNumber",
        img_id = "sample_id",
        colour_by = "CellTypeLabel_display",
        outline_by = NULL,
        exprs_values = "counts",
        colour = list(CellTypeLabel_display = display_cols),
        image_title = NULL,
        legend = list(colour_by.title.cex = 1, colour_by.labels.cex = 0.9,
                       colour_by.legend.cex = 1)
      )
    }, error = function(e) cat("Plot failed for", roi, ":", conditionMessage(e), "\n"))
    cat("\n\n")
  }
}
```

# Notes

- Same file-lookup and 13-category display grouping (12 curated lineages
  + Other, including the Patient10-driven macrophage artifact exclusion
  and "Lung Epithelial" rename) as the full breakdown version -- only the
  per-lineage highlight loop has been removed.
- ~54 ROIs total, one plot each -- much faster to knit than the full
  breakdown version (~700 plots).
- `TEST_MODE <- TRUE` restricts to 2 patients for a quick sanity check;
  switch to `FALSE` for the full cohort once that looks correct.
