---
title: "Tumor IMC - Cytomapper Cell Type Overlays"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    code_folding: show
    toc: true
    toc_float: true
  pdf_document:
    toc: true
editor_options:
  markdown:
    wrap: 72
---

Generates cytomapper overlays for one ROI each from Patient10 (Benign),
Patient12 (Higher stage, mixed LCNEC+Adenocarcinoma histology),adapted from 
the lymph node overlay code (Yusuke's approach) for the tumor project's current annotation.

For each ROI: an "All cells" plot colored by the full cell type label,
plus one highlight plot per individual cell type (`CellType` vs.
`NotCellType`, gold vs. gray). Each is rendered two ways -- true
segmentation-mask boundaries (`cytomapper::plotCells()`, what the
previous student used for the LN figures) and cell centroids as points
(`imcRtools::plotSpatial()`), since mask boundaries can get visually
dense across ~40 cell types.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# 1. Load Libraries

```{r libraries}
#.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(imcRtools)
library(cytomapper)
library(SpatialExperiment)
library(tidyverse)
library(RColorBrewer)
library(scales)
library(knitr)

set.seed(12345)
```

# 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"
steinbock_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/steinbock"

final_celltype_path <- file.path(analysis_path, "final_cell_annotation", "r2_annotation", "spe_final_celltype_r2.rds")

# Each patient's steinbock output lives in its own pXX_results folder, e.g.
# steinbock/p12_results/img, steinbock/p12_results/masks_deepcell
patient_result_dir <- function(patient) {
  patient_num <- gsub("[^0-9]", "", patient)
  file.path(steinbock_path, paste0("p", patient_num, "_results"))
}

overlay_out_dir <- file.path(analysis_path, "visualizations", "cytomapper_overlays")
dir.create(overlay_out_dir, recursive = TRUE, showWarnings = FALSE)
```

# 3. Load Finalized Annotated Object

```{r load-spe}
spe <- readRDS(final_celltype_path)

cat("Cells:", ncol(spe), "\n")
cat("Cell types:", nlevels(factor(spe$final_celltype)), "\n")

spe$CellTypeLabel <- as.character(spe$final_celltype)
```

## 3a. Sanity Check on Annotation

Same check as the sccomp script -- confirm no leftover numbered
prefixes or stray low-quality labels before generating overlays.

```{r sanity-check-annotation}
sort(unique(spe$CellTypeLabel))

cat("\nAny bare-number labels remaining:",
    any(grepl("^[0-9]+$", unique(spe$CellTypeLabel))), "\n")
```

# 4. Select ROIs of Interest

```{r select-rois}
patients_of_interest <- c(
  Patient10 = "Patient_10T_ROI_001_1",
  Patient12 = "Patient_12T_ROI_001_1"
)

data.frame(Patient = names(patients_of_interest),
           sample_id = unname(patients_of_interest))

img_ids <- unname(patients_of_interest)
```

```{r}
celltype_counts_by_roi <- table(spe$sample_id[spe$sample_id %in% img_ids],
                                 spe$CellTypeLabel[spe$sample_id %in% img_ids]) |>
  as.data.frame() |>
  setNames(c("sample_id", "CellType", "Count")) |>
  filter(Count > 0)

celltype_counts_wide <- celltype_counts_by_roi |>
  tidyr::pivot_wider(names_from = sample_id, values_from = Count, values_fill = 0) |>
  arrange(desc(Patient_12T_ROI_001_1))

print(celltype_counts_wide, n = Inf)

```

# 5. Load Images and Masks

Each patient's steinbock output lives in its own pXX_results folder (img/ and masks_deepcell/ subfolders) rather than a shared directory, so the exact TIFF path is built per-ROI using patient_result_dir() from Chunk 2.

```{r load-images}
img_paths <- file.path(patient_result_dir(names(patients_of_interest)), "img",
                        paste0(unname(patients_of_interest), ".tiff"))
mask_paths <- file.path(patient_result_dir(names(patients_of_interest)), "masks_deepcell",
                         paste0(unname(patients_of_interest), ".tiff"))

names(img_paths)  <- img_ids
names(mask_paths) <- img_ids

stopifnot(all(file.exists(img_paths)))
stopifnot(all(file.exists(mask_paths)))

images <- loadImages(img_paths)
masks  <- loadImages(mask_paths, as.is = TRUE)

names(images) <- img_ids
names(masks)  <- img_ids

mcols(images)$sample_id <- names(images)
mcols(masks)$sample_id  <- names(masks)

# Only valid if channel order matches the steinbock panel order -- verify against panel.csv
channelNames(images) <- rownames(spe)
```

# 6. Color Vector for Cell Types

Reuses the existing color scheme in `metadata(spe)$color_vectors` if
one is already stored, filling in any missing cell types with an
auto-generated palette; otherwise generates one from scratch.

```{r color-vector}
allCellTypes <- sort(unique(spe$CellTypeLabel))

if (!is.null(metadata(spe)$color_vectors$CellTypeLabel)) {
  CellTypeLabelCol <- metadata(spe)$color_vectors$CellTypeLabel
  missing_types <- setdiff(allCellTypes, names(CellTypeLabelCol))
  if (length(missing_types) > 0) {
    extra_cols <- hue_pal()(length(missing_types))
    names(extra_cols) <- missing_types
    CellTypeLabelCol <- c(CellTypeLabelCol, extra_cols)
  }
} else {
  CellTypeLabelCol <- setNames(hue_pal()(length(allCellTypes)), allCellTypes)
}

metadata(spe)$color_vectors$CellTypeLabel <- CellTypeLabelCol
```

# 7. Normalize Images

```{r normalize-images}
cur_images <- images[img_ids]
cur_images <- cytomapper::normalize(cur_images, separateImages = TRUE)
cur_images <- cytomapper::normalize(cur_images, inputRange = c(0, 0.5))

cur_masks <- masks[img_ids]
```


```{r selected-celltypes}
selected_types <- c(
  "Epithelial"          = "Epithelial (EpCAM+)",
  "Tumor Epithelial"     = "Tumor Epithelial (TTF1++)",
  "Proliferating Cells"  = "Proliferating Cells",
  "CD8+ T Cell"          = "CD8+ T Cell, Activated",
  "CD4+ T Cell"          = "CD4+ T Cell, Exhaustion-leaning",
  "Treg"                 = "Treg, Resting",
  "Dendritic Cell"       = "cDC1",
  "Macrophage"           = "Macrophage, M2-like (CD163+)",
  "Monocyte"             = "Monocyte-Macrophage Transitional (CD14+)",
  "B Cell/Plasmablast"   = "Transitional B Cell/Plasmablast (CD19+CD138+HLADRB+)",
  "Fibroblast"           = "Fibroblast/Stroma",
  "Mast Cell"            = "Mast Cell"
)
```

# 8. Mask Overlays (Segmentation Boundaries) {.tabset}

True cell-shape overlays via `cytomapper::plotCells()`, matching the
previous student's LN figures.

```{r}
# Build a display-only column: selected lineages get their friendly name,
# everything else collapses to "Other"
spe$CellTypeLabel_display <- "Other"
for (i in seq_along(selected_types)) {
  spe$CellTypeLabel_display[spe$CellTypeLabel == selected_types[i]] <- names(selected_types)[i]
}

display_levels <- c(names(selected_types), "Other")
spe$CellTypeLabel_display <- factor(spe$CellTypeLabel_display, levels = display_levels)

display_cols <- c(
  "Epithelial"          = "red",
  "Tumor Epithelial"     = "darkorange",
  "Proliferating Cells"  = "yellow",
  "CD8+ T Cell"          = "forestgreen",
  "CD4+ T Cell"          = "yellowgreen",
  "Treg"                 = "purple",
  "Dendritic Cell"       = "lightblue",
  "Macrophage"           = "dodgerblue",
  "Monocyte"             = "navy",
  "B Cell/Plasmablast"   = "pink",
  "Fibroblast"           = "violet",
  "Mast Cell"            = "deeppink",
  "Other"                = "gray85"
)
```

## 8a. All Cells, Colored by Full Cell Type Label {.tabset}

```{r mask-all-cells, results="asis", fig.width=16, fig.height=8}
for (roi in img_ids) {
  cat("### ", roi, " {-}\n\n")

  plotCells(
    mask = cur_masks[roi],
    object = spe,
    cell_id = "ObjectNumber",
    img_id = "sample_id",
    colour_by = "CellTypeLabel_display",
    outline_by = NULL,
    exprs_values = "counts",
    colour = list(CellTypeLabel_display = display_cols),
    image_title = NULL,
    legend = list(
      colour_by.title.cex = 1,
      colour_by.labels.cex = 1,
      colour_by.legend.cex = 1.2
    )
  )
  cat("\n\n")
}
```

## 8b. Per Cell Type {.tabset}

Skips any ROI/cell-type combination with zero cells present, since
looping across ~40 cell types will hit several empty combinations per
ROI.

```{r mask-per-celltype, results="asis", fig.width=14, fig.height=8}
for (roi in img_ids) {
  cat("### ", roi, " {.tabset .unnumbered}\n\n")

  for (i in seq_along(selected_types)) {
    lineage_name <- names(selected_types)[i]
    ct <- selected_types[i]

    spe$CellPopulation <- ifelse(spe$CellTypeLabel == ct, "CellType", "NotCellType")

    n_present <- sum(spe$CellPopulation[spe$sample_id == roi] == "CellType")
    if (n_present == 0) next

    cat("#### ", lineage_name, " {-}\n\n")

    plotCells(
      mask = cur_masks[roi],
      object = spe,
      cell_id = "ObjectNumber",
      img_id = "sample_id",
      colour_by = "CellPopulation",
      outline_by = NULL,
      exprs_values = "counts",
      colour = list(CellPopulation = c(`CellType` = "gold", `NotCellType` = "gray20")),
      image_title = NULL
    )
    cat("\n\n")
  }
}
```

# 9. Circle Overlays (Centroid Points) {.tabset}

Cell centroids as points via `imcRtools::plotSpatial()` instead of mask
boundaries -- useful when mask boundaries get too dense to read across
this many cell types, or for a quicker overview.

## All Cells, Colored by Full Cell Type Label

```{r circle-all-cells, results="asis", fig.width=8, fig.height=8}
for (roi in img_ids) {
  cat("### ", roi, " {-}\n\n")

  spe_roi <- spe[, spe$sample_id == roi]
  p <- plotSpatial(
    spe_roi,
    img_id = "sample_id",
    node_color_by = "CellTypeLabel_display",
    node_size_fix = 1,
    colPairName = NULL
  ) + scale_color_manual(values = display_cols) +
    ggtitle(paste0(roi, " -- All cells")) +
    theme(legend.position = "right",
          legend.text = element_text(size = 13),
          legend.title = element_text(size = 15),
          legend.key.size = unit(1, "lines"),
          legend.spacing.y = unit(0.3, "cm")) +
    guides(colour = guide_legend(override.aes = list(size = 4)))

  print(p)
  cat("\n\n")
}
```

# Notes

- `img`/`masks_deepcell` paths are built per-patient via `patient_result_dir()`
  (Section 2) -- update that helper if a patient's steinbock output moves.
- ROI selection (Section 4) is hardcoded to one specific ROI per patient
  (`Patient_10T_ROI_001_1`, `Patient_12T_ROI_001_1`) -- update
  `patients_of_interest` directly for a different ROI.
- Cell types shown are the curated `selected_types` shortlist (Section 8),
  not the full ~40-type annotation -- see the sccomp analysis and per-ROI
  count check for how these were chosen.