---
title: "B Cell/Plasma Cell Subclustering — Run5 Extraction"
author: "Tazein Shah"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(42)
```

## Overview

Extract B cell/plasma cell clusters (3, 10, 15, 16) from the
annotated run5 SPE (`h5_cluster_annot.rds`) and recluster on the
existing Harmony embedding (Louvain, k = 20).

---

## 1. Load Libraries

```{r libraries}
library(imcRtools)
library(scran)
library(bluster)
library(BiocParallel)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(scater)
library(patchwork)
library(RColorBrewer)
```

---

## 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

spe_path <- file.path(analysis_path, "final_cell_annotation", "h5_cluster_annot.rds")

bcell_clustering_apth <- file.path(analysis_path, "final_cell_annotation", "subclustering","bcells")

out_path <- file.path(bcell_clustering_apth, "bcell_subclustering.rds")
out_path_clean <- file.path(bcell_clustering_apth, "bcell_subclustering_clean.rds")
```

---

## 3. Load Annotated SPE

```{r load-data}
spe <- readRDS(spe_path)

print(table(spe$broad_clusters))
```

---

## 4. Extract B Cell/Plasma Clusters

```{r subset}
bcell_clusters <- c("3", "10", "15", "16")

spe_bcell <- spe[, as.character(spe$broad_clusters) %in% bcell_clusters]

cat("Cells in bcell subset:", ncol(spe_bcell), "\n")
cat("From cluster 3 (B Cell CD38+):", sum(spe_bcell$broad_clusters == "3"), "\n")
cat("From cluster 10 (Memory T/Plasma Cell):", sum(spe_bcell$broad_clusters == "10"), "\n")
cat("From cluster 15 (T Cell/Plasma Cell):", sum(spe_bcell$broad_clusters == "15"), "\n")
cat("From cluster 16 (Plasma Cell CD19+):", sum(spe_bcell$broad_clusters == "16"), "\n")
```

---

## 5. Recluster: Louvain on Existing Harmony Embedding (k = 20)

```{r recluster}
if (file.exists(out_path)) {
  cat("Output already exists, loading instead of reclustering:", out_path, "\n")
  spe_bcell <- readRDS(out_path)
} else {
  snn_graph <- buildSNNGraph(spe_bcell, use.dimred = "Harmony", k = 20,
                              BPPARAM = SerialParam())

  clusters <- igraph::cluster_louvain(snn_graph)$membership

  spe_bcell$bcell_subclusters <- factor(clusters)
}

cat("Number of subclusters:", nlevels(spe_bcell$bcell_subclusters), "\n")
print(table(spe_bcell$bcell_subclusters))
```

---

## 6. UMAP on the Subset

```{r umap-sub}
if ("UMAP_sub" %in% reducedDimNames(spe_bcell)) {
  cat("UMAP_sub already exists, skipping recomputation.\n")
} else {
  set.seed(220225)
  spe_bcell <- runUMAP(spe_bcell, dimred = "Harmony", name = "UMAP_sub")
}
```

```{r plot-umap-sub-original, fig.width=10, fig.height=7}
p_sub_original <- dittoDimPlot(spe_bcell, var = "broad_celltype",
                               reduction.use = "UMAP_sub",
                               size = 0.4) +
  ggtitle("B Cell/Plasma Subset: UMAP colored by Original Broad Label") +
  theme(legend.position = "right")
print(p_sub_original)
```

```{r plot-patient-composition, fig.width=10, fig.height=5}
patient_counts <- as.data.frame(table(patient_id = spe_bcell$patient_id))

p_patient_comp <- ggplot(patient_counts, aes(x = patient_id, y = Freq, fill = patient_id)) +
  geom_col() +
  labs(title = "B Cell/Plasma Subset: Cell Count by Patient",
       x = "Patient", y = "Cell Count") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
print(p_patient_comp)

cat("B cell/plasma subset composition by patient:\n")
print(patient_counts[order(-patient_counts$Freq), ])
```

---

## 7. Marker Sets for B Cell/Plasma Subtyping

```{r markers}
# core B cell/plasma lineage markers
lineage_markers <- c("CD19", "CD38", "CD138")

# activation / functional state markers
state_markers <- c("CD45RO", "MKI67", "PD1", "PDL1", "TIM3", "HLADRB")

# admixture / spillover flags (not B-lineage markers, but useful for
# catching doublets or spatial bleed from other lineages)
admixture_markers <- c("FoxP3", "CD1c", "CD15")

bcell_markers <- c(lineage_markers, state_markers, admixture_markers)
bcell_markers <- bcell_markers[bcell_markers %in% rownames(spe_bcell)]

cat("Markers found:", paste(bcell_markers, collapse = ", "), "\n")
```

---

## 8. Violin Plots per Subcluster

```{r make-violinplot}
make_violinplot <- function(spe, assay_name, cluster_col, marker_set, title, ncol = 3) {
  expr_mat   <- assay(spe, assay_name)[marker_set, , drop = FALSE]
  cluster_id <- spe[[cluster_col]]

  mean_expr <- sapply(levels(cluster_id), function(cl) {
    idx <- which(cluster_id == cl)
    rowMeans(expr_mat[, idx, drop = FALSE])
  })

  if (is.matrix(mean_expr) && nrow(mean_expr) > 1) {
    marker_order <- rownames(mean_expr)[hclust(dist(mean_expr))$order]
  } else {
    marker_order <- marker_set
  }

  plot_df <- as.data.frame(t(expr_mat)) |>
    mutate(cluster = cluster_id) |>
    pivot_longer(-cluster, names_to = "marker", values_to = "expr")

  plot_df$marker  <- factor(plot_df$marker, levels = marker_order)
  plot_df$cluster <- factor(plot_df$cluster, levels = levels(cluster_id))

  ggplot(plot_df, aes(x = cluster, y = expr, fill = cluster)) +
    geom_violin(scale = "width", trim = TRUE, linewidth = 0.2) +
    stat_summary(fun = median, geom = "point", size = 0.6, color = "black") +
    facet_wrap(~ marker, scales = "free_y", ncol = ncol) +
    labs(title = title, x = "Subcluster", y = "Z-score expression") +
    theme_bw(base_size = 11) +
    theme(
      axis.text.x      = element_text(angle = 0, hjust = 0.5, size = 9),
      strip.text        = element_text(face = "bold"),
      panel.grid.major.x = element_blank(),
      legend.position   = "none"
    )
}
```

```{r violin-bcell-global-lineage, fig.width=10, fig.height=5}
p_violin_lineage_global <- make_violinplot(spe_bcell, "scaled_0100", "bcell_subclusters",
                                           lineage_markers,
                                           "B Cell/Plasma Subset: Lineage Markers per Subcluster (Global Z-score)",
                                           ncol = 3)
print(p_violin_lineage_global)
```

```{r violin-bcell-global-state, fig.width=12, fig.height=8}
p_violin_state_global <- make_violinplot(spe_bcell, "scaled_0100", "bcell_subclusters",
                                         state_markers,
                                         "B Cell/Plasma Subset: State Markers per Subcluster (Global Z-score)",
                                         ncol = 3)
print(p_violin_state_global)
```

```{r violin-bcell-global-admixture, fig.width=10, fig.height=5}
p_violin_admixture_global <- make_violinplot(spe_bcell, "scaled_0100", "bcell_subclusters",
                                             admixture_markers,
                                             "B Cell/Plasma Subset: Admixture Markers per Subcluster (Global Z-score)",
                                             ncol = 3)
print(p_violin_admixture_global)
```

Subset-relative renormalization: since `scaled_0100` was capped and Z-scored
over the full run5 dataset, this recomputes the same cap+Z-score recipe
from raw counts using only this B cell/plasma subset's own distribution —
one normalization layer, just a different denominator.

```{r renorm-subset}
cap_99 <- function(mat) {
  t(apply(mat, 1, function(x) {
    q99 <- quantile(x, 0.99)
    pmin(x, q99)
  }))
}

zscore_rows <- function(mat) {
  t(scale(t(mat)))
}

mat_sub_0100 <- counts(spe_bcell) |> cap_99() |> zscore_rows()
assay(spe_bcell, "scaled_0100_sub") <- mat_sub_0100
```

```{r violin-bcell-subset-lineage, fig.width=10, fig.height=5}
p_violin_lineage_sub <- make_violinplot(spe_bcell, "scaled_0100_sub", "bcell_subclusters",
                                        lineage_markers,
                                        "B Cell/Plasma Subset: Lineage Markers per Subcluster (Subset-Renormalized Z-score)",
                                        ncol = 3)
print(p_violin_lineage_sub)
```

```{r violin-bcell-subset-state, fig.width=12, fig.height=8}
p_violin_state_sub <- make_violinplot(spe_bcell, "scaled_0100_sub", "bcell_subclusters",
                                      state_markers,
                                      "B Cell/Plasma Subset: State Markers per Subcluster (Subset-Renormalized Z-score)",
                                      ncol = 3)
print(p_violin_state_sub)
```

```{r violin-bcell-subset-admixture, fig.width=10, fig.height=5}
p_violin_admixture_sub <- make_violinplot(spe_bcell, "scaled_0100_sub", "bcell_subclusters",
                                          admixture_markers,
                                          "B Cell/Plasma Subset: Admixture Markers per Subcluster (Subset-Renormalized Z-score)",
                                          ncol = 3)
print(p_violin_admixture_sub)
```

---

## 9. Manual Subcluster Annotation

```{r annotate-sub, echo=TRUE}
bcell_subannotation <- c(
  "1" = "Plasmablast/Activated B Cell (CD38+)",  # CD38 sole top marker, sharp elevation
  "2" = "Unknown/Low Quality",  # No markers above logFC 1.0
  "3" = "Plasma Cell, Exhausted/Effector (CD138+PD1+TIM3+)",  # CD45RO, CD138, TIM3, PD1 co-elevated; FoxP3 also present but secondary to plasma cell signature, likely Treg-adjacent spatial admixture rather than driving identity
  "4" = "Transitional B Cell/Plasmablast (CD19+CD138+)",  # CD19 and CD138 co-elevated despite being largely mutually exclusive lineage markers -- flagging as transitional/mixed rather than assuming a clean state
  "5" = "Unknown/Low Quality",  # No markers above logFC 1.0
  "6" = "Plasma Cell, Memory-like (CD138+CD45RO+)",  # CD45RO, TIM3, CD138 elevated, no exhaustion markers beyond TIM3
  "7" = "Treg Admixture (FoxP3+)",  # FoxP3 dominant top marker by a wide margin; this is a Treg marker, not B-lineage
  "8" = "Unknown/Low Quality",  # No markers above logFC 1.0
  "9" = "B Cell (CD19+)"  # CD19 sole top marker, cleanest B cell signature
)

if (length(bcell_subannotation) > 0) {
  spe_bcell$sub_celltype <- bcell_subannotation[as.character(spe_bcell$bcell_subclusters)]
  cat("Subcluster annotation applied.\n")
  print(table(spe_bcell$sub_celltype, useNA = "ifany"))
} else {
  cat("No subcluster annotation yet. Fill in bcell_subannotation above.\n")
}
```

---

## 10. Treg Admixture (Subcluster 7): Origin

Subcluster 7 was called "Treg Admixture" above based on its dominant
FoxP3 signal in findMarkers. Before removing it, check where these
cells came from in the original broad clustering -- this tells us
whether the admixture traces back to one specific broad cluster
(likely spillover/doublets at the boundary of that cluster) or is
spread across several.

```{r treg-origin, fig.width=10, fig.height=5, results='asis'}
treg_counts <- table(spe_bcell$broad_clusters[spe_bcell$bcell_subclusters == "7"])
broad_totals <- table(spe_bcell$broad_clusters)

treg_origin <- data.frame(
  original_broad     = names(treg_counts),
  treg_cell_count     = as.integer(treg_counts),
  pct_of_treg_cluster = round(100 * as.integer(treg_counts) / sum(treg_counts), 1),
  pct_of_broad_cluster = round(100 * as.integer(treg_counts) /
                                as.integer(broad_totals[names(treg_counts)]), 1)
)
treg_origin <- treg_origin[treg_origin$treg_cell_count > 0, ]
treg_origin <- treg_origin[order(-treg_origin$pct_of_broad_cluster), ]

cat("Subcluster 7 (Treg admixture) origin:\n")
print(knitr::kable(treg_origin,
                   col.names = c("Original Broad Cluster", "Cell Count",
                                 "% of Subcluster 7", "% of That Broad Cluster"),
                   caption = "% of Subcluster 7 = share of Treg admixture cells coming from this broad cluster. % of That Broad Cluster = what fraction of the broad cluster's own cells ended up here."))
```

99.7% of subcluster 7 traces back to original broad cluster 15
(T Cell/Plasma Cell CD4+/CD138+), pulling out 8.4% of that cluster's
cells -- confirms this is spillover from cluster 15's known mixed
identity, not a genuine B cell subtype.

```{r treg-origin-patient, fig.width=10, fig.height=5}
treg_mask <- as.logical(spe_bcell$bcell_subclusters == "7")
treg_patient_ids <- as.character(as.vector(spe_bcell$patient_id))[treg_mask]

cat("Number of cells matching subcluster 7:", sum(treg_mask, na.rm = TRUE), "\n")
cat("Length of treg_patient_ids:", length(treg_patient_ids), "\n")

if (length(treg_patient_ids) == 0) {
  cat("No cells found in subcluster 7 for this run -- check table(spe_bcell$bcell_subclusters) in this same session; cluster numbering may have shifted, or spe_bcell was rebuilt differently than expected.\n")
} else {
  freq_table <- table(treg_patient_ids)
  treg_patient <- data.frame(
    patient = names(freq_table),
    Freq    = as.integer(freq_table),
    stringsAsFactors = FALSE
  )
  treg_patient <- treg_patient[treg_patient$Freq > 0, , drop = FALSE]

  p_treg_patient <- ggplot(treg_patient, aes(x = patient, y = Freq, fill = patient)) +
    geom_col() +
    labs(title = "Subcluster 7 (Treg Admixture): Patient Composition",
         x = "Patient", y = "Cell Count") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none")
  print(p_treg_patient)
}
```

---

## 11. Heatmap: Unknown/Low Quality Cluster Expression

Violin plots show one marker per panel, which makes it easy to miss a
modest but real signal spread across several markers at once. This
heatmap shows mean Z-score expression across the full marker panel for
all 9 subclusters side by side (clusters 2, 5, and 8 -- the low-signal
ones -- outlined in black), so any consistent low-grade pattern is
easier to spot in context than by looking at those three in isolation.

Findings from this check: cluster 2 showed real co-elevation across
CD45RO/CD138/CD38/FoxP3, cluster 5 was flat across every marker, and
cluster 8 showed a modest CD19 signal.

```{r unknown-heatmap, fig.width=10, fig.height=6}
all_clusters <- levels(spe_bcell$bcell_subclusters)

mean_expr_all <- sapply(all_clusters, function(cl) {
  idx <- which(spe_bcell$bcell_subclusters == cl)
  rowMeans(assay(spe_bcell, "scaled_0100")[bcell_markers, idx, drop = FALSE])
})

marker_order <- bcell_markers[hclust(dist(mean_expr_all))$order]

cluster_sizes <- table(spe_bcell$bcell_subclusters)
cluster_labels <- paste0(all_clusters, "\n(n=", cluster_sizes[all_clusters], ")")
names(cluster_labels) <- all_clusters

heatmap_df <- as.data.frame(mean_expr_all) |>
  rownames_to_column("marker") |>
  pivot_longer(-marker, names_to = "subcluster", values_to = "mean_expr")

heatmap_df$marker     <- factor(heatmap_df$marker, levels = marker_order)
heatmap_df$subcluster <- factor(heatmap_df$subcluster, levels = all_clusters)
heatmap_df$is_unknown <- heatmap_df$subcluster %in% c("2", "5", "8")

p_unknown_heatmap <- ggplot(heatmap_df, aes(x = subcluster, y = marker, fill = mean_expr)) +
  geom_tile(color = "white") +
  geom_tile(data = subset(heatmap_df, is_unknown), fill = NA, color = "black", linewidth = 0.8) +
  geom_text(aes(label = round(mean_expr, 2)), size = 3) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, name = "Mean\nZ-score") +
  scale_x_discrete(labels = cluster_labels) +
  labs(title = "Mean Marker Expression: All Subclusters (Unknown/Low Quality Outlined)",
       x = "Subcluster (cell count)", y = "Marker") +
  theme_bw(base_size = 11) +
  theme(panel.grid = element_blank())
print(p_unknown_heatmap)
```

---

## 12. Updated Manual Subcluster Annotation (Post-Heatmap Review)

The all-cluster heatmap above showed that clusters 2 and 8 aren't
actually flat -- they're weaker-intensity echoes of established
phenotypes (cluster 6's memory-like plasma cell profile, and cluster
9's B cell profile, respectively), just at lower magnitude. Cluster 5
is the only one of the three that holds up as genuinely low quality:
negative across every marker, several of the most negative values on
the whole heatmap. This overwrites `bcell_subannotation` and
`spe_bcell$sub_celltype` from chunk 9 with corrected calls for
clusters 2 and 8.

```{r annotate-sub-updated, echo=TRUE}
bcell_subannotation <- c(
  "1" = "Plasmablast/Activated B Cell (CD38+)",  # CD38 sole standout, sharp elevation
  "2" = "Plasma Cell, Memory-like (Weak) (CD138+CD45RO+)",  # CD45RO (0.85) and CD138 (0.59) both positive, same direction as cluster 6's memory-like profile (CD45RO 1.25, CD138 0.92, TIM3 0.75) -- a dimmer echo of that phenotype, not a distinct or flat population
  "3" = "Plasma Cell, Exhausted/Effector (CD138+PD1+TIM3+)",  # CD45RO, CD138, TIM3, PD1 co-elevated; FoxP3 also present but secondary to plasma cell signature, likely Treg-adjacent spatial admixture rather than driving identity
  "4" = "Transitional B Cell/Plasmablast (CD19+CD138+)",  # CD19 and CD138 co-elevated despite being largely mutually exclusive lineage markers -- flagging as transitional/mixed rather than assuming a clean state
  "5" = "Unknown/Low Quality",  # Negative across every single marker, several of the most negative values in the whole heatmap (CD1c -0.72, PDL1 -0.63, FoxP3 -0.58, HLADRB -0.51) -- genuinely flat, not just weak
  "6" = "Plasma Cell, Memory-like (CD138+CD45RO+)",  # CD45RO, TIM3, CD138 elevated, no exhaustion markers beyond TIM3
  "7" = "Treg Admixture (FoxP3+)",  # FoxP3 dominant, wide margin; this is a Treg marker, not B-lineage
  "8" = "B Cell, Weak (CD19+)",  # CD19 (0.55) positive in the same direction as cluster 9's CD19 signature (2.69), just much weaker -- a dimmer echo of that phenotype, not ambiguous/mixed
  "9" = "B Cell (CD19+)"  # CD19 sole standout, cleanest B cell signature
)

spe_bcell$sub_celltype <- bcell_subannotation[as.character(spe_bcell$bcell_subclusters)]
cat("Updated subcluster annotation applied.\n")
print(table(spe_bcell$sub_celltype, useNA = "ifany"))
```

---

## 13. Remove Treg Cells

```{r remove-treg}
spe_bcell_clean <- spe_bcell[, spe_bcell$bcell_subclusters != "7"]

cat("Cells before removal:", ncol(spe_bcell), "\n")
cat("Cells after removing Treg admixture (cluster 7):", ncol(spe_bcell_clean), "\n")
```

---

## 14. Remove Unknown/Low Quality Cluster (5)

Cluster 5 was negative across every marker on the all-cluster heatmap
(chunk 11) -- genuinely flat, unlike clusters 2 and 8 which turned out
to be weaker echoes of established phenotypes. Removing it from
`spe_bcell_clean` alongside the Treg admixture before reclustering.

```{r remove-unknown}
cells_before_unknown_removal <- ncol(spe_bcell_clean)

spe_bcell_clean <- spe_bcell_clean[, spe_bcell_clean$bcell_subclusters != "5"]

cat("Cells before removing Unknown/Low Quality (cluster 5):", cells_before_unknown_removal, "\n")
cat("Cells after removing Unknown/Low Quality (cluster 5):", ncol(spe_bcell_clean), "\n")
```

---

## 15. Recluster After Removal

Recluster on the existing Harmony embedding (same k = 20 as the
original subclustering) to check whether the remaining subclusters
hold up once the Treg admixture and Unknown/Low Quality cells are gone.

```{r recluster-after-removal}
if ("bcell_subclusters_clean" %in% colnames(colData(spe_bcell_clean))) {
  cat("bcell_subclusters_clean already exists, skipping recomputation.\n")
} else {
  set.seed(220225)
  snn_graph_clean <- buildSNNGraph(spe_bcell_clean, use.dimred = "Harmony", k = 20,
                                    BPPARAM = SerialParam())
  clusters_clean <- igraph::cluster_louvain(snn_graph_clean)$membership
  spe_bcell_clean$bcell_subclusters_clean <- factor(clusters_clean)
}

cat("Number of subclusters after removal:", nlevels(spe_bcell_clean$bcell_subclusters_clean), "\n")
print(table(spe_bcell_clean$bcell_subclusters_clean))
```

---

## 16. Violin Plots on Post-Removal Subclusters

```{r violin-clean, fig.width=14, fig.height=12}
p_violin_clean <- make_violinplot(spe_bcell_clean, "scaled_0100", "bcell_subclusters_clean",
                                  bcell_markers,
                                  "B Cell/Plasma Subset (Treg/Unknown-Removed): Marker Expression per Subcluster",
                                  ncol = 3)
print(p_violin_clean)
```

---

## 17. Heatmap: All Post-Removal Subclusters

Same approach as the pre-removal heatmap (chunk 11) -- mean Z-score
expression across the full marker panel for all 8 post-removal
subclusters side by side, to check clusters 2, 5, and 7 against the
full comparison context rather than eyeballing violins alone.

```{r heatmap-clean, fig.width=8, fig.height=5}
clean_clusters <- levels(spe_bcell_clean$bcell_subclusters_clean)

mean_expr_clean <- sapply(clean_clusters, function(cl) {
  idx <- which(spe_bcell_clean$bcell_subclusters_clean == cl)
  rowMeans(assay(spe_bcell_clean, "scaled_0100")[bcell_markers, idx, drop = FALSE])
})

marker_order_clean <- bcell_markers[hclust(dist(mean_expr_clean))$order]

cluster_sizes_clean <- table(spe_bcell_clean$bcell_subclusters_clean)
cluster_labels_clean <- paste0(clean_clusters, "\n(n=", cluster_sizes_clean[clean_clusters], ")")
names(cluster_labels_clean) <- clean_clusters

heatmap_clean_df <- as.data.frame(mean_expr_clean) |>
  rownames_to_column("marker") |>
  pivot_longer(-marker, names_to = "subcluster", values_to = "mean_expr")

heatmap_clean_df$marker     <- factor(heatmap_clean_df$marker, levels = marker_order_clean)
heatmap_clean_df$subcluster <- factor(heatmap_clean_df$subcluster, levels = clean_clusters)

p_heatmap_clean <- ggplot(heatmap_clean_df, aes(x = subcluster, y = marker, fill = mean_expr)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(mean_expr, 2)), size = 3) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, name = "Mean\nZ-score") +
  scale_x_discrete(labels = cluster_labels_clean) +
  labs(title = "Mean Marker Expression: All Post-Removal Subclusters",
       x = "Subcluster (cell count)", y = "Marker") +
  theme_bw(base_size = 11) +
  theme(panel.grid = element_blank())
print(p_heatmap_clean)
```

---

## 18. Manual Annotation of Post-Removal Subclusters

```{r annotate-sub-clean, echo=TRUE}
bcell_subannotation_clean <- c(
  "1" = "Plasmablast/Activated B Cell (CD38+)",
  "2" = "Plasma Cell, Memory-like (Weak) (CD138+CD45RO+)",  # Same direction as cluster 6, just weaker -- dimmer echo, not flat
  "3" = "Plasmablast, Strongly Activated (CD38++)",  # CD38 3.32, the single highest value on the heatmap -- distinct from cluster 1's more moderate CD38
  "4" = "Plasma Cell, Exhausted/Effector (CD138+CD45RO+TIM3+PD1+)",
  "5" = "Transitional B Cell/Plasmablast (CD19+CD138+)",
  "6" = "Plasma Cell, Memory-like (CD138+CD45RO+)",
  "7" = "B Cell, Weak (CD19+)",  # Same direction as cluster 8, just weaker -- dimmer echo, not flat
  "8" = "B Cell (CD19+)"
)

if (length(bcell_subannotation_clean) > 0) {
  spe_bcell_clean$sub_celltype_clean <- bcell_subannotation_clean[as.character(spe_bcell_clean$bcell_subclusters_clean)]
  cat("Post-removal subcluster annotation applied.\n")
  print(table(spe_bcell_clean$sub_celltype_clean, useNA = "ifany"))
} else {
  cat("No post-removal annotation yet. Fill in bcell_subannotation_clean above.\n")
}
```

---

## 19. Compare Original Annotation to Post-Removal Annotation

Cross-tab the original manual annotation (chunk 12, the updated
post-heatmap version, applied before Treg/cluster 5 removal) against
the new post-removal annotation (chunk 18). If each original label
maps cleanly onto one (or a small merge of) new labels, the original
annotation still holds. If a label splits across several new labels,
removing subclusters 7 and 5 changed the underlying structure enough
that the original call should be revisited.

```{r check-annotation-consistency, fig.width=10, fig.height=8}
if (!exists("spe_bcell_clean")) {
  cat("spe_bcell_clean doesn't exist -- run chunk 13 (Remove Treg Cells) onward first.\n")
} else if (!("bcell_subclusters_clean" %in% colnames(colData(spe_bcell_clean)))) {
  cat("bcell_subclusters_clean is missing -- run chunk 15 (Recluster After Removal).\n")
} else if (!("sub_celltype_clean" %in% colnames(colData(spe_bcell_clean)))) {
  cat("sub_celltype_clean is missing -- run chunk 18 (Manual Annotation of Post-Removal Subclusters).\n")
} else if (!("sub_celltype" %in% colnames(colData(spe_bcell_clean)))) {
  cat("sub_celltype is missing -- run chunk 12 (Updated Manual Subcluster Annotation) before removal.\n")
} else {
  comp_annotation <- table(original = spe_bcell_clean$sub_celltype,
                          post_removal = spe_bcell_clean$sub_celltype_clean)
  comp_annotation_props <- prop.table(comp_annotation, margin = 1)

  comp_df <- as.data.frame(comp_annotation_props)

  p_annotation_compare <- ggplot(comp_df, aes(x = post_removal, y = original, fill = Freq)) +
    geom_tile(color = "white") +
    geom_text(aes(label = paste0(round(Freq * 100, 0), "%")), size = 3.5) +
    scale_fill_gradient(low = "white", high = "firebrick", name = "Row %") +
    scale_x_discrete(labels = function(x) str_wrap(x, width = 18)) +
    scale_y_discrete(labels = function(x) str_wrap(x, width = 25)) +
    labs(title = "Original Annotation vs. Post-Removal Annotation",
         subtitle = "Row-normalized: each row (original label) sums to 100%",
         x = "Post-Removal Annotation", y = "Original Annotation") +
    theme_bw(base_size = 11) +
    theme(axis.text.x  = element_text(angle = 45, hjust = 1, color = "darkorange3"),
          axis.text.y  = element_text(color = "steelblue4"),
          axis.title.x = element_text(color = "darkorange3", face = "bold"),
          axis.title.y = element_text(color = "steelblue4", face = "bold"),
          panel.grid = element_blank())
  print(p_annotation_compare)
}
```

---

## 20. UMAP Colored by Subcluster Cell Type

```{r umap-sub-celltype, fig.width=10, fig.height=5}
if ("sub_celltype_clean" %in% colnames(colData(spe_bcell_clean))) {

  spe_bcell_clean$sub_celltype_clean <- factor(spe_bcell_clean$sub_celltype_clean, levels = bcell_subannotation_clean)

  # colors grouped by biological category rather than arbitrary rainbow:
  # blues = B cell, greens = plasmablast, oranges/red = plasma cell, purple = transitional
  subtype_cols <- c(
    "B Cell, Weak (CD19+)"                                       = "#A6CEE3",  # light blue
    "B Cell (CD19+)"                                             = "#1F78B4",  # dark blue
    "Plasmablast/Activated B Cell (CD38+)"                       = "#B2DF8A",  # light green
    "Plasmablast, Strongly Activated (CD38++)"                   = "#33A02C",  # dark green
    "Plasma Cell, Memory-like (Weak) (CD138+CD45RO+)"            = "#FDBF6F",  # light orange
    "Plasma Cell, Memory-like (CD138+CD45RO+)"                   = "#FF7F00",  # dark orange
    "Plasma Cell, Exhausted/Effector (CD138+CD45RO+TIM3+PD1+)"   = "#E31A1C",  # red
    "Transitional B Cell/Plasmablast (CD19+CD138+)"              = "#6A3D9A"   # purple
  )

  p_sub_celltype <- dittoDimPlot(spe_bcell_clean, var = "sub_celltype_clean",
                                 reduction.use = "UMAP_sub",
                                 size = 0.4) +
    scale_color_manual(values = subtype_cols, breaks = names(subtype_cols)) +
    ggtitle("B Cell/Plasma Subset (Treg-Removed): UMAP colored by Subcluster Cell Type") +
    theme(legend.position = "right")
  print(p_sub_celltype)

} else {
  cat("Run the annotate-sub-clean chunk first to apply post-removal annotations.\n")
}
```

---

## 21. Cell Type Composition by Patient

```{r celltype-patient-composition, fig.width=14, fig.height=16}
celltype_patient_counts <- as.data.frame(
  table(patient_id = spe_bcell_clean$patient_id,
        cell_type  = spe_bcell_clean$sub_celltype_clean)
)

p_celltype_patient <- ggplot(celltype_patient_counts, aes(x = patient_id, y = Freq, fill = patient_id)) +
  geom_col() +
  facet_wrap(~ cell_type, scales = "free", ncol = 2,
             labeller = labeller(cell_type = label_wrap_gen(width = 35))) +
  labs(title = "B Cell/Plasma Subset (Treg-Removed): Cell Type Composition by Patient",
       x = "Patient", y = "Cell Count") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        strip.text = element_text(face = "bold", size = 8),
        legend.position = "none")

print(p_celltype_patient)

# RStudio's Plots pane doesn't respect fig.width/fig.height when running
# chunks interactively (those only apply to a knitted document) -- save to
# file with fixed dimensions so the full plot is viewable without cutoff
celltype_patient_plot_path <- file.path(bcell_clustering_apth, "celltype_composition_by_patient.png")
ggsave(celltype_patient_plot_path, p_celltype_patient, width = 14, height = 16, dpi = 300)
cat("Saved full-size plot to:", celltype_patient_plot_path, "\n")

#cat("Cell type composition by patient:\n")
#celltype_patient_nonzero <- celltype_patient_counts[celltype_patient_counts$Freq > 0, ]
#print(celltype_patient_nonzero[order(celltype_patient_nonzero$cell_type, -celltype_patient_nonzero$Freq), ])
```

---

## 22. Save Subclustered/Annotated Object

```{r save}
dir.create(dirname(out_path), recursive = TRUE, showWarnings = FALSE)
saveRDS(spe_bcell, out_path)
cat("Saved B cell subcluster object (pre-removal, annotated) to:", out_path, "\n")

dir.create(dirname(out_path_clean), recursive = TRUE, showWarnings = FALSE)
saveRDS(spe_bcell_clean, out_path_clean)
cat("Saved B cell subcluster object (post-removal, annotated) to:", out_path_clean, "\n")
```
