---
title: "Subclustering: Macrophage/Monocyte (Clusters 8 & 14)"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 12, fig.height = 8)
set.seed(42)
```

## Overview

Subclustering the Macrophage (cluster 8) and Monocyte (cluster 14)
populations from the broad k=20 annotation. 
These two looked mixed together at the broad resolution and may
resolve into cleaner myeloid subtypes at finer resolution.

Myeloid: immune cell lineage that includes monocytes, macrophages, neutrophils/granulocytes, dendritic cells, and mast cells. 

---

## 1. Load Libraries

```{r libraries}
# Only run this if you start a new RStudio session
.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(imcRtools)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(scater)
library(scran)
library(bluster)
library(patchwork)
library(RColorBrewer)
library(ComplexHeatmap)
library(circlize)
```

---

## 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

spe_path <- file.path(analysis_path, "final_cell_annotation", "h5_cluster_annot.rds")
out_path <- file.path(analysis_path, "final_cell_annotation", "subclustering",
                      "myeloid_subcluster.rds")
findmarkers_path <- file.path(analysis_path, "final_cell_annotation", "subclustering",
                              "myeloid_subcluster_findmarkers.csv")
```

---

## 3. Load Annotated Data and Subset

```{r load-and-subset}
spe <- readRDS(spe_path)

myeloid_clusters <- c("8", "14")
spe_myeloid <- spe[, spe$broad_clusters %in% myeloid_clusters]

cat("Cells in myeloid subset:", ncol(spe_myeloid), "\n")
cat("From cluster 8 (Macrophage):", sum(spe_myeloid$broad_clusters == "8"), "\n")
cat("From cluster 14 (Monocyte):", sum(spe_myeloid$broad_clusters == "14"), "\n")
```

---

## 4. Re-cluster the Subset on the Existing Harmony Embedding

Using a smaller k (10 instead of 20) to encourage finer splits, since
the goal here is specifically to reveal subtypes that got merged at
the broader resolution.

```{r subcluster}
# this just pulls out the already corrected coordinates from the run5 spe object
# builds a fresh neighbor graph
harmony_sub <- reducedDim(spe_myeloid, "Harmony")
cat("Harmony embedding dimensions (subset):", dim(harmony_sub), "\n")

set.seed(220225)
g_sub     <- buildSNNGraph(t(harmony_sub), k = 10, type = "rank")
clust_sub <- igraph::cluster_louvain(g_sub)
spe_myeloid$sub_clusters <- factor(clust_sub$membership)

cat("Number of myeloid subclusters:", nlevels(spe_myeloid$sub_clusters), "\n")
print(table(spe_myeloid$sub_clusters))
```

---

## 5. UMAP on the Subset

```{r umap-sub}
set.seed(220225)
spe_myeloid <- runUMAP(spe_myeloid, dimred = "Harmony", name = "UMAP_sub")
```

```{r plot-umap-sub-clusters, fig.width=10, fig.height=7}
p_sub_clusters <- dittoDimPlot(spe_myeloid, var = "sub_clusters",
                               reduction.use = "UMAP_sub",
                               size = 0.4, do.label = TRUE, labels.size = 4) +
  ggtitle("Myeloid Subset: UMAP colored by Subcluster") +
  theme(legend.position = "right")
print(p_sub_clusters)
```

```{r plot-umap-sub-original, fig.width=10, fig.height=7}
p_sub_original <- dittoDimPlot(spe_myeloid, var = "broad_celltype",
                               reduction.use = "UMAP_sub",
                               size = 0.4) +
  ggtitle("Myeloid Subset: UMAP colored by Original Broad Label") +
  theme(legend.position = "right")
print(p_sub_original)
```
When comparing this with the subcluster UMAP, seems like subcluster 7 has mixed composition (both macrophage and myeloid). It also has CD14 as its only top positive marker.

---

## 6. Marker Sets for Myeloid Subtyping

```{r markers}
myeloid_markers <- c(
  "CD14", "CD11c", "CD68", "CD86", "CD163", "CD1c", "MKI67",
  "PDL1", "PD1", "TIM3", "LAMP3", "CD15", "HLADRB", "CD38"
)
myeloid_markers <- myeloid_markers[myeloid_markers %in% rownames(spe_myeloid)]
cat("Markers found:", paste(myeloid_markers, collapse = ", "), "\n")
```

---

## 7. Violin Plots per Subcluster

```{r make-violinplot}
make_violinplot <- function(spe, assay_name, cluster_col, marker_set, title, ncol = 3) {
  expr_mat   <- assay(spe, assay_name)[marker_set, , drop = FALSE]
  cluster_id <- spe[[cluster_col]]

  mean_expr <- sapply(levels(cluster_id), function(cl) {
    idx <- which(cluster_id == cl)
    rowMeans(expr_mat[, idx, drop = FALSE])
  })

  if (is.matrix(mean_expr) && nrow(mean_expr) > 1) {
    marker_order <- rownames(mean_expr)[hclust(dist(mean_expr))$order]
  } else {
    marker_order <- marker_set
  }

  plot_df <- as.data.frame(t(expr_mat)) |>
    mutate(cluster = cluster_id) |>
    pivot_longer(-cluster, names_to = "marker", values_to = "expr")

  plot_df$marker  <- factor(plot_df$marker, levels = marker_order)
  plot_df$cluster <- factor(plot_df$cluster, levels = levels(cluster_id))

  ggplot(plot_df, aes(x = cluster, y = expr, fill = cluster)) +
    geom_violin(scale = "width", trim = TRUE, linewidth = 0.2) +
    stat_summary(fun = median, geom = "point", size = 0.6, color = "black") +
    facet_wrap(~ marker, scales = "free_y", ncol = ncol) +
    labs(title = title, x = "Subcluster", y = "Z-score expression") +
    theme_bw(base_size = 11) +
    theme(
      axis.text.x      = element_text(angle = 0, hjust = 0.5, size = 9),
      strip.text        = element_text(face = "bold"),
      panel.grid.major.x = element_blank(),
      legend.position   = "none"
    )
}
```

```{r violin-myeloid, fig.width=16, fig.height=12}
p_violin_myeloid <- make_violinplot(spe_myeloid, "scaled_0100", "sub_clusters",
                                    myeloid_markers,
                                    "Myeloid Subset: Marker Expression per Subcluster",
                                    ncol = 3)
print(p_violin_myeloid)
```

---

## 8. Statistical Marker Discovery (findMarkers) on Subclusters

```{r findmarkers}
markers_de <- findMarkers(spe_myeloid,
                          groups     = spe_myeloid$sub_clusters,
                          assay.type = "scaled_0100",
                          subset.row = myeloid_markers)

all_markers <- lapply(levels(spe_myeloid$sub_clusters), function(cl) {
  df <- as.data.frame(markers_de[[cl]])
  data.frame(
    subcluster = cl,
    marker     = rownames(df),
    Top        = df$Top,
    FDR        = df$FDR,
    logFC      = df$summary.logFC
  )
})

all_markers_df <- do.call(rbind, all_markers)
dir.create(dirname(findmarkers_path), recursive = TRUE, showWarnings = FALSE)
write.csv(all_markers_df, findmarkers_path, row.names = FALSE)

top1_summary <- do.call(rbind, lapply(levels(spe_myeloid$sub_clusters), function(cl) {
  sub <- all_markers_df[all_markers_df$subcluster == cl & all_markers_df$logFC > 1.0, ]
  sub <- sub[order(sub$Top, -sub$logFC), ]
  if (nrow(sub) == 0) {
    data.frame(subcluster = cl, top_markers = "none above logFC 1.0")
  } else {
    data.frame(subcluster = cl,
               top_markers = paste(head(sub$marker, 5), collapse = ", "))
  }
}))

cat("Top positive markers (logFC > 1.0) per subcluster:\n")
print(knitr::kable(top1_summary, col.names = c("Subcluster", "Top Positive Markers")))
```

Subclusters 4, 10, and 11 (none above logFC 1.0): not patient-driven, their ambiguity is most likely a marker-resolution problem (weakly differentiated cells sitting between clean states).

Subcluster 4 is the largest subcluster and maps almost entirely to Monocyte. Could be a less-activated monocyte population.
---

## 9. Subcluster Composition By Original Broad Cluster

```{r composition-broad-heatmap, fig.width=12, fig.height=6}

comp_broad <- table(subcluster = spe_myeloid$sub_clusters,
                    original_broad = spe_myeloid$broad_clusters)
comp_broad_props <- prop.table(comp_broad, margin = 1)

col_fun <- colorRamp2(c(0, 0.5, 1), rev(brewer.pal(3, "RdBu")))

ht <- Heatmap(
  comp_broad_props,
  name              = "Proportion",
  col               = col_fun,
  cluster_rows      = FALSE,
  cluster_columns   = FALSE,
  column_title      = "Broad Cluster",
  column_title_side = "bottom",
  row_title         = "Subcluster",
  column_names_rot  = 45,
  cell_fun = function(j, i, x, y, width, height, fill) {
    val <- comp_broad_props[i, j]
    text_col <- if (val < 0.2 || val > 0.8) "white" else "black"
    grid.text(sprintf("%.2f", val), x, y, gp = gpar(col = text_col, fontsize = 7))
  }
)

draw(ht, column_title = "Subcluster composition by original broad cluster (row-normalized %)")
```
Subcluster 7 has 58% macrophage and 42% monocyte. Could be a transitional population.

Subcluster 11 has 87% monocyte and 13% macrophage.

The rest of the subclusters either are entirely in Cluster 8 (Macrophage) or Cluster 14 (Monocyte), which means that the split we see with the broad clusters still holds true.

---

## 10: Subcluster Composition by Patient 

```{r composition-patient-heatmap, fig.width=12, fig.height=6}
comp_patient <- table(subcluster = spe_myeloid$sub_clusters,
                      patient = spe_myeloid$patient_id)
comp_patient_props <- prop.table(comp_patient, margin = 1)

col_fun_patient <- colorRamp2(c(0, 0.5, 1), rev(brewer.pal(3, "RdBu")))

ht_patient <- Heatmap(
  comp_patient_props,
  name              = "Proportion",
  col               = col_fun_patient,
  cluster_rows      = FALSE,
  cluster_columns   = FALSE,
  column_title      = "Patient",
  column_title_side = "bottom",
  row_title         = "Subcluster",
  column_names_rot  = 45,
  cell_fun = function(j, i, x, y, width, height, fill) {
    val <- comp_patient_props[i, j]
    text_col <- if (val < 0.2 || val > 0.8) "white" else "black"
    grid.text(sprintf("%.2f", val), x, y, gp = gpar(col = text_col, fontsize = 7))
  }
)

draw(ht_patient, column_title = "Subcluster composition by patient (row-normalized %)")
```
Subclusters 2 (45% Patient 10) and 3 (48% Patient 10) have roughly 4x their expected share.
---

## 11. Manual Subcluster Annotation

```{r annotate-sub, echo=TRUE}
myeloid_subannotation <- c(
  "1"  = "1. Macrophage, Antigen-Presenting (HLADRB+)", # Multi-marker
  "2"  = "2. Macrophage, Proliferating (MKI67+)",  # Real signature, but 45% Patient 10
  "3"  = "3. Monocyte (PD1+)",
  "4"  = "4. Monocyte (Generic)",  # Largest subcluster, flat patient composition, likely an undifferentiated monocyte population
  "5"  = "5. Macrophage, M2-like (CD163+)", # CD163 is an M2/anti-inflammatory macrophage marker
  "6"  = "6. Macrophage (CD68+)",
  "7"  = "7. Monocyte-Macrophage Transitional (CD14+)",  # This subcluster has the 58/42 blend and sits spatially between the two 
  "8"  = "8. Macrophage, Mixed Signal (PDL1+)", 
  "9"  = "9. Macrophage, Proliferating (HLADRB+)", # Different from subcluster 2 because it has MKI67
  "10"  = "10. Monocyte, Unresolved", # Has no marker above 1.0, weak signal
  "11"  = "11. Monocyte-Macrophage, Unresolved", # Has no marker above 1.0, 87% Monocyte / 13% Macrophage 
  "12"  = "12. Macrophage (CD11c+)"
)

if (length(myeloid_subannotation) > 0) {
  spe_myeloid$sub_celltype <- myeloid_subannotation[as.character(spe_myeloid$sub_clusters)]
  cat("Subcluster annotation applied.\n")
  print(table(spe_myeloid$sub_celltype, useNA = "ifany"))
} else {
  cat("No subcluster annotation yet. Fill in myeloid_subannotation above.\n")
}
```

## 12. UMAP colored by subcluster cell type

```{r umap-sub-celltype, fig.width=10, fig.height=7}
if ("sub_celltype" %in% colnames(colData(spe_myeloid))) {

  spe_myeloid$sub_celltype <- factor(spe_myeloid$sub_celltype, levels = myeloid_subannotation)

  n_types <- nlevels(spe_myeloid$sub_celltype)
  subtype_cols <- setNames(
    colorRampPalette(brewer.pal(12, "Paired"))(n_types),
    levels(spe_myeloid$sub_celltype)
  )

  p_sub_celltype <- dittoDimPlot(spe_myeloid, var = "sub_celltype",
                                 reduction.use = "UMAP_sub",
                                 size = 0.4, do.label = TRUE, labels.size = 3.5) +
    scale_color_manual(values = subtype_cols) +
    ggtitle("Myeloid Subset: UMAP colored by Subcluster Cell Type") +
    theme(legend.position = "right")
  print(p_sub_celltype)

} else {
  cat("Run the annotate-sub chunk first to apply subcluster annotations.\n")
}
```
---

## 13. Save Subclustered/Annotated Object

```{r save}
dir.create(dirname(out_path), recursive = TRUE, showWarnings = FALSE)
saveRDS(spe_myeloid, out_path)
cat("Saved myeloid subcluster object to:", out_path, "\n")
```


## Notes

- 10 of the 12 subclusters are purely either Macrophage or Monocyte. The other two subclusters (7 and 11) have evidence of a blend.

- Subcluster 7 has a 58% macrophage and 42% monocyte blend and a CD14 only signature, sitting spatially between two territories on the UMAP. Looks like a genuine transitional population, or cells caught in the process of differentiating from monocyte into macrophage. This is probably what is mixed together within these two groups.

- Subclusters 2 and 3 (both ~45-48% Patient 10) could be because that patient's tissue is producing a cell state that doesn't look like anyone else's, which makese sense as it is a benign hamartoma patient. 

