---
title: "Final Cell Type Annotation Merge (v2 -- T Cell Subclustering Update)"
author: "Vaidehi Gupta"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

Final merge of all subclustering results (myeloid, T cell, B cell, 
DC/mregDC) into a single annotated `spe` object. This supersedes the
previous merge notebook: cluster 15 is no longer merged on its own,
it's folded into the new joint T cell subclustering (cluster 15 cleaned +
broad clusters 5/6/7).

**Important:** cluster 15's already-cleaned survivors are in
`spe_tcell`, but the ~12,085 cells previously identified as low
quality/artifact (original cluster 15 subclusters 2, 4, 7, 8) were
*removed* before that subclustering ran -- they aren't covered by any
subclustering object anymore. This notebook explicitly flags and excludes
them again here, rather than letting them silently fall back to their
stale pre-subclustering broad label.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# 1. Load Libraries

```{r libraries}
#.libPaths(c("/restricted/projectnb/camplab/home/vgupta7/R_libs", .libPaths()))

library(tidyverse)
library(SingleCellExperiment)
library(SpatialExperiment)
library(scater)
```

# 2. Paths

```{r paths}
analysis_path <- "/restricted/projectnb/camplab/projects/20220504_Suzuki_LN/TumorIMC/analysis"

spe_path <- file.path(analysis_path, "final_cell_annotation", "final_broad_cluster", "h5_cluster_annot.rds")

subclustering_path <- file.path(analysis_path, "final_cell_annotation", "subclustering")

myeloid_path   <- file.path(subclustering_path, "myeloid_cells", "myeloid_subcluster.rds")
tcell_path     <- file.path(subclustering_path, "tcells", "tcell_subclustering.rds")
bcell_path     <- file.path(subclustering_path, "bcells", "bcell_wo_15", "bcell_subclustering_wo15.rds")
dc_mregdc_path <- file.path(subclustering_path, "dc_mregdc", "dc_mregdc_subclustering.rds")

# Original cluster 15 object, needed only to identify which cells were
# excluded before the joint T cell subclustering, not used for labels
cluster15_original_path <- file.path(subclustering_path, "cluster15", "cluster15_subcluster.rds")

final_out_path <- file.path(analysis_path, "final_cell_annotation", "r2_annotation", "spe_final_celltype_r2.rds")
```

# 3. Load Broad-Annotated SPE and Subclustering Results

```{r load-intermediates}
spe <- readRDS(spe_path)

myeloid_obj    <- readRDS(myeloid_path)
tcell_obj      <- readRDS(tcell_path)
bcell_obj      <- readRDS(bcell_path)
dc_mregdc_obj  <- readRDS(dc_mregdc_path)
cluster15_orig <- readRDS(cluster15_original_path)

cat("Broad-annotated cells:", ncol(spe), "\n")
cat("Myeloid subset cells:", ncol(myeloid_obj), "\n")
cat("T cell subset cells:", ncol(tcell_obj), "\n")
cat("B cell/plasma subset cells:", ncol(bcell_obj), "\n")
cat("DC/mregDC subset cells:", ncol(dc_mregdc_obj), "\n")
```

# 4. Identify Cluster 15 Cells Excluded Before T Cell Subclustering

These are the original cluster 15 cells (subclusters 2, 4, 7 -- low
quality/patient-driven artifact, and 8, confirmed Patient20 CD19
technical artifact) that were removed prior to the joint T cell
subclustering, so they aren't present in `tcell_obj` at all.

```{r identify-excluded-cluster15}
cluster15_excluded_ids <- setdiff(colnames(cluster15_orig), colnames(tcell_obj))
cat("Cluster 15 cells excluded before T cell subclustering:",
    length(cluster15_excluded_ids), "\n")
```

# 5. Final Merge

```{r final-merge}
spe$final_celltype <- as.character(spe$broad_celltype)

# Myeloid subcluster labels
spe$final_celltype[match(colnames(myeloid_obj), colnames(spe))] <- as.character(myeloid_obj$sub_celltype)

# Joint T cell subcluster labels (covers cleaned cluster 15 + broad clusters 5/6/7)
spe$final_celltype[match(colnames(tcell_obj), colnames(spe))] <- as.character(tcell_obj$sub_celltype)

# Flag cluster 15's previously-excluded low-quality/artifact cells explicitly,
# so they get caught by the exclusion step below rather than silently
# keeping their stale pre-subclustering broad label
spe$final_celltype[colnames(spe) %in% cluster15_excluded_ids] <- "Unresolved/Low Quality"

# B cell (rerun without cluster 15) subcluster labels
spe$final_celltype[match(colnames(bcell_obj), colnames(spe))] <- as.character(bcell_obj$sub_celltype)

# DC/mregDC subcluster labels
spe$final_celltype[match(colnames(dc_mregdc_obj), colnames(spe))] <- as.character(dc_mregdc_obj$sub_celltype)

spe$final_celltype <- factor(spe$final_celltype)

cat("Cell types after merge:", nlevels(spe$final_celltype), "\n")
cat("NAs:", sum(is.na(spe$final_celltype)), "\n")
```

# 6. Strip Tracking-Number Prefixes from Labels

```{r strip-label-numbers}
spe$final_celltype <- factor(gsub("^\\d+\\.\\s*", "", as.character(spe$final_celltype)))

levels(spe$final_celltype)
```

# 7. Exclude Confirmed Low-Quality/Artifact Clusters

- **Broad cluster 17**: confirmed low quality at the broad level.
- **"Unresolved/Low Quality"**: cluster 15's previously-excluded
  low-quality/Patient20-artifact cells, flagged explicitly in chunk 5
  above (this is NOT optional -- without this, ~12,085 already-diagnosed
  bad cells would silently re-enter the final annotation under their
  stale broad label).

```{r exclude-low-quality}
n_before <- ncol(spe)

spe <- spe[, spe$broad_clusters != "17"]

low_quality_labels <- c("Unresolved/Low Quality")
spe <- spe[, !(spe$final_celltype %in% low_quality_labels)]

n_after <- ncol(spe)

cat("Cells before exclusion:", n_before, "\n")
cat("Cells after excluding low-quality/artifact clusters:", n_after, "\n")
cat("Cells removed:", n_before - n_after,
    sprintf("(%.1f%% of dataset)\n", 100 * (n_before - n_after) / n_before))

spe$final_celltype <- droplevels(spe$final_celltype)
```

# 8. Final Cell Type Counts

```{r final-celltype-table}
spe$final_celltype |>
  table(useNA = "ifany") |>
  as.data.frame() |>
  setNames(c("CellType", "Count")) |>
  arrange(desc(Count)) |>
  knitr::kable(caption = "Final cell type counts after merge and QC exclusion")
```

# 9. Validation: Heatmap of Marker Expression by Final Cell Type

Violin plots will be handled separately in a separate notebook. 

```{r validate-heatmap, fig.width=16, fig.height=12}
all_markers <- rownames(spe)

celltype_levels <- levels(spe$final_celltype)

mean_expr <- sapply(celltype_levels, function(ct) {
  idx <- which(spe$final_celltype == ct)
  rowMeans(assay(spe, "scaled_0100")[all_markers, idx, drop = FALSE])
})

marker_order <- all_markers[hclust(dist(mean_expr))$order]

celltype_sizes <- table(spe$final_celltype)
celltype_labels <- paste0(celltype_levels, "\n(n=", celltype_sizes[celltype_levels], ")")
names(celltype_labels) <- celltype_levels

heatmap_df <- as.data.frame(mean_expr) |>
  rownames_to_column("marker") |>
  pivot_longer(-marker, names_to = "celltype", values_to = "mean_expr")

heatmap_df$marker   <- factor(heatmap_df$marker, levels = marker_order)
heatmap_df$celltype <- factor(heatmap_df$celltype, levels = celltype_levels)

ggplot(heatmap_df, aes(x = celltype, y = marker, fill = mean_expr)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, name = "Mean\nZ-score") +
  scale_x_discrete(labels = celltype_labels) +
  theme_bw(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
        panel.grid = element_blank()) +
  labs(x = "Final Cell Type (n)", y = "Marker",
       title = "Final Annotation Validation: Mean Marker Expression by Cell Type")
```

# 10. Save Finalized Object

```{r save-final}
saveRDS(spe, final_out_path)
cat("Saved final annotated object to:", final_out_path, "\n")
```

# Handoff Notes

- This merge replaces cluster 15 as a standalone piece with the joint T
  cell subclustering (`tcell_subclustering.rds`), which covers cluster 15's
  cleaned survivors + broad clusters 5 (Treg), 6 (CD8+ T Cell), 7 (Memory
  T Cell).
- Cluster 15's originally-excluded low-quality/Patient20-artifact cells
  (~12,085) are explicitly flagged and excluded in chunks 4/7 -- do not
  remove this step even though "only cluster 17" was the stated intent,
  since skipping it would silently reintroduce already-diagnosed bad cells.
- Violin plots intentionally omitted -- see partner's separate notebook.
- Given ~35-40 final cell types, the full-legend UMAP is best used as an
  internal QC check (confirming each subcluster occupies a distinct region)
  rather than a presentation figure. Consider faceting by broad lineage
  (T cell / B cell / Myeloid / Epithelial) for anything shown to Campbell.
